### 2023-03-06 paper/paper.Rmd
started intro paper

---
title: |
  Introducing STRAIN: A simple and interpretable pass rush metric inspired by materials science
short_title: STRAIN
authors: 
- name: "Author 1"
  affiliation: "Address"
  city: "City"
  country: "Country"
  email: "email1\\@example.com"
- name: "Author 2"
  affiliation: "Address2"
  city: "City2"
  country: "Country2"
  email: "email2\\@example.com"
abstract: |
  Abstract goes here
keywords: tracking data, Bornn legacy
bibliography: "bibliography"
output: rticles::rss_article
---

# Introduction

A paragraph on tracking data

<!--
TODO: update citations after they're published
-->

See @Baumer2023Big, @Macdonald2020Recreating, and @Kovalchik2023Player for more details

Importance of linemen

Previous metrics: Greg's lit review, their flaws

Our contribution

# Data

# Methods

# Results

example play

positional curves

rankings (indiv, add more for team-level offense/defense)

corr with pressure rate = (hits + sacks + hurries) per snap

meta metrics: discriminant, stability, independence

@Franks2016Meta

### 2023-03-06 paper/paper.Rmd
paper sketch + everything from Kaggle notebook

@@
 
-# Introduction
+**PAPER SKETCH**
 
-A paragraph on tracking data
+1. Introduction
 
@@
 
+Tracking data, big data bowl
 See @Baumer2023Big, @Macdonald2020Recreating, and @Kovalchik2023Player for more details
 
-Importance of linemen
+Importance of linemen. Did all hard work. QB
 
@@
 
-# Data
+2. Data
 
-# Methods
+talk about data
 
-# Results
+3. Methods
+
+4. Results
 
@@
 
-meta metrics: discriminant, stability, independence
+meta metrics: discriminant, stability, independence @Franks2016Meta
 
-@Franks2016Meta
+5. Discussion
+
+# Introduction
+
+<!--
+
+Greg's long lit review
+
+The theme of Big Data Bowl 2023 is about evaluating performance of linemen, both offensive and defensive, on passing plays.
+While the ideas presented here could be extended to offensive linemen, we have chosen to focus on measuring the performance of defensive linemen.
+Officially tracked by the NFL since 1982, the most well known statistic for evaluating defensive linemen on pass rushing plays is the sack, which occurs when a defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass.
+Other counting statistics have been tracked by various entities such as hurries and hits as well as pressures, which is defined as the sum of hurries, hits, and sacks.
+More recently metrics such as Time In Pocket (TIP) and Pass Rusher Win Rate (PRWR) have also been proposed and are substantial improvements over the less sophisticated counting statistics.
+While PRWR is certainly a step in the right direction, it takes continuous data and converts it to a binary win/loss, which is generally not recommended in statistical analysis, thus, the metric we propose here is computed continuously over time just as the data is actually observed.
+
+The metric we propose is based on the following simple question: What is considered a success for a pass rusher on a pass play?
+The simplest answer to this question, and the first defensive lineman statistics to be tracked by the NFL, is a sack.
+In terms of tracking data, one can think of a sack as reducing the distance between the pass rusher and the quarterback to 0.
+Therefore, a decent starting point for a continuous measure of pass rushing effectiveness shoould be based on how close the pass rusher is able to get to the quarterback with smaller distances indicating amore successful pass rush.
+However, a good measure of pass rushing effectiveness cannot simply be based on distance to the quarterback; Consider the following case: a pass rusher gets within two yards of the quarterback, but is then stopped and fails to get any closer for the rest of the play.
+This is not a successful rush because the pass rusher is not continuing to reduce the distance between themselves and the quarterback.
+Therefore, a good metric should consider the rate at which this distance is being reduced.
+However, just like distance, rate by itself is inadequate in assessing success of a pass rusher.
+Consider the following case: A defensive player is 50 yards away and moving towards the quarterback very rapidly.
+Their moving rate is essentially unimportant due to a significantly far distance away from the quarterback.
+
+This means that in order to appropriately evaluate rusher performance, the distance from the quarterback and the rate at which this distance is being reduced must both be considered.
+Therefore, we propose STRAIN, a new measure for pass rusher effectiveness, which is computed as the ratio of the rate that the pass rusher is reducing the distance between themselve and the quarterback and distance between the pass rusher and the quarterback.
+In performing some research about this proposed quantity, we discovered that this concept actually already exists in materials science and is called the "strain rate" (which is where the name of our metric is derived).
+Given this connection, one can view the players as particles in some material and the defensive "particles" are attempting to exert pressure on the pocket in an attempt to compress and collapse this pocket around the quarterback.
+We think this analogy is just delightful.
+
+What is considered a success for a pass-rusher on a pass play?
+The simplest answer to this question is a sack.
+In terms of tracking data, one can think of a sack as reducing the distance between the pass-rusher and quarterback to 0.
+More generally, a good starting point for a continuous measure of pass-rushing effectiveness should be based on the **distance** between the rusher and quarterback, with smaller distances indicating greater success.
+However, this naive approach has major drawbacks.
+Consider the case where a rusher is within two yards of the quarterback, but is then stopped and fails to get any closer for the rest of the play.
+This is not a successful rush because the pass-rusher is not continuing to reduce their distance to the quarterback.
+Thus, another approach for pass-rushing assessment is to measure the **rate** at which the rusher is getting to the quarterback.
+Yet, the rate by itself is also inadequate.
+If a defender is 50 yards away and moving towards the quarterback, their moving rate is essentially unimportant due to a significantly far distance.
+
+Therefore, in order to appropriately evaluate rusher performance, the **distance** from the quarterback *and* the **rate** at which this distance is being reduced must both be considered.
+We propose **STRAIN**, a new measure for pass-rusher effectiveness, which is computed as the ratio of the aforementioned moving rate and distance quantities.
+Our proposed metric has a nice interpretation and is inspired by the concept of "strain rate" in materials science.
+
+-->
+
+# Data
+
+introduce data
+
+# Methods
+
+## Strain rate in materials science
+
+In [*materials science*](https://en.wikipedia.org/wiki/Materials_science), [*strain*](https://en.wikipedia.org/wiki/Strain_(materials_science)) is the deformation of a material from stress.
+The stretching of a rubber band is an example of strain: the longer the stretch, the larger the strain.
+Formally, let $L(t)$ be the distance between two points of interest at time $t$, and $L_0$ be the initial distance between those two points.
+Then the strain at time $t$ is defined as $\displaystyle s(t)={\frac  {L(t)-L_{0}}{L_{0}}}$.
+Notice that this measure is unitless as a ratio of two quantities having the same dimension.
+Furthermore, the [*strain rate*](https://en.wikipedia.org/wiki/Strain_rate) is the derivative of strain.
+That is, ${\displaystyle {{s}'}(t)={\frac {ds}{dt}} = {\frac {v(t)}{L_{0}}}}$, where $v(t)$ is the rate at which the two points of interest are moving away from/towards each other.
+Whereas strain is unitless, the strain rate is measured in inverse of time, usually $\text{seconds}^{-1}$.
+
+## Application to pass-rushing in football
+
+Motivated by its scientific definition, we draw an analogy between strain and pass-rushing in football.
+**Just as strain rate is a measure of deformation in materials science, a pass-rusher's efforts involve the application of force/deformation against the offensive line, with the ultimate goal of breaking through the protection to reach the quarterback.
+The players can be viewed as "particles" in some material and the defensive "particles" are attempting to exert pressure on the pocket with the aim of compressing and collapsing this pocket around the quarterback.**
+
+In order to apply strain rate to measure NFL pass-rusher effectiveness, we make modifications to how this concept is traditionally defined.
+
+Let $(x_{ijt}, y_{ijt})$ be the $(x, y)$ location on the field of player $j = 1, \cdots, J$ at frame $t = 1, \cdots, T_i$ for play $i = 1, \cdots, n$; and $(x^{QB}_{it}, y^{QB}_{it})$ be the $(x, y)$ location of the quarterback at frame $t$ during play $i$.
+
+* The distance between player $j$ and the quarterback at frame $t$ during play $i$ is $f_{d_{ij}}(t) = \sqrt{(x_{ijt} - x^{QB}_{it})^2 + (y_{ijt} - y^{QB}_{it})^2}$.
+* The rate at which player $j$ is moving towards the quarterback at frame $t$ during play $i$ is $\displaystyle f'_{d_{ij}}(t) = \frac{df_{d_{ij}}(t)}{dt}$.
+* The STRAIN for player $j$ at frame $t$ during play $i$ is
+
+$$\text{STRAIN}_{ij}(t) = \frac{-f'_{d_{ij}}(t)}{f_{d_{ij}}(t)} \,.$$
+
+(*Note that to distinguish our metric from strain and strain rate in materials science, we write it in capital letters* (STRAIN) *for the remainder of this report.*)
+
+Recall that by its materials science property, strain is generally defined to increase as the distance between two points **increases**.
+In our football setting, the two points of interest are the pass-rusher and the quarterback, and we want our metric to increase as the distance between the pass-rusher and the quarterback **decreases**.
+Thus, the negative sign in the numerator of our formula accounts for this. 
+Additionally, rather than keeping the initial distance ($L_0$ as previously denoted) between two points constant over time, we update the initial position to be the player locations at the beginning of each frame.
+This effectively gives us the STRAIN for each frame throughout a play.
+
+Since we only observe these functions discretely in increments of 10 frames/second, an estimate for our proposed metric STRAIN is
+
+$$\widehat{\text{STRAIN}}_{ij}(t) = \cfrac{-\cfrac{f_{d_{ij}}(t) - f_{d_{ij}}(t - 1)}{0.1}}{f_{d_{ij}}(t)} \,.$$
+
+Notice that this quantity increases in two ways: 1) the rate at which the rusher is moving towards the quarterback increases, and 2) the distance between the rusher and the quarterback gets smaller.
+Both of these are indications of pass-rusher effectiveness.
+Finally, our statistic STRAIN is measured in inverse seconds, similar to strain rate.
+(There is also an interesting interpretation of the reciprocal of our metric, namely, the amount of time required for the rusher to get to the quarterback at the current location and rate at any given time $t$.)
+
+# Results
+
+In the forthcoming analysis, we rely on the provided data of all passing plays from the first 8 weeks of the 2021 NFL season.
+We include only the frames between the ball snap and when a pass forward or QB sack is recorded for each play.
+We also remove all plays with multiple quarterbacks on the field, since we need a uniquely defined quarterback to calculate our measure.
+Using PFF scouting data, we only consider defenders with role "Pass rush" for evaluation.
+
+## Example play
+
+To illustrate STRAIN for pass-rusher evaluation, we use a play from the 2021 NFL regular season week 6 matchup between the Las Vegas Raiders and Denver Broncos, which ended with [Broncos QB Teddy Bridgewater getting sacked by Raiders DE Maxx Crosby](https://www.raiders.com/video/de-maxx-crosby-sacks-qb-teddy-bridgewater-for-a-loss-of-6-yards-nfl).
+The figure below consists of two elements:
+at top, the locations of every Las Vegas (black) and Denver (orange) player on the field during the play, with Maxx Crosby highlighted in blue;
+and at bottom, a line graph showing how Crosby's distance from the QB, velocity, and STRAIN change continuously throughout the play.
+Note that the point size for each Las Vegas defender corresponds to the estimated STRAIN in each frame as the play progresses.
+
+For the first 2.5 seconds, Crosby is being covered by a Denver player and unable to get close to the quarterback, hence the corresponding STRAIN values are virtually zero.
+Suddenly, the STRAIN increases after Crosby is freed up and charges towards Bridgewater.
+At 4 seconds after the snap, Crosby's STRAIN is 2.30, which means at his current moving rate, it will take Crosby about 0.43 (1/2.30) seconds to make the distance between him and the quarterback 0 (i.e.
+essentially sack the quarterback).
+This is consistent with the final outcome of the play, as the sack takes place at the very last frame (4.4 seconds) where the estimated STRAIN for Crosby reaches its peak at 3.96.
+
+## Positional STRAIN curves
+
+The following plot displays the average STRAIN by position for the first 40 frames (4 seconds) after the snap.
+We observe a clear difference in STRAIN between edge rushers (OLB's and DE's) and interior linemen (DT's and NT's).
+Specifically, edge rushers have higher STRAIN than interior linemen on average, as they are more easily able to approach the quarterback on the edge of the pocket versus, for instance, a nose tackle attacking the line head on.
+
+On average, STRAIN appears to increase for the first 0.5 seconds of a play, followed by a decline in the next second.
+STRAIN then increases again until around 2.5-3 seconds after the snap before trending down towards the end of the play.
+In context, this reflects the actions that a pass-rusher initially moves towards the quarterback, but is then stopped by the offensive line while the quarterback drops back.
+When the quarterback stops dropping, the rusher closes the gap and increases STRAIN, before slowing down later on.
+
+## Ranking the best pass-rushers
+
+<!--
+
+To determine the most effective pass rushers, we aggregate across all frames played and calculate the expected STRAIN (xSTRAIN) for each player, which is computed by simply averaging the STRAIN across all the frames of a play.
+The following tables show the top performing pass rushers (with at least 100 plays) rated by xSTRAIN for the first 8 weeks of the 2021 season.
+
+The top interior rusher for the first 8 weeks of the 2021 season according to our xSTRAIN metric was Aaron Donald, almost universally considered the best interior rusher in the NFL currently with an xSTRAIN of 1.67.
+Through the first 8 weeks of the 2021 season, Donald had an impressive 5 sacks and 12 quarterback hits.
+Donald Was followed by two Raiders: Solomon Thomas and Quinton Jefferson with xSTRAIN values of 1.51 and 1.46, respectively.
+Rounding out the top 5 were Chris Jones with an xSTRAIN of 1.42 and DeForest Buckner with an xSTRAIN of 1.26.
+
+As for edge rushers, Rashan Gary with an xSTRAIN value of 2.82 is the top of our list for the first 8 weeks of the 2021 season.
+Over this time period, Gary had 3.5 sacks and 15 quarterback hits.
+Second on our list of edge rushers is Leonard Floyd with an xSTRAIN value of 2.80 who had 6.5 sacks and 9 quarterback hits over the time period observed.
+Rounding out the top 5 are Justin Houston (2.78), Myles Garrett (2.75), and Von Miller (2.75).
+The sixth player on our rankings of edge rushers over the first 8 weeks on the 2021 season is the 2021 Defensive Player of the Year, T.J.
+Watt (2.71) who had 11.5 sacks and 18 quarterback hits in those first 8 weeks.
+Notably their is considerable overlap between our top 20 list over the first 8 weeks of the 2021 season and the Pro Football Focus edge rusher rankings prior to the 2022 season.
+
+-->
+
+To determine the most effective pass-rushers, we compute $\overline{\text{STRAIN}}$, which represents the average STRAIN across all frames played for each player.
+Based on the clearly distinct patterns for different positions as previously observed, we evaluate interior pass-rushers (NT's and DT's) separately from edge rushers (OLB's and DE's).
+The following tables show the top performing pass-rushers (with at least 100 plays) rated by $\overline{\text{STRAIN}}$ for the first 8 weeks of the 2021 season.
+
+Our results show many true positives with no surprises.
+Myles Garrett and TJ Watt are known as top-tier edge rushers, while Aaron Donald is undoubtedly the best interior defender in football.
+Rashan Gary, a young rusher with huge potential, ranks first among all pass-rushers according to our metric.
+Our leaderboards largely match the rankings of experts in the field (for instance, PFF's [edge](https://www.pff.com/news/nfl-2022-edge-rusher-rankings-tiers) and [interior](https://www.pff.com/news/nfl-interior-defensive-line-rankings-tiers-2022) rusher rankings after the 2021 season), lending credibility to our proposed metric as a measure of pass-rushing effectiveness.
+
+# Discussion
+
+<!--
+
+We have proposed a simple, effective, and easily interpretable metric for evaluating the performance of pass rushers in the NFL that we call STRAIN.
+Higher values of STRAIN indicate two important aspects of a pass rush: 1) the distance between the pass rusher and the quarterback is small and 2) the rate at which the pass rusher is moving towards the quarterback is high.
+
+One of the major advantages of STRAIN over existing measures is that it gives a measure of effectiveness of a pass rush for every single play at a much more granular level than simply the play resulted in a sack or not or the play was a "success" or not.
+It is important to have a quantitative measure of pass rushing effectiveness if one wants to objectively evaluate the performance of pass rushers.
+STRAIN is just such a measure that objectively measures the pressure that pass rushers put on a quarterback.
+
+In this report, we simply propose the new metric STRAIN; However, there are innumerable future projects that could be based on STRAIN.
+For example, STRAIN could be used to collectively evaluate the performance of a pass rushing unit by looking at the total amount of STRAIN during a play, which could, in turn, be used to evaluate the effectiveness of different types of blitzes.
+In addtion, STRAIN could be included as a covariate in models trying to predict quantities such as passing yards or completion probability.
+Notably, I believe there is very interesting work that could be done with functional regression (i.e.
+treating the STRAIN curve over time as a function) in modeling how STRAIN affects offensive production though is non-trivial as the length in time of plays varies from one play to another, and we leave this as future work.
+
+In addtion, while we focused solely on evaluating the performance of pass rushers, this concept can be easily extended to the evaluation of pass blockers as a unit by simply looking at a quantity such as the total amount of STRAIN on a play or the maximum amount of STRAIN per frame aggregated across the entire play.
+Further, STRAIN could be extended to individual offensive pass blockers provided that a method of matching blockers to rushers was devised.
+
+While STRAIN is a big step forward, there are still some notable limitations to this metric.
+For instance, STRAIN, as currently devised, cannot distinguish between a pass rusher who has to face a single blocker versus a double team.
+This problem could be addressed, at least somewhat, by devising a method of matching pass rushers to blockers as previously mentioned.
+Another limitation of this metric occurs when a quarterback scrambles out of the pocket.
+If you are a pass rusher on the left side and the quarterback scrambles out to the left, the STRAIN value for that pass rusher can drop considerably as the distance between the quarterback and the pass rusher gets larger, which the pass rushers shouldn't necessarily be negatively affected by in a measure of pass rushing effectiveness.
+
+Finally, we believe there are other concepts in materials science that could be applied to evaluating pass rushing and pass blocking.
+One potential idea is a quantity called stress, which measures force over an area.
+Recall from intro physics that force is mass times acceleration, so this quantity would take into account the physical size of a pass rusher in the computation of force.
+You could then somehow divide this force over the "area" of the pocket formed by pass blockers to compute stress.
+There is also a quantity measuring the ratio of stress and strain called the elastic modulus which measures "an object's or substance's resistance to being deformed elastically".
+It is easy to imagine the pocket as an object and evaluating the succes of the pocket based on it's "resistance to being deformed".
+
+-->
+
+We have proposed STRAIN, a simple and interpretable statistic for evaluating pass-rushers, with higher STRAIN corresponds to greater pass-rushing ability.
+Compared to existing metrics such as [ESPN's Pass Rusher Win Rate (PRWR)](https://www.espn.com/nfl/story/_/id/24892208), STRAIN measures pass-rush effectiveness for every play continuously over time, which is at a much more granular level than considering whether the play resulted in a sack or converting continuous data to a binary win/loss.
+
+As for future directions, STRAIN could be utilized to collectively evaluate the performance of a pass-rushing unit by considering the total amount of STRAIN during a play, which could, in turn, be used to examine the effectiveness of different types of blitzes.
+Moreover, while we focused solely on pass-rushers, STRAIN can be easily extended to the assessment of pass-blockers *as a unit* by looking at quantities such as the total STRAIN or the maximum STRAIN per frame aggregated across the entire play.
+It is also possible to apply STRAIN to assess individual offensive pass-blockers, provided that a method of matching blockers to rushers was developed.
+
+Our metric, however, is subject to several limitations.
+As currently devised, STRAIN cannot distinguish between a pass-rusher who has to face a single blocker versus a double team.
+This problem could be addressed, at least somewhat, by having a method of matching pass-rushers to blockers as previously mentioned.
+Additionally, there are a few situations such as quarterback out-of-pocket scrambling where STRAIN does not appear to be useful.
+If a pass-rusher is on the left side and the quarterback scrambles out to the left, the STRAIN for that pass-rusher can drop considerably as the distance between the quarterback and pass-rusher increases, which the defender should not necessarily be negatively affected by in a measure of pass-rushing effectiveness.

### 2023-03-08 paper/paper.Rmd
some template updates
data and more sketches

@@
   Abstract goes here
-keywords: tracking data, Bornn legacy
+keywords: tracking data, Bornn legacy, too much Canadian
 bibliography: "bibliography"
@@
 
+\newpage
+
 **PAPER SKETCH**
 
-1. Introduction
+i. Introduction
 
@@
 
-Tracking data, big data bowl
+Tracking data in recent years, NFL
+
+examples of past nfl tracking data projects
+
 See @Baumer2023Big, @Macdonald2020Recreating, and @Kovalchik2023Player for more details
 
-Importance of linemen. Did all hard work. QB
+Importance of D-linemen, pass rush (Maybe talk about role of each position, edge vs interior, defensive formation)
 
@@
 
-2. Data
+ii. Methods
 
-talk about data
+a. data
 
-3. Methods
+2023 bdb [@bdb2023]
 
-4. Results
+first 8 weeks 2021 season
+
+tracking: locations of players and ball, 10 fps,
+
+pre-processing standardize
+
+PFF scouting: pass rush role
+
+sample restriction: remove plays w multiple QB
+
+frames between snap and QB outcome (pass forward, sack)
+
+sample size
+
+b. strain rate
+
+c. applying to pass rush
+
+iii. Results
 
@@
 
-5. Discussion
+iv. Discussion
+
+\newpage
 
 # Introduction
+
+## Previous Metrics for Pass Rush Evaluation
+
+## Our Contribution
 
@@
 
-# Data
-
-introduce data
+The remainder of this paper is outlined as follows.
 
 # Methods
+
+## Data
+
+bdb
 
@@
 
-* The distance between player $j$ and the quarterback at frame $t$ during play $i$ is $f_{d_{ij}}(t) = \sqrt{(x_{ijt} - x^{QB}_{it})^2 + (y_{ijt} - y^{QB}_{it})^2}$.
-* The rate at which player $j$ is moving towards the quarterback at frame $t$ during play $i$ is $\displaystyle f'_{d_{ij}}(t) = \frac{df_{d_{ij}}(t)}{dt}$.
+* The distance between player $j$ and the quarterback at frame $t$ during play $i$ is
+
+$$f_{d_{ij}}(t) = \sqrt{\left(x_{ijt} - x^{QB}_{it}\right)^2 + \left(y_{ijt} - y^{QB}_{it}\right)^2}$$.
+
+* The rate at which player $j$ is moving towards the quarterback at frame $t$ during play $i$ is
+
+$$\displaystyle f'_{d_{ij}}(t) = \frac{df_{d_{ij}}(t)}{dt}$$.
+
 * The STRAIN for player $j$ at frame $t$ during play $i$ is

### 2023-03-09 paper/kaggle.Rmd
more clean-up; moved Kaggle nb to separate file; started intro

--- 2023-03-06 paper/paper.Rmd
@@
 ---
-title: |
-  Introducing STRAIN: A simple and interpretable pass rush metric inspired by materials science
-short_title: STRAIN
-authors: 
-- name: "Author 1"
-  affiliation: "Address"
-  city: "City"
-  country: "Country"
-  email: "email1\\@example.com"
-- name: "Author 2"
-  affiliation: "Address2"
-  city: "City2"
-  country: "Country2"
-  email: "email2\\@example.com"
-abstract: |
-  Abstract goes here
-keywords: tracking data, Bornn legacy
-bibliography: "bibliography"
-output: rticles::rss_article
+output: html_document
 ---
-
-**PAPER SKETCH**
-
-1. Introduction
-
-<!--
-TODO: update citations after they're published
--->
-
-Tracking data, big data bowl
-See @Baumer2023Big, @Macdonald2020Recreating, and @Kovalchik2023Player for more details
-
-Importance of linemen. Did all hard work. QB
-
-Previous metrics: Greg's lit review, their flaws
-
-Our contribution
-
-2. Data
-
-talk about data
-
-3. Methods
-
-4. Results
-
-example play
-
-positional curves
-
-rankings (indiv, add more for team-level offense/defense)
-
-corr with pressure rate = (hits + sacks + hurries) per snap
-
-meta metrics: discriminant, stability, independence @Franks2016Meta
-
-5. Discussion
 
@@
 <!--
-
 Greg's long lit review
-
 The theme of Big Data Bowl 2023 is about evaluating performance of linemen, both offensive and defensive, on passing plays.
@@
 While PRWR is certainly a step in the right direction, it takes continuous data and converts it to a binary win/loss, which is generally not recommended in statistical analysis, thus, the metric we propose here is computed continuously over time just as the data is actually observed.
-
 The metric we propose is based on the following simple question: What is considered a success for a pass rusher on a pass play?
@@
 Their moving rate is essentially unimportant due to a significantly far distance away from the quarterback.
-
 This means that in order to appropriately evaluate rusher performance, the distance from the quarterback and the rate at which this distance is being reduced must both be considered.
@@
 We think this analogy is just delightful.
+-->
 
@@
 Our proposed metric has a nice interpretation and is inspired by the concept of "strain rate" in materials science.
-
--->
-
-# Data
-
-introduce data
 
@@
 
-# Results
+# Analysis
 
@@
 <!--
-
 To determine the most effective pass rushers, we aggregate across all frames played and calculate the expected STRAIN (xSTRAIN) for each player, which is computed by simply averaging the STRAIN across all the frames of a play.
 The following tables show the top performing pass rushers (with at least 100 plays) rated by xSTRAIN for the first 8 weeks of the 2021 season.
-
 The top interior rusher for the first 8 weeks of the 2021 season according to our xSTRAIN metric was Aaron Donald, almost universally considered the best interior rusher in the NFL currently with an xSTRAIN of 1.67.
@@
 Rounding out the top 5 were Chris Jones with an xSTRAIN of 1.42 and DeForest Buckner with an xSTRAIN of 1.26.
-
 As for edge rushers, Rashan Gary with an xSTRAIN value of 2.82 is the top of our list for the first 8 weeks of the 2021 season.
@@
 Notably their is considerable overlap between our top 20 list over the first 8 weeks of the 2021 season and the Pro Football Focus edge rusher rankings prior to the 2022 season.
-
 -->
 
-To determine the most effective pass-rushers, we compute $\overline{\text{STRAIN}}$, which represents the average STRAIN across all frames played for each player.
+To determine the most effective pass-rushers, we compute <span style="text-decoration:overline; font-weight:bold">STRAIN</span>, which represents the average STRAIN across all frames played for each player.
 Based on the clearly distinct patterns for different positions as previously observed, we evaluate interior pass-rushers (NT's and DT's) separately from edge rushers (OLB's and DE's).
-The following tables show the top performing pass-rushers (with at least 100 plays) rated by $\overline{\text{STRAIN}}$ for the first 8 weeks of the 2021 season.
+The following tables show the top performing pass-rushers (with at least 100 plays) rated by <span style="text-decoration:overline; font-weight:bold">STRAIN</span> for the first 8 weeks of the 2021 season.
 
@@
 <!--
-
 We have proposed a simple, effective, and easily interpretable metric for evaluating the performance of pass rushers in the NFL that we call STRAIN.
 Higher values of STRAIN indicate two important aspects of a pass rush: 1) the distance between the pass rusher and the quarterback is small and 2) the rate at which the pass rusher is moving towards the quarterback is high.
-
 One of the major advantages of STRAIN over existing measures is that it gives a measure of effectiveness of a pass rush for every single play at a much more granular level than simply the play resulted in a sack or not or the play was a "success" or not.
@@
 STRAIN is just such a measure that objectively measures the pressure that pass rushers put on a quarterback.
-
 In this report, we simply propose the new metric STRAIN; However, there are innumerable future projects that could be based on STRAIN.
@@
 treating the STRAIN curve over time as a function) in modeling how STRAIN affects offensive production though is non-trivial as the length in time of plays varies from one play to another, and we leave this as future work.
-
 In addtion, while we focused solely on evaluating the performance of pass rushers, this concept can be easily extended to the evaluation of pass blockers as a unit by simply looking at a quantity such as the total amount of STRAIN on a play or the maximum amount of STRAIN per frame aggregated across the entire play.
 Further, STRAIN could be extended to individual offensive pass blockers provided that a method of matching blockers to rushers was devised.
-
 While STRAIN is a big step forward, there are still some notable limitations to this metric.
@@
 If you are a pass rusher on the left side and the quarterback scrambles out to the left, the STRAIN value for that pass rusher can drop considerably as the distance between the quarterback and the pass rusher gets larger, which the pass rushers shouldn't necessarily be negatively affected by in a measure of pass rushing effectiveness.
-
 Finally, we believe there are other concepts in materials science that could be applied to evaluating pass rushing and pass blocking.
@@
 It is easy to imagine the pocket as an object and evaluating the succes of the pocket based on it's "resistance to being deformed".
-
 -->

### 2023-03-09 paper/paper.Rmd
more clean-up; moved Kaggle nb to separate file; started intro

--- 2023-03-06 paper/paper.Rmd
@@
   Abstract goes here
-keywords: tracking data, Bornn legacy
+keywords: tracking data, Bornn legacy, too much Canadian
 bibliography: "bibliography"
 output: rticles::rss_article
+header-includes:
+  \usepackage{setspace}
+  \usepackage{booktabs}
 ---
 
+\newpage
+
 # Introduction
-
-A paragraph on tracking data
 
@@
 
-See @Baumer2023Big, @Macdonald2020Recreating, and @Kovalchik2023Player for more details
+In recent years, tracking data have replaced traditional box-score statistics and play-by-play data as the state-of-the-art tool for analyzing and deeper understanding different sports.
 
-Importance of linemen
+examples of past nfl tracking data projects
+
+See @Baumer2023Big, @Macdonald2020Recreating, and @Kovalchik2023Player for more complete reviews on how tracking data have transformed sports analytics.
+
+NFL, next gen stats https://operations.nfl.com/gameday/technology/nfl-next-gen-stats/
+
+Importance of D-linemen, pass rush (Maybe talk about role of each position, edge vs interior, defensive formation)
 
@@
 
-Our contribution
+## Previous Pass Rush Metrics
 
-# Data
+\begin{table}
+\caption{A summary of previously-existed pass rush metrics. \label{tab:metrics}}
+\centering
+\begin{tabular}{ll}
+\toprule
+\textbf{Metric} & \textbf{Description}\\
+\midrule
+Sacks & a\\
+Hits & a\\
+Hurries & a\\
+QB Pressures & a\\
+Pass-Rush Productivity (PRP) & a\\
+Time In Pocket (TIP) & a\\
+Pass Rush Win Rate (PRWR) & Rate at which pass rusher beats pass block within 2.5 seconds after snap\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+Table \ref{tab:metrics}
+
+Sacks, hits, hurries, pressures
+
+Officially tracked by the NFL since 1982, the most well known statistic for evaluating defensive linemen on pass rushing plays is the sack, which occurs when a defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass.
+
+https://www.pro-football-reference.com/about/advanced_stats.htm
+
+Pressures: QB pressures (hurries + knockdowns + sack plays). Note that "sack plays" in this instance include both sacks and assists.
+
+Hurried: Times QBs were hurried - defined as a situation when the QB is forced to throw the ball earlier than intended or is chased around/out of the pocket as the result of defensive pressure
+
+Hit: Times QBs were hit - occurs when the QB hits the ground after a throw, not marked on a sack play
+
+PFF's Pass-Rush Productivity (PRP)
+
+https://www.pff.com/news/pro-signature-stats-spotlight-defensive-line
+
+More recently metrics such as Time In Pocket (TIP) and Pass Rush Win Rate (PRWR) have also been proposed and are substantial improvements over the less sophisticated counting statistics.
+
+Pocket Time: Number of seconds the quarterback was in the pocket between the snap of the ball and a pass or the collapse of the pocket
+
+@Burke2018Created
+
+https://www.espn.com/nfl/story/_/id/24892208
+
+2.5 seconds, arbitrary
+
+While PRWR is certainly a step in the right direction, it takes continuous data and converts it to a binary win/loss, which is generally not recommended in statistical analysis
+
+## Contributions of STRAIN
+
+The remainder of this paper is outlined as follows.
 
@@
 
+## Data
+
+2023 bdb [@Howard2023NFL]
+
+first 8 weeks 2021 season
+
+tracking: locations of players and ball, 10 fps,
+
+pre-processing standardize
+
+PFF scouting: pass rush role
+
+sample restriction: remove plays w multiple QB
+
+frames between snap and QB outcome (pass forward, sack)
+
+sample size
+
+## Strain rate in materials science
+
+## Application to pass-rushing in football
+
 # Results
 
-example play
+## Example play
 
-positional curves
+## Positional STRAIN curves
 
-rankings (indiv, add more for team-level offense/defense)
+## Ranking the best pass-rushers
 
-corr with pressure rate = (hits + sacks + hurries) per snap
+## Corr with pressure rate = (hits + sacks + hurries) per snap
 
-meta metrics: discriminant, stability, independence
+## Statistical Properties of STRAIN
 
-@Franks2016Meta
+From @Franks2016Meta (github https://github.com/afranks86/meta-analytics)
+
+Discrimination: Does the metric reliably differentiate between players?
+
+Stability: Does the metric measure a quantity which is stable over time?
+
+Independence: Does the metric provide new information?
+
+# Discussion
+
+# References

### 2023-03-09 paper/paper.Rmd
tracking data intro

@@
 
-In recent years, tracking data have replaced traditional box-score statistics and play-by-play data as the state-of-the-art tool for analyzing and deeper understanding different sports.
+In recent years, tracking data have replaced traditional box-score statistics and play-by-play data as the state-of-the-art tool in sports analytics.
+Numerous sports have been collecting and releasing data on player and ball locations on the playing surface over the course of a game.
+This spatio-temporal multi-resolution source of data have provided exceptional opportunities for researchers to perform advanced studies at a more complex level to deepen the understanding of different sports.
+For complete surveys on how tracking data have transformed sports analytics, see @Macdonald2020Recreating, @Baumer2023Big, and @Kovalchik2023Player.
 
-examples of past nfl tracking data projects
+The National Football League (NFL) introduced their player tracking system known as Next Gen Stats (NGS) in 2016
+BDB
+https://operations.nfl.com/gameday/technology/nfl-next-gen-stats/
 
-See @Baumer2023Big, @Macdonald2020Recreating, and @Kovalchik2023Player for more complete reviews on how tracking data have transformed sports analytics.
-
-NFL, next gen stats https://operations.nfl.com/gameday/technology/nfl-next-gen-stats/
-
+Previous research bdb
+this year: linemen
 Importance of D-linemen, pass rush (Maybe talk about role of each position, edge vs interior, defensive formation)
 
-Previous metrics: Greg's lit review, their flaws
+## Previous Pass Rush Metrics
 
-## Previous Pass Rush Metrics
+Limitations
 

### 2023-03-10 paper/paper.Rmd
built figures + pipeline

@@
 ---
+
+```{r}
+#| label: setup
+#| include: false
+knitr::opts_chunk$set(
+  echo = FALSE,
+  message = FALSE,
+  warning = FALSE,
+  fig.align = "center"
+)
+library(here)
+```
 
@@
 
+```{r}
+#| label: fig_field
+#| fig.cap: "snapshots"
+#| fig.width: 7
+#| fig.height: 9
+source(here("scripts", "fig_field.R"))
+fig_field
+```
+
+Figure \ref{fig:fig_field} shows snapshots
+
+```{r}
+#| label: fig_crosby_curves
+#| fig.cap: "crosby"
+#| fig.width: 5.1
+#| fig.height: 3
+source(here("scripts", "fig_crosby_curves.R"))
+fig_crosby_curves
+```
+
+Figure \ref{fig:fig_crosby_curves} shows Crosby throughout the play
+
 ## Positional STRAIN curves
+
+```{r}
+#| label: fig_pos_curves
+#| fig.cap: "STRAIN curves for different positions"
+#| fig.width: 4.5
+#| fig.height: 3
+source(here("scripts", "fig_pos_curves.R"))
+fig_pos_curves
+```
+
+Figure \ref{fig:fig_pos_curves} shows pos curves
 
@@
 ## Corr with pressure rate = (hits + sacks + hurries) per snap
+
+```{r}
+#| label: fig_cor_pressure
+#| fig.cap: "Cor with"
+#| fig.width: 4.2
+#| fig.height: 3
+source(here("scripts", "fig_cor_pressure.R"))
+fig_cor_pressure
+```
+
+Figure \ref{fig:fig_cor_pressure} shows cor with pressure
+
+Fairly strong correlation (r = 0.6254943) between average STRAIN & hurries + sacks + hits per snaps
 
@@
 
+```{r}
+#| label: fig_stability
+#| fig.cap: "Metric stability"
+#| fig.width: 4.2
+#| fig.height: 3
+source(here("scripts", "fig_stability.R"))
+fig_stability
+```
+
+0.8544865
+
+Figure \ref{fig:fig_stability} shows
+
+Average STRAIN across all frames played in first and last 4 weeks
+
+Strong correlation shows great metric stability
+
 # Discussion

### 2023-03-10 paper/paper.tex
built figures + pipeline

--- 2023-03-09 paper/paper.Rmd
@@
----
-title: |
-  Introducing STRAIN: A simple and interpretable pass rush metric inspired by materials science
-short_title: STRAIN
-authors: 
-- name: "Author 1"
-  affiliation: "Address"
-  city: "City"
-  country: "Country"
-  email: "email1\\@example.com"
-- name: "Author 2"
-  affiliation: "Address2"
-  city: "City2"
-  country: "Country2"
-  email: "email2\\@example.com"
-abstract: |
+\documentclass{statsoc}
+\usepackage{graphicx}
+\usepackage{listings}
+\usepackage{color}
+\usepackage{amssymb, amsmath, geometry}
+\usepackage{natbib}
+\usepackage{hyperref}
+
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+
+\title[STRAIN]{Introducing STRAIN: A simple and interpretable pass rush
+metric inspired by materials science}
+
+\author[]{Author 1}
+\address{Address,
+City,
+Country}
+\email{email1@example.com}
+
+\author[]{Author 2}
+\address{Address2,
+City2,
+Country2}
+\email{email2@example.com}
+
+% BIBLIOGRAPHY
+\usepackage[authoryear]{natbib}
+\bibpunct{(}{)}{;}{a}{}{,}
+
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+\usepackage{setspace} \usepackage{booktabs}
+
+\begin{document}
+
+\begin{abstract}
 Abstract goes here
-keywords: tracking data, Bornn legacy, too much Canadian
-bibliography: "bibliography"
-output: rticles::rss_article
-header-includes:
-  \usepackage{setspace}
-  \usepackage{booktabs}
----
+\end{abstract}
+\keywords{tracking data, Bornn legacy, too much Canadian}
 
@@
 
-# Introduction
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
-<!--
-TODO: update citations after they're published
--->
+In recent years, tracking data have replaced traditional box-score
+statistics and play-by-play data as the state-of-the-art tool in sports
+analytics. Numerous sports have been collecting and releasing data on
+player and ball locations on the playing surface over the course of a
+game. This spatio-temporal multi-resolution source of data have provided
+exceptional opportunities for researchers to perform advanced studies at
+a more complex level to deepen the understanding of different sports.
+For complete surveys on how tracking data have transformed sports
+analytics, see \citet{Macdonald2020Recreating}, \citet{Baumer2023Big},
+and \citet{Kovalchik2023Player}.
 
-In recent years, tracking data have replaced traditional box-score statistics and play-by-play data as the state-of-the-art tool in sports analytics.
-Numerous sports have been collecting and releasing data on player and ball locations on the playing surface over the course of a game.
-This spatio-temporal multi-resolution source of data have provided exceptional opportunities for researchers to perform advanced studies at a more complex level to deepen the understanding of different sports.
-For complete surveys on how tracking data have transformed sports analytics, see @Macdonald2020Recreating, @Baumer2023Big, and @Kovalchik2023Player.
+The National Football League (NFL) introduced their player tracking
+system known as Next Gen Stats (NGS) in 2016 BDB
+\url{https://operations.nfl.com/gameday/technology/nfl-next-gen-stats/}
 
-The National Football League (NFL) introduced their player tracking system known as Next Gen Stats (NGS) in 2016
-BDB
-https://operations.nfl.com/gameday/technology/nfl-next-gen-stats/
+Previous research bdb this year: linemen Importance of D-linemen, pass
+rush (Maybe talk about role of each position, edge vs interior,
+defensive formation)
 
-Previous research bdb
-this year: linemen
-Importance of D-linemen, pass rush (Maybe talk about role of each position, edge vs interior, defensive formation)
-
-## Previous Pass Rush Metrics
+\hypertarget{previous-pass-rush-metrics}{%
+\subsection{Previous Pass Rush
+Metrics}\label{previous-pass-rush-metrics}}
 
@@
 
-Officially tracked by the NFL since 1982, the most well known statistic for evaluating defensive linemen on pass rushing plays is the sack, which occurs when a defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass.
+Officially tracked by the NFL since 1982, the most well known statistic
+for evaluating defensive linemen on pass rushing plays is the sack,
+which occurs when a defensive player tackles the quarterback behind the
+line of scrimmage before the quarterback is able to throw a pass.
 
-https://www.pro-football-reference.com/about/advanced_stats.htm
+\url{https://www.pro-football-reference.com/about/advanced_stats.htm}
 
-Pressures: QB pressures (hurries + knockdowns + sack plays). Note that "sack plays" in this instance include both sacks and assists.
+Pressures: QB pressures (hurries + knockdowns + sack plays). Note that
+``sack plays'' in this instance include both sacks and assists.
 
-Hurried: Times QBs were hurried - defined as a situation when the QB is forced to throw the ball earlier than intended or is chased around/out of the pocket as the result of defensive pressure
+Hurried: Times QBs were hurried - defined as a situation when the QB is
+forced to throw the ball earlier than intended or is chased around/out
+of the pocket as the result of defensive pressure
 
-Hit: Times QBs were hit - occurs when the QB hits the ground after a throw, not marked on a sack play
+Hit: Times QBs were hit - occurs when the QB hits the ground after a
+throw, not marked on a sack play
 
@@
 
-https://www.pff.com/news/pro-signature-stats-spotlight-defensive-line
+\url{https://www.pff.com/news/pro-signature-stats-spotlight-defensive-line}
 
-More recently metrics such as Time In Pocket (TIP) and Pass Rush Win Rate (PRWR) have also been proposed and are substantial improvements over the less sophisticated counting statistics.
+More recently metrics such as Time In Pocket (TIP) and Pass Rush Win
+Rate (PRWR) have also been proposed and are substantial improvements
+over the less sophisticated counting statistics.
 
-Pocket Time: Number of seconds the quarterback was in the pocket between the snap of the ball and a pass or the collapse of the pocket
+Pocket Time: Number of seconds the quarterback was in the pocket between
+the snap of the ball and a pass or the collapse of the pocket
 
-@Burke2018Created
+\citet{Burke2018Created}
 
-https://www.espn.com/nfl/story/_/id/24892208
+\url{https://www.espn.com/nfl/story/_/id/24892208}
 
@@
 
-While PRWR is certainly a step in the right direction, it takes continuous data and converts it to a binary win/loss, which is generally not recommended in statistical analysis
+While PRWR is certainly a step in the right direction, it takes
+continuous data and converts it to a binary win/loss, which is generally
+not recommended in statistical analysis
 
-## Contributions of STRAIN
+\hypertarget{contributions-of-strain}{%
+\subsection{Contributions of STRAIN}\label{contributions-of-strain}}
 
@@
 
-# Methods
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
 
-## Data
+\hypertarget{data}{%
+\subsection{Data}\label{data}}
 
-2023 bdb [@Howard2023NFL]
+2023 bdb \citep{Howard2023NFL}
 
@@
 
-## Strain rate in materials science
+\hypertarget{strain-rate-in-materials-science}{%
+\subsection{Strain rate in materials
+science}\label{strain-rate-in-materials-science}}
 
-## Application to pass-rushing in football
+\hypertarget{application-to-pass-rushing-in-football}{%
+\subsection{Application to pass-rushing in
+football}\label{application-to-pass-rushing-in-football}}
 
-# Results
+\hypertarget{results}{%
+\section{Results}\label{results}}
 
-## Example play
+\hypertarget{example-play}{%
+\subsection{Example play}\label{example-play}}
 
-## Positional STRAIN curves
+\begin{figure}
 
-## Ranking the best pass-rushers
+{\centering \includegraphics{paper_files/figure-latex/fig_field-1}
 
-## Corr with pressure rate = (hits + sacks + hurries) per snap
+}
 
-## Statistical Properties of STRAIN
+\caption{snapshots}\label{fig:fig_field}
+\end{figure}
 
-From @Franks2016Meta (github https://github.com/afranks86/meta-analytics)
+Figure \ref{fig:fig_field} shows snapshots
+
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-latex/fig_crosby_curves-1}
+
+}
+
+\caption{crosby}\label{fig:fig_crosby_curves}
+\end{figure}
+
+Figure \ref{fig:fig_crosby_curves} shows Crosby throughout the play
+
+\hypertarget{positional-strain-curves}{%
+\subsection{Positional STRAIN curves}\label{positional-strain-curves}}
+
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-latex/fig_pos_curves-1}
+
+}
+
+\caption{STRAIN curves for different positions}\label{fig:fig_pos_curves}
+\end{figure}
+
+Figure \ref{fig:fig_pos_curves} shows pos curves
+
+\hypertarget{ranking-the-best-pass-rushers}{%
+\subsection{Ranking the best
+pass-rushers}\label{ranking-the-best-pass-rushers}}
+
+\hypertarget{corr-with-pressure-rate-hits-sacks-hurries-per-snap}{%
+\subsection{Corr with pressure rate = (hits + sacks + hurries) per
+snap}\label{corr-with-pressure-rate-hits-sacks-hurries-per-snap}}
+
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-latex/fig_cor_pressure-1}
+
+}
+
+\caption{Cor with}\label{fig:fig_cor_pressure}
+\end{figure}
+
+Figure \ref{fig:fig_cor_pressure} shows cor with pressure
+
+Fairly strong correlation (r = 0.6254943) between average STRAIN \&
+hurries + sacks + hits per snaps
+
+\hypertarget{statistical-properties-of-strain}{%
+\subsection{Statistical Properties of
+STRAIN}\label{statistical-properties-of-strain}}
+
+From \citet{Franks2016Meta} (github
+\url{https://github.com/afranks86/meta-analytics})
 
@@
 
-# Discussion
+\begin{figure}
 
-# References
+{\centering \includegraphics{paper_files/figure-latex/fig_stability-1}
+
+}
+
+\caption{Metric stability}\label{fig:fig_stability}
+\end{figure}
+
+0.8544865
+
+Figure \ref{fig:fig_stability} shows
+
+Average STRAIN across all frames played in first and last 4 weeks
+
+Strong correlation shows great metric stability
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+\bibliographystyle{rss}
+\bibliography{bibliography}
+
+\end{document}

### 2023-03-10 scripts/fig_cor_pressure.R

@@ -1,0 +1,53 @@
+library(tidyverse)
+library(here)
+
+pass_rush_pos <- read_rds(here("data", "pass_rush_pos.rds"))
+hsh <- pass_rush_pos |>
+  distinct(nflId, gameId, pff_hurry, pff_sack, pff_hit) |>
+  group_by(nflId) |>
+  summarize(hurries = sum(pff_hurry, na.rm = TRUE),
+            sacks = sum(pff_sack, na.rm = TRUE),
+            hits = sum(pff_hit, na.rm = TRUE))
+
+test <- pass_rush_pos |> 
+  filter(team == defensiveTeam & !is.na(strain))
+test$strain <- test$slope_distance_from_QB/test$distance_from_QB
+test_summed <- test |> 
+  filter(!is.infinite(strain)) |> 
+  group_by(nflId) |>
+  summarize(
+    strain = sum(strain),
+    name = head(displayName, 1),
+    pos = head(officialPosition, 1),
+    n = n(),
+    n_plays = length(unique(playId))
+  ) |>
+  mutate(strain_rate = 10 * strain / n) |> 
+  arrange(-strain_rate) |> 
+  ungroup()
+
+players <- read_csv(here("data", "players.csv"))
+
+fig_cor_pressure <- test_summed |> 
+  left_join(hsh) |> 
+  filter(n_plays >= 100 & !is.infinite(strain)) |> 
+  mutate(pres = (hurries + sacks + hits) / n_plays) |>
+  mutate(pos = factor(pos, levels = c("OLB", "DE", "DT", "NT"))) |> 
+  ggplot(aes(strain_rate, pres)) +
+  geom_smooth(method = "lm", se = FALSE, linewidth = 1.5, color = "black", alpha = 0.5) +
+  geom_point(aes(color = pos, group = name), size = 2, alpha = 0.8) +
+  expand_limits(x = 3, y = c(0, 0.12)) +
+  scale_y_continuous(breaks = seq(0, 0.12, 0.04)) +
+  scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107", "#004D40")) +
+  labs(y = "(Hurries + Sacks + Hits) per snaps",
+       x = "Average STRAIN across all frames",
+       color = "Position") +
+  theme_light() +
+  theme(axis.title = element_text(size = rel(1)),
+        legend.title = element_text(size = rel(1)),
+        legend.text = element_text(size = rel(0.8)),
[…]

### 2023-03-10 scripts/fig_crosby_curves.R

@@ -1,0 +1,50 @@
+library(here)
+source(here("scripts", "fig_field.R"))
+pff <- read_csv(here("data", "pffScoutingData.csv"))
+
+crosby <- play_lv_den |> 
+  ungroup() |> 
+  filter(nflId == 47889) |>
+  select(frameId_snap_corrected, distance_from_QB, slope_distance_from_QB, strain)
+
+
+fig_crosby_curves <- play_lv_den |> 
+  left_join(pff) |>
+  filter(pff_role == "Pass Rush", nflId != 47889) |> 
+  select(name = displayName, frameId_snap_corrected, value = strain) |> 
+  mutate(name = str_c("strain_", name)) |> 
+  bind_rows(pivot_longer(crosby, !frameId_snap_corrected)) |> 
+  mutate(
+    name = case_when(
+      name == "distance_from_QB" ~ "Distance from QB (yards)",
+      name == "slope_distance_from_QB" ~ "Velocity (yards/second)",
+      name == "strain" ~ "STRAIN (1/second)",
+      TRUE ~ as.character(name)
+    ),
+    name_fct = ifelse(str_detect(name, "strain_"), "STRAIN of other rushers", name),
+    name_fct = factor(
+      name_fct,
+      levels = c(
+        "Distance from QB (yards)",
+        "Velocity (yards/second)",
+        "STRAIN (1/second)",
+        "STRAIN of other rushers"
+      )
+    )
+  ) |> 
+  ggplot() +
+  geom_line(aes(frameId_snap_corrected, value, color = name_fct, alpha = name_fct, group = name), linewidth = 1.2) +
+  labs(x = "Time since snap (seconds)",
+       y = "Feature value") +
+  scale_x_continuous(breaks = seq(0, 40, 10), labels = 0:4) +
+  scale_color_manual(values = c("red", "#FFCC33" , "#1143E2", "#CDCDCD"),
+                     guide = guide_legend(order = 1)) +
+  scale_alpha_manual(values = c(1, 1, 1, 0.4),
+                     guide = guide_legend(order = 1)) +
+  labs(color = NULL,
+       alpha = NULL) +
+  theme_light() +
+  theme(axis.title = element_text(size = rel(1)),
+        axis.text = element_text(size = rel(0.8)),
+        legend.text = element_text(size = rel(0.8)),
+        legend.title = element_text(size = rel(1)))

### 2023-03-10 scripts/fig_field.R

@@ -1,0 +1,108 @@
+library(tidyverse)
+library(here)
+
+# df_cleaned <- read_csv(here("data", "df_cleaned.csv.zip"))
+
+# play_lv_den <- df_cleaned |> 
+#   filter(gameId == 2021101709 & playId == 1444)
+# write_rds(play_lv_den, here("data", "play_lv_den.rds"))
+
+play_lv_den <- read_rds(here("data", "play_lv_den.rds"))
+
+plays <- read_csv(here("data", "plays.csv"))
+desc <- plays |>
+  filter(gameId == 2021101709 & playId == 1444) |>
+  pull(playDescription) |> 
+  str_replace("\\)\\.", "\\)")
+
+lv_strain_range <- play_lv_den |> 
+  filter(team == "LV") |> 
+  pull(strain) |> 
+  range()
+
+play_lv_den <- play_lv_den |>
+  mutate(point_size = ifelse(
+    team != "LV",
+    1.5,
+    scales::rescale(strain, to = c(1.5, 7.5), from = lv_strain_range)
+  ))
+
+plot_field <- function(frame) {
+  plot_df <- filter(play_lv_den, frameId_snap_corrected == frame)
+  field <- ggplot() + 
+    annotate("text", 
+             x = seq(40, 80, 10),
+             y = 12,
+             color = "#bebebe",
+             size = rel(4),
+             label = 10 * c(3:5, 4:3)) +
+    annotate("text", 
+             x = seq(40, 80, 10),
+             y = 38.35,
+             color = "#bebebe",
+             size = rel(4),
+             label = 10 * c(3:5, 4:3)) +
+    annotate("text", 
+             x = setdiff(seq(35, 85, 1), seq(35, 85, 5)),
+             y = 0,
+             color = "#bebebe",
+             label = "|",
+             vjust = 1.5) +
+    annotate("text", 
+             x = setdiff(seq(35, 85, 1), seq(35, 85, 5)),
+             y = 48.7,
+             color = "#bebebe",
+             label = "∣",
+             vjust = -0.5) +
+    annotate("text", 
+             x = setdiff(seq(35, 85, 1), seq(35, 85, 5)),
+             y = 23.36667,
+             color = "#bebebe",
+             label = "|",
+             size = rel(2)) +
+    annotate("text", 
+             x = setdiff(seq(35, 85, 1), seq(35, 85, 5)),
+             y = 29.96667,
+             color = "#bebebe",
+             label = "|",
[…]

### 2023-03-10 scripts/fig_pos_curves.R

@@ -1,0 +1,45 @@
+library(tidyverse)
+library(here)
+
+# df_cleaned <- read_csv(here("data", "df_cleaned.csv.zip"))
+
+pff <- read_csv(here("data", "pffScoutingData.csv"))
+
+# pass_rush_pos <- df_cleaned |> 
+#   left_join(pff) |>
+#   filter(pff_role == "Pass Rush")
+# write_rds(pass_rush_pos, here("data", "pass_rush_pos.rds"))
+
+pass_rush_pos <- read_rds(here("data", "pass_rush_pos.rds"))
+
+avg_strain_all <- pass_rush_pos |> 
+  filter(team == defensiveTeam,
+         frameId_snap_corrected <= 40) |> 
+  group_by(frameId_snap_corrected) |> 
+  summarise(mn = mean(strain, na.rm = TRUE))
+
+
+fig_pos_curves <- pass_rush_pos |> 
+  filter(team == defensiveTeam,
+         officialPosition %in% c("OLB", "NT", "DE", "DT"),
+         frameId_snap_corrected <= 40) |> 
+  group_by(officialPosition, frameId_snap_corrected) |> 
+  summarise(mn = mean(strain, na.rm = TRUE)) |> 
+  bind_rows(mutate(avg_strain_all, officialPosition = "Average")) |>
+  mutate(officialPosition = factor(officialPosition, levels = c("OLB", "DE", "DT", "NT", "Average"))) |> 
+  ggplot(aes(frameId_snap_corrected, mn, 
+             color = officialPosition, 
+             group = officialPosition, 
+             linetype = officialPosition)) +
+  geom_smooth(se = FALSE, span = 0.3, linewidth = 1) +
+  scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107", "#004D40", "gray"), name = "Position") +
+  scale_x_continuous(breaks = seq(0, 40, 10), labels = 0:4) +
+  scale_linetype_manual(values = c(rep("solid", 4), "dotted"), name = "Position") +
+  labs(y = "STRAIN",
+       x = "Time since snap (seconds)") +
+  theme_light() +
+  theme(legend.key.width = unit(0.9, "cm"),
+        axis.title = element_text(size = rel(1)),
+        axis.text = element_text(size = rel(0.8)),
+        legend.title = element_text(size = rel(1)),
+        legend.text = element_text(size = rel(0.8)))

### 2023-03-10 scripts/fig_stability.R

@@ -1,0 +1,55 @@
+games <- read_csv(here("data", "games.csv"))
+source(here("scripts", "fig_cor_pressure.R"))
+
+min_100_snaps <- test |> 
+  select(nflId, team, displayName, officialPosition) |> 
+  distinct() |> 
+  right_join(test_summed) |> 
+  filter(n_plays >= 100 & !is.infinite(strain)) |> 
+  pull(displayName)
+
+first <- test |>
+  left_join(select(games, gameId, week)) |> 
+  filter(displayName %in% min_100_snaps, week %in% 1:4, !is.infinite(strain)) |> 
+  group_by(nflId) |>
+  summarize(
+    strain = sum(strain),
+    name = head(displayName, 1),
+    pos = head(officialPosition, 1),
+    n = n()
+  ) |>
+  mutate(strain_rate = 10 * strain / n) |> 
+  filter(pos %in% c("OLB", "DE", "DT", "NT")) |> 
+  select(name, pos, first_strain_rate = strain_rate)
+
+last <- test |>
+  left_join(select(games, gameId, week)) |> 
+  filter(displayName %in% min_100_snaps, week %in% 5:8, !is.infinite(strain)) |> 
+  group_by(nflId) |>
+  summarize(
+    strain = sum(strain),
+    name = head(displayName, 1),
+    pos = head(officialPosition, 1),
+    n = n()
+  ) |>
+  mutate(strain_rate = 10 * strain / n) |> 
+  filter(pos %in% c("OLB", "DE", "DT", "NT")) |> 
+  arrange(-strain_rate) |> 
+  select(name, pos, last_strain_rate = strain_rate)
+
+fig_stability <- first |> 
+  full_join(last) |>
+  mutate(pos = factor(pos, levels = c("OLB", "DE", "DT", "NT"))) |> 
+  ggplot(aes(first_strain_rate, last_strain_rate)) +
+  geom_smooth(method = "lm", se = FALSE, linewidth = 1.5, color = "black", alpha = 0.5) +
+  geom_point(aes(color = pos, group = name), size = 2, alpha = 0.8) +
+  scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107", "#004D40")) +
+  labs(x = "Average STRAIN in first 4 weeks",
+       y = "Average STRAIN in last 4 weeks",
+       color = "Position") +
+    theme_light() +
+    theme(axis.title = element_text(size = rel(1)),
+          legend.title = element_text(size = rel(1)),
[…]

### 2023-03-10 scripts/fig_cor_pressure.R
dealt with large files

@@ -1,15 +1,15 @@
 library(tidyverse)
 library(here)
 
-pass_rush_pos <- read_rds(here("data", "pass_rush_pos.rds"))
-hsh <- pass_rush_pos |>
+pass_rush <- read_csv(here("data", "pass_rush.csv.gz"))
+hsh <- pass_rush |>
   distinct(nflId, gameId, pff_hurry, pff_sack, pff_hit) |>
   group_by(nflId) |>
   summarize(hurries = sum(pff_hurry, na.rm = TRUE),
             sacks = sum(pff_sack, na.rm = TRUE),
             hits = sum(pff_hit, na.rm = TRUE))
 
-test <- pass_rush_pos |> 
+test <- pass_rush |> 
   filter(team == defensiveTeam & !is.na(strain))
 test$strain <- test$slope_distance_from_QB/test$distance_from_QB
 test_summed <- test |> 
@@ -36,7 +36,8 @@ fig_cor_pressure <- test_summed |>
   ggplot(aes(strain_rate, pres)) +
   geom_smooth(method = "lm", se = FALSE, linewidth = 1.5, color = "black", alpha = 0.5) +
   geom_point(aes(color = pos, group = name), size = 2, alpha = 0.8) +
-  expand_limits(x = 3, y = c(0, 0.12)) +
+  expand_limits(x = 2.7, y = c(0.02, 0.12)) +
+  scale_x_continuous(breaks = seq(0.5, 2.5, 1)) +
   scale_y_continuous(breaks = seq(0, 0.12, 0.04)) +
   scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107", "#004D40")) +
   labs(y = "(Hurries + Sacks + Hits) per snaps",
@@ -48,6 +49,3 @@ fig_cor_pressure <- test_summed |>
         legend.text = element_text(size = rel(0.8)),
         axis.text = element_text(size = rel(0.8)),
         panel.grid.minor = element_blank())
-
-fig_cor_pressure
-

### 2023-03-10 scripts/fig_field.R

@@ -1,13 +1,21 @@
 library(tidyverse)
 library(here)
 
-# df_cleaned <- read_csv(here("data", "df_cleaned.csv.zip"))
+# We have a cleaned data file from BDB
+# This data file is too large to push to github
+# So I stored it on dropbox (as a zipped file)
+# Download and save it in the "data" folder
+# https://www.dropbox.com/s/haojedm4kgq3atu/df_cleaned.csv.gz?dl=1
+# What's great about readr::read_csv() is that
+# it allows direct importing of .csv.gz
 
-# play_lv_den <- df_cleaned |> 
+# df_cleaned <- read_csv(here("data", "df_cleaned.csv.gz"))
+
+# play_lv_den <- df_cleaned |>
 #   filter(gameId == 2021101709 & playId == 1444)
-# write_rds(play_lv_den, here("data", "play_lv_den.rds"))
+# write_csv(play_lv_den, here("data", "play_lv_den.csv.gz"))
 
-play_lv_den <- read_rds(here("data", "play_lv_den.rds"))
+play_lv_den <- read_csv(here("data", "play_lv_den.csv.gz"))
 
 plays <- read_csv(here("data", "plays.csv"))
 desc <- plays |>
@@ -106,3 +114,4 @@ fig_field <- plot_grid(
   ncol = 1
 )
 
+fig_field

### 2023-03-10 scripts/fig_pos_curves.R

@@ -1,25 +1,25 @@
 library(tidyverse)
 library(here)
 
-# df_cleaned <- read_csv(here("data", "df_cleaned.csv.zip"))
+# df_cleaned <- read_csv(here("data", "df_cleaned.csv.gz"))
 
 pff <- read_csv(here("data", "pffScoutingData.csv"))
 
-# pass_rush_pos <- df_cleaned |> 
+# pass_rush <- df_cleaned |> 
 #   left_join(pff) |>
 #   filter(pff_role == "Pass Rush")
-# write_rds(pass_rush_pos, here("data", "pass_rush_pos.rds"))
+# write_csv(pass_rush, here("data", "pass_rush.csv.gz"))
 
-pass_rush_pos <- read_rds(here("data", "pass_rush_pos.rds"))
+pass_rush <- read_csv(here("data", "pass_rush.csv.gz"))
 
-avg_strain_all <- pass_rush_pos |> 
+avg_strain_all <- pass_rush |> 
   filter(team == defensiveTeam,
          frameId_snap_corrected <= 40) |> 
   group_by(frameId_snap_corrected) |> 
   summarise(mn = mean(strain, na.rm = TRUE))
 
 
-fig_pos_curves <- pass_rush_pos |> 
+fig_pos_curves <- pass_rush |> 
   filter(team == defensiveTeam,
          officialPosition %in% c("OLB", "NT", "DE", "DT"),
          frameId_snap_corrected <= 40) |> 

### 2023-03-10 paper/paper.Rmd
renamed files

--- 2023-03-09 paper/paper.Rmd
@@
 
+```{r}
+#| label: fig_field
+#| fig.cap: "snapshots"
+#| fig.width: 7
+#| fig.height: 9
+source(here("scripts", "01_fig_field.R"))
+fig_field
+```
+
+Figure \ref{fig:fig_field} shows snapshots
+
+```{r}
+#| label: fig_crosby_curves
+#| fig.cap: "crosby"
+#| fig.width: 5
+#| fig.height: 3
+source(here("scripts", "02_fig_crosby_curves.R"))
+fig_crosby_curves
+```
+
+Figure \ref{fig:fig_crosby_curves} shows Crosby throughout the play
+
 ## Positional STRAIN curves
+
+```{r}
+#| label: fig_pos_curves
+#| fig.cap: "STRAIN curves for different positions"
+#| fig.width: 4.5
+#| fig.height: 3
+source(here("scripts", "03_fig_pos_curves.R"))
+fig_pos_curves
+```
+
+Figure \ref{fig:fig_pos_curves} shows pos curves
 
@@
 ## Corr with pressure rate = (hits + sacks + hurries) per snap
+
+```{r}
+#| label: fig_cor_pressure
+#| fig.cap: "Cor with"
+#| fig.width: 4
+#| fig.height: 3
+source(here("scripts", "04_fig_cor_pressure.R"))
+fig_cor_pressure
+```
+
+Figure \ref{fig:fig_cor_pressure} shows cor with pressure
+
+Fairly strong correlation (r = 0.6254943) between average STRAIN & hurries + sacks + hits per snaps
 
@@
 
+```{r}
+#| label: fig_stability
+#| fig.cap: "Metric stability"
+#| fig.width: 4
+#| fig.height: 3
+source(here("scripts", "05_fig_stability.R"))
+fig_stability
+```
+
+0.8544865
+
+Figure \ref{fig:fig_stability} shows
+
+Average STRAIN across all frames played in first and last 4 weeks
+
+Strong correlation shows great metric stability
+
 # Discussion

### 2023-03-10 scripts/00_data_prep.R

@@ -1,0 +1,154 @@
+# data from BDB
+# https://www.kaggle.com/competitions/nfl-big-data-bowl-2023/data
+# save them in the "data" folder
+
+# a lot of base R from Greg here
+
+library(tidyverse)
+library(here)
+games <- read_csv(here("data", "games.csv"))
+plays <- read_csv(here("data", "plays.csv"))
+players <- read_csv(here("data", "players.csv"))
+
+#This play is a duplicate.
+plays <- plays[!(plays$gameId == 2021110100 &
+                   plays$playId == 1267 & 
+                   plays$absoluteYardlineNumber == 71), ]
+
+week <- list()
+for (i in 1:8) {
+  print(i)
+  week[[i]] <- read_csv(here("data", paste0("week", i, ".csv")))
+  #merge on position
+  week[[i]] <- merge(
+    week[[i]],
+    players[, c("nflId", "officialPosition")],
+    by.x = "nflId",
+    by.y = "nflId",
+    all.x = TRUE
+  )
+  #merge on play outcome
+  week[[i]] <- merge(
+    week[[i]],
+    plays[, c("gameId", "playId", "playResult")],
+    by.x = c("gameId", "playId"),
+    by.y = c("gameId", "playId"),
+    all.x = TRUE
+  )
+  
+  week[[i]]$y[week[[i]]$y > 160/3] <- 160/3
+  week[[i]]$y[week[[i]]$y < 0] <- 0
+  
+  #convert all plays to be moving left
+  week[[i]]$x[week[[i]]$playDirection == "right"] <-
+    120 - week[[i]]$x[week[[i]]$playDirection == "right"]
+  week[[i]]$y[week[[i]]$playDirection == "right"] <-
+    160/3 - week[[i]]$y[week[[i]]$playDirection == "right"]
+  week[[i]]
+}
+
+
+for (i in 1:length(week)) {
+  week[[i]]$week <- i
+}
+
+
+track <- do.call(rbind, week)
+track <-
+  track[!duplicated(paste0(track$gameId, track$playId, track$nflId, track$frameId)), ]
+#track[track$gameId == 2021110100 & track$playId == 1267,]
+
+#Now pull out the QB rows and then merge that on.
+qb <- track[track$officialPosition == "QB", ]
+qb <- qb[!is.na(qb$gameId), ]
[…]

### 2023-03-10 scripts/01_fig_field.R

@@ -50,18 +50,18 @@ plot_field <- function(frame) {
              color = "#bebebe",
              size = rel(4),
              label = 10 * c(3:5, 4:3)) +
-    annotate("text", 
+    annotate("segment", 
              x = setdiff(seq(35, 85, 1), seq(35, 85, 5)),
-             y = 0,
-             color = "#bebebe",
-             label = "|",
-             vjust = 1.5) +
-    annotate("text", 
+             xend = setdiff(seq(35, 85, 1), seq(35, 85, 5)),
+             y = -Inf,
+             yend = 0,
+             color = "#bebebe") +
+    annotate("segment", 
              x = setdiff(seq(35, 85, 1), seq(35, 85, 5)),
+             xend = setdiff(seq(35, 85, 1), seq(35, 85, 5)),
              y = 48.7,
-             color = "#bebebe",
-             label = "∣",
-             vjust = -0.5) +
+             yend = Inf,
+             color = "#bebebe") +
     annotate("text", 
              x = setdiff(seq(35, 85, 1), seq(35, 85, 5)),
              y = 23.36667,
@@ -115,3 +115,4 @@ fig_field <- plot_grid(
 )
 
 fig_field
+

### 2023-03-10 scripts/02_fig_crosby_curves.R

@@ -1,5 +1,5 @@
 library(here)
-source(here("scripts", "fig_field.R"))
+source(here("scripts", "01_fig_field.R"))
 pff <- read_csv(here("data", "pffScoutingData.csv"))
 
 crosby <- play_lv_den |> 

### 2023-03-10 scripts/05_fig_stability.R

@@ -1,5 +1,5 @@
 games <- read_csv(here("data", "games.csv"))
-source(here("scripts", "fig_cor_pressure.R"))
+source(here("scripts", "04_fig_cor_pressure.R"))
 
 min_100_snaps <- test |> 
   select(nflId, team, displayName, officialPosition) |> 

### 2023-03-10 scripts/00_data_prep.R
labeling; notes

@@ -1,3 +1,6 @@
+# FYI: no need to run this script
+# data needed to figures/paper have already been saved
+
 # data from BDB
 # https://www.kaggle.com/competitions/nfl-big-data-bowl-2023/data
 # save them in the "data" folder

### 2023-03-10 paper/00-main.Rmd
separate files for each section

---
title: |
  Introducing STRAIN: A simple and interpretable pass rush metric inspired by materials science
short_title: STRAIN
authors: 
- name: "Author1"
  affiliation: "Address"
  city: "City"
  country: "Country"
  email: "email1\\@example.com"
- name: "Author2"
  affiliation: "Address2"
  city: "City2"
  country: "Country2"
  email: "email2\\@example.com"
- name: "Author3"
  affiliation: "Address3"
  city: "City3"
  country: "Country3"
  email: "email3\\@example.com"
abstract: |
  Abstract goes here
keywords: tracking data, Bornn legacy, too much Canadian
header-includes:
  \usepackage{booktabs}
output: rticles::rss_article
bibliography: bibliography.bib
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
library(here)
```

\newpage

```{r child = "01-introduction.Rmd"}
```

```{r child = "02-methods.Rmd"}
```

```{r child = "03-results.Rmd"}
```

```{r child = "04-discussion.Rmd"}
```

# Acknowledgements {-}

Big data bowl, Mike

# References

### 2023-03-10 paper/01-introduction.Rmd
separate files for each section

--- 2023-03-09 paper/paper.Rmd
@@
 ---
-title: |
-  Introducing STRAIN: A simple and interpretable pass rush metric inspired by materials science
-short_title: STRAIN
-authors: 
-- name: "Author 1"
-  affiliation: "Address"
-  city: "City"
-  country: "Country"
-  email: "email1\\@example.com"
-- name: "Author 2"
-  affiliation: "Address2"
-  city: "City2"
-  country: "Country2"
-  email: "email2\\@example.com"
-abstract: |
-  Abstract goes here
-keywords: tracking data, Bornn legacy, too much Canadian
-bibliography: "bibliography"
 output: rticles::rss_article
-header-includes:
-  \usepackage{setspace}
-  \usepackage{booktabs}
+bibliography: bibliography.bib
 ---
 
-\newpage
-
-# Introduction
+# Introduction {#sec:introduction}
 
@@
 TODO: update citations after they're published
+-->
+
+<!--
+textual citation (\citet): @Baumer2023Big
+parenthetical citation (\citep): [@Baumer2023Big]
 -->
@@
 
+STRAIN is the best etc.
+
 The remainder of this paper is outlined as follows.
 
-# Methods
+Section \ref{sec:methods}
 
-## Data
+Section \ref{sec:results}
 
-2023 bdb [@Howard2023NFL]
-
-first 8 weeks 2021 season
-
-tracking: locations of players and ball, 10 fps,
-
-pre-processing standardize
-
-PFF scouting: pass rush role
-
-sample restriction: remove plays w multiple QB
-
-frames between snap and QB outcome (pass forward, sack)
-
-sample size
-
-## Strain rate in materials science
-
-## Application to pass-rushing in football
-
-# Results
-
-## Example play
-
-## Positional STRAIN curves
-
-## Ranking the best pass-rushers
-
-## Corr with pressure rate = (hits + sacks + hurries) per snap
-
-## Statistical Properties of STRAIN
-
-From @Franks2016Meta (github https://github.com/afranks86/meta-analytics)
-
-Discrimination: Does the metric reliably differentiate between players?
-
-Stability: Does the metric measure a quantity which is stable over time?
-
-Independence: Does the metric provide new information?
-
-# Discussion
-
-# References
+Section \ref{sec:discussion}

### 2023-03-10 paper/02-methods.Rmd
separate files for each section

--- 2023-03-09 paper/paper.Rmd
@@
 ---
-title: |
-  Introducing STRAIN: A simple and interpretable pass rush metric inspired by materials science
-short_title: STRAIN
-authors: 
-- name: "Author 1"
-  affiliation: "Address"
-  city: "City"
-  country: "Country"
-  email: "email1\\@example.com"
-- name: "Author 2"
-  affiliation: "Address2"
-  city: "City2"
-  country: "Country2"
-  email: "email2\\@example.com"
-abstract: |
-  Abstract goes here
-keywords: tracking data, Bornn legacy, too much Canadian
-bibliography: "bibliography"
 output: rticles::rss_article
-header-includes:
-  \usepackage{setspace}
-  \usepackage{booktabs}
+bibliography: bibliography.bib
 ---
 
-\newpage
-
-# Introduction
-
-<!--
-TODO: update citations after they're published
--->
-
-In recent years, tracking data have replaced traditional box-score statistics and play-by-play data as the state-of-the-art tool in sports analytics.
-Numerous sports have been collecting and releasing data on player and ball locations on the playing surface over the course of a game.
-This spatio-temporal multi-resolution source of data have provided exceptional opportunities for researchers to perform advanced studies at a more complex level to deepen the understanding of different sports.
-For complete surveys on how tracking data have transformed sports analytics, see @Macdonald2020Recreating, @Baumer2023Big, and @Kovalchik2023Player.
-
-The National Football League (NFL) introduced their player tracking system known as Next Gen Stats (NGS) in 2016
-BDB
-https://operations.nfl.com/gameday/technology/nfl-next-gen-stats/
-
-Previous research bdb
-this year: linemen
-Importance of D-linemen, pass rush (Maybe talk about role of each position, edge vs interior, defensive formation)
-
-## Previous Pass Rush Metrics
-
-Limitations
-
-\begin{table}
-\caption{A summary of previously-existed pass rush metrics. \label{tab:metrics}}
-\centering
-\begin{tabular}{ll}
-\toprule
-\textbf{Metric} & \textbf{Description}\\
-\midrule
-Sacks & a\\
-Hits & a\\
-Hurries & a\\
-QB Pressures & a\\
-Pass-Rush Productivity (PRP) & a\\
-Time In Pocket (TIP) & a\\
-Pass Rush Win Rate (PRWR) & Rate at which pass rusher beats pass block within 2.5 seconds after snap\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-Table \ref{tab:metrics}
-
-Sacks, hits, hurries, pressures
-
-Officially tracked by the NFL since 1982, the most well known statistic for evaluating defensive linemen on pass rushing plays is the sack, which occurs when a defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass.
-
-https://www.pro-football-reference.com/about/advanced_stats.htm
-
-Pressures: QB pressures (hurries + knockdowns + sack plays). Note that "sack plays" in this instance include both sacks and assists.
-
-Hurried: Times QBs were hurried - defined as a situation when the QB is forced to throw the ball earlier than intended or is chased around/out of the pocket as the result of defensive pressure
-
-Hit: Times QBs were hit - occurs when the QB hits the ground after a throw, not marked on a sack play
-
-PFF's Pass-Rush Productivity (PRP)
-
-https://www.pff.com/news/pro-signature-stats-spotlight-defensive-line
-
-More recently metrics such as Time In Pocket (TIP) and Pass Rush Win Rate (PRWR) have also been proposed and are substantial improvements over the less sophisticated counting statistics.
-
-Pocket Time: Number of seconds the quarterback was in the pocket between the snap of the ball and a pass or the collapse of the pocket
-
-@Burke2018Created
-
-https://www.espn.com/nfl/story/_/id/24892208
-
-2.5 seconds, arbitrary
-
-While PRWR is certainly a step in the right direction, it takes continuous data and converts it to a binary win/loss, which is generally not recommended in statistical analysis
-
-## Contributions of STRAIN
-
-The remainder of this paper is outlined as follows.
-
-# Methods
+# Methods {#sec:methods}
 
@@
 
+math
+
 ## Application to pass-rushing in football
 
-# Results
-
-## Example play
-
-## Positional STRAIN curves
-
-## Ranking the best pass-rushers
-
-## Corr with pressure rate = (hits + sacks + hurries) per snap
-
-## Statistical Properties of STRAIN
-
-From @Franks2016Meta (github https://github.com/afranks86/meta-analytics)
-
-Discrimination: Does the metric reliably differentiate between players?
-
-Stability: Does the metric measure a quantity which is stable over time?
-
-Independence: Does the metric provide new information?
-
-# Discussion
-
-# References
+analogy

### 2023-03-10 paper/03-results.Rmd
separate files for each section

--- 2023-03-10 paper/paper.Rmd
@@
 ---
-title: |
-  Introducing STRAIN: A simple and interpretable pass rush metric inspired by materials science
-short_title: STRAIN
-authors: 
-- name: "Author 1"
-  affiliation: "Address"
-  city: "City"
-  country: "Country"
-  email: "email1\\@example.com"
-- name: "Author 2"
-  affiliation: "Address2"
-  city: "City2"
-  country: "Country2"
-  email: "email2\\@example.com"
-abstract: |
-  Abstract goes here
-keywords: tracking data, Bornn legacy, too much Canadian
-bibliography: "bibliography"
 output: rticles::rss_article
-header-includes:
-  \usepackage{setspace}
-  \usepackage{booktabs}
+bibliography: bibliography.bib
 ---
 
-```{r}
-#| label: setup
-#| include: false
-knitr::opts_chunk$set(
-  echo = FALSE,
-  message = FALSE,
-  warning = FALSE,
-  fig.align = "center"
-)
-library(here)
-```
-
-\newpage
-
-# Introduction
-
-<!--
-TODO: update citations after they're published
--->
-
-In recent years, tracking data have replaced traditional box-score statistics and play-by-play data as the state-of-the-art tool in sports analytics.
-Numerous sports have been collecting and releasing data on player and ball locations on the playing surface over the course of a game.
-This spatio-temporal multi-resolution source of data have provided exceptional opportunities for researchers to perform advanced studies at a more complex level to deepen the understanding of different sports.
-For complete surveys on how tracking data have transformed sports analytics, see @Macdonald2020Recreating, @Baumer2023Big, and @Kovalchik2023Player.
-
-The National Football League (NFL) introduced their player tracking system known as Next Gen Stats (NGS) in 2016
-BDB
-https://operations.nfl.com/gameday/technology/nfl-next-gen-stats/
-
-Previous research bdb
-this year: linemen
-Importance of D-linemen, pass rush (Maybe talk about role of each position, edge vs interior, defensive formation)
-
-## Previous Pass Rush Metrics
-
-Limitations
-
-\begin{table}
-\caption{A summary of previously-existed pass rush metrics. \label{tab:metrics}}
-\centering
-\begin{tabular}{ll}
-\toprule
-\textbf{Metric} & \textbf{Description}\\
-\midrule
-Sacks & a\\
-Hits & a\\
-Hurries & a\\
-QB Pressures & a\\
-Pass-Rush Productivity (PRP) & a\\
-Time In Pocket (TIP) & a\\
-Pass Rush Win Rate (PRWR) & Rate at which pass rusher beats pass block within 2.5 seconds after snap\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-Table \ref{tab:metrics}
-
-Sacks, hits, hurries, pressures
-
-Officially tracked by the NFL since 1982, the most well known statistic for evaluating defensive linemen on pass rushing plays is the sack, which occurs when a defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass.
-
-https://www.pro-football-reference.com/about/advanced_stats.htm
-
-Pressures: QB pressures (hurries + knockdowns + sack plays). Note that "sack plays" in this instance include both sacks and assists.
-
-Hurried: Times QBs were hurried - defined as a situation when the QB is forced to throw the ball earlier than intended or is chased around/out of the pocket as the result of defensive pressure
-
-Hit: Times QBs were hit - occurs when the QB hits the ground after a throw, not marked on a sack play
-
-PFF's Pass-Rush Productivity (PRP)
-
-https://www.pff.com/news/pro-signature-stats-spotlight-defensive-line
-
-More recently metrics such as Time In Pocket (TIP) and Pass Rush Win Rate (PRWR) have also been proposed and are substantial improvements over the less sophisticated counting statistics.
-
-Pocket Time: Number of seconds the quarterback was in the pocket between the snap of the ball and a pass or the collapse of the pocket
-
-@Burke2018Created
-
-https://www.espn.com/nfl/story/_/id/24892208
-
-2.5 seconds, arbitrary
-
-While PRWR is certainly a step in the right direction, it takes continuous data and converts it to a binary win/loss, which is generally not recommended in statistical analysis
-
-## Contributions of STRAIN
-
-The remainder of this paper is outlined as follows.
-
-# Methods
-
-## Data
-
-2023 bdb [@Howard2023NFL]
-
-first 8 weeks 2021 season
-
-tracking: locations of players and ball, 10 fps,
-
-pre-processing standardize
-
-PFF scouting: pass rush role
-
-sample restriction: remove plays w multiple QB
-
-frames between snap and QB outcome (pass forward, sack)
-
-sample size
-
-## Strain rate in materials science
-
-## Application to pass-rushing in football
-
-# Results
+# Results {#sec:results}
 
@@
 
+Do we want gt tables or traditional tex table
+
 ## Corr with pressure rate = (hits + sacks + hurries) per snap
@@
 
+<!--
+
 Discrimination: Does the metric reliably differentiate between players?
 
-Stability: Does the metric measure a quantity which is stable over time?
+Independence: Does the metric provide new information?
 
-Independence: Does the metric provide new information?
+-->
+
+Stability: Does the metric measure a quantity which is stable over time?  (section 5.1 of openWAR has a similar idea)
 
@@
 Strong correlation shows great metric stability
-
-# Discussion
-
-# References

### 2023-03-10 paper/00-main.Rmd
TAS template

@@
 title: |
-  Introducing STRAIN: A simple and interpretable pass rush metric inspired by materials science
-short_title: STRAIN
+  STRAIN: A Continuous-Time Within-Play Statistic for Evaluating Pass Rush in American Football with Tracking Data
+
+# to produce blinded version set to 1
+blinded: 0
+
 authors: 
-- name: "Author1"
-  affiliation: "Address"
-  city: "City"
-  country: "Country"
-  email: "email1\\@example.com"
-- name: "Author2"
-  affiliation: "Address2"
-  city: "City2"
-  country: "Country2"
-  email: "email2\\@example.com"
-- name: "Author3"
-  affiliation: "Address3"
-  city: "City3"
-  country: "Country3"
-  email: "email3\\@example.com"
+- name: Quang Nguyen
+  # thanks: uncomment to add corresponding author's email
+  affiliation: |
+    | Department of Statistics & Data Science
+    | Carnegie Mellon University
+- name: Ronald Yurko
+  affiliation: |
+    | Department of Statistics & Data Science
+    | Carnegie Mellon University
+- name: Gregory J. Matthews
+  affiliation: |
+    | Department of Mathematics and Statistics
+    | Loyola University Chicago
+
+keywords:
+- tracking data
+- football
+- pass rush
+- statistics in sports
+
 abstract: |
-  Abstract goes here
-keywords: tracking data, Bornn legacy, too much Canadian
-header-includes:
-  \usepackage{booktabs}
-output: rticles::rss_article
+  The text of your abstract.  200 or fewer words.
+
 bibliography: bibliography.bib
+output: rticles::asa_article
 ---

### 2023-03-10 paper/01-introduction.Rmd
TAS template

@@
 ---
-output: rticles::rss_article
 bibliography: bibliography.bib
+output: rticles::asa_article
 ---
@@
 \caption{A summary of previously-existed pass rush metrics. \label{tab:metrics}}
-\centering
+\begin{center}
 \begin{tabular}{ll}
-\toprule
-\textbf{Metric} & \textbf{Description}\\
-\midrule
+\textbf{Metric} & \textbf{Description}\\\hline
 Sacks & a\\
@@
 Time In Pocket (TIP) & a\\
-Pass Rush Win Rate (PRWR) & Rate at which pass rusher beats pass block within 2.5 seconds after snap\\
-\bottomrule
+Pass Rush Win Rate (PRWR) & Rate at which pass rusher beats pass block within 2.5 seconds after snap\\\hline
 \end{tabular}
+\end{center}
 \end{table}

### 2023-03-10 paper/03-results.Rmd
TAS template
simplified sub-files; removed headers

@@
----
-output: rticles::rss_article
-bibliography: bibliography.bib
----
-
 # Results {#sec:results}
@@
 #| fig.cap: "snapshots"
-#| fig.width: 7
-#| fig.height: 9
+#| fig.width: 6
+#| fig.height: 8
 source(here("scripts", "01_fig_field.R"))

### 2023-03-11 paper/00-main.Rmd
more work on methods

@@
 - name: Quang Nguyen
-  # thanks: uncomment to add corresponding author's email
+  thanks: Corresponding author's email
   affiliation: |
@@
 
+\bigskip
+\begin{center}
+{\large\bf SUPPLEMENTARY MATERIAL}
+\end{center}
+
+**GitHub repository:** https://github.com/getstrained/intro-strain.
+
+**NFL Big Data Bowl 2023 Data:**: https://www.kaggle.com/competitions/nfl-big-data-bowl-2023/data.
+
 # References

### 2023-03-11 paper/02-methods.Rmd
TAS template
simplified sub-files; removed headers
more work on methods

@@
----
-output: rticles::rss_article
-bibliography: bibliography.bib
----
-
 # Methods {#sec:methods}
@@
 
-2023 bdb [@Howard2023NFL]
-
-first 8 weeks 2021 season
+We rely on Next Gen Stats data provided by the 2023 NFL Big Data Bowl competition [@Howard2023NFL] for our investigation.
+The data span the first eight weeks of the 2021 NFL regular season, consisting of a total of 122 games.
+Also included are general information about the players, games, and plays within each game
+At a more granular level, we are given information on the player and football locations recorded at 10 frames per second for each play.
+Also included are scouting data supplied by Pro Football Focus, which collects player-level information at the play level.
 
@@
 
-math
+In materials science, strain is the deformation of a material from stress, showing the rate of change in a material's length relative to its original length.
+Formally, let $L(t)$ be the distance between any given two points of interest within a material at time $t$, and $L_0$ be the initial distance between those two points.
+The **strain** for a material at time $t$ is defined as 
+\begin{equation}
+s(t)= \frac{L(t)-L_{0}}{L_{0}}.
+\end{equation}
+Notice that this measure is unitless due to being a ratio of two quantities having the same dimension.
+
+According, the **strain rate** of a material measures the change in its deformation with respect to time.
+Mathematically, the strain rate of a material can be expressed as the derivative of its strain.
+That is,
+\begin{equation}
+s'(t) = \frac{ds}{dt} = \frac{v(t)}{L_{0}},
+\end{equation}
+where $v(t)$ is the velocity at which the two points of interest within the material are moving away from or towards each other.
+Whereas strain has no units, the strain rate is measured in inverse of time, usually 1/second.
 
@@
 
-analogy
+Motivated by its scientific definition, we draw a delightful analogy between strain rate and pass-rushing in football.
+Just as strain rate is a measure of deformation in materials science, a pass-rusher's efforts involve the application of deformation against the offensive line, with the ultimate goal of breaking through the protection to reach the quarterback.
+The players can be viewed as "particles" in some material and the defensive "particles" are attempting to exert pressure on the pocket with the aim of compressing and collapsing this pocket around the quarterback.
+
+In order to apply strain rate to measure NFL pass-rusher effectiveness, we make modifications to how this concept is traditionally defined.
+
+Let $(x_{ijt}, y_{ijt})$ be the $(x, y)$ location on the field of player $j = 1, \cdots, J$ at frame $t = 1, \cdots, T_i$ for play $i = 1, \cdots, n$; and $(x^{QB}_{it}, y^{QB}_{it})$ be the $(x, y)$ location of the quarterback at frame $t$ during play $i$.
+
+*   The distance between player $j$ and the quarterback at frame $t$ during play $i$ is 
+\begin{equation}
+f_{d_{ij}}(t) = \sqrt{(x_{ijt} - x^{QB}_{it})^2 + (y_{ijt} - y^{QB}_{it})^2}.
+\end{equation}
+*   The velocity at which player $j$ is moving towards the quarterback at frame $t$ during play $i$ is 
+\begin{equation}
+f'_{d_{ij}}(t) = \frac{df_{d_{ij}}(t)}{dt}.
+\end{equation}
+*   The STRAIN for player $j$ at frame $t$ during play $i$ is 
+\begin{equation}
+\text{STRAIN}_{ij}(t) = \frac{-f'_{d_{ij}}(t)}{f_{d_{ij}}(t)}.
+\end{equation}
+Note that to distinguish our metric from strain and strain rate in materials science, we write it in capital letters (STRAIN) for the remainder of this manuscript.
+
+Recall that based on its materials science property, an increase in strain rate is associated with an increase in the distance between two points.
+In our football setting, the two points of interest are the pass-rusher and the quarterback, and we expect our metric to increase as the distance between the pass-rusher and the quarterback decreases.
+Thus, the negative sign in the numerator of our formula effectively accounts for this. 
+Additionally, rather than keeping the initial distance ($L_0$ as previously denoted) between two points constant over time, we update the initial position to be the player locations at the beginning of each frame.
+This effectively gives us the STRAIN for each frame throughout a play.
+
+Since we only observe the distance and velocity quantities discretely in increments of 10 frames/second, a point estimate for our proposed metric STRAIN for player $j$ at frame $t$ during play $i$ is 
+\begin{equation}
+\widehat{\text{STRAIN}}_{ij}(t) = \cfrac{-\cfrac{f_{d_{ij}}(t) - f_{d_{ij}}(t - 1)}{0.1}}{f_{d_{ij}}(t)}.
+\end{equation}
+Notice that this quantity increases in two ways: 1) the rate at which the rusher is moving towards the quarterback increases, and 2) the distance between the rusher and the quarterback gets smaller.
+Both of these are indications of pass-rusher effectiveness.
+Finally, our statistic STRAIN is measured in inverse second, similar to strain rate.
+In addition, the reciprocal of our metric (1/STRAIN) has a straightforward interpretation.
+It illustrates the amount of time required for the rusher to get to the quarterback at the current location and rate at any given time $t$.

### 2023-03-11 scripts/05_fig_stability.R

@@ -1,3 +1,6 @@
+library(tidyverse)
+library(here)
+
 games <- read_csv(here("data", "games.csv"))
 source(here("scripts", "04_fig_cor_pressure.R"))
 
@@ -8,7 +11,7 @@ min_100_snaps <- test |>
   filter(n_plays >= 100 & !is.infinite(strain)) |> 
   pull(displayName)
 
-first <- test |>
+strain_first <- test |>
   left_join(select(games, gameId, week)) |> 
   filter(displayName %in% min_100_snaps, week %in% 1:4, !is.infinite(strain)) |> 
   group_by(nflId) |>
@@ -22,7 +25,7 @@ first <- test |>
   filter(pos %in% c("OLB", "DE", "DT", "NT")) |> 
   select(name, pos, first_strain_rate = strain_rate)
 
-last <- test |>
+strain_last <- test |>
   left_join(select(games, gameId, week)) |> 
   filter(displayName %in% min_100_snaps, week %in% 5:8, !is.infinite(strain)) |> 
   group_by(nflId) |>
@@ -37,8 +40,8 @@ last <- test |>
   arrange(-strain_rate) |> 
   select(name, pos, last_strain_rate = strain_rate)
 
-fig_stability <- first |> 
-  full_join(last) |>
+fig_stability <- strain_first |> 
+  full_join(strain_last) |>
   mutate(pos = factor(pos, levels = c("OLB", "DE", "DT", "NT"))) |> 
   ggplot(aes(first_strain_rate, last_strain_rate)) +
   geom_smooth(method = "lm", se = FALSE, linewidth = 1.5, color = "black", alpha = 0.5) +
@@ -53,3 +56,4 @@ fig_stability <- first |>
           legend.text = element_text(size = rel(0.8)),
           axis.text = element_text(size = rel(0.8)),
           panel.grid.minor = element_blank())
+

### 2023-03-11 paper/03-results.Rmd
added rankings tables

@@
 
-Do we want gt tables or traditional tex table
+\begin{table}
+\caption{Top 15 edge rushers (with at least 100 snaps played) according to the average STRAIN across all frames played. \label{tab:edge}}
+\centering
+\begin{tabular}{rrlllrr}
+\hline
+ Rank & Player & Team & Position & Snaps & Average STRAIN \\ 
+\hline
+1 & Rashan Gary & GB & OLB & 176 & 2.82 \\ 
+2 & Leonard Floyd & LA & OLB & 185 & 2.80 \\ 
+3 & Justin Houston & BAL & OLB & 132 & 2.78 \\ 
+4 & Myles Garrett & CLE & DE & 197 & 2.75 \\ 
+5 & Von Miller & DEN & OLB & 145 & 2.75 \\ 
+6 & T.J. Watt & PIT & OLB & 147 & 2.71 \\ 
+7 & Yannick Ngakoue & LV & DE & 175 & 2.70 \\ 
+8 & Alex Highsmith & PIT & OLB & 129 & 2.65 \\ 
+9 & Preston Smith & GB & OLB & 124 & 2.61 \\ 
+10 & Randy Gregory & DAL & DE & 134 & 2.58 \\ 
+11 & Joey Bosa & LAC & OLB & 160 & 2.58 \\ 
+12 & Darrell Taylor & SEA & DE & 107 & 2.57 \\ 
+13 & Josh Sweat & PHI & DE & 159 & 2.57 \\ 
+14 & Maxx Crosby & LV & DE & 198 & 2.56 \\ 
+15 & Markus Golden & ARI & OLB & 164 & 2.50 \\ 
+\hline
+\end{tabular}
+\end{table}
+
+\begin{table}
+\caption{Top 15 interior rushers (with at least 100 snaps played) according to the average STRAIN across all frames played. \label{tab:interior}}
+\centering
+\begin{tabular}{rlllrr}
+\hline
+Rank & Player & Team & Position & Snaps & Average STRAIN \\ 
+\hline
+1 & Aaron Donald & LA & DT & 239 & 1.67 \\ 
+2 & Solomon Thomas & LV & DT & 115 & 1.51 \\ 
+3 & Quinton Jefferson & LV & DT & 144 & 1.46 \\ 
+4 & Chris Jones & KC & DT & 139 & 1.42 \\ 
+5 & DeForest Buckner & IND & DT & 198 & 1.26 \\ 
+6 & Cameron Heyward & PIT & DT & 188 & 1.25 \\ 
+7 & Javon Hargrave & PHI & DT & 156 & 1.24 \\ 
+8 & Jerry Tillery & LAC & DT & 171 & 1.16 \\ 
+9 & Ed Oliver & BUF & DT & 133 & 1.15 \\ 
+10 & Osa Odighizuwa & DAL & DT & 162 & 1.13 \\ 
+11 & Greg Gaines & LA & NT & 111 & 1.11 \\ 
+12 & Leonard Williams & NYG & DT & 226 & 1.03 \\ 
+13 & Christian Barmore & NE & DT & 166 & 1.02 \\ 
+14 & Vita Vea & TB & NT & 184 & 1.01 \\ 
+15 & B.J. Hill & CIN & DT & 123 & 0.96 \\ 
+\hline
+\end{tabular}
+\end{table}
+
+\ref{tab:edge}
+
+\ref{tab:interior}
 

### 2023-03-11 scripts/06_rankings.R

@@ -1,0 +1,66 @@
+library(tidyverse)
+library(here)
+
+pass_rush <- read_csv(here("data", "pass_rush.csv.gz"))
+
+test <- pass_rush |> 
+  filter(team == defensiveTeam & !is.na(strain))
+test$strain <- test$slope_distance_from_QB/test$distance_from_QB
+test_summed <- test |> 
+  group_by(nflId) |>
+  summarize(
+    strain = sum(strain),
+    name = head(displayName, 1),
+    pos = head(officialPosition, 1),
+    n = n(),
+    n_plays = length(unique(playId))
+  ) |>
+  mutate(strain_rate = 10 * strain / n) |> 
+  arrange(-strain_rate) |> 
+  ungroup()
+
+info <- test |> 
+  ungroup() |> 
+  select(nflId, team, displayName, officialPosition) |> 
+  distinct()
+
+out <- test_summed |> 
+  left_join(info, by = "nflId") |> 
+  filter(n_plays >= 100 & !is.infinite(strain)) |> 
+  mutate(rank = row_number()) |> 
+  filter(strain_rate > 0) |> 
+  select(rank, displayName, 
+         officialPosition,
+         team, 
+         n_plays, 
+         strain, 
+         strain_rate)
+
+library(xtable)
+# edge 
+edge <- out |>
+  filter(officialPosition %in% c("OLB", "DE")) |> 
+  mutate(rank = row_number()) |> 
+  select(Rank = rank, 
+         Player = displayName, 
+         Team = team, 
+         Position = officialPosition,
+         Snaps = n_plays, 
+         "Average STRAIN" = strain_rate) |> 
+  slice_head(n = 15)
+
+print(xtable(edge), include.rownames = FALSE)
+
+# interior
+interior <- out |>
+  filter(officialPosition %in% c("DT", "NT")) |> 
+  mutate(rank = row_number()) |> 
+  select(Rank = rank, 
+         Player = displayName, 
+         Team = team, 
+         Position = officialPosition,
+         Snaps = n_plays, 
+         "Average STRAIN" = strain_rate) |> 
+  slice_head(n = 15)
+
+print(xtable(interior), include.rownames = FALSE)

### 2023-03-11 paper/02-methods.Rmd
predictability + play outcome curves

@@
 
-## Application to pass-rushing in football
+## Application to pass rush in football
 
-Motivated by its scientific definition, we draw a delightful analogy between strain rate and pass-rushing in football.
-Just as strain rate is a measure of deformation in materials science, a pass-rusher's efforts involve the application of deformation against the offensive line, with the ultimate goal of breaking through the protection to reach the quarterback.
+Motivated by its scientific definition, we draw a delightful analogy between strain rate and pass rushing in football.
+Just as strain rate is a measure of deformation in materials science, a pass rusher's efforts involve the application of deformation against the offensive line, with the ultimate goal of breaking through the protection to reach the quarterback.
 The players can be viewed as "particles" in some material and the defensive "particles" are attempting to exert pressure on the pocket with the aim of compressing and collapsing this pocket around the quarterback.
 
-In order to apply strain rate to measure NFL pass-rusher effectiveness, we make modifications to how this concept is traditionally defined.
+In order to apply strain rate to measure NFL pass rusher effectiveness, we make modifications to how this concept is traditionally defined.
 
-Let $(x_{ijt}, y_{ijt})$ be the $(x, y)$ location on the field of player $j = 1, \cdots, J$ at frame $t = 1, \cdots, T_i$ for play $i = 1, \cdots, n$; and $(x^{QB}_{it}, y^{QB}_{it})$ be the $(x, y)$ location of the quarterback at frame $t$ during play $i$.
+Let $(x_{ijt}, y_{ijt})$ be the $(x, y)$ location on the field of pass rusher $j = 1, \cdots, J$ at frame $t = 1, \cdots, T_i$ for play $i = 1, \cdots, n$; and $(x^{QB}_{it}, y^{QB}_{it})$ be the $(x, y)$ location of the quarterback at frame $t$ during play $i$.
 
-*   The distance between player $j$ and the quarterback at frame $t$ during play $i$ is 
+*   The distance between pass rusher $j$ and the quarterback at frame $t$ during play $i$ is 
 \begin{equation}
@@
 \end{equation}
-*   The velocity at which player $j$ is moving towards the quarterback at frame $t$ during play $i$ is 
+*   The velocity at which pass ruher $j$ is moving towards the quarterback at frame $t$ during play $i$ is 
 \begin{equation}
@@
 \end{equation}
-*   The STRAIN for player $j$ at frame $t$ during play $i$ is 
+*   The STRAIN for pass rusher $j$ at frame $t$ during play $i$ is 
 \begin{equation}
@@
 Recall that based on its materials science property, an increase in strain rate is associated with an increase in the distance between two points.
-In our football setting, the two points of interest are the pass-rusher and the quarterback, and we expect our metric to increase as the distance between the pass-rusher and the quarterback decreases.
+In our football setting, the two points of interest are the pass rusher and the quarterback, and we expect our metric to increase as the distance between the pass rusher and the quarterback decreases.
 Thus, the negative sign in the numerator of our formula effectively accounts for this. 
@@
 
-Since we only observe the distance and velocity quantities discretely in increments of 10 frames/second, a point estimate for our proposed metric STRAIN for player $j$ at frame $t$ during play $i$ is 
+Since we only observe the distance and velocity quantities discretely in increments of 10 frames/second, a point estimate for our proposed metric STRAIN for pass rusher $j$ at frame $t$ during play $i$ is
 \begin{equation}
@@
 Notice that this quantity increases in two ways: 1) the rate at which the rusher is moving towards the quarterback increases, and 2) the distance between the rusher and the quarterback gets smaller.
-Both of these are indications of pass-rusher effectiveness.
+Both of these are indications of pass rusher effectiveness.
 Finally, our statistic STRAIN is measured in inverse second, similar to strain rate.

### 2023-03-11 scripts/07_fig_predictability.R

@@ -1,0 +1,95 @@
+library(tidyverse)
+library(here)
+
+
+games <- read_csv(here("data", "games.csv"))
+players <- read_csv(here("data", "players.csv"))
+source(here("scripts", "05_fig_stability.R"))
+
+
+min_100_snaps_id <- test |> 
+  select(nflId, team, displayName, officialPosition) |> 
+  distinct() |> 
+  right_join(test_summed) |> 
+  filter(n_plays >= 100 & !is.infinite(strain)) |> 
+  pull(nflId)
+
+hsh <- pass_rush |>
+  distinct(nflId, gameId, pff_hurry, pff_sack, pff_hit) |> 
+  left_join(select(games, gameId, week))
+
+
+hsh_first <- hsh |> 
+  filter(nflId %in% min_100_snaps_id, week %in% 1:4) |> 
+  group_by(nflId) |>
+  summarize(hurries = sum(pff_hurry, na.rm = TRUE),
+            sacks = sum(pff_sack, na.rm = TRUE),
+            hits = sum(pff_hit, na.rm = TRUE)) |> 
+  full_join(
+    test |>
+      left_join(select(games, gameId, week)) |> 
+      filter(displayName %in% min_100_snaps, week %in% 1:4) |> 
+      group_by(nflId) |>
+      summarize(n_plays = length(unique(playId)))
+  ) |> 
+  transmute(nflId, first_hsh = (hurries + sacks + sacks) / n_plays)
+
+hsh_last <- hsh |> 
+  filter(nflId %in% min_100_snaps_id, week %in% 5:8) |> 
+  group_by(nflId) |>
+  summarize(hurries = sum(pff_hurry, na.rm = TRUE),
+            sacks = sum(pff_sack, na.rm = TRUE),
+            hits = sum(pff_hit, na.rm = TRUE)) |> 
+  full_join(
+    test |>
+      left_join(select(games, gameId, week)) |> 
+      filter(displayName %in% min_100_snaps, week %in% 5:8) |> 
+      group_by(nflId) |>
+      summarize(n_plays = length(unique(playId)))
+  ) |> 
+  transmute(nflId, last_hsh = (hurries + sacks + sacks) / n_plays)
+
+
+
+#0.09651732
+hsh_first |> 
+  full_join(hsh_last) |>
+  left_join(select(players, nflId, officialPosition)) |>
[…]

### 2023-03-11 scripts/08_fig_outcome_curves.R

@@ -1,0 +1,133 @@
+
+pass_rush
+
+
+pff <- read_csv(here("data", "pffScoutingData.csv"))
+
+outcomes <- pff |> 
+  group_by(gameId, playId, nflId) |> 
+  summarize(is_hit = sum(pff_hit, na.rm = TRUE) > 0,
+            is_hurry = sum(pff_hurry, na.rm = TRUE) > 0 ,
+            is_sack = sum(pff_sack, na.rm = TRUE) > 0)
+
+outcomes |> 
+  count(gameId, playId, nflId, sort = TRUE)
+
+
+hit <- pass_rush |> 
+  inner_join(outcomes |> 
+               filter(is_hit == TRUE)) |> 
+  filter(team == defensiveTeam,
+         officialPosition %in% c("OLB", "NT", "DE", "DT"),
+         frameId_snap_corrected <= 40) |> 
+  group_by(officialPosition, frameId_snap_corrected) |> 
+  summarise(mn = mean(strain, na.rm = TRUE)) |> 
+  mutate(officialPosition = factor(officialPosition, levels = c("OLB", "DE", "DT", "NT"))) |> 
+  ggplot(aes(frameId_snap_corrected, mn, 
+             color = officialPosition, 
+             group = officialPosition)) +
+  geom_smooth(se = FALSE, span = 0.3, linewidth = 1) +
+  scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107", "#004D40"), name = "Position") +
+  scale_x_continuous(breaks = seq(0, 40, 10), labels = 0:4) +
+  scale_linetype_manual(values = c(rep("solid", 4), "dotted"), name = "Position") +
+  expand_limits(y = 2.3) +
+  labs(y = "STRAIN",
+       x = "Time since snap (seconds)",
+       title = "Hit") +
+  theme_light() +
+  theme(legend.key.width = unit(0.9, "cm"),
+        legend.position = "none",
+        axis.title = element_text(size = rel(1)),
+        axis.text = element_text(size = rel(0.8)),
+        legend.title = element_text(size = rel(1)),
+        legend.text = element_text(size = rel(0.8)))
+
+sack <- pass_rush |> 
+  inner_join(outcomes |> 
+               filter(is_sack == TRUE)) |> 
+  filter(team == defensiveTeam,
+         officialPosition %in% c("OLB", "NT", "DE", "DT"),
+         frameId_snap_corrected <= 40) |> 
+  group_by(officialPosition, frameId_snap_corrected) |> 
[…]

### 2023-03-11 scripts/99_scratch.R

@@ -1,0 +1,22 @@
+
+# correlation between weight and average STRAIN
+# 0.-8185309
+test_summed |> 
+  filter(n_plays >= 100) |> 
+  left_join(select(players, nflId, weight)) |>
+  mutate(pos = factor(pos, levels = c("OLB", "DE", "DT", "NT")),
+         class = ifelse(pos == "DT" | pos == "NT", "Interior", "Edge")) |> 
+  ggplot(aes(weight, strain_rate)) +
+  geom_smooth(method = "lm", se = FALSE, linewidth = 1.5, color = "black", alpha = 0.5) +
+  geom_point(aes(color = pos, group = name), size = 2, alpha = 0.8) +
+  scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107", "#004D40")) +
+  labs(y = "Average STRAIN across all frames",
+       x = "Weight (lbs)",
+       color = "Position") +
+  facet_wrap(~ class, scales = "free") +
+  theme_light() +
+  theme(axis.title = element_text(size = rel(1)),
+        legend.title = element_text(size = rel(1)),
+        legend.text = element_text(size = rel(0.8)),
+        axis.text = element_text(size = rel(0.8)),
+        panel.grid.minor = element_blank())

### 2023-03-12 paper/00-main.Rmd
added unsuccess pass rush play

@@
   thanks: Corresponding author's email
-  affiliation: |
-    | Department of Statistics & Data Science
-    | Carnegie Mellon University
+  affiliation: Department of Statistics & Data Science, Carnegie Mellon University
 - name: Ronald Yurko
-  affiliation: |
-    | Department of Statistics & Data Science
-    | Carnegie Mellon University
+  affiliation: Department of Statistics & Data Science, Carnegie Mellon University
 - name: Gregory J. Matthews
-  affiliation: |
-    | Department of Mathematics and Statistics
-    | Loyola University Chicago
+  affiliation: Department of Mathematics and Statistics, Loyola University Chicago
 
@@
 
+# abstract from BDB submission form
+# rewrite this
 abstract: |
-  The text of your abstract.  200 or fewer words.
+  This paper evaluates pass rusher performance in the National Football League.
+We propose STRAIN, a new measure for pass rusher effectiveness, which accounts for the distance between the pass rusher and quarterback and the rate at which this distance is being reduced.
+STRAIN is a simple and interpretable statistic, with higher STRAIN corresponds to greater pass rushing ability.
+Our metric is closely related to the concept of "strain rate" in materials science.
+We present the mathematical motivations for our metric, followed by an illustration with an example play.
+The analysis concludes with a comparison of STRAIN for different positions and our rankings of NFL's best pass-rushers according to our metric.
 
@@
 output: rticles::asa_article
+header-includes:
+  \renewcommand{\arraystretch}{0.9}
 ---

### 2023-03-12 paper/02-methods.Rmd
added unsuccess pass rush play

@@
 
-According, the **strain rate** of a material measures the change in its deformation with respect to time.
+Accordingly, the **strain rate** of a material measures the change in its deformation with respect to time.
 Mathematically, the strain rate of a material can be expressed as the derivative of its strain.
@@
 Notice that this quantity increases in two ways: 1) the rate at which the rusher is moving towards the quarterback increases, and 2) the distance between the rusher and the quarterback gets smaller.
-Both of these are indications of pass rusher effectiveness.
+Both of these are indications of an effective pass rush attempt.
 Finally, our statistic STRAIN is measured in inverse second, similar to strain rate.
-In addition, the reciprocal of our metric (1/STRAIN) has a straightforward interpretation.
-It illustrates the amount of time required for the rusher to get to the quarterback at the current location and rate at any given time $t$.
+In addition, the reciprocal of our metric (1/STRAIN) has a straightforward interpretation:
+it illustrates the amount of time required for the rusher to get to the quarterback at the current location and rate at any given time $t$.

### 2023-03-12 paper/03-results.Rmd
added unsuccess pass rush play

@@
 #| fig.cap: "snapshots"
-#| fig.width: 6
-#| fig.height: 8
+#| fig.width: 6.6
+#| fig.height: 4.4
 source(here("scripts", "01_fig_field.R"))
@@
 #| fig.cap: "crosby"
-#| fig.width: 5
-#| fig.height: 3
+#| fig.width: 4.8
+#| fig.height: 2.7
 source(here("scripts", "02_fig_crosby_curves.R"))
@@
 
-\ref{tab:edge}
+Table \ref{tab:edge}
 
-\ref{tab:interior}
+Table \ref{tab:interior}
 

### 2023-03-12 scripts/01_fig_field.R

@@ -31,8 +31,8 @@ lv_strain_range <- play_lv_den |>
 play_lv_den <- play_lv_den |>
   mutate(point_size = ifelse(
     team != "LV",
-    1.5,
-    scales::rescale(strain, to = c(1.5, 7.5), from = lv_strain_range)
+    0.5,
+    scales::rescale(strain, to = c(0.5, 5), from = lv_strain_range)
   ))
 
 plot_field <- function(frame) {
@@ -42,13 +42,14 @@ plot_field <- function(frame) {
              x = seq(40, 80, 10),
              y = 12,
              color = "#bebebe",
-             size = rel(4),
+             size = rel(3),
              label = 10 * c(3:5, 4:3)) +
     annotate("text", 
              x = seq(40, 80, 10),
              y = 38.35,
              color = "#bebebe",
-             size = rel(4),
+             size = rel(3),
+             angle = 180,
              label = 10 * c(3:5, 4:3)) +
     annotate("segment", 
              x = setdiff(seq(35, 85, 1), seq(35, 85, 5)),
@@ -95,23 +96,24 @@ plot_field <- function(frame) {
                shape = 20,
                data = filter(plot_df, team == "football")) +
     scale_size_identity() +
-    labs(title = str_c(frame/10, " seconds since snap")) +
+    labs(title = str_c(frame/10, "s since snap")) +
     theme_minimal() +
     theme(axis.text = element_blank(),
-          plot.title = element_text(size = rel(1), face = "bold", hjust = 0.5),
+          plot.title = element_text(size = rel(0.8), face = "bold", hjust = 0.5),
           panel.grid = element_blank(),
           axis.title = element_blank(),
-          axis.ticks = element_blank())
+          axis.ticks = element_blank(),
+          plot.margin = unit(c(0.25, 0, 0, -0.25), "cm"))
   return(field)
 }
 
 
 library(cowplot)
 fig_field <- plot_grid(
+  plot_field(10),
   plot_field(20),
   plot_field(30),
-  plot_field(40),
-  ncol = 1
+  plot_field(40)
 )
 
 fig_field

### 2023-03-12 scripts/09_fig_play_unsuccess.R

@@ -1,0 +1,65 @@
+library(tidyverse)
+library(here)
+
+plays <- read_csv(here("data", "plays.csv"))
+players <- read_csv(here("data", "players.csv"))
+pff <- read_csv(here("data", "pffScoutingData.csv"))
+
+desc <- plays |>
+  filter(gameId == 2021101709 & playId == 1113) |>
+  pull(playDescription) |> 
+  str_replace("\\)\\.", "\\)")
+
+# df_cleaned <- read_csv(here("data", "df_cleaned.csv.gz"))
+
+play_unsucess <- df_cleaned |>
+  filter(gameId == 2021101709 & playId == 1113)
+# write_csv(play_unsucess, here("data", "play_unsucess.csv.gz"))
+
+# https://www.nfl.com/videos/teddy-bridgewater-courtland-sutton-barely-miss-out-on-chance-at-62-yard-td-bomb
+
+crosby <- play_unsucess |> 
+  filter(nflId == 47889) |>
+  select(frameId_snap_corrected, distance_from_QB, slope_distance_from_QB, strain)
+
+
+fig_crosby_curves_unsucess <- play_unsucess |> 
+  left_join(pff) |>
+  filter(pff_role == "Pass Rush", nflId != 47889) |> 
+  select(name = displayName, frameId_snap_corrected, value = strain) |> 
+  mutate(name = str_c("strain_", name)) |> 
+  bind_rows(pivot_longer(crosby, !frameId_snap_corrected)) |> 
+  mutate(
+    name = case_when(
+      name == "distance_from_QB" ~ "Distance from QB (yards)",
+      name == "slope_distance_from_QB" ~ "Velocity (yards/second)",
+      name == "strain" ~ "STRAIN (1/second)",
+      TRUE ~ as.character(name)
+    ),
+    name_fct = ifelse(str_detect(name, "strain_"), "STRAIN of other rushers", name),
+    name_fct = factor(
+      name_fct,
+      levels = c(
+        "Distance from QB (yards)",
+        "Velocity (yards/second)",
+        "STRAIN (1/second)",
+        "STRAIN of other rushers"
+      )
+    )
+  ) |> 
+  ggplot() +
+  geom_line(aes(frameId_snap_corrected, value, color = name_fct, alpha = name_fct, group = name), linewidth = 1.2) +
+  labs(x = "Time since snap (seconds)",
+       y = "Feature value") +
+  scale_x_continuous(breaks = seq(0, 40, 10), labels = 0:4) +
[…]

### 2023-03-16 paper/01-introduction.Rmd
simplified sub-files; removed headers
Small updates.  Added some citations.

@@
----
-bibliography: bibliography.bib
-output: rticles::asa_article
----
-
 # Introduction {#sec:introduction}
@@
 @Burke2018Created
+@AlamarGoldner2013: They talk about how there is no data available to measure offensive line performance.
+1259 plays, 20 teams, 2010 season.
+Success of each offensive lineman was evaluated using the best judgement of the observer.
+Collected time in pocket.
+They collected a bunch of binary variables and did this all by HAND!
+Estimate a survival model for each position on the offensive line.
+They don't rank individuals, but rather positions on a team.
+They then use the predicted probability as a covariate in a model to predict completion.
+
+@AlamarGould2008
 

### 2023-03-17 paper/01-introduction.Rmd
Add bdb intro with text about previous work to transition into pass rush metrics

@@
+---
+output:
+  html_document: default
+  pdf_document: default
+---
 # Introduction {#sec:introduction}
@@
 
-The National Football League (NFL) introduced their player tracking system known as Next Gen Stats (NGS) in 2016
-BDB
-https://operations.nfl.com/gameday/technology/nfl-next-gen-stats/
+We contribute to the growing literature of research based on player tracking data in American football.
+The National Football League (NFL) introduced their player tracking system known as [Next Gen Stats (NGS)](https://operations.nfl.com/gameday/technology/nfl-next-gen-stats/) in 2016, using radio frequency identification (RFID) chips placed in players' shoulder pads (and in the ball) to collect data 10 times per second.
+These NGS tracking data contain information about the on-field locations, speeds, and angles of all 22 players on the field (as well as the ball).
+While these data were initially only available for teams, media, and vendors, in December 2018 the NFL launched the inaugural edition of their annual [Big Data Bowl (BDB)](https://operations.nfl.com/gameday/analytics/big-data-bowl/past-big-data-bowl-recaps/) competition.
+
+The first BDB led to several contributions largely focused offensive performance evaluation.
+For example.
+finalists @deshpande2020expected introduced an approach for modeling the hypothetical completional probability of a pass aiding in the evalaution of quarterback (QB) decision making.
+The winners of the inaugural BDB focused on identifying receiver routes via clustering techniques [@chu2020route] and convolutional neural networks [@sterken2019routenet].
+Along with the BDB competition entries, the public release of NGS data allowed researchers to tackle a variety of other problems such as revisiting fourth down decision making [@lopez2020bigger], annotating pass coverage with gaussian mixture models [@dutta2020unsupervised], and introducing a continuous-time framework to estimate within-play value [@yurko2020going].
+Since its inception, the BDB has changed themes each year leading to new insight about evaluating running backs, defensive backs, and special teams performance.
+
+In this work, we present a novel approach to measure the performance of pass rushers (defensive players charging the QB) that was selected as a finalist for the [2023 edition of the NFL BDB](https://www.kaggle.com/competitions/nfl-big-data-bowl-2023), which asked participants to evaluate linemen on passing plays.
+Relative to other aspects of American football, such as QB evaluation [@burke2019deepqb, @reyers2021quarterback], the literature on evaluating pass rushers is scarce.
+Below, we provide a brief overview of existing pass rush metrics.
 

### 2023-03-20 paper/00-main.Rmd
lmer fit

@@
 title: |
-  STRAIN: A Continuous-Time Within-Play Statistic for Evaluating Pass Rush in American Football with Tracking Data
+  | Bearing the Strain: Analyzing Pass Rush in American Football with Player Tracking Data
+
+# Deforming the line:
+# Straining the line:
+# Unleashing the Rush
+# straining for the takedown
+# straining for success
+# straining forward
+# straining for pressure
+# Unleashing the Strain
+# Strain on the Gridiron
+# from strain to gain
+# Bearing the Strain:
+# Straining the trenches:
 
@@
   thanks: Corresponding author's email
-  affiliation: Department of Statistics & Data Science, Carnegie Mellon University
+  affiliation: |
+    | Department of Statistics & Data Science
+    | Carnegie Mellon University
 - name: Ronald Yurko
-  affiliation: Department of Statistics & Data Science, Carnegie Mellon University
+  affiliation: |
+    | Department of Statistics & Data Science
+    | Carnegie Mellon University
 - name: Gregory J. Matthews
-  affiliation: Department of Mathematics and Statistics, Loyola University Chicago
+  affiliation: |
+    | Department of Mathematics and Statistics
+    | Loyola University Chicago
 

### 2023-03-20 scripts/10_multilevel_mod.R

@@ -1,0 +1,176 @@
+library(tidyverse)
+library(here)
+
+# df_cleaned <- read_csv(here("data", "df_cleaned.csv.gz"))
+# 
+# pass_rush_snaps <- df_cleaned |>
+#   left_join(pff) |>
+#   filter(pff_role == "Pass Rush") |> 
+#   group_by(nflId) |>
+#   summarize(n_plays = n_distinct(playId))
+# write_rds(pass_rush_snaps, here("data", "pass_rush_snaps.rds"))
+# 
+# pass_snaps <- df_cleaned |>
+#   left_join(pff) |>
+#   filter(pff_role == "Pass") |> 
+#   group_by(nflId) |>
+#   summarize(n_plays = n_distinct(playId))
+# write_rds(pass_snaps, here("data", "pass_snaps.rds"))
+# 
+# pass_block_snaps <- df_cleaned |>
+#   left_join(pff) |>
+#   filter(pff_role == "Pass Block") |> 
+#   group_by(nflId) |>
+#   summarize(n_plays = n_distinct(playId))
+# write_rds(pass_block_snaps, here("data", "pass_block_snaps.rds"))
+
+plays <- read_csv(here("data", "plays.csv"))
+players <- read_csv(here("data", "players.csv"))
+pff <- read_csv(here("data", "pffScoutingData.csv"))
+
+# snap counts
+pass_rush_snaps <- read_rds(here("data", "pass_rush_snaps.rds"))
+pass_snaps <- read_rds(here("data", "pass_snaps.rds"))
+pass_block_snaps <- read_rds(here("data", "pass_block_snaps.rds"))
+
+# df_cleaned |>
+#   filter(event == "ball_snap") |>
+#   write_csv(here("data", "ball_snap.csv.gz"))
+
+ball_snap <- read_csv(here("data", "ball_snap.csv.gz"))
+
+pass_rush <- read_csv(here("data", "pass_rush.csv.gz"))
+
+# check distribution of average strain
+# 6 obs with avg strain > 2, else pretty normal
+get_avg_strain <- pass_rush |> 
+  filter(team == defensiveTeam, !is.infinite(strain)) |> 
+  group_by(gameId, playId, nflId) |> 
+  summarize(avg_strain = mean(strain, na.rm = TRUE)) |>
+  ungroup()
+
+get_avg_strain |> 
+  ggplot(aes(avg_strain)) +
+  geom_histogram()
+
[…]

### 2023-03-20 paper/02-data.Rmd
created data section

--- 2023-03-11 paper/02-methods.Rmd
@@
-# Methods {#sec:methods}
-
-## Data
+# Data {#sec:data}
 
@@
 sample size
-
-## Strain rate in materials science
-
-In materials science, strain is the deformation of a material from stress, showing the rate of change in a material's length relative to its original length.
-Formally, let $L(t)$ be the distance between any given two points of interest within a material at time $t$, and $L_0$ be the initial distance between those two points.
-The **strain** for a material at time $t$ is defined as 
-\begin{equation}
-s(t)= \frac{L(t)-L_{0}}{L_{0}}.
-\end{equation}
-Notice that this measure is unitless due to being a ratio of two quantities having the same dimension.
-
-According, the **strain rate** of a material measures the change in its deformation with respect to time.
-Mathematically, the strain rate of a material can be expressed as the derivative of its strain.
-That is,
-\begin{equation}
-s'(t) = \frac{ds}{dt} = \frac{v(t)}{L_{0}},
-\end{equation}
-where $v(t)$ is the velocity at which the two points of interest within the material are moving away from or towards each other.
-Whereas strain has no units, the strain rate is measured in inverse of time, usually 1/second.
-
-## Application to pass-rushing in football
-
-Motivated by its scientific definition, we draw a delightful analogy between strain rate and pass-rushing in football.
-Just as strain rate is a measure of deformation in materials science, a pass-rusher's efforts involve the application of deformation against the offensive line, with the ultimate goal of breaking through the protection to reach the quarterback.
-The players can be viewed as "particles" in some material and the defensive "particles" are attempting to exert pressure on the pocket with the aim of compressing and collapsing this pocket around the quarterback.
-
-In order to apply strain rate to measure NFL pass-rusher effectiveness, we make modifications to how this concept is traditionally defined.
-
-Let $(x_{ijt}, y_{ijt})$ be the $(x, y)$ location on the field of player $j = 1, \cdots, J$ at frame $t = 1, \cdots, T_i$ for play $i = 1, \cdots, n$; and $(x^{QB}_{it}, y^{QB}_{it})$ be the $(x, y)$ location of the quarterback at frame $t$ during play $i$.
-
-*   The distance between player $j$ and the quarterback at frame $t$ during play $i$ is 
-\begin{equation}
-f_{d_{ij}}(t) = \sqrt{(x_{ijt} - x^{QB}_{it})^2 + (y_{ijt} - y^{QB}_{it})^2}.
-\end{equation}
-*   The velocity at which player $j$ is moving towards the quarterback at frame $t$ during play $i$ is 
-\begin{equation}
-f'_{d_{ij}}(t) = \frac{df_{d_{ij}}(t)}{dt}.
-\end{equation}
-*   The STRAIN for player $j$ at frame $t$ during play $i$ is 
-\begin{equation}
-\text{STRAIN}_{ij}(t) = \frac{-f'_{d_{ij}}(t)}{f_{d_{ij}}(t)}.
-\end{equation}
-Note that to distinguish our metric from strain and strain rate in materials science, we write it in capital letters (STRAIN) for the remainder of this manuscript.
-
-Recall that based on its materials science property, an increase in strain rate is associated with an increase in the distance between two points.
-In our football setting, the two points of interest are the pass-rusher and the quarterback, and we expect our metric to increase as the distance between the pass-rusher and the quarterback decreases.
-Thus, the negative sign in the numerator of our formula effectively accounts for this. 
-Additionally, rather than keeping the initial distance ($L_0$ as previously denoted) between two points constant over time, we update the initial position to be the player locations at the beginning of each frame.
-This effectively gives us the STRAIN for each frame throughout a play.
-
-Since we only observe the distance and velocity quantities discretely in increments of 10 frames/second, a point estimate for our proposed metric STRAIN for player $j$ at frame $t$ during play $i$ is 
-\begin{equation}
-\widehat{\text{STRAIN}}_{ij}(t) = \cfrac{-\cfrac{f_{d_{ij}}(t) - f_{d_{ij}}(t - 1)}{0.1}}{f_{d_{ij}}(t)}.
-\end{equation}
-Notice that this quantity increases in two ways: 1) the rate at which the rusher is moving towards the quarterback increases, and 2) the distance between the rusher and the quarterback gets smaller.
-Both of these are indications of pass-rusher effectiveness.
-Finally, our statistic STRAIN is measured in inverse second, similar to strain rate.
-In addition, the reciprocal of our metric (1/STRAIN) has a straightforward interpretation.
-It illustrates the amount of time required for the rusher to get to the quarterback at the current location and rate at any given time $t$.

### 2023-03-20 paper/03-methods.Rmd
created data section

--- 2023-03-12 paper/02-methods.Rmd
@@
 # Methods {#sec:methods}
-
-## Data
-
-We rely on Next Gen Stats data provided by the 2023 NFL Big Data Bowl competition [@Howard2023NFL] for our investigation.
-The data span the first eight weeks of the 2021 NFL regular season, consisting of a total of 122 games.
-Also included are general information about the players, games, and plays within each game
-At a more granular level, we are given information on the player and football locations recorded at 10 frames per second for each play.
-Also included are scouting data supplied by Pro Football Focus, which collects player-level information at the play level.
-
-tracking: locations of players and ball, 10 fps,
-
-pre-processing standardize
-
-PFF scouting: pass rush role
-
-sample restriction: remove plays w multiple QB
-
-frames between snap and QB outcome (pass forward, sack)
-
-sample size
 
@@
 it illustrates the amount of time required for the rusher to get to the quarterback at the current location and rate at any given time $t$.
+
+## Average STRAIN
+
+## Multilevel Model

### 2023-03-20 paper/04-results.Rmd
created data section

--- 2023-03-12 paper/03-results.Rmd
@@
 #| label: fig_field
-#| fig.cap: "snapshots"
+#| fig.cap: "A display of the player tracking data for a play during the Las Vegas Raiders (defense, in black) versus Denver Broncos (offense, in orange) NFL game on October 17, 2021.
+Raiders DE Maxx Crosby is highlighted in blue.
+For each Raiders defender, the point size indicates their individual STRAIN value, with larger points suggesting larger STRAIN.
+Snapshots are captured at 1, 2, 3, and 4 seconds after the ball snap."
 #| fig.width: 6.6
@@
 ```
-
-Figure \ref{fig:fig_field} shows snapshots
 
@@
 
-Figure \ref{fig:fig_crosby_curves} shows Crosby throughout the play
+To illustrate our proposed metric STRAIN for pass rush evaluation, we use a play from the 2021 NFL regular season week 6 matchup between the Las Vegas Raiders and Denver Broncos, which ended with Broncos QB Teddy Bridgewater getting sacked by Raiders DE Maxx Crosby[^crosbysack].
+Figure \ref{fig:fig_field} presents the locations of every Las Vegas (black) and Denver (orange) player on the field at particular time points in this play, with our focused player Maxx Crosby highlighted in blue.
+Note that the point size for each Las Vegas defender corresponds to the estimated STRAIN in the selected frames as the play progresses.
+In addition, Figure \ref{fig:fig_field} is accompanied by Figure \ref{fig:fig_crosby_curves}, which is a line graph showing how Crosby's distance from the quarterback, velocity, and STRAIN change continuously throughout the play.
+
+[^crosbysack]: An real-game footage of this play is available at https://www.raiders.com/video/de-maxx-crosby-sacks-qb-teddy-bridgewater-for-a-loss-of-6-yards-nfl
+
+We observe that for the first two seconds, Crosby is being blocked by a Denver offensive lineman and unable to get close to the quarterback, hence the corresponding STRAIN values are virtually zero. 
+Suddenly at around three seconds after the snap, the STRAIN for Crosby starts to increase after the Raiders DE is freed up and charges towards Bridgewater. 
+At 4 seconds after the snap, Crosby's STRAIN is 2.30, which means at his current moving rate, it will take Crosby about 0.43 (1/2.30) seconds to make the distance between him and the quarterback 0 (i.e.
+essentially sack the quarterback).
+This matches well with the final outcome of the play, as the sack takes place at the very last frame (4.4 seconds) where the estimated STRAIN for Crosby reaches its peak at 3.96.
+
+Moreover, Figure \ref{fig:fig_crosby_curves} clearly demonstrates the interactions between the features for Maxx Crosby.
+Here, a higher STRAIN generally corresponds to faster moving rate towards the quarterback.
+STRAIN also increases as the distance between the pass rusher and quarterback is being reduced.
+Both of these relationships suggest an overall successful pass rush by Maxx Crosby.
+
+(bring in the unsuccessful pass rush play)
+
+(talk a bit about other rushers, maybe connect this with positional curves broken down by outcome - sack.)
 
@@
 ## Ranking the best pass-rushers
+
+Since STRAIN is observed continuously for every play in our data, this allows us to aggregate across all frames played and compute the average STRAIN for NFL pass rushers over the course of the eight-week sample size.
+
+(maybe some math)
+
+Based on the clearly distinct patterns for different positions as previously observed, we evaluate interior pass-rushers (NT's and DT's) separately from edge rushers (OLB's and DE's).
+Tables \ref{tab:edge} and \ref{tab:interior} are leaderboards for the NFL's best edge and interior rushers (with at least 100 plays) rated by the average STRAIN across all frames for the first eight weeks of the 2021 regular season.
+
+Our results show many true positives with no surprises.
+Myles Garrett and TJ Watt are known as top-tier edge rushers, while Aaron Donald is undoubtedly the best interior defender in football.
+Rashan Gary, a young rusher with huge potential, ranks first among all pass-rushers according to our metric.
+Our leaderboards largely match the rankings of experts in the field (for instance, PFF's edge and interior rusher rankings after the 2021 season), lending credibility to our proposed metric as a measure of pass-rushing effectiveness.
+
+https://www.pff.com/news/nfl-2022-edge-rusher-rankings-tiers
+
+https://www.pff.com/news/nfl-interior-defensive-line-rankings-tiers-2022
 
@@
 
-Table \ref{tab:edge}
-
-Table \ref{tab:interior}
-
 ## Corr with pressure rate = (hits + sacks + hurries) per snap

### 2023-03-20 scripts/10_multilevel_mod.R

@@ -141,7 +141,7 @@ tibble(nflId = as.double(rownames(fit_out_rush)),
   arrange(-intercept) |> 
   left_join(players) |> 
   filter(nflId %in% filter(pass_rush_snaps, n_plays >= 100)$nflId) |> 
-  filter(officialPosition == "DT")
+  filter(officialPosition == "OLB")
 
 
 fit_out_qb <- strain_fit |> 
@@ -156,10 +156,12 @@ get_passer_id <- players |>
 
 tibble(passer = rownames(fit_out_qb), 
        intercept = fit_out_qb$`(Intercept)`) |>
-  arrange(intercept) |> 
+  arrange(-intercept) |> 
   left_join(get_passer_id) |> 
-  filter(nflId %in% filter(pass_snaps, n_plays >= 100)$nflId)
+  filter(nflId %in% filter(pass_snaps, n_plays >= 100)$nflId) |> 
+  View()
   
+# https://cran.r-project.org/web/packages/merTools/index.html
 
 fit_out_block <- strain_fit |> 
   ranef() |> 
@@ -170,7 +172,11 @@ tibble(nflId = as.double(rownames(fit_out_block)),
   arrange(intercept) |> 
   left_join(players) |> 
   filter(nflId %in% filter(pass_block_snaps, n_plays >= 100)$nflId) |> 
-  filter(officialPosition == "C")
+  filter(officialPosition == "G")
+
+
+# fixed eff
+# var partition
 
 
 # some plays are missing - Greg?

### 2023-03-22 paper/01-introduction.Rmd
created data section
Modify into to remove finalist mention

@@
 
-In recent years, tracking data have replaced traditional box-score statistics and play-by-play data as the state-of-the-art tool in sports analytics.
-Numerous sports have been collecting and releasing data on player and ball locations on the playing surface over the course of a game.
+In recent years, tracking data have replaced traditional box-score statistics and play-by-play data as the state-of-the-art tool in sports analytics.Numerous sports have been collecting and releasing data on player and ball locations on the playing surface over the course of a game.
 This spatio-temporal multi-resolution source of data have provided exceptional opportunities for researchers to perform advanced studies at a more complex level to deepen the understanding of different sports.
@@
 Along with the BDB competition entries, the public release of NGS data allowed researchers to tackle a variety of other problems such as revisiting fourth down decision making [@lopez2020bigger], annotating pass coverage with gaussian mixture models [@dutta2020unsupervised], and introducing a continuous-time framework to estimate within-play value [@yurko2020going].
-Since its inception, the BDB has changed themes each year leading to new insight about evaluating running backs, defensive backs, and special teams performance.
+Since its inception, the BDB has changed themes each year leading to new insight about evaluating different positions such as running backs, defensive backs, and special teams.
+The [2023 edition of the NFL BDB](https://www.kaggle.com/competitions/nfl-big-data-bowl-2023) continued this revolving theme by asking participants to evaluate lineman on passing plays.
 
-In this work, we present a novel approach to measure the performance of pass rushers (defensive players charging the QB) that was selected as a finalist for the [2023 edition of the NFL BDB](https://www.kaggle.com/competitions/nfl-big-data-bowl-2023), which asked participants to evaluate linemen on passing plays.
+In this work, using data made available in the 2023 BDB, we present a novel approach to measure the performance of pass rushers (defensive players charging the QB).
 Relative to other aspects of American football, such as QB evaluation [@burke2019deepqb, @reyers2021quarterback], the literature on evaluating pass rushers is scarce.
@@
 
+Section \ref{sec:data}
+
 Section \ref{sec:methods}

### 2023-03-22 paper/01-introduction.Rmd
begin dumping in text for data section to clean up and modify later

@@
 Since its inception, the BDB has changed themes each year leading to new insight about evaluating different positions such as running backs, defensive backs, and special teams.
-The [2023 edition of the NFL BDB](https://www.kaggle.com/competitions/nfl-big-data-bowl-2023) continued this revolving theme by asking participants to evaluate lineman on passing plays.
+The [2023 edition of the NFL BDB](https://www.kaggle.com/competitions/nfl-big-data-bowl-2023) continued this revolving theme by asking participants to evaluate lineman on passing plays [@Howard2023NFL].
 

### 2023-03-22 paper/02-data.Rmd
begin dumping in text for data section to clean up and modify later

@@
 
-We rely on Next Gen Stats data provided by the 2023 NFL Big Data Bowl competition [@Howard2023NFL] for our investigation.
-The data span the first eight weeks of the 2021 NFL regular season, consisting of a total of 122 games.
+We use the tracking data made available in the 2023 NFL BDB competition for our investigation. The data span the first eight weeks of the 2021 NFL regular season, consisting of a total of 122 games.
+
 Also included are general information about the players, games, and plays within each game
@@
 Also included are scouting data supplied by Pro Football Focus, which collects player-level information at the play level.
+
+In December 2018, the NFL became the first North American professional sports league to release a portion of their tracking data to the public when temporarily made available a subset of this data from the first 6 weeks of the 2017 season for the inaugural “Big Data Bowl” competition.[2]
+
+The NFL’s tracking data collected as follows: Two radio frequency identification (RFID) chips are placed in each player’s shoulder pads (and in the ball).
+The RFID chips emit a signal to sensors in each stadium, which triangulate the location of the chip on the field.
+The data is collected at a rate of 10 Hz, so that the on-field location, speed, and angle of each player (and the ball) is recorded 10 times per second.
+Event annotations (e.g.
+ball snapped, first contact, pass thrown, etc) are recorded by the NFL for each play.
+In total, the dataset contains 1,075,720 unique frames across 14,167 plays, each of which records the locations and trajectories (speed, angle) of all twenty-two players (and the ball) on the field.
+
+Table 1 shows an example of this data for a 47-yard touchdown run by WR Cordarrelle Patterson, which occurred in a Week 6 game between the Los Angeles Chargers and Oakland Raiders in the 2017 season.
+Four frames from this play are displayed in Figure 2 displaying the coordinates of the offense (blue), defense (orange), and the ball-carrier (black) at particular events in the play.
+We will visualize the player tracking data in this manner for the remainder of the manuscript.
+
+This data can easily be joined to existing play-by-play data from the NFL’s API (e.g.
+via the nflscrapR package), which contains additional information about each play (Horowitz et al.
+2017).
+For the models in Section 4, we identified all ball-carrier sequences for running plays, which includes designed runs and QB scrambles.
+While the tracking data records the location of the ball in addition to the players, it does not identify who is the ball-carrier for a particular frame.
+We first identified the ball-carriers for every type of play (pass attempts, runs, returns, etc.) based on the information available from the NFL’s API via nflscrapR, which denotes who was directly involved in each play.
+Given the roles a player can have (passer, runner, receiver, interceptor, or returner), we used the provided event annotations to determine when a player became the ball-carrier.
+Since, for simplicity, we focus our attention on running plays in this manuscript, we identify the beginning of the ball-carrier sequence when the runner received the ball by either a handoff, lateral, or direct snap.
+The end of the ball-carrier sequence was marked when either the player was tackled, ran out of bounds, fumbled, or scored a touchdown.
+We excluded all plays missing the necessary information from the NFL API, as well as plays where the snap of the play was missing in the tracking data, and any ball-carrier sequences where either the starting or ending events were missing.
+After further pre-processing for the covariates described in Section 4, our final modeling dataset consisted of 154,908 frames from 4502 unique ball-carrier sequences on running plays.
+Figure 3A displays the distribution of the length of these ball-carrier sequences, revealing that majority of ball-carrier sequences are between two to five seconds in length, while Figure 3B displays the observed change in field position from the ball-carrier’s current location that will be modeled, as discussed in 3.3.
+
+The NFL Big Data Bowl Dataset contains tracking, play, and game data from all 91 games
+in the the first 6 weeks of the 2017 NFL season. The tracking data is at the granularity of
+every tenth of a second per play. For each player on the field (and the ball) for a given play,
+5
+the data contains: time stamp of play (time, yyyy-mm-dd, hh:mm:ss), player position along
+the long axis of the field (0 - 120 yards), player position along the short axis of the field
+(0 - 53.3 yards), speed in yards/second, distance traveled from prior time point (in yards),
+angle of player motion (0 - 360 degrees), tagged play details, (including moment of ball snap,
+pass release, pass catch, tackle, etc), player identification number (unique across players),
+player name, jersey number of player, team (away or home) of corresponding player, frame
+identifier for each play (starting at 1), unique game identifier, and play identifier (not unique
+across games). From the tracking data we are able to calculate eadch player’s distance to
+the ball and to other players as well as their cumulative distance run in the play and in the
+game. The play data contains (not an exhaustive list): game quarter, time on game clock at
+the start of the play (counting down from 15:00, MM:SS), down, distance needed for a first
+down, yard line at line-of-scrimmage, home team score prior to the play, visiting team score
+prior to the play, home team points at the end of the play, visiting team points at the end
+of the play, indicator for penalty called on play, indicator for special teams play, pass length
+(in yards), result of pass play (caught, incomplete, intercepted, run, sack), result of play in
+yards, and a description of the play. The play data allows us to calculate the difference in
+score and incorporate the timing of the play in our models. The game data contains game
+specific information like final score, temperature, humidity, and wind. We did not use any
+of this information in our models. More detailed information on the data can be found at
+the Big Data Bowl GitHub page
 

### 2023-03-23 paper/02-data.Rmd
Initial draft of data section with table and figure ideas

--- 2023-03-20 paper/02-data.Rmd
@@
 
-We rely on Next Gen Stats data provided by the 2023 NFL Big Data Bowl competition [@Howard2023NFL] for our investigation.
-The data span the first eight weeks of the 2021 NFL regular season, consisting of a total of 122 games.
-Also included are general information about the players, games, and plays within each game
-At a more granular level, we are given information on the player and football locations recorded at 10 frames per second for each play.
-Also included are scouting data supplied by Pro Football Focus, which collects player-level information at the play level.
+We use the tracking data made available in the 2023 NFL BDB competition for our investigation.
+The data corresponds to _INSERT NUMBER_ passing plays across 122 games in the first eight weeks of the 2021 NFL regular season.
+The aforementioned NGS data provides the on-field location, speed, direction, and orientation of each player on the field (and the ball) at a rate of 10 Hz, along with event annotations (e.g., ball snapped, pass thrown, etc.).
+In this particular dataset, there are _INSERT NUMBER_ unique frames corresponding to moments of time from the start of the play at snap until the moment the quarterback either throws the pass or is sacked.
+(_VERIFY THIS STATEMENT MAKES SENSE_)
+
+_Table X_ displays example data for a play from the 2021 NFL regular season week six matchup between the Las Vegas Raiders and Denver Broncos, which ended with Broncos QB Teddy Bridgewater getting sacked by Raiders DE Maxx Crosby[^crosbysack].
+_Figure X_ presents the locations of every Las Vegas (black) and Denver (orange) player on the field at four particular moments of time, with Maxx Crosby highlighted (blue).
+
+_INSERT TABLE_
+
+_INSERT FIGURE_
+
+Along with the tracking information, the 2023 BDB includes scouting data provided by Pro Football Focus (PFF).
+PFF's scouting data includes manually collected player-level information at the play level, such as the player's role (e.g., if they are a pass rusher) and credited events (e.g., player is credited with hitting the QB on the play).
+In this manuscript, we use the PFF data to identify _INSERT NUMBER_ unique players designated in the "pass rush" role across all plays.
+For context, _Figure X_ displays the distribution of the number of observed pass rushers involved in a play, ranging from _INSERT NUMBER_ to _INSERT NUMBER_ with _INSERT NUMBER_ pass rushers as the most common value.
+We also leverage the PFF data to count how many hits, hurries, and sacks each pass rusher is credited with across span of collected data.
+Additionally, we use the PFF player roles to identify which players were labeled as blocking each pass rusher to adjust for opponent strength, as discussed in Section _INSERT NUMBER_.
+
+INSERT FIGURE
 

### 2023-03-23 paper/03-methods.Rmd
Add draft for section on multilevel model

@@
 ## Multilevel Model
+
+While average STRAIN appears to be positively correlated with pass rusher ability, the observed average STRAIN on a single play is likely due to a convolution of factors.
+Outside of the pass rusher's ability, there is variability in the opposing strength of pass blockers across plays a pass rusher is involved.
+Additionally, based on previous work by PFF, the opposing QB plays an important role on observed pressure in passing plays [@eager2019pressure].
+Thus, we need to appropriately divide the credit of an observed average STRAIN across the different players involved and other factors.
+
+We use a multilevel model (i.e., hierarchical, random-effects, or mixed effects model) to evaluate pass rushers' impact on the average STRAIN observed in a play, while accounting for their pass blocker, QB, and play-level information.
+We identify the pass blocker linked with the pass rusher of interest using the data provided by PFF.
+Since there can be multiple blockers linked to a rusher, for simplicity we only consider the nearest blocker positioned to the pass rusher at the start of the play.
+We use random intercepts for the three player groups: (1) pass rushers as $R$, (2) nearest pass blockers as $B$, and (3) QBs as $Q$.
+The coefficients $\beta$ for additional information $X_{j,i}$ about pass rusher $j$ in play $i$ are estimated as fixed-effects.
+Our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows:
+
+$$
+\begin{aligned}
+\overline{STRAIN}_{j,i} &\sim Normal(R_{r[j,i]} + B_{b[j,i]} + Q_{q[i]} + X_{j,i} \cdot \beta, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
+R_{r} &\sim Normal(\mu_R, \sigma^2_R), \text{ for } r = 1, \cdots, \text{ \# of pass rushers}, \\
+B_{b} &\sim Normal(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
+Q_{q} &\sim Normal(\mu_Q, \sigma^2_Q), \text{ for } q = 1, \cdots, \text{ \# of QBs}.
+\end{aligned}
+$$
+
+We use a Normal distribution to pull the random intercepts for each player toward their respective position group means.
+This provides us with average individual effects on STRAIN for each player while applying shrinkage, which is a useful property in this problem's context since we do not observe the same number of plays for each player.
+With regards to the fixed-effects about pass rusher $j$ in play $i$, we include the length of the play, the pass rusher position (INSERT POSITIONS), the nearest pass blocker's position (INSERT POSITIONS), and the number of pass blockers assigned to the pass rusher according to PFF.
+We fit the model using penalized likelihood via the `lme4` package in `R` [@bateslme4].

### 2023-03-28 paper/01-introduction.Rmd
Editing.

--- 2023-03-22 paper/01-introduction.Rmd
@@
 In recent years, tracking data have replaced traditional box-score statistics and play-by-play data as the state-of-the-art tool in sports analytics.Numerous sports have been collecting and releasing data on player and ball locations on the playing surface over the course of a game.
+
 This spatio-temporal multi-resolution source of data have provided exceptional opportunities for researchers to perform advanced studies at a more complex level to deepen the understanding of different sports.
+
 For complete surveys on how tracking data have transformed sports analytics, see @Macdonald2020Recreating, @Baumer2023Big, and @Kovalchik2023Player.
@@
 
-The first BDB led to several contributions largely focused offensive performance evaluation.
-For example.
-finalists @deshpande2020expected introduced an approach for modeling the hypothetical completional probability of a pass aiding in the evalaution of quarterback (QB) decision making.
+The first BDB led to several contributions largely focused on offensive performance evaluation.
+For example, one group of finalists introduced an approach for modeling the hypothetical completion probability of a pass aiding in the evaluation of quarterback (QB) decision making [@deshpande2020expected].
 The winners of the inaugural BDB focused on identifying receiver routes via clustering techniques [@chu2020route] and convolutional neural networks [@sterken2019routenet].
-Along with the BDB competition entries, the public release of NGS data allowed researchers to tackle a variety of other problems such as revisiting fourth down decision making [@lopez2020bigger], annotating pass coverage with gaussian mixture models [@dutta2020unsupervised], and introducing a continuous-time framework to estimate within-play value [@yurko2020going].
-Since its inception, the BDB has changed themes each year leading to new insight about evaluating different positions such as running backs, defensive backs, and special teams.
-The [2023 edition of the NFL BDB](https://www.kaggle.com/competitions/nfl-big-data-bowl-2023) continued this revolving theme by asking participants to evaluate lineman on passing plays.
+Along with the BDB competition entries, the public release of NGS data allowed researchers to tackle (pun intended) a variety of other problems such as revisiting fourth down decision making [@lopez2020bigger], annotating pass coverage with Gaussian mixture models [@dutta2020unsupervised], and introducing a continuous-time framework to estimate within-play value [@yurko2020going].
+Since its inception, the BDB has chosen a different theme each year leading to new insight about evaluating different positions such as running backs, defensive backs, and special teams.
+The [2023 edition of the NFL BDB](https://www.kaggle.com/competitions/nfl-big-data-bowl-2023) asked participants to evaluate lineman on passing plays [@Howard2023NFL].
 
-In this work, using data made available in the 2023 BDB, we present a novel approach to measure the performance of pass rushers (defensive players charging the QB).
+In this work, using data made available in the 2023 BDB, we present a novel approach to measure the performance of pass rushers (defensive players whose role on a given play is to put pressure on the QB).
 Relative to other aspects of American football, such as QB evaluation [@burke2019deepqb, @reyers2021quarterback], the literature on evaluating pass rushers is scarce.
@@
 ## Previous Pass Rush Metrics
+The most widely know metric for tracking pass rusher performance is the sack.
+A sack is defined as follows: "to tackle (the quarterback) behind the line of scrimmage before the quarterback is able to throw a pass" (https://www.dictionary.com/browse/sack#:~:text=Football.%20to%20tackle%20(the%20quarterback,Verb%20Phrases) and has been officially tracked in the NFL since 1982 (citation?).
+Sacks represent a simple binary measure of success for a rusher on a pass play.
+However, there are many intermediate outcomes for a rusher that can also be considered successes that don't result in a sack.
+Measures such a hits and hurries are collected by various outlets where hits counts the number of times a QB is hit while passing and hurries counts how often a QB has to throw the ball earlier than they would have liked to.
+
+The sum of sacks, hits, and hurries has been defined as "pressures" (https://www.pro-football-reference.com/years/2022/advanced.htm) and is an improvement over simply sacks, but suffers from problems of subjectivity (i.e.
+was the play actually a hurry?) and are still binary outcomes (i.e.
+There is a continuum of plays that aren't hurries that all counted as simply "not hurries")
+
+hurries hits and sacks are "pressures"
 
@@
 Officially tracked by the NFL since 1982, the most well known statistic for evaluating defensive linemen on pass rushing plays is the sack, which occurs when a defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass.
+
+Sack definition:
 

### 2023-03-28 scripts/10_multilevel_mod.R
added number of blockers to model; computed vpc & icc; random effects uncertainty

@@ -102,8 +102,8 @@ mod_df |>
   count(block_pos)
 
 # some rushers are assigned 4 blockers 
-mod_df |> 
-  count(gameId, playId, nflId, sort = TRUE)
+n_blockers <- mod_df |> 
+  count(gameId, playId, nflId, name = "n_blockers")
 
 loc_rush <- mod_df |> 
   select(gameId, playId, nflId) |> 
@@ -121,12 +121,13 @@ mod_df_final <- mod_df |>
          y_block = loc_block$y,
          dis = sqrt((x_rush - x_block) ^ 2 + (y_rush - y_block) ^ 2)) |>
   group_by(gameId, playId, nflId) |> 
-  slice_min(dis, n = 1)
+  slice_min(dis, n = 1) |> 
+  left_join(n_blockers)
 
 
 library(lme4)  
 strain_fit <- lmer(
-  avg_strain ~ (1 | nflId) + (1 | blocker_id) + (1 | passer) + play_len + rush_pos + block_pos,
+  avg_strain ~ (1 | nflId) + (1 | blocker_id) + (1 | passer) + play_len + rush_pos + block_pos + n_blockers,
   data = mod_df_final
 )
 
@@ -158,8 +159,7 @@ tibble(passer = rownames(fit_out_qb),
        intercept = fit_out_qb$`(Intercept)`) |>
   arrange(-intercept) |> 
   left_join(get_passer_id) |> 
-  filter(nflId %in% filter(pass_snaps, n_plays >= 100)$nflId) |> 
-  View()
+  filter(nflId %in% filter(pass_snaps, n_plays >= 100)$nflId)
   
 # https://cran.r-project.org/web/packages/merTools/index.html
 
@@ -172,11 +172,39 @@ tibble(nflId = as.double(rownames(fit_out_block)),
   arrange(intercept) |> 
   left_join(players) |> 
   filter(nflId %in% filter(pass_block_snaps, n_plays >= 100)$nflId) |> 
-  filter(officialPosition == "G")
+  filter(officialPosition == "C")
 
 
 # fixed eff
 # var partition
 
+
+strain_fit |> 
+  VarCorr() |> 
+  as_tibble() |> 
+  mutate(icc = vcov / sum(vcov)) |> 
+  select(grp, icc)
+
+
+library(merTools)
+
+strain_eff <- REsim(strain_fit)
+
+strain_eff |> 
+  plotREsim()
+
+strain_eff |> 
[…]

### 2023-04-03 paper/01-introduction.Rmd
More intro edits.

@@
 ## Previous Pass Rush Metrics
-The most widely know metric for tracking pass rusher performance is the sack.
-A sack is defined as follows: "to tackle (the quarterback) behind the line of scrimmage before the quarterback is able to throw a pass" (https://www.dictionary.com/browse/sack#:~:text=Football.%20to%20tackle%20(the%20quarterback,Verb%20Phrases) and has been officially tracked in the NFL since 1982 (citation?).
+Officially tracked by the NFL since 1982, the most well known statistic for evaluating defensive linemen on pass rushing plays is the sack, which occurs when a defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass.
 Sacks represent a simple binary measure of success for a rusher on a pass play.
@@
 
-The sum of sacks, hits, and hurries has been defined as "pressures" (https://www.pro-football-reference.com/years/2022/advanced.htm) and is an improvement over simply sacks, but suffers from problems of subjectivity (i.e.
+The sum of sacks, hits, and hurries has been defined as "pressures" (https://www.pro-football-reference.com/years/2022/advanced.htm) and is an improvement over sacks, but suffers from problems of subjectivity (e.g.
 was the play actually a hurry?) and are still binary outcomes (i.e.
-There is a continuum of plays that aren't hurries that all counted as simply "not hurries")
+There is a continuum of plays that aren't hurries that all counted as simply "not hurries").
+
+Pro football focus (PFF) defines pass-rush productivity as: 
+$$
+\frac{\frac{(\# Hurries + \# Hits)}{2} + \# Sacks}{\#PassRush Snaps}
+$$.
+This metric gives twice as much weight to a sack relative to hurries and hits, which is a slight improvement, but the choice of weights is ad-hoc an still measures binary outcomes.
+(https://www.pro-football-reference.com/about/advanced_stats.htm)
+
+More recently metrics such as Time In Pocket (TIP) and Pass Rush Win Rate (PRWR) have also been proposed and are substantial improvements over the less sophisticated counting statistics.
+However, these are still imperfect measures as TIP is highly context dependent (e.g.
+deep vs shallow passing routes).
+PRWR is based on whether a pass rusher is able to beat their block before a fixed time from the snap, but this is highly dependent on the rather arbitrary choice of time used to define a "win".
+
+All of these measures represent different versions of "the best we can do with the available data".
+Without more sophisticated data (i.e.
+player tracking data), it is difficult to measure effectiveness of a pass rusher.
+
+Pass Rush productivity 
+Time in pocket
+Pass Rush win rate
 

### 2023-04-10 paper/01-introduction.Rmd
Updated intro.

@@
 
-Previous research bdb
-this year: linemen
-Importance of D-linemen, pass rush (Maybe talk about role of each position, edge vs interior, defensive formation)
+<!-- Previous research bdb -->
+<!-- this year: linemen -->
+<!-- Importance of D-linemen, pass rush (Maybe talk about role of each position, edge vs interior, defensive formation) -->
 
@@
 $$.
-This metric gives twice as much weight to a sack relative to hurries and hits, which is a slight improvement, but the choice of weights is ad-hoc an still measures binary outcomes.
+
+This metric gives twice as much weight to a sack relative to hurries and hits, which is a slight improvement, but the choice of weights is ad-hoc an still only measures binary outcomes similar to the shortcomings of previously discussed metrics.
 (https://www.pro-football-reference.com/about/advanced_stats.htm)
 
-More recently metrics such as Time In Pocket (TIP) and Pass Rush Win Rate (PRWR) have also been proposed and are substantial improvements over the less sophisticated counting statistics.
+More recently metrics such as Time In Pocket (TIP) and Pass Rush Win Rate [@Burke2018Created] (PRWR) have also been proposed and are substantial improvements over the less sophisticated counting statistics.
 However, these are still imperfect measures as TIP is highly context dependent (e.g.
 deep vs shallow passing routes).
-PRWR is based on whether a pass rusher is able to beat their block before a fixed time from the snap, but this is highly dependent on the rather arbitrary choice of time used to define a "win".
+PRWR is based on whether a pass rusher is able to beat their block before a fixed time (ESPN chooses 2.5 seconds) from the snap, but this is highly dependent on the rather arbitrary choice of time used to define a "win".
+In addition, once a cut-off is chosen, this metric becomes binary like most of the previously discussed measures.
+
+(https://www.espn.com/nfl/story/_/id/34536376/2022-nfl-pass-rushing-run-stopping-blocking-leaderboard-win-rate-rankings-top-players-teams)
 
@@
 Without more sophisticated data (i.e.
-player tracking data), it is difficult to measure effectiveness of a pass rusher.
+player tracking data), it is difficult to measure effectiveness of a pass rusher so many simple metrics of success have been proposed.
+Table \ref{tab:metrics} gives a summary of existing pass-rush metrics.
+
+\begin{table}
+\caption{A summary of previously-existing pass rush metrics. \label{tab:metrics}}
+\begin{center}
+\begin{tabular}{ll}
+\textbf{Metric} & \textbf{Description}\\\hline
+Sacks & A defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass\\
+Hits & A defensive player tackles the quarterback behind the line of scrimmage after the quarterback is able to throw a pass.   \\
+Hurries & A defensive player pressures the quarterback behind the line of scrimmage such that the quarterback must release the ball prior sooner than they wanted to. \\
+QB Pressures & A sack, hit, or hurry.  \\
+Pass-Rush Productivity (PRP) & $\frac{\frac{(\# Hurries + \# Hits)}{2} + \# Sacks}{\#PassRush Snaps}$\\
+Time In Pocket (TIP) & The time from snap to throw or pocket collapse for a quarterback (https://www.pff.com/news/pro-pff-signature-statistics-a-glossary#:~:text=Time%20in%20Pocket,time%20until%20a%20quarterback%20scrambles)\\
+Pass Rush Win Rate (PRWR) & Rate at which pass rusher beats pass block within 2.5 seconds after snap (https://www.espn.com/nfl/story/_/id/34536376/2022-nfl-pass-rushing-run-stopping-blocking-leaderboard-win-rate-rankings-top-players-teams)\\\hline
+\end{tabular}
+\end{center}
+\end{table}
+
+## Previous Pass-Rush Measure in academic literaure 
+@AlamarGould2008: Abstract: Protecting the quarterback is an integral part of the passing game in the National Football league, yet the relationship between the abilities of an individual lineman and the effectiveness of a passing game remains unexplored.
+One of the principal reasons for this lack of study is the absence of publicly available data that is needed in order to track the performance of a specific lineman.
+In order to create the relevant data, the first 3 games of the 2007 NFL season for seven different teams were charted.
+The performance of each lineman was recorded on every pass play, as well as the amount of undisturbed time the quarterback was given (time in the pocket) to make a throw.
+These data were used in a series of regressions to determine how likely a lineman was to successfully hold his block in relation to the time it took for the quarterback to throw the ball, for each lineman in the sample.
+These data were also used to estimate the correlation between successful blocking and completion rate.
+The results of these regressions were then used to simulate the effects that different linemen have on the passing game.
+The trade in the offseason between the New York Jets and Washington Redskins which sent left guard Pete Kendall to Washington was examined.
+The analysis finds that the Jets lost approximately 3 percentage points on their completion rate due to the trade.
+
+@AlamarGoldner2011: They talk about how there is no data available to measure offensive line performance.
+1259 plays, 20 teams, 2010 season.
+Success of each offensive lineman was evaluated using the best judgement of the observer.
+Collected time in pocket.
+They collected a bunch of binary variables and did this all by HAND!
+Estimate a survival model for each position on the offensive line.
+They don't rank individuals, but rather positions on a team.
+They then use the predicted probability as a covariate in a model to predict completion.
 
@@
 Limitations
-
-\begin{table}
-\caption{A summary of previously-existed pass rush metrics. \label{tab:metrics}}
-\begin{center}
-\begin{tabular}{ll}
-\textbf{Metric} & \textbf{Description}\\\hline
-Sacks & a\\
-Hits & a\\
-Hurries & a\\
-QB Pressures & a\\
-Pass-Rush Productivity (PRP) & a\\
-Time In Pocket (TIP) & a\\
-Pass Rush Win Rate (PRWR) & Rate at which pass rusher beats pass block within 2.5 seconds after snap\\\hline
-\end{tabular}
-\end{center}
-\end{table}
-
-Table \ref{tab:metrics}
 

### 2023-04-10 paper/01-introduction.Rmd
First Draft of intro.

@@
 ## Previous Pass-Rush Measure in academic literaure
-@AlamarGould2008: Abstract: Protecting the quarterback is an integral part of the passing game in the National Football league, yet the relationship between the abilities of an individual lineman and the effectiveness of a passing game remains unexplored.
+The academic literature on the performance of linemen, either offensive or defensive in football is scant.
+@AlamarGould2008 studied the the relationship between pass completion rate and performance of offensive linemen in terms of how successful they were at blocking.
+The data for this study was collecting essentially by hand by watching and charting the performance of each lineman on a play.
+This was done using games from the beginning 2007 season for seven teams and their first three games each.
+The authors note that there is not much research on linemen saying "One of the principal reasons for this lack of study is the absence of publicly available data that is needed in order to track the performance of a specific lineman."
+
+Later, @AlamarGoldner2011 again studied offensive linemen performance.
+They analyzed 20 teams during the 2010 season for a total of 1,259 plays.
+They don't estimate the performance of an individual lineman, but rather positions on a team (e.g.
+Chicago Bears' left tackle, Pittsburgh Steelers' center, etc.).
+They then go on to use a survival analysis model to estimate time in pocket.
+The average probability of success from the survival model is then used as a covariate in a model to estimate the probability of competing a pass, and they show that there is a significant relationship between these two probabilities.
+
+## Contributions of STRAIN
+All the aforementioned methods suffered from not having the data that was actually necessary to develop a meaningful metric for linemen in football, either offensive or defensive.
+Now though, thanks to the 2023 Big Data Bowl, we have a publicly available tracking data not only for the linemen, but for every player on the field!
+With this data performance of offensive and defensive linemen can be meaningfully assessed, though we focus solely on the performance of defensive linemen here.
+This manuscript presents a metric we refer to as STRAIN based on a concept of the same name from materials science.
+STRAIN gives a continuous measure of pressure for each of the defensive linemen over the course of an entire play allowing for the measurement of the level of success of the pass rush even on plays that don't result in an easily observed outcome like a sack, hit, or hurry.
+We view this as a major step forward in the accurate evaluation of defensive linemen performance.
+Further, we note than while we focus exclusively on defensive linemen here, the proposed metric STRAIN is relatively easily extended to the evaluation of offensive linemen performance.
+
+The remainder of this paper is outlined as follows.
+Section \ref{sec:data} describes the data set from the 2023 Big Data Bowl.
+Section \ref{sec:methods} introduces the formal definition of STRAIN.
+Section \ref{sec:results} studies the performance of the proposed metric.
+We end with Section \ref{sec:discussion} where we present a discussion of future directions for STRAIN.
+
+<!-- : They talk about how there is no data available to measure offensive line performance.
+1259 plays, 20 teams, 2010 season.
+Success of each offensive lineman was evaluated using the best judgement of the observer.
+Collected time in pocket.
+They collected a bunch of binary variables and did this all by HAND!
+Estimate a survival model for each position on the offensive line.
+They don't rank individuals, but rather positions on a team.
+They then use the predicted probability as a covariate in a model to predict completion.
+-->
+
+<!-- @AlamarGould2008: Abstract: Protecting the quarterback is an integral part of the passing game in the National Football league, yet the relationship between the abilities of an individual lineman and the effectiveness of a passing game remains unexplored.
 One of the principal reasons for this lack of study is the absence of publicly available data that is needed in order to track the performance of a specific lineman.
@@
 The analysis finds that the Jets lost approximately 3 percentage points on their completion rate due to the trade.
+-->
 
-@AlamarGoldner2011: They talk about how there is no data available to measure offensive line performance.
+<!-- Is this interesting? https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2938948/ -->
+
+<!-- Evaluating offensive line salray: https://sites.duke.edu/djepapers/files/2016/10/froelich-dje.pdf -->
+
+<!-- Read this:  -->
+<!-- https://www.researchgate.net/profile/Gita-Sukthankar/publication/220814393_A_Real-Time_Opponent_Modeling_System_for_Rush_Football/links/00b7d528a16b30fc12000000/A-Real-Time-Opponent-Modeling-System-for-Rush-Football.pdf -->
+
+<!-- Pass Rush productivity  -->
+<!-- Time in pocket -->
+<!-- Pass Rush win rate -->
+
+<!-- hurries hits and sacks are "pressures" -->
+
+<!-- Limitations -->
+
+<!-- Sacks, hits, hurries, pressures -->
+
+<!-- Officially tracked by the NFL since 1982, the most well known statistic for evaluating defensive linemen on pass rushing plays is the sack, which occurs when a defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass.
+-->
+
+<!-- Sack definition:  -->
+
+<!-- https://www.pro-football-reference.com/about/advanced_stats.htm -->
+
+<!-- Pressures: QB pressures (hurries + knockdowns + sack plays). Note that "sack plays" in this instance include both sacks and assists. -->
+
+<!-- Hurried: Times QBs were hurried - defined as a situation when the QB is forced to throw the ball earlier than intended or is chased around/out of the pocket as the result of defensive pressure -->
+
+<!-- Hit: Times QBs were hit - occurs when the QB hits the ground after a throw, not marked on a sack play -->
+
+<!-- PFF's Pass-Rush Productivity (PRP) -->
+
+<!-- https://www.pff.com/news/pro-signature-stats-spotlight-defensive-line -->
+
+<!-- More recently metrics such as Time In Pocket (TIP) and Pass Rush Win Rate (PRWR) have also been proposed and are substantial improvements over the less sophisticated counting statistics.  -->
+
+<!-- Pocket Time: Number of seconds the quarterback was in the pocket between the snap of the ball and a pass or the collapse of the pocket -->
+
+<!-- @Burke2018Created -->
+<!-- @AlamarGoldner2013: They talk about how there is no data available to measure offensive line performance.
 1259 plays, 20 teams, 2010 season.
@@
 They then use the predicted probability as a covariate in a model to predict completion.
+-->
 
-Pass Rush productivity 
-Time in pocket
-Pass Rush win rate
+<!-- @AlamarGould2008 -->
 
-hurries hits and sacks are "pressures"
+<!-- https://www.espn.com/nfl/story/_/id/24892208 -->
 
-Limitations
+<!-- 2.5 seconds, arbitrary -->
 
-Sacks, hits, hurries, pressures
-
-Officially tracked by the NFL since 1982, the most well known statistic for evaluating defensive linemen on pass rushing plays is the sack, which occurs when a defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass.
-
-Sack definition:
-
-https://www.pro-football-reference.com/about/advanced_stats.htm
-
-Pressures: QB pressures (hurries + knockdowns + sack plays). Note that "sack plays" in this instance include both sacks and assists.
-
-Hurried: Times QBs were hurried - defined as a situation when the QB is forced to throw the ball earlier than intended or is chased around/out of the pocket as the result of defensive pressure
-
-Hit: Times QBs were hit - occurs when the QB hits the ground after a throw, not marked on a sack play
-
-PFF's Pass-Rush Productivity (PRP)
-
-https://www.pff.com/news/pro-signature-stats-spotlight-defensive-line
-
-More recently metrics such as Time In Pocket (TIP) and Pass Rush Win Rate (PRWR) have also been proposed and are substantial improvements over the less sophisticated counting statistics.
-
-Pocket Time: Number of seconds the quarterback was in the pocket between the snap of the ball and a pass or the collapse of the pocket
-
-@Burke2018Created
-@AlamarGoldner2013: They talk about how there is no data available to measure offensive line performance.
-1259 plays, 20 teams, 2010 season.
-Success of each offensive lineman was evaluated using the best judgement of the observer.
-Collected time in pocket.
-They collected a bunch of binary variables and did this all by HAND!
-Estimate a survival model for each position on the offensive line.
-They don't rank individuals, but rather positions on a team.
-They then use the predicted probability as a covariate in a model to predict completion.
-
-@AlamarGould2008
-
-https://www.espn.com/nfl/story/_/id/24892208
-
-2.5 seconds, arbitrary
-
-While PRWR is certainly a step in the right direction, it takes continuous data and converts it to a binary win/loss, which is generally not recommended in statistical analysis
-
-## Contributions of STRAIN
-
-STRAIN is the best etc.
-
-The remainder of this paper is outlined as follows.
-
-Section \ref{sec:data}
-
-Section \ref{sec:methods}
-
-Section \ref{sec:results}
-
-Section \ref{sec:discussion}
+<!-- While PRWR is certainly a step in the right direction, it takes continuous data and converts it to a binary win/loss, which is generally not recommended in statistical analysis -->

### 2023-04-10 paper/01-introduction.Rmd
fix knitting issues

@@
 ---
+
 # Introduction {#sec:introduction}
@@
 
-In recent years, tracking data have replaced traditional box-score statistics and play-by-play data as the state-of-the-art tool in sports analytics.Numerous sports have been collecting and releasing data on player and ball locations on the playing surface over the course of a game.
-
+In recent years, tracking data have replaced traditional box-score statistics and play-by-play data as the state-of-the-art tool in sports analytics.
+Numerous sports have been collecting and releasing data on player and ball locations on the playing surface over the course of a game.
 This spatio-temporal multi-resolution source of data have provided exceptional opportunities for researchers to perform advanced studies at a more complex level to deepen the understanding of different sports.
-
 For complete surveys on how tracking data have transformed sports analytics, see @Macdonald2020Recreating, @Baumer2023Big, and @Kovalchik2023Player.
@@
 ## Previous Pass Rush Metrics
+
 Officially tracked by the NFL since 1982, the most well known statistic for evaluating defensive linemen on pass rushing plays is the sack, which occurs when a defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass.
@@
 \textbf{Metric} & \textbf{Description}\\\hline
-Sacks & A defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass\\
+Sacks & A defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass \\
 Hits & A defensive player tackles the quarterback behind the line of scrimmage after the quarterback is able to throw a pass. \\
@@
 Pass-Rush Productivity (PRP) & $\frac{\frac{(\# Hurries + \# Hits)}{2} + \# Sacks}{\#PassRush Snaps}$\\
-Time In Pocket (TIP) & The time from snap to throw or pocket collapse for a quarterback (https://www.pff.com/news/pro-pff-signature-statistics-a-glossary#:~:text=Time%20in%20Pocket,time%20until%20a%20quarterback%20scrambles)\\
-Pass Rush Win Rate (PRWR) & Rate at which pass rusher beats pass block within 2.5 seconds after snap (https://www.espn.com/nfl/story/_/id/34536376/2022-nfl-pass-rushing-run-stopping-blocking-leaderboard-win-rate-rankings-top-players-teams)\\\hline
+Time In Pocket (TIP) & The time from snap to throw or pocket collapse for a quarterback \\
+Pass Rush Win Rate (PRWR) & Rate at which pass rusher beats pass block within 2.5 seconds after snap\\\hline
 \end{tabular}
@@
 
+(https://www.pff.com/news/pro-pff-signature-statistics-a-glossary#:~:text=Time%20in%20Pocket,time%20until%20a%20quarterback%20scrambles)
+
+(https://www.espn.com/nfl/story/_/id/34536376/2022-nfl-pass-rushing-run-stopping-blocking-leaderboard-win-rate-rankings-top-players-teams)
+
 ## Previous Pass-Rush Measure in academic literaure
+
 The academic literature on the performance of linemen, either offensive or defensive in football is scant.
@@
 ## Contributions of STRAIN
+
 All the aforementioned methods suffered from not having the data that was actually necessary to develop a meaningful metric for linemen in football, either offensive or defensive.

### 2023-04-11 paper/00-main.Rmd
created data section
added more figures

@@
 title: |
-  | Bearing the Strain: Analyzing Pass Rush in American Football with Player Tracking Data
+  | Bearing the STRAIN: Analyzing Pass Rush in American Football with Player Tracking Data
 
@@
 - name: Quang Nguyen
-  thanks: Corresponding author's email
+  # thanks: Corresponding author's email
   affiliation: |
@@
 
-```{r child = "02-methods.Rmd"}
+```{r child = "02-data.Rmd"}
 ```
 
-```{r child = "03-results.Rmd"}
+```{r child = "03-methods.Rmd"}
 ```
 
-```{r child = "04-discussion.Rmd"}
+```{r child = "04-results.Rmd"}
+```
+
+```{r child = "05-discussion.Rmd"}
 ```

### 2023-04-11 paper/04-results.Rmd
added more figures

@@
 
-Figure \ref{fig:fig_pos_curves} shows pos curves
+```{r}
+#| label: fig_outcome_curves
+#| fig.cap: "STRAIN curves for different positions by outcome"
+#| fig.width: 6
+#| fig.height: 6
+source(here("scripts", "08_fig_outcome_curves.R"))
+fig_outcome_curves
+```
+
+Figure \ref{fig:fig_pos_curves} displays the average STRAIN by position for the first 40 frames (4 seconds) after the snap.
+We observe a clear difference in STRAIN between edge rushers (OLB's and DE's) and interior linemen (DT's and NT's).
+Specifically, edge rushers have higher STRAIN than interior linemen on average, as they are more easily able to approach the quarterback on the edge of the pocket versus, for instance, a nose tackle attacking the line head on.
+
+On average, STRAIN appears to increase for the first 0.5 seconds of a play, followed by a decline in the next second.
+STRAIN then increases again until around 2.5-3 seconds after the snap before trending down towards the end of the play.
+In context, this reflects the actions that a pass-rusher initially moves towards the quarterback, but is then stopped by the offensive line while the quarterback drops back.
+When the quarterback stops dropping, the rusher closes the gap and increases STRAIN, before slowing down later on.
+
+Further, Figure \ref{fig:fig_outcome_curves} shows the positional STRAIN curves by whether a play's outcome is a hit, sack, hurry, or none of those.
 
@@
 
+## Statistical Properties of STRAIN
+
+<!-- From @Franks2016Meta (github https://github.com/afranks86/meta-analytics) -->
+
+<!--
+
+Discrimination: Does the metric reliably differentiate between players?
+
+Independence: Does the metric provide new information?
+
+-->
+
 ## Corr with pressure rate = (hits + sacks + hurries) per snap
@@
 
-Fairly strong correlation (r = 0.6254943) between average STRAIN & hurries + sacks + hits per snaps
-
-## Statistical Properties of STRAIN
-
-From @Franks2016Meta (github https://github.com/afranks86/meta-analytics)
-
-<!--
-
-Discrimination: Does the metric reliably differentiate between players?
-
-Independence: Does the metric provide new information?
-
--->
+We observe a fairly strong correlation (r = 0.625) between average STRAIN and pressure rate.
 
@@
 
-0.8544865
+Figure \ref{fig:fig_stability} is a scatterplot of average STRAIN across all frames played in first and last 4 weeks.
+It is apparent that there is a strong correlation (r = 0.854), illustrating great metric stability.
 
-Figure \ref{fig:fig_stability} shows
+## Multilevel Model
 
-Average STRAIN across all frames played in first and last 4 weeks
+coefficient table
 
-Strong correlation shows great metric stability
+icc
+rusher 0.0574
+blocker 0.0121
+passer 0.0233
+Residual 0.907
+
+```{r}
+#| label: fig_rankings_boot
+#| fig.cap: "steve"
+#| fig.width: 6
+#| fig.height: 8
+source(here("scripts", "10_multilevel_mod.R"))
+fig_rankings_boot
+```

### 2023-04-11 scripts/08_fig_outcome_curves.R

@@ -1,5 +1,7 @@
+library(tidyverse)
+library(here)
 
-pass_rush
+pass_rush <- read_csv(here("data", "pass_rush.csv.gz"))
 
 
 pff <- read_csv(here("data", "pffScoutingData.csv"))
@@ -10,9 +12,6 @@ outcomes <- pff |>
             is_hurry = sum(pff_hurry, na.rm = TRUE) > 0 ,
             is_sack = sum(pff_sack, na.rm = TRUE) > 0)
 
-outcomes |> 
-  count(gameId, playId, nflId, sort = TRUE)
-
 
 hit <- pass_rush |> 
   inner_join(outcomes |> 
@@ -22,25 +21,7 @@ hit <- pass_rush |>
          frameId_snap_corrected <= 40) |> 
   group_by(officialPosition, frameId_snap_corrected) |> 
   summarise(mn = mean(strain, na.rm = TRUE)) |> 
-  mutate(officialPosition = factor(officialPosition, levels = c("OLB", "DE", "DT", "NT"))) |> 
-  ggplot(aes(frameId_snap_corrected, mn, 
-             color = officialPosition, 
-             group = officialPosition)) +
-  geom_smooth(se = FALSE, span = 0.3, linewidth = 1) +
-  scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107", "#004D40"), name = "Position") +
-  scale_x_continuous(breaks = seq(0, 40, 10), labels = 0:4) +
-  scale_linetype_manual(values = c(rep("solid", 4), "dotted"), name = "Position") +
-  expand_limits(y = 2.3) +
-  labs(y = "STRAIN",
-       x = "Time since snap (seconds)",
-       title = "Hit") +
-  theme_light() +
-  theme(legend.key.width = unit(0.9, "cm"),
-        legend.position = "none",
-        axis.title = element_text(size = rel(1)),
-        axis.text = element_text(size = rel(0.8)),
-        legend.title = element_text(size = rel(1)),
-        legend.text = element_text(size = rel(0.8)))
+  mutate(outcome = "Hit")
 
 sack <- pass_rush |> 
   inner_join(outcomes |> 
@@ -50,25 +31,7 @@ sack <- pass_rush |>
          frameId_snap_corrected <= 40) |> 
   group_by(officialPosition, frameId_snap_corrected) |> 
   summarise(mn = mean(strain, na.rm = TRUE)) |> 
[…]

### 2023-04-11 scripts/10_multilevel_mod.R

@@ -49,9 +49,9 @@ get_avg_strain <- pass_rush |>
   summarize(avg_strain = mean(strain, na.rm = TRUE)) |>
   ungroup()
 
-get_avg_strain |> 
-  ggplot(aes(avg_strain)) +
-  geom_histogram()
+# get_avg_strain |> 
+#   ggplot(aes(avg_strain)) +
+#   geom_histogram()
 
 # get qb and play info from nflfastR
 d <- nflfastR::load_pbp(2021)
@@ -95,11 +95,11 @@ mod_df <- pff |>
   mutate(block_pos = ifelse(!block_pos %in% c("C", "G", "T"), "other", block_pos))
 
 # check to make sure
-mod_df |> 
-  count(rush_pos)
-
-mod_df |> 
-  count(block_pos)
+# mod_df |> 
+#   count(rush_pos)
+# 
+# mod_df |> 
+#   count(block_pos)
 
 # some rushers are assigned 4 blockers 
 n_blockers <- mod_df |> 
@@ -131,48 +131,47 @@ strain_fit <- lmer(
   data = mod_df_final
 )
 
-strain_fit
 
-fit_out_rush <- strain_fit |> 
-  ranef() |> 
-  pluck("nflId")
-
-tibble(nflId = as.double(rownames(fit_out_rush)), 
-       intercept = fit_out_rush$`(Intercept)`) |> 
-  arrange(-intercept) |> 
-  left_join(players) |> 
-  filter(nflId %in% filter(pass_rush_snaps, n_plays >= 100)$nflId) |> 
-  filter(officialPosition == "OLB")
-
-
-fit_out_qb <- strain_fit |> 
-  ranef() |> 
-  pluck("passer")
-
-get_passer_id <- players |> 
-  filter(officialPosition == "QB") |> 
-  separate(displayName, into = c("fn", "ln")) |> 
-  transmute(passer = str_c(str_sub(fn, 1, 1), ".", ln),
-            nflId)
-
-tibble(passer = rownames(fit_out_qb), 
-       intercept = fit_out_qb$`(Intercept)`) |>
-  arrange(-intercept) |> 
-  left_join(get_passer_id) |> 
-  filter(nflId %in% filter(pass_snaps, n_plays >= 100)$nflId)
+# fit_out_rush <- strain_fit |> 
+#   ranef() |> 
+#   pluck("nflId")
+# 
+# tibble(nflId = as.double(rownames(fit_out_rush)), 
[…]

### 2023-04-11 paper/02-data.Rmd
added example tracking data table

@@
 We use the tracking data made available in the 2023 NFL BDB competition for our investigation.
-The data corresponds to _INSERT NUMBER_ passing plays across 122 games in the first eight weeks of the 2021 NFL regular season.
+The data corresponds to 8557 passing plays across 122 games in the first eight weeks of the 2021 NFL regular season.
 The aforementioned NGS data provides the on-field location, speed, direction, and orientation of each player on the field (and the ball) at a rate of 10 Hz, along with event annotations (e.g., ball snapped, pass thrown, etc.).
@@
 
-_Table X_ displays example data for a play from the 2021 NFL regular season week six matchup between the Las Vegas Raiders and Denver Broncos, which ended with Broncos QB Teddy Bridgewater getting sacked by Raiders DE Maxx Crosby[^crosbysack].
+Table \ref{tab:track} displays example data for a play from the 2021 NFL regular season week six matchup between the Las Vegas Raiders and Denver Broncos, which ended with Broncos QB Teddy Bridgewater getting sacked by Raiders DE Maxx Crosby[^crosbysack].
 _Figure X_ presents the locations of every Las Vegas (black) and Denver (orange) player on the field at four particular moments of time, with Maxx Crosby highlighted (blue).
 
-_INSERT TABLE_
+\begin{table}
+\caption{track. \label{tab:track}}
+\centering
+\begin{tabular}{rrrlrrrrrrrl}
+\hline
+frameId & jerseyNumber & team & x & y & s & a & dis & o & dir & event \\ 
+\hline
+7 & 98 & LV & 67.68 & 29.89 & 0.34 & 1.57 & 0.04 & 124.86 & 88.21 & ball\_snap \\ 
+8 & 98 & LV & 67.76 & 29.89 & 0.69 & 2.13 & 0.08 & 124.07 & 89.59 & None \\ 
+\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\ 
+50 & 98 & LV & 73.67 & 25.06 & 4.19 & 2.62 & 0.42 & 134.21 & 125.26 & qb\_sack \\ 
+\hline
+\end{tabular}
+\end{table}
 
@@
 
-tracking: locations of players and ball, 10 fps,
+(sample restriction)
 
-pre-processing standardize
-
-PFF scouting: pass rush role
-
-sample restriction: remove plays w multiple QB
-
-frames between snap and QB outcome (pass forward, sack)
-
-sample size
+We include only the frames between the ball snap and when a pass forward or QB sack is recorded for each play.
+We also remove all plays with multiple quarterbacks on the field, since we need a uniquely defined quarterback to calculate our measure.
+Using PFF scouting data, we only consider defenders with role "Pass rush" for evaluation.

### 2023-04-11 paper/04-results.Rmd
added example tracking data table

@@
 
-## Corr with pressure rate = (hits + sacks + hurries) per snap
-
 ```{r}
 #| label: fig_cor_pressure
-#| fig.cap: "Cor with"
+#| fig.cap: "Corr with pressure rate = (hits + sacks + hurries) per snap"
 #| fig.width: 4
@@
 Figure \ref{fig:fig_cor_pressure} shows cor with pressure
-
 We observe a fairly strong correlation (r = 0.625) between average STRAIN and pressure rate.
-
-Stability: Does the metric measure a quantity which is stable over time?  (section 5.1 of openWAR has a similar idea)
 
@@
 Figure \ref{fig:fig_stability} is a scatterplot of average STRAIN across all frames played in first and last 4 weeks.
-It is apparent that there is a strong correlation (r = 0.854), illustrating great metric stability.
+It is apparent that there is a strong correlation (r = 0.854), illustrating great metric stability over time.
 

### 2023-04-11 scripts/01_fig_field.R

@@ -17,6 +17,11 @@ library(here)
 
 play_lv_den <- read_csv(here("data", "play_lv_den.csv.gz"))
 
+# play_lv_den |> filter(nflId == 47889) |> 
+#   select(frameId, jerseyNumber, team, x:event) |> 
+#   filter(frameId %in% c(7, 8, 50)) |> 
+#   xtable::xtable()
+
 plays <- read_csv(here("data", "plays.csv"))
 desc <- plays |>
   filter(gameId == 2021101709 & playId == 1444) |>

### 2023-04-11 scripts/10_multilevel_mod.R

@@ -177,6 +177,8 @@ strain_fit <- lmer(
 # fixed eff
 # var partition
 
+summary(strain_fit)$coef
+broom.mixed::tidy(strain_fit, conf.int=TRUE)
 
 strain_fit |> 
   VarCorr() |> 

### 2023-04-11 paper/04-results.Rmd
outlined stability section

@@
 Further, Figure \ref{fig:fig_outcome_curves} shows the positional STRAIN curves by whether a play's outcome is a hit, sack, hurry, or none of those. 
+It is striking that for plays with sack outcome,
 
@@
 Since STRAIN is observed continuously for every play in our data, this allows us to aggregate across all frames played and compute the average STRAIN for NFL pass rushers over the course of the eight-week sample size.
-
-(maybe some math)
-
 Based on the clearly distinct patterns for different positions as previously observed, we evaluate interior pass-rushers (NT's and DT's) separately from edge rushers (OLB's and DE's).
@@
 
+Next, we examine the predictability of STRAIN as a measure of pressure.
+We first look at how well strain correlates with a simple measure of pressure rate, defined as the total hits, sacks and hurries per snap.
+Figure \ref{fig:fig_cor_pressure} shows a scatterplot of average STRAIN and pressure rate.
+We observe a fairly strong correlation (r = 0.625)...
+
+Furthermore, first 4 weeks’ avg STRAIN is more predictive of last 4 weeks’ hurries sacks hits rate than first 4 weeks’ HSH rate
+
+(INSERT FIGURE)
+
 ```{r}
@@
 ```
-
-Figure \ref{fig:fig_cor_pressure} shows cor with pressure
-We observe a fairly strong correlation (r = 0.625) between average STRAIN and pressure rate.
 

### 2023-04-11 paper/04-results.Rmd
Begin initial description of multilevel model results

@@
 
+There are a variety of quantities we're interested in from the multilevel model fit in Section [INSERT SECTION REF].
+First, Table [FIXED EFFECTS REF] displays the results for the fixed effects terms.
+[INSERT CONCLUSIONS ABOUT THESE VARIABLES]  Next, Table [INSERT ICC TABLE REF] displays the intraclass correlation coefficients (ICC) for the three different grouped terms, describing the proportion variance explained between the group terms in comparison to the residual variance.
+While the residual variance is unsurprisingly the largest value, between the three group terms, we observe the largest ICC for pass rushers, followed by the QB, and the blocker.
+This emphasizes how STRAIN is mostly attributable to pass rushers, but is necessary to adjust for opposition and other factors.
+
+We display the random intercepts for pass rusher in Table [INSERT TABLE REFERENCE] alongside their unadjusted STRAIN values.
+We observe similar rankings based on the random intercepts in comparison to the average STRAIN, with Myles Garrett (DE), Aaron Donald (DT), Vita Vea (NT), and T.J.
+Watt (OLB) as the top pass rushers.
+In addition to adjusting for opposition, our multilevel model allows us to provide some notion of uncertainty about the pass rusher's effect on the play level STRAIN.
+Figure \ref{fig:fig_rankings_boot} displays the mean varying intercept for the top ten pass rushers for each position along with 95% confidence intervals generated by sampling from the posterior distributions via parametric bootstrapping using the `merTools` package [@knowlesmerTools, @gelman2006data].
+We observe different levels of variability across these top players, with the only a subset of players with intervals strictly above zero.
+[NEED TO THINK ABOUT WHAT MORE TO SAY HERE]
+
 coefficient table
 
+MAKE THIS A TABLE
 icc

### 2023-05-01 paper/01-introduction.Rmd
Some edits.

@@
 $$
-\frac{\frac{(\# Hurries + \# Hits)}{2} + \# Sacks}{\#PassRush Snaps}
+\frac{\frac{1}{2}(\# Hurries + \# Hits) + \# Sacks}{\#PassRush Snaps}
 $$.
@@
 
-(https://www.pff.com/news/pro-pff-signature-statistics-a-glossary#:~:text=Time%20in%20Pocket,time%20until%20a%20quarterback%20scrambles)
+<!-- (https://www.pff.com/news/pro-pff-signature-statistics-a-glossary#:~:text=Time%20in%20Pocket,time%20until%20a%20quarterback%20scrambles) -->
 
-(https://www.espn.com/nfl/story/_/id/34536376/2022-nfl-pass-rushing-run-stopping-blocking-leaderboard-win-rate-rankings-top-players-teams)
+<!-- (https://www.espn.com/nfl/story/_/id/34536376/2022-nfl-pass-rushing-run-stopping-blocking-leaderboard-win-rate-rankings-top-players-teams) -->
 

### 2023-05-01 paper/02-data.Rmd
Some edits.

@@
 # Data {#sec:data}
+```{r eval = FALSE, echo = FALSE}
+pass_rush <- read_csv(here("data", "pass_rush.csv.gz"))
+pass_rush %>% select(gameId, playId, frameId) %>% n_distinct()
+```
 
@@
 The aforementioned NGS data provides the on-field location, speed, direction, and orientation of each player on the field (and the ball) at a rate of 10 Hz, along with event annotations (e.g., ball snapped, pass thrown, etc.).
-In this particular dataset, there are _INSERT NUMBER_ unique frames corresponding to moments of time from the start of the play at snap until the moment the quarterback either throws the pass or is sacked.
+In this particular dataset, there are _INSERT NUMBER_ (our cleaned dataset has 251060 unique frames) unique frames corresponding to moments of time from the start of the play at snap until the moment the quarterback either throws the pass or is sacked.
 (_VERIFY THIS STATEMENT MAKES SENSE_)

### 2023-05-01 scripts/00_data_prep.R

@@ -6,6 +6,7 @@
 # save them in the "data" folder
 
 # a lot of base R from Greg here
+# You call it "base R".  I call it "OG R".  
 
 library(tidyverse)
 library(here)

### 2023-05-08 paper/01-introduction.Rmd
edits/comments on intro section

@@
 
-We contribute to the growing literature of research based on player tracking data in American football.
-The National Football League (NFL) introduced their player tracking system known as [Next Gen Stats (NGS)](https://operations.nfl.com/gameday/technology/nfl-next-gen-stats/) in 2016, using radio frequency identification (RFID) chips placed in players' shoulder pads (and in the ball) to collect data 10 times per second.
+<!-- We contribute to the growing literature of research based on player tracking data in American football.  -->
+
+In an attempt to foster analytics and innovate the game, the National Football League (NFL) introduced their player tracking system known as [Next Gen Stats (NGS)](https://operations.nfl.com/gameday/technology/nfl-next-gen-stats/) in 2016, using radio frequency identification (RFID) chips placed in players' shoulder pads (and in the ball) to collect data 10 times per second.
 These NGS tracking data contain information about the on-field locations, speeds, and angles of all 22 players on the field (as well as the ball).
@@
 The [2023 edition of the NFL BDB](https://www.kaggle.com/competitions/nfl-big-data-bowl-2023) asked participants to evaluate lineman on passing plays [@Howard2023NFL].
+
+[I think we need a couple of paragraphs here to introduce the defensive linemen position, and the role/significance of linemen, why is it important to analyze them]
 

### 2023-05-08 paper/01-introduction.Rmd
better section title

@@
 
-## Previous Pass-Rush Measure in academic literaure
+<!-- ## Previous Pass-Rush Measure in academic literaure -->
 
-The academic literature on the performance of linemen, either offensive or defensive in football is scant.
+## Previous Research on Football Linemen
+
+The literature on the performance of linemen, either offensive or defensive in football is scant.
 @AlamarGould2008 studied the the relationship between pass completion rate and performance of offensive linemen in terms of how successful they were at blocking.

### 2023-05-08 paper/01-introduction.Rmd
added figures to the data section

@@
 
-[I think we need a couple of paragraphs here to introduce the defensive linemen position, and the role/significance of linemen, why is it important to analyze them]
+[I think we need a couple of paragraphs here to introduce the defensive linemen position, and the role/significance of linemen (kinda similar to section 2 of rishav's paper https://arxiv.org/pdf/1906.11373.pdf)]
 

### 2023-05-08 paper/02-data.Rmd
added figures to the data section

--- 2023-04-11 paper/02-data.Rmd
@@
 
-We use the tracking data made available in the 2023 NFL BDB competition for our investigation.
+```{r eval = FALSE, echo = FALSE}
+pass_rush <- read_csv(here("data", "pass_rush.csv.gz"))
+pass_rush %>% select(gameId, playId, frameId) %>% n_distinct()
+```
+
+In the forthcoming analysis, we rely on the data from the 2023 NFL Big Data Bowl provided by the NFL Next Gen Stats tracking system.
 The data corresponds to 8557 passing plays across 122 games in the first eight weeks of the 2021 NFL regular season. 
-The aforementioned NGS data provides the on-field location, speed, direction, and orientation of each player on the field (and the ball) at a rate of 10 Hz, along with event annotations (e.g., ball snapped, pass thrown, etc.).
-In this particular dataset, there are _INSERT NUMBER_ unique frames corresponding to moments of time from the start of the play at snap until the moment the quarterback either throws the pass or is sacked.
-(_VERIFY THIS STATEMENT MAKES SENSE_)
+For each play, we have information on the on-field location, speed, angle, direction, and orientation of each player on the field and the football at a rate of 10 Hz, along with event annotations for each frame such as ball snap, pass forward, and quarterback sack, to name a few.
+For our investigation, we consider only the frames between the ball snap and when a pass forward or quarterback sack is recorded for each play.
+We also remove all plays with multiple quarterbacks on the field, since we need a uniquely defined quarterback to compute our metric.
+After preprocessing, there are 251,060 unique frames corresponding to moments of time from the start of the play at snap until the moment the quarterback either throws the pass or is sacked.
 
-Table \ref{tab:track} displays example data for a play from the 2021 NFL regular season week six matchup between the Las Vegas Raiders and Denver Broncos, which ended with Broncos QB Teddy Bridgewater getting sacked by Raiders DE Maxx Crosby[^crosbysack].
-_Figure X_ presents the locations of every Las Vegas (black) and Denver (orange) player on the field at four particular moments of time, with Maxx Crosby highlighted (blue).
+Table \ref{tab:track} displays a tracking data example for a play from the 2021 NFL regular season week six matchup between the Las Vegas Raiders and Denver Broncos, which ended with Broncos QB Teddy Bridgewater getting sacked by Raiders DE Maxx Crosby[^crosbysack].
+In addition, Figure \ref{fig:fig_field_raw} presents the locations of every Las Vegas (in black) and Denver (in orange) player on the field from this play at 1, 2, 3 and 4 seconds after the ball snap, with Maxx Crosby highlighted in blue.
 
 \begin{table}
-\caption{track. \label{tab:track}}
+\caption{Example of tracking data for the play that resulted in a Maxx Crosby's sack. \label{tab:track}}
 \centering
-\begin{tabular}{rrrlrrrrrrrl}
+\begin{tabular}{ccccccccccc}
 \hline
@@
 
-_INSERT FIGURE_
+```{r}
+#| label: fig_field_raw
+#| fig.cap: "A display of the player tracking data for a play during the Las Vegas Raiders (defense, in black) versus Denver Broncos (offense, in orange) NFL game on October 17, 2021.
+Raiders DE Maxx Crosby is highlighted in blue.
+Snapshots are captured at 1, 2, 3, and 4 seconds after the ball snap."
+#| fig.width: 6.6
+#| fig.height: 4.4
+source(here("scripts", "11_fig_field_raw.R"))
+fig_field_raw
+```
 
 Along with the tracking information, the 2023 BDB includes scouting data provided by Pro Football Focus (PFF). 
-PFF's scouting data includes manually collected player-level information at the play level, such as the player's role (e.g., if they are a pass rusher) and credited events (e.g., player is credited with hitting the QB on the play).
-In this manuscript, we use the PFF data to identify _INSERT NUMBER_ unique players designated in the "pass rush" role across all plays.
-For context, _Figure X_ displays the distribution of the number of observed pass rushers involved in a play, ranging from _INSERT NUMBER_ to _INSERT NUMBER_ with _INSERT NUMBER_ pass rushers as the most common value.
+PFF's scouting data includes manually collected player-level information at the play level, such as the player's role (e.g., whether they are a pass rusher) and credited events (e.g., player is credited with hitting the QB on the play).
+In this manuscript, we use the PFF data to identify 36,362 unique players designated in the "pass rush" role across all plays.
+For context, Figure \ref{fig:pass_rushers_count} displays the distribution of the number of observed pass rushers involved in a play, ranging from 1 to 8 with 4 pass rushers as the most common value.
 We also leverage the PFF data to count how many hits, hurries, and sacks each pass rusher is credited with across span of collected data. 
-Additionally, we use the PFF player roles to identify which players were labeled as blocking each pass rusher to adjust for opponent strength, as discussed in Section _INSERT NUMBER_.
+Additionally, we use the PFF player roles to identify which players were labeled as blocking each pass rusher to adjust for opponent strength, as discussed in Section \ref{sec:methods}.
 
-INSERT FIGURE
-
-(sample restriction)
-
-We include only the frames between the ball snap and when a pass forward or QB sack is recorded for each play.
-We also remove all plays with multiple quarterbacks on the field, since we need a uniquely defined quarterback to calculate our measure.
-Using PFF scouting data, we only consider defenders with role "Pass rush" for evaluation.
+```{r}
+#| label: pass_rushers_count
+#| fig.cap: "Distribution of the number of pass rushers in a play."
+#| fig.width: 4.5
+#| fig.height: 3
+source(here("scripts", "12_fig_pass_rushers_count.R"))
+pass_rushers_count
+```

### 2023-05-08 scripts/11_fig_field_raw.R

@@ -1,0 +1,97 @@
+library(tidyverse)
+library(here)
+
+play_lv_den <- read_csv(here("data", "play_lv_den.csv.gz"))
+
+# play_lv_den |> filter(nflId == 47889) |>
+#   select(frameId, jerseyNumber, team, x:event) |>
+#   filter(frameId %in% c(7, 8, 50)) |>
+#   xtable::xtable()
+
+plays <- read_csv(here("data", "plays.csv"))
+desc <- plays |>
+  filter(gameId == 2021101709 & playId == 1444) |>
+  pull(playDescription) |> 
+  str_replace("\\)\\.", "\\)")
+
+plot_field_raw <- function(frame) {
+  plot_df <- filter(play_lv_den, frameId_snap_corrected == frame)
+  field <- ggplot() + 
+    annotate("text", 
+             x = seq(40, 80, 10),
+             y = 12,
+             color = "#bebebe",
+             size = rel(3),
+             label = 10 * c(3:5, 4:3)) +
+    annotate("text", 
+             x = seq(40, 80, 10),
+             y = 38.35,
+             color = "#bebebe",
+             size = rel(3),
+             angle = 180,
+             label = 10 * c(3:5, 4:3)) +
+    annotate("segment", 
+             x = setdiff(seq(35, 85, 1), seq(35, 85, 5)),
+             xend = setdiff(seq(35, 85, 1), seq(35, 85, 5)),
+             y = -Inf,
+             yend = 0,
+             color = "#bebebe") +
+    annotate("segment", 
+             x = setdiff(seq(35, 85, 1), seq(35, 85, 5)),
+             xend = setdiff(seq(35, 85, 1), seq(35, 85, 5)),
+             y = 48.7,
+             yend = Inf,
+             color = "#bebebe") +
+    annotate("text", 
+             x = setdiff(seq(35, 85, 1), seq(35, 85, 5)),
+             y = 23.36667,
+             color = "#bebebe",
+             label = "|",
+             size = rel(2)) +
+    annotate("text", 
+             x = setdiff(seq(35, 85, 1), seq(35, 85, 5)),
+             y = 29.96667,
+             color = "#bebebe",
+             label = "|",
+             size = rel(2)) +
+    annotate("segment", 
+             x = 35,
+             xend = 85,
+             y = c(-Inf, Inf),
+             yend = c(-Inf, Inf),
+             color = "#bebebe") +
+    geom_vline(xintercept = seq(35, 85, 5), color = "#bebebe") +
[…]

### 2023-05-08 scripts/12_fig_pass_rushers_count.R

@@ -1,0 +1,19 @@
+library(tidyverse)
+theme_set(theme_light())
+library(here)
+
+pff_full <- read_csv(here("data", "pffScoutingData.csv"))
+
+# pff_full |> 
+#   filter(pff_role == "Pass Rush") |> 
+#   nrow()
+
+pass_rushers_count <- pff_full |> 
+  filter(pff_role == "Pass Rush") |> 
+  count(gameId, playId) |> 
+  count(n) |> 
+  ggplot(aes(n, nn)) +
+  geom_col() +
+  scale_x_continuous(breaks = 1:8) +
+  labs(x = "Number of pass rushers",
+       y = "Count")

### 2023-05-08 paper/02-data.Rmd
tracking data table caption

@@
 \begin{table}
-\caption{Example of tracking data for the play that resulted in a Maxx Crosby's sack. \label{tab:track}}
+\caption{Example of tracking data for a play during the Las Vegas Raiders versus Denver Broncos NFL game on October 17, 2021.
+The data shown here is for Raiders DE Maxx Crosby, and the frames included are between the ball snap and when the sack by Crosby is recorded.
+\label{tab:track}}
 \centering

### 2023-05-08 paper/03-methods.Rmd
outlined stability section
average strain, unsuccessful pass rush

@@
 
-## Average STRAIN
+Moreover, since we observe STRAIN continuously at every tenth of a second within each play, we can then compute the average STRAIN across all frames played for every pass rusher.
+This can be helpful in regards to player evaluation, as we can determine the most effective pass-rushers based on their average STRAIN values.
+Formally, the average STRAIN, denoted by $\overline{\text{STRAIN}}$, for pass rusher $i$ over $n$ total plays with $T$ total frames is
+
+[CHECK THIS - is notation consistent with what we have above?]
+\begin{equation}
+\overline{\text{STRAIN}}_{i} = \frac{1}{n} \sum_{t=1}^T {\text{STRAIN}}_{i}(t).
+\end{equation}
 

### 2023-05-08 paper/04-results.Rmd
added figures to the data section
average strain, unsuccessful pass rush

@@
 
-To illustrate our proposed metric STRAIN for pass rush evaluation, we use a play from the 2021 NFL regular season week 6 matchup between the Las Vegas Raiders and Denver Broncos, which ended with Broncos QB Teddy Bridgewater getting sacked by Raiders DE Maxx Crosby[^crosbysack].
-Figure \ref{fig:fig_field} presents the locations of every Las Vegas (black) and Denver (orange) player on the field at particular time points in this play, with our focused player Maxx Crosby highlighted in blue.
-Note that the point size for each Las Vegas defender corresponds to the estimated STRAIN in the selected frames as the play progresses.
+To illustrate our proposed metric STRAIN for pass rush evaluation, we use the same play from the Las Vegas Raiders and Denver Broncos game as mentioned in Section \ref{sec:data}.
+Figure \ref{fig:fig_field} shows an updated version of Figure \ref{fig:fig_field_raw}, with the point size for each Las Vegas defender corresponding to the estimated STRAIN in the selected frames as the play progresses.
 In addition, Figure \ref{fig:fig_field} is accompanied by Figure \ref{fig:fig_crosby_curves}, which is a line graph showing how Crosby's distance from the quarterback, velocity, and STRAIN change continuously throughout the play.
@@
 
+```{r}
+#| label: fig_crosby_curves_unsucess
+#| fig.cap: "strain curve for unsuccessful play"
+#| fig.width: 4.8
+#| fig.height: 2.7
+source(here("scripts", "09_fig_play_unsuccess.R"))
+fig_crosby_curves_unsucess
+```
+
 (bring in the unsuccessful pass rush play)
-
-(talk a bit about other rushers, maybe connect this with positional curves broken down by outcome - sack.)
 
@@
 #| fig.cap: "STRAIN curves for different positions by outcome"
-#| fig.width: 6
-#| fig.height: 6
+#| fig.width: 6.5
+#| fig.height: 2.7
 source(here("scripts", "08_fig_outcome_curves.R"))
@@
 
-There are a variety of quantities we're interested in from the multilevel model fit in Section [INSERT SECTION REF].
+There are a variety of quantities we're interested in from the multilevel model fit in Section \ref{sec:methods}.
+
+[INSERT SECTION REF].
 First, Table [FIXED EFFECTS REF] displays the results for the fixed effects terms.

### 2023-05-08 scripts/08_fig_outcome_curves.R

@@ -72,8 +72,9 @@ fig_outcome_curves <- hit |>
   labs(y = "STRAIN",
        x = "Time since snap (seconds)") +
   theme_light() +
-  facet_wrap(~ outcome) +
+  facet_wrap(~ outcome, nrow = 1) +
   theme(legend.key.width = unit(0.9, "cm"),
+        legend.position = "bottom",
         axis.title = element_text(size = rel(1)),
         axis.text = element_text(size = rel(0.8)),
         legend.title = element_text(size = rel(1)),

### 2023-05-08 scripts/09_fig_play_unsuccess.R

@@ -12,11 +12,13 @@ desc <- plays |>
 
 # df_cleaned <- read_csv(here("data", "df_cleaned.csv.gz"))
 
-play_unsucess <- df_cleaned |>
-  filter(gameId == 2021101709 & playId == 1113)
+# play_unsucess <- df_cleaned |>
+#   filter(gameId == 2021101709 & playId == 1113)
 # write_csv(play_unsucess, here("data", "play_unsucess.csv.gz"))
 
 # https://www.nfl.com/videos/teddy-bridgewater-courtland-sutton-barely-miss-out-on-chance-at-62-yard-td-bomb
+
+play_unsucess <- read_csv(here("data", "play_unsucess.csv.gz"))
 
 crosby <- play_unsucess |> 
   filter(nflId == 47889) |>

### 2023-05-08 scripts/11_fig_field_raw.R

@@ -62,16 +62,20 @@ plot_field_raw <- function(frame) {
              color = "#bebebe") +
     geom_vline(xintercept = seq(35, 85, 5), color = "#bebebe") +
     geom_point(aes(x = x, y = y),
+               size = 0.5,
                color = "black",
                data = filter(plot_df, team == "LV" & nflId != 47889)) +
     geom_point(aes(x = x, y = y),
                color = "#1143E2",
+               size = 0.5,
                data = filter(plot_df, team == "LV" & nflId == 47889)) +
     geom_point(aes(x = x, y = y),
                color = "#FB4F14",
+               size = 0.5,
                data = filter(plot_df, team == "DEN")) +
     geom_point(aes(x = x, y = y),
                color = "#803621",
+               size = 0.5,
                shape = 20,
                data = filter(plot_df, team == "football")) +
     scale_size_identity() +

### 2023-05-08 paper/04-results.Rmd
added more figures and tables for the results section

@@
 
-(bring in the unsuccessful pass rush play)
+unsuccessful pass rush play: https://www.nfl.com/videos/teddy-bridgewater-courtland-sutton-barely-miss-out-on-chance-at-62-yard-td-bomb
+
+Figure \ref{fig:fig_crosby_curves_unsucess}
 
@@
 
+```{r}
+#| label: fig_predictability
+#| fig.cap: "strain is more predictive of pressure rate than pressure rate itself."
+#| fig.width: 4
+#| fig.height: 3
+#| fig.show: "hold"
+source(here("scripts", "07_fig_predictability.R"))
+hsh_hsh
+strain_hsh
+```
+
 Figure \ref{fig:fig_stability} is a scatterplot of average STRAIN across all frames played in first and last 4 weeks.
 It is apparent that there is a strong correlation (r = 0.854), illustrating great metric stability over time.
+
+Figure \ref{fig:fig_predictability} (0.09651732 vs 0.3216974)
 
@@
 There are a variety of quantities we're interested in from the multilevel model fit in Section \ref{sec:methods}.
-
-[INSERT SECTION REF].
-First, Table [FIXED EFFECTS REF] displays the results for the fixed effects terms.
-[INSERT CONCLUSIONS ABOUT THESE VARIABLES]  Next, Table [INSERT ICC TABLE REF] displays the intraclass correlation coefficients (ICC) for the three different grouped terms, describing the proportion variance explained between the group terms in comparison to the residual variance.
+First, Table \ref{tab:fixedeff} displays the results for the fixed effects terms.
+[INSERT CONCLUSIONS ABOUT THESE VARIABLES]
+Next, Table \ref{tab:icc} displays the intraclass correlation coefficients (ICC) for the three different grouped terms, describing the proportion variance explained between the group terms in comparison to the residual variance.
 While the residual variance is unsurprisingly the largest value, between the three group terms, we observe the largest ICC for pass rushers, followed by the QB, and the blocker.
@@
 
-We display the random intercepts for pass rusher in Table [INSERT TABLE REFERENCE] alongside their unadjusted STRAIN values.
-We observe similar rankings based on the random intercepts in comparison to the average STRAIN, with Myles Garrett (DE), Aaron Donald (DT), Vita Vea (NT), and T.J.
-Watt (OLB) as the top pass rushers.
+\begin{table}
+\caption{Fixed effects for the multilevel model. \label{tab:fixedeff}}
+\centering
+\begin{tabular}{rrrrrr}
+  \hline
+ & estimate & se & df & $t$-statistic & $p$-value \\ 
+  \hline
+Intercept & 0.355 & 0.010 & 332.288 & 35.688 & 0.000 \\ 
+  play\_len & -0.004 & 0.000 & 30653.793 & -38.214 & 0.000 \\ 
+  rush\_posDT & -0.081 & 0.008 & 369.713 & -10.340 & 0.000 \\ 
+  rush\_posinteriorLB & 0.058 & 0.011 & 770.441 & 5.332 & 0.000 \\ 
+  rush\_posNT & -0.106 & 0.012 & 355.999 & -9.173 & 0.000 \\ 
+  rush\_posOLB & 0.062 & 0.008 & 388.506 & 7.695 & 0.000 \\ 
+  rush\_possecondary & 0.145 & 0.012 & 1734.779 & 12.328 & 0.000 \\ 
+  block\_posG & 0.011 & 0.006 & 123.672 & 1.867 & 0.064 \\ 
+  block\_posother & 0.010 & 0.008 & 500.451 & 1.261 & 0.208 \\ 
+  block\_posT & 0.023 & 0.006 & 161.556 & 3.625 & 0.000 \\ 
+  n\_blockers & -0.064 & 0.003 & 30949.514 & -24.788 & 0.000 \\ 
+   \hline
+\end{tabular}
+\end{table}
+
+\begin{table}[ht]
+\caption{ICC. \label{tab:icc}}
+\centering
+\begin{tabular}{rlr}
+\hline
+Group & ICC \\ 
+\hline
+nflId & 0.057 \\ 
+blocker\_id & 0.012 \\ 
+passer & 0.023 \\ 
+residual & 0.907 \\ 
+\hline
+\end{tabular}
+\end{table}
+
+<!-- We display the random intercepts for pass rusher in Table [INSERT TABLE REFERENCE] alongside their unadjusted STRAIN values.  -->
+
 In addition to adjusting for opposition, our multilevel model allows us to provide some notion of uncertainty about the pass rusher's effect on the play level STRAIN.
 Figure \ref{fig:fig_rankings_boot} displays the mean varying intercept for the top ten pass rushers for each position along with 95% confidence intervals generated by sampling from the posterior distributions via parametric bootstrapping using the `merTools` package [@knowlesmerTools, @gelman2006data].
-We observe different levels of variability across these top players, with the only a subset of players with intervals strictly above zero.
-[NEED TO THINK ABOUT WHAT MORE TO SAY HERE]
-
-coefficient table
-
-MAKE THIS A TABLE
-icc
-rusher 0.0574
-blocker 0.0121
-passer 0.0233
-Residual 0.907
+We observe similar rankings based on the random intercepts in comparison to the average STRAIN, with Myles Garrett (DE), Aaron Donald (DT), Vita Vea (NT), and T.J. Watt (OLB) as the top pass rushers. 
+We observe different levels of variability across these top players, with the only a subset of players with intervals strictly above zero.  [NEED TO THINK ABOUT WHAT MORE TO SAY HERE]
 

### 2023-05-08 scripts/07_fig_predictability.R

@@ -52,7 +52,7 @@ hsh_last <- hsh |>
 
 
 #0.09651732
-hsh_first |> 
+hsh_hsh <- hsh_first |> 
   full_join(hsh_last) |>
   left_join(select(players, nflId, officialPosition)) |>
   mutate(officialPosition = factor(officialPosition, levels = c("OLB", "DE", "DT", "NT"))) |> 
@@ -73,7 +73,7 @@ hsh_first |>
 
 #0.3216974
 
-hsh_last |> 
+strain_hsh <- hsh_last |> 
   left_join(select(players, nflId, officialPosition, displayName)) |> 
   select(name = displayName, pos = officialPosition, last_hsh) |> 
   full_join(strain_first) |>

### 2023-05-08 scripts/10_multilevel_mod.R

@@ -131,6 +131,9 @@ strain_fit <- lmer(
   data = mod_df_final
 )
 
+# library(lmerTest)
+# xtable(summary(strain_fit)$coef, digits = rep(3, 6))
+
 
 # fit_out_rush <- strain_fit |> 
 #   ranef() |> 

### 2023-05-09 paper/01-introduction.Rmd
figures + tables formatting

@@
 
-Pro football focus (PFF) defines pass-rush productivity as: 
-$$
-\frac{\frac{1}{2}(\# Hurries + \# Hits) + \# Sacks}{\#PassRush Snaps}
-$$.
-
+Pro football focus (PFF) defines pass-rush productivity as 
+$$[ \ (\text{Hurries} + \text{Hits})/2 + \text{Sacks} \ ]\  / \text{ Pass Rush Snaps}$$
 This metric gives twice as much weight to a sack relative to hurries and hits, which is a slight improvement, but the choice of weights is ad-hoc an still only measures binary outcomes similar to the shortcomings of previously discussed metrics.
@@
 \caption{A summary of previously-existing pass rush metrics. \label{tab:metrics}}
-\begin{center}
-\begin{tabular}{ll}
-\textbf{Metric} & \textbf{Description}\\\hline
-Sacks & A defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass \\
-Hits & A defensive player tackles the quarterback behind the line of scrimmage after the quarterback is able to throw a pass. \\
-Hurries & A defensive player pressures the quarterback behind the line of scrimmage such that the quarterback must release the ball prior sooner than they wanted to. \\
-QB Pressures & A sack, hit, or hurry. \\
-Pass-Rush Productivity (PRP) & $\frac{\frac{(\# Hurries + \# Hits)}{2} + \# Sacks}{\#PassRush Snaps}$\\
-Time In Pocket (TIP) & The time from snap to throw or pocket collapse for a quarterback \\
-Pass Rush Win Rate (PRWR) & Rate at which pass rusher beats pass block within 2.5 seconds after snap\\\hline
+\centering
+\begin{tabular}{p{0.27\textwidth}p{0.67\textwidth}}
+  \hline
+Metric & Description \\ 
+  \hline
+Sacks & A defender tackles the QB behind the line of scrimmage before the QB is able to throw a pass \\
+Hits & A defender tackles the QB behind the line of scrimmage after the QB is able to throw a pass. \\
+Hurries & A defender pressures the QB behind the line of scrimmage such that the QB must release the ball prior sooner than they wanted to. \\
+Pressures & $\text{Hurries} + \text{ Hits } + \text{ Sacks}$ \\
+Pass-Rush Productivity & $[\ (\text{Hurries} + \text{Hits})/2 + \text{Sacks}\ ]\  / \text{ Pass Rush Snaps}$\\
+Time In Pocket & Time (in seconds) between ball snap and throw or pocket collapse for a QB \\
+Pass Rush Win Rate & Rate at which pass rusher beats pass block within 2.5 seconds after ball snap\\
+   \hline
 \end{tabular}
-\end{center}
 \end{table}
+
+[Pro football ref has different QB pressures formula? https://www.pro-football-reference.com/about/advanced_stats.htm]
 

### 2023-05-09 paper/04-results.Rmd
figures + tables formatting

@@
 #| fig.cap: "strain is more predictive of pressure rate than pressure rate itself."
-#| fig.width: 4
+#| fig.width: 6.5
 #| fig.height: 3
@@
 source(here("scripts", "07_fig_predictability.R"))
-hsh_hsh
-strain_hsh
+fig_predictability
 ```

### 2023-05-09 scripts/07_fig_predictability.R

@@ -67,6 +67,7 @@ hsh_hsh <- hsh_first |>
   theme(axis.title = element_text(size = rel(1)),
         legend.title = element_text(size = rel(1)),
         legend.text = element_text(size = rel(0.8)),
+        legend.position = "none",
         axis.text = element_text(size = rel(0.8)),
         panel.grid.minor = element_blank())
 
@@ -83,7 +84,7 @@ strain_hsh <- hsh_last |>
   geom_point(aes(color = pos, group = name), size = 2, alpha = 0.8) +
   scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107", "#004D40")) +
   labs(x = "STRAIN in first 4 weeks",
-       y = "(Hurries + Sacks + Hits) per snap in last 4 weeks",
+       y = "(Hurries + Sacks + Hits) per snap\nin last 4 weeks",
        color = "Position") +
   theme_light() +
   theme(axis.title = element_text(size = rel(1)),
@@ -91,5 +92,7 @@ strain_hsh <- hsh_last |>
         legend.text = element_text(size = rel(0.8)),
         axis.text = element_text(size = rel(0.8)),
         panel.grid.minor = element_blank())
+
+fig_predictability <- cowplot::plot_grid(hsh_hsh, strain_hsh, rel_widths = c(2.4, 3.6))
 
 

### 2023-05-09 paper/00-main.Rmd
figures + tables formatting
more work

@@
 bibliography: bibliography.bib
+biblio-style: chicago
 output: rticles::asa_article
-header-includes:
-  \renewcommand{\arraystretch}{0.9}
+# header-includes:
+#   \renewcommand{\arraystretch}{0.8}
 ---
@@
   warning = FALSE,
-  fig.align = "center"
+  fig.align = "center",
+  cache = TRUE
 )

### 2023-05-09 paper/01-introduction.Rmd
more work

@@
 
-Officially tracked by the NFL since 1982, the most well known statistic for evaluating defensive linemen on pass rushing plays is the sack, which occurs when a defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass.
+Table \ref{tab:metrics} gives a summary of existing pass-rush metrics.
+
+Officially tracked by the NFL since 1982, the most well-known statistic for evaluating defensive linemen on pass rushing plays is the sack, which occurs when a defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass.
 Sacks represent a simple binary measure of success for a rusher on a pass play.
@@
 
-The sum of sacks, hits, and hurries has been defined as "pressures" (https://www.pro-football-reference.com/years/2022/advanced.htm) and is an improvement over sacks, but suffers from problems of subjectivity (e.g.
+<!-- (https://www.pro-football-reference.com/years/2022/advanced.htm) -->
+
+[Pro football ref has a slightly different QB pressures formula than the formula below https://www.pro-football-reference.com/about/advanced_stats.htm]
+
+The sum of sacks, hits, and hurries has been defined as "pressures" and is an improvement over sacks, but suffers from problems of subjectivity (e.g.
 was the play actually a hurry?) and are still binary outcomes (i.e.
@@
 
-Pro football focus (PFF) defines pass-rush productivity as 
-$$[ \ (\text{Hurries} + \text{Hits})/2 + \text{Sacks} \ ]\  / \text{ Pass Rush Snaps}$$
-This metric gives twice as much weight to a sack relative to hurries and hits, which is a slight improvement, but the choice of weights is ad-hoc an still only measures binary outcomes similar to the shortcomings of previously discussed metrics.
-(https://www.pro-football-reference.com/about/advanced_stats.htm)
+<!-- $$\frac{(\text{Hurries} + \text{Hits})/2 + \text{Sacks}}{\text{ Pass Rush Snaps}}$$ -->
 
-More recently metrics such as Time In Pocket (TIP) and Pass Rush Win Rate [@Burke2018Created] (PRWR) have also been proposed and are substantial improvements over the less sophisticated counting statistics.
-However, these are still imperfect measures as TIP is highly context dependent (e.g.
-deep vs shallow passing routes).
+Pro football focus (PFF) comes up with a metric called "pass-rush productivity" (PRP), which is a minor modification from the aforementioned pressures metric (see Table \ref{tab:metrics}).
+In particular, PRP gives twice as much weight to a sack relative to hurries and hits, which is a slight improvement to pressures.
+However, the choice of weights is ad-hoc and still only measures binary outcomes similar to the shortcomings of previously discussed metrics.
+
+<!-- (https://www.pro-football-reference.com/about/advanced_stats.htm) -->
+
+More recently metrics such as Time In Pocket (TIP) and Pass Rush Win Rate (PRWR) [@Burke2018Created] have also been proposed and are substantial improvements over the less sophisticated counting statistics.
+Nevertheless, these are still imperfect measures.
+TIP is highly context dependent, as it can be influenced by the type of passing route (e.g. deep versus shallow).
 PRWR is based on whether a pass rusher is able to beat their block before a fixed time (ESPN chooses 2.5 seconds) from the snap, but this is highly dependent on the rather arbitrary choice of time used to define a "win".
@@
 
-(https://www.espn.com/nfl/story/_/id/34536376/2022-nfl-pass-rushing-run-stopping-blocking-leaderboard-win-rate-rankings-top-players-teams)
+<!-- (https://www.espn.com/nfl/story/_/id/34536376/2022-nfl-pass-rushing-run-stopping-blocking-leaderboard-win-rate-rankings-top-players-teams) -->
 
 All of these measures represent different versions of "the best we can do with the available data".
-Without more sophisticated data (i.e.
-player tracking data), it is difficult to measure effectiveness of a pass rusher so many simple metrics of success have been proposed.
-Table \ref{tab:metrics} gives a summary of existing pass-rush metrics.
+Without granular data, it is difficult to measure effectiveness of a pass rusher so many simple metrics of success have been proposed.
 
@@
 Pressures & $\text{Hurries} + \text{ Hits } + \text{ Sacks}$ \\
-Pass-Rush Productivity & $[\ (\text{Hurries} + \text{Hits})/2 + \text{Sacks}\ ]\  / \text{ Pass Rush Snaps}$\\
+Pass-Rush Productivity & $\displaystyle \frac{(\text{Hurries} + \text{Hits})/2 + \text{Sacks}}{\text{ Pass Rush Snaps}}$\\
 Time In Pocket & Time (in seconds) between ball snap and throw or pocket collapse for a QB \\
@@
 \end{table}
-
-[Pro football ref has different QB pressures formula? https://www.pro-football-reference.com/about/advanced_stats.htm]
 
@@
 The literature on the performance of linemen, either offensive or defensive in football is scant.
-@AlamarGould2008 studied the the relationship between pass completion rate and performance of offensive linemen in terms of how successful they were at blocking.
-The data for this study was collecting essentially by hand by watching and charting the performance of each lineman on a play.
-This was done using games from the beginning 2007 season for seven teams and their first three games each.
-The authors note that there is not much research on linemen saying "One of the principal reasons for this lack of study is the absence of publicly available data that is needed in order to track the performance of a specific lineman."
+@AlamarGould2008 examine the the relationship between pass completion and offensive linemen's performance, in terms of how successful they were at blocking.
+The data for this study are manually collected for the first three weeks of the 2007 NFL season, recording whether a lineman holds a block and the time in pocket for the quarterback.
 
-Later, @AlamarGoldner2011 again studied offensive linemen performance.
-They analyzed 20 teams during the 2010 season for a total of 1,259 plays.
-They don't estimate the performance of an individual lineman, but rather positions on a team (e.g.
-Chicago Bears' left tackle, Pittsburgh Steelers' center, etc.).
+Later, @AlamarGoldner2011 use manually-tracked data during the 2010 season to estimate lineman performance for different team-positions instead of individual defenders (e.g.
+Chicago Bears' left tackle, Pittsburgh Steelers' center).
+[this next part needs re-wording]
 They then go on to use a survival analysis model to estimate time in pocket.

### 2023-05-09 paper/03-methods.Rmd
more work

@@
 
-While average STRAIN appears to be positively correlated with pass rusher ability, the observed average STRAIN on a single play is likely due to a convolution of factors.
+<!-- While average STRAIN appears to be positively correlated with pass rusher ability -->
+
+While average STRAIN is a simple first step for evaluating the pass rush, the observed average STRAIN on a single play is likely due to a convolution of factors.
 Outside of the pass rusher's ability, there is variability in the opposing strength of pass blockers across plays a pass rusher is involved.
@@
 
-We use a multilevel model (i.e., hierarchical, random-effects, or mixed effects model) to evaluate pass rushers' impact on the average STRAIN observed in a play, while accounting for their pass blocker, QB, and play-level information.
+To this end, we fit a multilevel model to evaluate pass rushers' impact on the average STRAIN observed in a play, while accounting for their pass blocker, QB, and play-level information.
 We identify the pass blocker linked with the pass rusher of interest using the data provided by PFF.
@@
 The coefficients $\beta$ for additional information $X_{j,i}$ about pass rusher $j$ in play $i$ are estimated as fixed-effects. 
-Our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows:
+Our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows.
+\begin{equation}
+\begin{aligned}
+\overline{\text{STRAIN}}_{j,i} &\sim N(R_{r[j,i]} + B_{b[j,i]} + Q_{q[i]} + X_{j,i} \cdot \beta, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
+R_{r} &\sim N(\mu_R, \sigma^2_R), \text{ for } r = 1, \cdots, \text{ \# of pass rushers}, \\
+B_{b} &\sim N(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
+Q_{q} &\sim N(\mu_Q, \sigma^2_Q), \text{ for } q = 1, \cdots, \text{ \# of QBs}.
+\end{aligned}
+\end{equation}
 
-$$
-\begin{aligned}
-\overline{STRAIN}_{j,i} &\sim Normal(R_{r[j,i]} + B_{b[j,i]} + Q_{q[i]} + X_{j,i} \cdot \beta, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
-R_{r} &\sim Normal(\mu_R, \sigma^2_R), \text{ for } r = 1, \cdots, \text{ \# of pass rushers}, \\
-B_{b} &\sim Normal(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
-Q_{q} &\sim Normal(\mu_Q, \sigma^2_Q), \text{ for } q = 1, \cdots, \text{ \# of QBs}.
-\end{aligned}
-$$
+We consider a normal distribution to obtain the random intercepts for each player toward their respective position group means.
+This provides us with average individual effects on STRAIN for each player while applying shrinkage, which is a useful property in this problem's context since we do not observe the same number of plays for each player.
+With regards to the fixed-effects about pass rusher $j$ in play $i$, we include the length of the play, the pass rusher position, the nearest pass blocker's position, and the number of pass blockers assigned to the pass rusher based on the PFF data.
+Table \ref{tab:pos} shows our categorization of position for the pass rush and pass block roles.
+We implement the model using penalized likelihood via the `lme4` package in `R` [@bateslme4; @R2023Language].
 
-We use a Normal distribution to pull the random intercepts for each player toward their respective position group means.
-This provides us with average individual effects on STRAIN for each player while applying shrinkage, which is a useful property in this problem's context since we do not observe the same number of plays for each player.
-With regards to the fixed-effects about pass rusher $j$ in play $i$, we include the length of the play, the pass rusher position (INSERT POSITIONS), the nearest pass blocker's position (INSERT POSITIONS), and the number of pass blockers assigned to the pass rusher according to PFF.
-We fit the model using penalized likelihood via the `lme4` package in `R` [@bateslme4].
+\begin{table}
+\caption{Position groupings for pass rushers and blockers. \label{tab:pos}}
+\centering
+\begin{tabular}{ll}
+\hline
+Role & Position \\ 
+\hline
+Pass rush & Defensive end \\ 
+Pass rush & Defensive tackle \\ 
+Pass rush & Nose tackle \\ 
+Pass rush & Outside linebacker \\ 
+Pass rush & Interior linebacker (middle and inside linebackers) \\ 
+Pass rush & Secondary (cornerback, free safety, and strong safety) \\
+\hline
+Pass block & Center \\ 
+Pass block & Guard \\ 
+Pass block & Tackle \\ 
+Pass block & Other (tight end, running back, fullback, wide receiver) \\ 
+\hline
+\end{tabular}
+\end{table}
+
+```{r, eval=FALSE}
+rush <- tibble(
+  Role = "Pass rush",
+  Position = c("Defensive end", "Defensive tackle", "Nose tackle", "Outside linebacker",
+               "Interior linebacker (middle and inside linebackers)", 
+               "Secondary (cornerback, free safety, and strong safety)")
+)
+block <- tibble(
+  Role = "Pass block",
+  Position = c("Center", "Guard", "Tackle", "Other (tight end, running back, fullback, wide receiver)")
+)
+rush |> 
+  bind_rows(block) |> 
+  xtable()
+```

### 2023-05-09 paper/04-results.Rmd
more work

@@
 Figure \ref{fig:fig_field} shows an updated version of Figure \ref{fig:fig_field_raw}, with the point size for each Las Vegas defender corresponding to the estimated STRAIN in the selected frames as the play progresses.
-In addition, Figure \ref{fig:fig_field} is accompanied by Figure \ref{fig:fig_crosby_curves}, which is a line graph showing how Crosby's distance from the quarterback, velocity, and STRAIN change continuously throughout the play.
+This is accompanied by Figure \ref{fig:fig_crosby_curves}, which is a line graph showing how Crosby's distance from the quarterback, velocity, and STRAIN change continuously throughout the play.
 

### 2023-05-09 paper/01-introduction.Rmd
figure size, literature

@@
 The literature on the performance of linemen, either offensive or defensive in football is scant.
-@AlamarGould2008 examine the the relationship between pass completion and offensive linemen's performance, in terms of how successful they were at blocking.
-The data for this study are manually collected for the first three weeks of the 2007 NFL season, recording whether a lineman holds a block and the time in pocket for the quarterback.
-
+@AlamarGould2008 find an association between pass completion rate and successful pass blocking by offensive linemen.
+The data for this study are collected for the first three weeks of the 2007 NFL season, manually recording whether a lineman holds a block and the time it took for quarterback to throw the football.
 Later, @AlamarGoldner2011 use manually-tracked data during the 2010 season to estimate lineman performance for different team-positions instead of individual defenders (e.g.
 Chicago Bears' left tackle, Pittsburgh Steelers' center).
-[this next part needs re-wording]
-They then go on to use a survival analysis model to estimate time in pocket.
+The paper uses survival analysis to model time in pocket and completion percentage for quarterbacks, before ultimately measures a lineman's contribution to the overall offense in terms of yards gained.
+
+<!-- [this next part needs re-wording] -->
+<!-- They then go on to use a survival analysis model to estimate time in pocket.
 The average probability of success from the survival model is then used as a covariate in a model to estimate the probability of competing a pass, and they show that there is a significant relationship between these two probabilities.
+-->
 

### 2023-05-09 scripts/01_fig_field.R

@@ -36,8 +36,8 @@ lv_strain_range <- play_lv_den |>
 play_lv_den <- play_lv_den |>
   mutate(point_size = ifelse(
     team != "LV",
-    0.5,
-    scales::rescale(strain, to = c(0.5, 5), from = lv_strain_range)
+    1,
+    scales::rescale(strain, to = c(1, 6), from = lv_strain_range)
   ))
 
 plot_field <- function(frame) {

### 2023-05-09 scripts/11_fig_field_raw.R

@@ -62,20 +62,20 @@ plot_field_raw <- function(frame) {
              color = "#bebebe") +
     geom_vline(xintercept = seq(35, 85, 5), color = "#bebebe") +
     geom_point(aes(x = x, y = y),
-               size = 0.5,
+               size = 1,
                color = "black",
                data = filter(plot_df, team == "LV" & nflId != 47889)) +
     geom_point(aes(x = x, y = y),
                color = "#1143E2",
-               size = 0.5,
+               size = 1,
                data = filter(plot_df, team == "LV" & nflId == 47889)) +
     geom_point(aes(x = x, y = y),
                color = "#FB4F14",
-               size = 0.5,
+               size = 1,
                data = filter(plot_df, team == "DEN")) +
     geom_point(aes(x = x, y = y),
                color = "#803621",
-               size = 0.5,
+               size = 1,
                shape = 20,
                data = filter(plot_df, team == "football")) +
     scale_size_identity() +

### 2023-05-09 paper/04-results.Rmd
more results writing

@@
 
-[^crosbysack]: An real-game footage of this play is available at https://www.raiders.com/video/de-maxx-crosby-sacks-qb-teddy-bridgewater-for-a-loss-of-6-yards-nfl
-
 We observe that for the first two seconds, Crosby is being blocked by a Denver offensive lineman and unable to get close to the quarterback, hence the corresponding STRAIN values are virtually zero. 
@@
 
+In contrast, Figure \ref{fig:fig_crosby_curves_unsucess} shows the feature curves for an unsuccessful pass rush attempt by the Raiders defense from the same game.
+In this play, Broncos QB Teddy Bridgewater is well-protected by the offensive line and is able to release a long pass[^unsuccess].
+We see that the STRAIN generated by Crosby and other pass rushers are relative small for this play, compared to the previous play which resulted in a sack.
+
+[^unsuccess]: https://www.nfl.com/videos/teddy-bridgewater-courtland-sutton-barely-miss-out-on-chance-at-62-yard-td-bomb
+
 ```{r}
@@
 ```
-
-unsuccessful pass rush play: https://www.nfl.com/videos/teddy-bridgewater-courtland-sutton-barely-miss-out-on-chance-at-62-yard-td-bomb
-
-Figure \ref{fig:fig_crosby_curves_unsucess}
 
@@
 Rashan Gary, a young rusher with huge potential, ranks first among all pass-rushers according to our metric.
-Our leaderboards largely match the rankings of experts in the field (for instance, PFF's edge and interior rusher rankings after the 2021 season), lending credibility to our proposed metric as a measure of pass-rushing effectiveness.
-
-https://www.pff.com/news/nfl-2022-edge-rusher-rankings-tiers
-
-https://www.pff.com/news/nfl-interior-defensive-line-rankings-tiers-2022
+Our leaderboards largely match the rankings of experts in the field -- for instance, PFF's edge [@Monson2022nfl] and interior [@Linsey2022nfl] rusher rankings after the 2021 season, lending credibility to our proposed metric as a measure of pass-rushing effectiveness.
 
@@
 
+some sentences
+
 Next, we examine the predictability of STRAIN as a measure of pressure.
 We first look at how well strain correlates with a simple measure of pressure rate, defined as the total hits, sacks and hurries per snap.
-Figure \ref{fig:fig_cor_pressure} shows a scatterplot of average STRAIN and pressure rate.
-We observe a fairly strong correlation (r = 0.625)...
+Figure \ref{fig:fig_cor_pressure} is a scatterplot of average STRAIN and pressure rate which reveals a fairly strong correlation (r = 0.625) between the quantities.
+Note that we only consider players with at least 100 snaps played.
+Furthermore, the average STRAIN for the first 4 weeks of the 2021 season is more predictive of the last 4 weeks' pressure rate ($r=0.322$) than the first 4 weeks' pressure rate ($r=0.097$), as illustrated in Figure \ref{fig:fig_predictability}.
 
-Furthermore, first 4 weeks’ avg STRAIN is more predictive of last 4 weeks’ hurries sacks hits rate than first 4 weeks’ HSH rate
-
-(INSERT FIGURE)
+We also examine the stability of STRAIN over time by comparing the average STRAIN across all frames played in first and last four weeks of the 2021 NFL regular season.
+It is apparent from Figure \ref{fig:fig_stability} that there is a strong correlation $(r = 0.854)$ for this relationship, illustrating that STRAIN is a stable quantity over the provided 8-week window.
 
@@
 #| label: fig_cor_pressure
-#| fig.cap: "Corr with pressure rate = (hits + sacks + hurries) per snap"
+#| fig.cap: "Relationship between  with pressure rate = (hits + sacks + hurries) per snap"
 #| fig.width: 4
@@
 fig_cor_pressure
-```
-
-```{r}
-#| label: fig_stability
-#| fig.cap: "Metric stability"
-#| fig.width: 4
-#| fig.height: 3
-source(here("scripts", "05_fig_stability.R"))
-fig_stability
 ```
@@
 
-Figure \ref{fig:fig_stability} is a scatterplot of average STRAIN across all frames played in first and last 4 weeks.
-It is apparent that there is a strong correlation (r = 0.854), illustrating great metric stability over time.
-
-Figure \ref{fig:fig_predictability} (0.09651732 vs 0.3216974)
+```{r}
+#| label: fig_stability
+#| fig.cap: "Metric stability"
+#| fig.width: 4
+#| fig.height: 3
+source(here("scripts", "05_fig_stability.R"))
+fig_stability
+```
 

### 2023-05-09 scripts/04_fig_cor_pressure.R

@@ -40,7 +40,7 @@ fig_cor_pressure <- test_summed |>
   scale_x_continuous(breaks = seq(0.5, 2.5, 1)) +
   scale_y_continuous(breaks = seq(0, 0.12, 0.04)) +
   scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107", "#004D40")) +
-  labs(y = "(Hurries + Sacks + Hits) per snaps",
+  labs(y = "Pressure rate",
        x = "Average STRAIN across all frames",
        color = "Position") +
   theme_light() +

### 2023-05-09 scripts/07_fig_predictability.R

@@ -60,8 +60,8 @@ hsh_hsh <- hsh_first |>
   geom_smooth(method = "lm", se = FALSE, linewidth = 1.5, color = "black", alpha = 0.5) +
   geom_point(aes(color = officialPosition, group = nflId), size = 2, alpha = 0.8) +
   scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107", "#004D40")) +
-  labs(x = "(Hurries + Sacks + Hits) per snap in first 4 weeks",
-       y = "(Hurries + Sacks + Hits) per snap in last 4 weeks",
+  labs(x = "Pressure rate in first 4 weeks",
+       y = "Pressure rate per snap in last 4 weeks",
        color = "Position") +
   theme_light() +
   theme(axis.title = element_text(size = rel(1)),
@@ -83,8 +83,8 @@ strain_hsh <- hsh_last |>
   geom_smooth(method = "lm", se = FALSE, linewidth = 1.5, color = "black", alpha = 0.5) +
   geom_point(aes(color = pos, group = name), size = 2, alpha = 0.8) +
   scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107", "#004D40")) +
-  labs(x = "STRAIN in first 4 weeks",
-       y = "(Hurries + Sacks + Hits) per snap\nin last 4 weeks",
+  labs(x = "Average STRAIN in first 4 weeks",
+       y = "Pressure rate in last 4 weeks",
        color = "Position") +
   theme_light() +
   theme(axis.title = element_text(size = rel(1)),

### 2023-05-10 scripts/04_fig_cor_pressure.R
figures labeling

@@ -41,7 +41,7 @@ fig_cor_pressure <- test_summed |>
   scale_y_continuous(breaks = seq(0, 0.12, 0.04)) +
   scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107", "#004D40")) +
   labs(y = "Pressure rate",
-       x = "Average STRAIN across all frames",
+       x = "Average STRAIN",
        color = "Position") +
   theme_light() +
   theme(axis.title = element_text(size = rel(1)),

### 2023-05-10 scripts/05_fig_stability.R

@@ -47,8 +47,8 @@ fig_stability <- strain_first |>
   geom_smooth(method = "lm", se = FALSE, linewidth = 1.5, color = "black", alpha = 0.5) +
   geom_point(aes(color = pos, group = name), size = 2, alpha = 0.8) +
   scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107", "#004D40")) +
-  labs(x = "Average STRAIN in first 4 weeks",
-       y = "Average STRAIN in last 4 weeks",
+  labs(x = "Average STRAIN (first 4 weeks)",
+       y = "Average STRAIN (last 4 weeks)",
        color = "Position") +
     theme_light() +
     theme(axis.title = element_text(size = rel(1)),

### 2023-05-10 scripts/07_fig_predictability.R

@@ -60,8 +60,8 @@ hsh_hsh <- hsh_first |>
   geom_smooth(method = "lm", se = FALSE, linewidth = 1.5, color = "black", alpha = 0.5) +
   geom_point(aes(color = officialPosition, group = nflId), size = 2, alpha = 0.8) +
   scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107", "#004D40")) +
-  labs(x = "Pressure rate in first 4 weeks",
-       y = "Pressure rate per snap in last 4 weeks",
+  labs(x = "Pressure rate (first 4 weeks)",
+       y = "Pressure rate (last 4 weeks)",
        color = "Position") +
   theme_light() +
   theme(axis.title = element_text(size = rel(1)),
@@ -83,16 +83,16 @@ strain_hsh <- hsh_last |>
   geom_smooth(method = "lm", se = FALSE, linewidth = 1.5, color = "black", alpha = 0.5) +
   geom_point(aes(color = pos, group = name), size = 2, alpha = 0.8) +
   scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107", "#004D40")) +
-  labs(x = "Average STRAIN in first 4 weeks",
-       y = "Pressure rate in last 4 weeks",
+  labs(x = "Average STRAIN (first 4 weeks)",
+       # y = "Pressure rate (last 4 weeks)",
+       y = NULL,
        color = "Position") +
   theme_light() +
   theme(axis.title = element_text(size = rel(1)),
+        axis.text.y = element_blank(),
         legend.title = element_text(size = rel(1)),
         legend.text = element_text(size = rel(0.8)),
         axis.text = element_text(size = rel(0.8)),
         panel.grid.minor = element_blank())
 
-fig_predictability <- cowplot::plot_grid(hsh_hsh, strain_hsh, rel_widths = c(2.4, 3.6))
-
-
+fig_predictability <- cowplot::plot_grid(hsh_hsh, strain_hsh, rel_widths = c(2.1, 2.4))

### 2023-05-10 paper/00-main.Rmd
figure captions and more

@@
   fig.align = "center",
+  fig.pos = "p",
   cache = TRUE

### 2023-05-10 paper/01-introduction.Rmd
figure captions and more

@@
 Table \ref{tab:metrics} gives a summary of existing pass-rush metrics.
+We now highlight what these quantities measure as well as their limitations.
 
-Officially tracked by the NFL since 1982, the most well-known statistic for evaluating defensive linemen on pass rushing plays is the sack, which occurs when a defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass.
+Officially tracked by the NFL since 1982, the most well-known statistic for evaluating defensive linemen on pass rushing plays is the sack.
+A sack is recorded when a defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass.
 Sacks represent a simple binary measure of success for a rusher on a pass play.
-However, there are many intermediate outcomes for a rusher that can also be considered successes that don't result in a sack.
+However, there are many intermediate outcomes for a rusher that can also be considered successes that do not result in a sack.
 Measures such a hits and hurries are collected by various outlets where hits counts the number of times a QB is hit while passing and hurries counts how often a QB has to throw the ball earlier than they would have liked to.
@@
 
-[Pro football ref has a slightly different QB pressures formula than the formula below https://www.pro-football-reference.com/about/advanced_stats.htm]
+<!-- [Pro football ref has a slightly different QB pressures formula than the formula below https://www.pro-football-reference.com/about/advanced_stats.htm] -->
 
@@
 All the aforementioned methods suffered from not having the data that was actually necessary to develop a meaningful metric for linemen in football, either offensive or defensive.
-Now though, thanks to the 2023 Big Data Bowl, we have a publicly available tracking data not only for the linemen, but for every player on the field!
-With this data performance of offensive and defensive linemen can be meaningfully assessed, though we focus solely on the performance of defensive linemen here.
-This manuscript presents a metric we refer to as STRAIN based on a concept of the same name from materials science.
-STRAIN gives a continuous measure of pressure for each of the defensive linemen over the course of an entire play allowing for the measurement of the level of success of the pass rush even on plays that don't result in an easily observed outcome like a sack, hit, or hurry.
+Now though, thanks to the 2023 Big Data Bowl, we have a publicly available tracking data not only for the linemen, but for every player on the field.
+This allows us to meaningfully assess the performance of linemen in football.
+In this paper, we focus on defensive linemen, and propose STRAIN, a metric for pass rush evaluation inspired by the concept of strain rate in materials science.
+STRAIN gives a continuous measure of pressure for each of the defensive linemen over the course of an entire play.
+This allows for the measurement of pass rush success even on plays that do not result in an observed outcome like a sack, hit, or hurry.
 We view this as a major step forward in the accurate evaluation of defensive linemen performance.
@@
 The remainder of this paper is outlined as follows.
-Section \ref{sec:data} describes the data set from the 2023 Big Data Bowl.
-Section \ref{sec:methods} introduces the formal definition of STRAIN.
-Section \ref{sec:results} studies the performance of the proposed metric.
-We end with Section \ref{sec:discussion} where we present a discussion of future directions for STRAIN.
+We first describe the player tracking data provided by 2023 Big Data Bowl in Section \ref{sec:data}.
+We then introduces the formal definition of our metric STRAIN and our modeling approach in Section \ref{sec:methods}.
+We present an real-game illustration of the proposed metric and study its performance in Section \ref{sec:results}.
+We close with a discussion of future directions for STRAIN in Section \ref{sec:discussion}.
 

### 2023-05-10 paper/02-data.Rmd
figures + tables formatting
figure captions and more

@@
 \caption{Example of tracking data for a play during the Las Vegas Raiders versus Denver Broncos NFL game on October 17, 2021.
-The data shown here is for Raiders DE Maxx Crosby, and the frames included are between the ball snap and when the sack by Crosby is recorded.
+The data shown here are for Raiders DE Maxx Crosby, and the frames included are between the ball snap and when the sack by Crosby is recorded.
 \label{tab:track}}
 \centering
-\begin{tabular}{ccccccccccc}
+\begin{tabular}{ccccccccc}
 \hline
-frameId & jerseyNumber & team & x & y & s & a & dis & o & dir & event \\ 
+frameId & x & y & s & a & dis & o & dir & event \\ 
 \hline
-7 & 98 & LV & 67.68 & 29.89 & 0.34 & 1.57 & 0.04 & 124.86 & 88.21 & ball\_snap \\ 
-8 & 98 & LV & 67.76 & 29.89 & 0.69 & 2.13 & 0.08 & 124.07 & 89.59 & None \\ 
-\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\ 
-50 & 98 & LV & 73.67 & 25.06 & 4.19 & 2.62 & 0.42 & 134.21 & 125.26 & qb\_sack \\ 
+7 & 67.68 & 29.89 & 0.34 & 1.57 & 0.04 & 124.86 & 88.21 & ball\_snap \\ 
+8 & 67.76 & 29.89 & 0.69 & 2.13 & 0.08 & 124.07 & 89.59 & None \\ 
+\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\ 
+50 & 73.67 & 25.06 & 4.19 & 2.62 & 0.42 & 134.21 & 125.26 & qb\_sack \\ 
 \hline

### 2023-05-10 paper/03-methods.Rmd
more results writing
figure captions and more

@@
 Moreover, since we observe STRAIN continuously at every tenth of a second within each play, we can then compute the average STRAIN across all frames played for every pass rusher.
-This can be helpful in regards to player evaluation, as we can determine the most effective pass-rushers based on their average STRAIN values.
 Formally, the average STRAIN, denoted by $\overline{\text{STRAIN}}$, for pass rusher $i$ over $n$ total plays with $T$ total frames is
-
 [CHECK THIS - is notation consistent with what we have above?]
 \begin{equation}
-\overline{\text{STRAIN}}_{i} = \frac{1}{n} \sum_{t=1}^T {\text{STRAIN}}_{i}(t).
+\overline{\text{STRAIN}}_{i} = \frac{1}{n} \sum_{t=1}^T {\text{STRAIN}}_{it}.
 \end{equation}
+This can be helpful for player evaluation, as we determine the most effective pass-rushers based on their average STRAIN values in Section \ref{sec:strainbar}.
+We also use average STRAIN to assess statistical properties of our metric in Section \ref{sec:statprop}.
 
@@
 \begin{aligned}
-\overline{\text{STRAIN}}_{j,i} &\sim N(R_{r[j,i]} + B_{b[j,i]} + Q_{q[i]} + X_{j,i} \cdot \beta, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
+\overline{\text{STRAIN}}_{ij} &\sim N(R_{r[ij]} + B_{b[ij]} + Q_{q[i]} + X_{ij} \cdot \beta, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
 R_{r} &\sim N(\mu_R, \sigma^2_R), \text{ for } r = 1, \cdots, \text{ \# of pass rushers}, \\
@@
 With regards to the fixed-effects about pass rusher $j$ in play $i$, we include the length of the play, the pass rusher position, the nearest pass blocker's position, and the number of pass blockers assigned to the pass rusher based on the PFF data.
-Table \ref{tab:pos} shows our categorization of position for the pass rush and pass block roles.
+Table \ref{tab:pos} shows our positional categorization for the pass rush and pass block roles.
 We implement the model using penalized likelihood via the `lme4` package in `R` [@bateslme4; @R2023Language].
@@
 Pass rush & Outside linebacker \\ 
-Pass rush & Interior linebacker (middle and inside linebackers) \\ 
-Pass rush & Secondary (cornerback, free safety, and strong safety) \\
+Pass rush & Interior linebacker (middle linebacker, inside linebacker) \\ 
+Pass rush & Secondary (cornerback, free safety, strong safety) \\
 \hline

### 2023-05-10 paper/04-results.Rmd
figure captions and more

@@
 #| label: fig_crosby_curves
-#| fig.cap: "crosby"
+#| fig.cap: "The change in STRAIN, velocity, and distance from the QB for Maxx Crosby over the course of a unsuccessful pass rush play that results in a sack."
 #| fig.width: 4.8
@@
 Both of these relationships suggest an overall successful pass rush by Maxx Crosby.
+It is also notable that Crosby, who is credited with a sack, generates more STRAIN than other pass rushers during this play.
 
@@
 #| label: fig_crosby_curves_unsucess
-#| fig.cap: "strain curve for unsuccessful play"
+#| fig.cap: "The change in STRAIN, velocity, and distance from the QB for Maxx Crosby over the course of an unsuccessful pass rush play."
 #| fig.width: 4.8
@@
 Further, Figure \ref{fig:fig_outcome_curves} shows the positional STRAIN curves by whether a play's outcome is a hit, sack, hurry, or none of those. 
-It is striking that for plays with sack outcome,
+We clearly see that players tend to generate more STRAIN when a play ends in a hit, sack, or hurry, compared to no outcome.
+Within the three pressure metrics, it is not surprised that a sack corresponds to the highest amount of STRAIN.
+[Can we say anything about why NT has highest curve when there's a sack?]
 
-## Ranking the best pass-rushers
+## Ranking the best pass-rushers {#sec:strainbar}
 
@@
 
-## Statistical Properties of STRAIN
+## Statistical Properties of STRAIN {#sec:statprop}
 
@@
 
-some sentences
+Next, we examine different statistical properties of our proposed metric STRAIN.
+Our focus here is to understand the stability and predictability of STRAIN, and we pose the following question:
+How much does our metric vary from week to week?
+In other words, is previous performance predictive of future performance based on our metric?
+Below, we attempt the answer these questions, and the following results are for pass rushers with at least 100 snaps played during the first 8 weeks of the 2021 NFL regular season.
 
-Next, we examine the predictability of STRAIN as a measure of pressure.
-We first look at how well strain correlates with a simple measure of pressure rate, defined as the total hits, sacks and hurries per snap.
-Figure \ref{fig:fig_cor_pressure} is a scatterplot of average STRAIN and pressure rate which reveals a fairly strong correlation (r = 0.625) between the quantities.
-Note that we only consider players with at least 100 snaps played.
+We first investigate the predictability of STRAIN as a measure of pressure.
+In particular, we look at how well our metric correlates with a simple measure of pressure rate, defined as the total hits, sacks and hurries per snap.
+Figure \ref{fig:fig_cor_pressure} is a scatterplot of average STRAIN and pressure rate over the course the provided 8-game sample, which reveals a fairly strong correlation ($r =0.625$) between the quantities.
 Furthermore, the average STRAIN for the first 4 weeks of the 2021 season is more predictive of the last 4 weeks' pressure rate ($r=0.322$) than the first 4 weeks' pressure rate ($r=0.097$), as illustrated in Figure \ref{fig:fig_predictability}.
-
-We also examine the stability of STRAIN over time by comparing the average STRAIN across all frames played in first and last four weeks of the 2021 NFL regular season.
+We also analyze the stability of STRAIN over time by comparing the average STRAIN across all frames played in first and last four weeks of the 2021 NFL regular season.
 It is apparent from Figure \ref{fig:fig_stability} that there is a strong correlation $(r = 0.854)$ for this relationship, illustrating that STRAIN is a stable quantity over the provided 8-week window.
@@
 #| label: fig_cor_pressure
-#| fig.cap: "Relationship between  with pressure rate = (hits + sacks + hurries) per snap"
+#| fig.cap: "Relationship between average STRAIN and pressure rate (total hits, sacks, and hurries per snap) over the first 8 weeks of the 2021 NFL season.
+There is a moderately strong association between average STRAIN and pressure rate ($r =0.625$).
+Results shown here are for pass rushers with at least 100 snaps play over the 8-week data."
 #| fig.width: 4
@@
 #| label: fig_predictability
-#| fig.cap: "strain is more predictive of pressure rate than pressure rate itself."
+#| fig.cap: "Relationships between pressure rate for the last 4 weeks of the 2021 NFL season and first 4 weeks' pressure rate (left) and average STRAIN (right).
+STRAIN is more predictive ($r=0.322$) of future pressure rate than previous pressure rate itself ($r=0.097$).
+Results shown here are for pass rushers with at least 100 snaps play over the 8-week data."
 #| fig.width: 6.5
@@
 #| label: fig_stability
-#| fig.cap: "Metric stability"
+#| fig.cap: "Relationships between average STRAIN for the last and first 4 weeks of the 2021 NFL season.
+A strong linear correlation ($r = 0.854$) demonstrates that STRAIN is a highly stable metric over time.
+Results shown here are for pass rushers with at least 100 snaps play over the 8-week data."
 #| fig.width: 4
@@
 
-There are a variety of quantities we're interested in from the multilevel model fit in Section \ref{sec:methods}.
+There are a variety of quantities we are interested in from the multilevel model fit mentioned in Section \ref{sec:methods}.
 First, Table \ref{tab:fixedeff} displays the results for the fixed effects terms.
@@
 \begin{table}
-\caption{Fixed effects for the multilevel model. \label{tab:fixedeff}}
+\caption{Fixed-effects coefficients estimates for the multilevel model. \label{tab:fixedeff}}
 \centering
@@
 \hline
-nflId & 0.057 \\ 
-blocker\_id & 0.012 \\ 
-passer & 0.023 \\ 
-residual & 0.907 \\ 
+Pass rusher & 0.057 \\ 
+Pass blocker & 0.012 \\ 
+Quarterback & 0.023 \\ 
+Residual & 0.907 \\ 
 \hline
@@
 We observe similar rankings based on the random intercepts in comparison to the average STRAIN, with Myles Garrett (DE), Aaron Donald (DT), Vita Vea (NT), and T.J. Watt (OLB) as the top pass rushers. 
-We observe different levels of variability across these top players, with the only a subset of players with intervals strictly above zero.  [NEED TO THINK ABOUT WHAT MORE TO SAY HERE]
+We observe different levels of variability across these top players, with the only a subset of players with intervals strictly above zero.
+[NEED TO THINK ABOUT WHAT MORE TO SAY HERE]
 
@@
 #| label: fig_rankings_boot
-#| fig.cap: "steve"
+#| fig.cap: "Top 10 pass rushers for each position (DE, DT, NT, OLB) by mean varying intercept."
 #| fig.width: 6

### 2023-05-10 scripts/08_fig_outcome_curves.R

@@ -60,11 +60,12 @@ fig_outcome_curves <- hit |>
   bind_rows(sack) |> 
   bind_rows(hurry) |> 
   bind_rows(none) |> 
-  mutate(officialPosition = factor(officialPosition, levels = c("OLB", "DE", "DT", "NT"))) |> 
+  mutate(officialPosition = factor(officialPosition, levels = c("OLB", "DE", "DT", "NT")),
+         outcome = factor(outcome, levels = c("Sack", "Hit", "Hurry", "None"))) |> 
   ggplot(aes(frameId_snap_corrected, mn, 
              color = officialPosition, 
              group = officialPosition)) +
-  geom_smooth(se = FALSE, span = 0.3, linewidth = 1) +
+  geom_smooth(se = FALSE, span = 0.4, linewidth = 1) +
   scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107", "#004D40"), name = "Position") +
   scale_x_continuous(breaks = seq(0, 40, 10), labels = 0:4) +
   scale_linetype_manual(values = c(rep("solid", 4), "dotted"), name = "Position") +

### 2023-05-10 paper/02-data.Rmd
more captions

@@
 
-Table \ref{tab:track} displays a tracking data example for a play from the 2021 NFL regular season week six matchup between the Las Vegas Raiders and Denver Broncos, which ended with Broncos QB Teddy Bridgewater getting sacked by Raiders DE Maxx Crosby[^crosbysack].
+Table \ref{tab:track} displays a tracking data example for a play from the 2021 NFL regular season week six matchup between the Las Vegas Raiders and Denver Broncos, which ends with Broncos QB Teddy Bridgewater getting sacked by Raiders DE Maxx Crosby[^crosbysack].
 In addition, Figure \ref{fig:fig_field_raw} presents the locations of every Las Vegas (in black) and Denver (in orange) player on the field from this play at 1, 2, 3 and 4 seconds after the ball snap, with Maxx Crosby highlighted in blue.

### 2023-05-10 paper/04-results.Rmd
more captions

@@
 #| label: fig_crosby_curves
-#| fig.cap: "The change in STRAIN, velocity, and distance from the QB for Maxx Crosby over the course of a unsuccessful pass rush play that results in a sack."
+#| fig.cap: "Changes in STRAIN, velocity, and distance from the QB for Maxx Crosby over the course of a successful pass rush play that results in a sack."
 #| fig.width: 4.8
@@
 In this play, Broncos QB Teddy Bridgewater is well-protected by the offensive line and is able to release a long pass[^unsuccess].
-We see that the STRAIN generated by Crosby and other pass rushers are relative small for this play, compared to the previous play which resulted in a sack.
+We see that the STRAIN generated by Crosby is relatively small over the course of this play, compared to the previous play shown in Figure \ref{fig:fig_crosby_curves} which results in a sack.
 
@@
 #| label: fig_crosby_curves_unsucess
-#| fig.cap: "The change in STRAIN, velocity, and distance from the QB for Maxx Crosby over the course of an unsuccessful pass rush play."
+#| fig.cap: "Changes in STRAIN, velocity, and distance from the QB for Maxx Crosby over the course of an unsuccessful pass rush play."
 #| fig.width: 4.8
@@
 #| label: fig_pos_curves
-#| fig.cap: "STRAIN curves for different positions"
+#| fig.cap: "STRAIN curves for different positions.
+Edge rushers (OLB and DE) tend to generate more STRAIN than interior rushers (DT and NT).
+The dotted gray line represents the average STRAIN curve for all players without accounting for position."
 #| fig.width: 4.5
@@
 #| label: fig_outcome_curves
-#| fig.cap: "STRAIN curves for different positions by outcome"
+#| fig.cap: "Positional STRAIN curves by play outcome (sack, hit, hurry, and none)."
 #| fig.width: 6.5
@@
 #| label: fig_stability
-#| fig.cap: "Relationships between average STRAIN for the last and first 4 weeks of the 2021 NFL season.
+#| fig.cap: "Relationship between average STRAIN for the last and first 4 weeks of the 2021 NFL season.
 A strong linear correlation ($r = 0.854$) demonstrates that STRAIN is a highly stable metric over time.
@@
 Figure \ref{fig:fig_rankings_boot} displays the mean varying intercept for the top ten pass rushers for each position along with 95% confidence intervals generated by sampling from the posterior distributions via parametric bootstrapping using the `merTools` package [@knowlesmerTools, @gelman2006data].
-We observe similar rankings based on the random intercepts in comparison to the average STRAIN, with Myles Garrett (DE), Aaron Donald (DT), Vita Vea (NT), and T.J. Watt (OLB) as the top pass rushers. 
+We observe similar rankings based on the random intercepts in comparison to the average STRAIN results in Tables \ref{tab:edge} and \ref{tab:interior}, with Myles Garrett (DE), Aaron Donald (DT), Vita Vea (NT), and T.J.
+Watt (OLB) as the top pass rusher for their respective position.
 We observe different levels of variability across these top players, with the only a subset of players with intervals strictly above zero.

### 2023-05-10 paper/04-results.Rmd
added hits, sacks, hurries to rankings

@@
 Tables \ref{tab:edge} and \ref{tab:interior} are leaderboards for the NFL's best edge and interior rushers (with at least 100 plays) rated by the average STRAIN across all frames for the first eight weeks of the 2021 regular season.
+The tables also consist of the total number of hits, hurries, and sacks for each defender.
 
@@
 Rashan Gary, a young rusher with huge potential, ranks first among all pass-rushers according to our metric.
-Our leaderboards largely match the rankings of experts in the field -- for instance, PFF's edge [@Monson2022nfl] and interior [@Linsey2022nfl] rusher rankings after the 2021 season, lending credibility to our proposed metric as a measure of pass-rushing effectiveness.
+Moreover, our leaderboards largely match the rankings of experts in the field.
+For instance, there is considerable overlap between our lists and PFF's edge [@Monson2022nfl] and interior [@Linsey2022nfl] rusher rankings released after the 2021 season.
+This ultimately lends credibility to our proposed metric STRAIN as a measure of pass-rushing effectiveness.
 
@@
 \centering
-\begin{tabular}{rrlllrr}
+\begin{tabular}{rlllrrrrr}
 \hline
- Rank & Player & Team & Position & Snaps & Average STRAIN \\ 
+Rank & Player & Team & Position & Snaps & Hits & Hurries & Sacks & $\overline{\text{STRAIN}}$ \\
 \hline
-1 & Rashan Gary & GB & OLB & 176 & 2.82 \\ 
-2 & Leonard Floyd & LA & OLB & 185 & 2.80 \\ 
-3 & Justin Houston & BAL & OLB & 132 & 2.78 \\ 
-4 & Myles Garrett & CLE & DE & 197 & 2.75 \\ 
-5 & Von Miller & DEN & OLB & 145 & 2.75 \\ 
-6 & T.J. Watt & PIT & OLB & 147 & 2.71 \\ 
-7 & Yannick Ngakoue & LV & DE & 175 & 2.70 \\ 
-8 & Alex Highsmith & PIT & OLB & 129 & 2.65 \\ 
-9 & Preston Smith & GB & OLB & 124 & 2.61 \\ 
-10 & Randy Gregory & DAL & DE & 134 & 2.58 \\ 
-11 & Joey Bosa & LAC & OLB & 160 & 2.58 \\ 
-12 & Darrell Taylor & SEA & DE & 107 & 2.57 \\ 
-13 & Josh Sweat & PHI & DE & 159 & 2.57 \\ 
-14 & Maxx Crosby & LV & DE & 198 & 2.56 \\ 
-15 & Markus Golden & ARI & OLB & 164 & 2.50 \\ 
+1 & Rashan Gary & GB & OLB & 176 & 10 & 25 & 5 & 2.82 \\ 
+2 & Leonard Floyd & LA & OLB & 185 & 2 & 25 & 8 & 2.80 \\ 
+3 & Justin Houston & BAL & OLB & 132 & 8 & 8 & 4 & 2.78 \\ 
+4 & Myles Garrett & CLE & DE & 197 & 9 & 29 & 12 & 2.75 \\ 
+5 & Von Miller & DEN & OLB & 145 & 4 & 21 & 5 & 2.75 \\ 
+6 & T.J. Watt & PIT & OLB & 147 & 6 & 9 & 8 & 2.71 \\ 
+7 & Yannick Ngakoue & LV & DE & 175 & 6 & 20 & 4 & 2.70 \\ 
+8 & Alex Highsmith & PIT & OLB & 129 & 4 & 7 & 2 & 2.65 \\ 
+9 & Preston Smith & GB & OLB & 124 & 4 & 8 & 2 & 2.61 \\ 
+10 & Randy Gregory & DAL & DE & 134 & 7 & 19 & 5 & 2.58 \\ 
+11 & Joey Bosa & LAC & OLB & 160 & 5 & 21 & 4 & 2.58 \\ 
+12 & Darrell Taylor & SEA & DE & 107 & 5 & 9 & 3 & 2.57 \\ 
+13 & Josh Sweat & PHI & DE & 159 & 4 & 14 & 5 & 2.57 \\ 
+14 & Maxx Crosby & LV & DE & 198 & 12 & 30 & 7 & 2.56 \\ 
+15 & Markus Golden & ARI & OLB & 164 & 5 & 14 & 5 & 2.50 \\ 
 \hline
@@
 \centering
-\begin{tabular}{rlllrr}
+\begin{tabular}{rlllrrrrr}
 \hline
-Rank & Player & Team & Position & Snaps & Average STRAIN \\ 
+Rank & Player & Team & Position & Snaps & Hits & Hurries & Sacks & $\overline{\text{STRAIN}}$ \\
 \hline
-1 & Aaron Donald & LA & DT & 239 & 1.67 \\ 
-2 & Solomon Thomas & LV & DT & 115 & 1.51 \\ 
-3 & Quinton Jefferson & LV & DT & 144 & 1.46 \\ 
-4 & Chris Jones & KC & DT & 139 & 1.42 \\ 
-5 & DeForest Buckner & IND & DT & 198 & 1.26 \\ 
-6 & Cameron Heyward & PIT & DT & 188 & 1.25 \\ 
-7 & Javon Hargrave & PHI & DT & 156 & 1.24 \\ 
-8 & Jerry Tillery & LAC & DT & 171 & 1.16 \\ 
-9 & Ed Oliver & BUF & DT & 133 & 1.15 \\ 
-10 & Osa Odighizuwa & DAL & DT & 162 & 1.13 \\ 
-11 & Greg Gaines & LA & NT & 111 & 1.11 \\ 
-12 & Leonard Williams & NYG & DT & 226 & 1.03 \\ 
-13 & Christian Barmore & NE & DT & 166 & 1.02 \\ 
-14 & Vita Vea & TB & NT & 184 & 1.01 \\ 
-15 & B.J. Hill & CIN & DT & 123 & 0.96 \\ 
+1 & Aaron Donald & LA & DT & 239 & 8 & 24 & 6 & 1.67 \\ 
+2 & Solomon Thomas & LV & DT & 115 & 7 & 11 & 3 & 1.51 \\ 
+3 & Quinton Jefferson & LV & DT & 144 & 6 & 8 & 3 & 1.46 \\ 
+4 & Chris Jones & KC & DT & 139 & 3 & 18 & 3 & 1.42 \\ 
+5 & DeForest Buckner & IND & DT & 198 & 4 & 18 & 4 & 1.26 \\ 
+6 & Cameron Heyward & PIT & DT & 188 & 2 & 22 & 3 & 1.25 \\ 
+7 & Javon Hargrave & PHI & DT & 156 & 6 & 15 & 5 & 1.24 \\ 
+8 & Jerry Tillery & LAC & DT & 171 & 4 & 7 & 3 & 1.16 \\ 
+9 & Ed Oliver & BUF & DT & 133 & 4 & 12 & 1 & 1.15 \\ 
+10 & Osa Odighizuwa & DAL & DT & 162 & 3 & 18 & 3 & 1.13 \\ 
+11 & Greg Gaines & LA & NT & 111 & 2 & 13 & 2 & 1.11 \\ 
+12 & Leonard Williams & NYG & DT & 226 & 4 & 14 & 6 & 1.03 \\ 
+13 & Christian Barmore & NE & DT & 166 & 5 & 17 & 1 & 1.02 \\ 
+14 & Vita Vea & TB & NT & 184 & 6 & 12 & 1 & 1.01 \\ 
+15 & B.J. Hill & CIN & DT & 123 & 3 & 4 & 3 & 0.96 \\ 
 \hline

### 2023-05-10 paper/05-discussion.Rmd
added hits, sacks, hurries to rankings

--- 2023-03-09 paper/kaggle.Rmd
@@
----
-output: html_document
----
+# Discussion {#sec:discussion}
 
-# Introduction
+Summary
+
+Critique
+
+Future work
 
 <!--
-Greg's long lit review
-The theme of Big Data Bowl 2023 is about evaluating performance of linemen, both offensive and defensive, on passing plays.
-While the ideas presented here could be extended to offensive linemen, we have chosen to focus on measuring the performance of defensive linemen.
-Officially tracked by the NFL since 1982, the most well known statistic for evaluating defensive linemen on pass rushing plays is the sack, which occurs when a defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass.
-Other counting statistics have been tracked by various entities such as hurries and hits as well as pressures, which is defined as the sum of hurries, hits, and sacks.
-More recently metrics such as Time In Pocket (TIP) and Pass Rusher Win Rate (PRWR) have also been proposed and are substantial improvements over the less sophisticated counting statistics.
-While PRWR is certainly a step in the right direction, it takes continuous data and converts it to a binary win/loss, which is generally not recommended in statistical analysis, thus, the metric we propose here is computed continuously over time just as the data is actually observed.
-The metric we propose is based on the following simple question: What is considered a success for a pass rusher on a pass play?
-The simplest answer to this question, and the first defensive lineman statistics to be tracked by the NFL, is a sack.
-In terms of tracking data, one can think of a sack as reducing the distance between the pass rusher and the quarterback to 0.
-Therefore, a decent starting point for a continuous measure of pass rushing effectiveness shoould be based on how close the pass rusher is able to get to the quarterback with smaller distances indicating amore successful pass rush.
-However, a good measure of pass rushing effectiveness cannot simply be based on distance to the quarterback; Consider the following case: a pass rusher gets within two yards of the quarterback, but is then stopped and fails to get any closer for the rest of the play.
-This is not a successful rush because the pass rusher is not continuing to reduce the distance between themselves and the quarterback.
-Therefore, a good metric should consider the rate at which this distance is being reduced.
-However, just like distance, rate by itself is inadequate in assessing success of a pass rusher.
-Consider the following case: A defensive player is 50 yards away and moving towards the quarterback very rapidly.
-Their moving rate is essentially unimportant due to a significantly far distance away from the quarterback.
-This means that in order to appropriately evaluate rusher performance, the distance from the quarterback and the rate at which this distance is being reduced must both be considered.
-Therefore, we propose STRAIN, a new measure for pass rusher effectiveness, which is computed as the ratio of the rate that the pass rusher is reducing the distance between themselve and the quarterback and distance between the pass rusher and the quarterback.
-In performing some research about this proposed quantity, we discovered that this concept actually already exists in materials science and is called the "strain rate" (which is where the name of our metric is derived).
-Given this connection, one can view the players as particles in some material and the defensive "particles" are attempting to exert pressure on the pocket in an attempt to compress and collapse this pocket around the quarterback.
-We think this analogy is just delightful.
--->
 
-What is considered a success for a pass-rusher on a pass play?
-The simplest answer to this question is a sack.
-In terms of tracking data, one can think of a sack as reducing the distance between the pass-rusher and quarterback to 0.
-More generally, a good starting point for a continuous measure of pass-rushing effectiveness should be based on the **distance** between the rusher and quarterback, with smaller distances indicating greater success.
-However, this naive approach has major drawbacks.
-Consider the case where a rusher is within two yards of the quarterback, but is then stopped and fails to get any closer for the rest of the play.
-This is not a successful rush because the pass-rusher is not continuing to reduce their distance to the quarterback.
-Thus, another approach for pass-rushing assessment is to measure the **rate** at which the rusher is getting to the quarterback.
-Yet, the rate by itself is also inadequate.
-If a defender is 50 yards away and moving towards the quarterback, their moving rate is essentially unimportant due to a significantly far distance.
+(Stuff from BDB draft)
 
-Therefore, in order to appropriately evaluate rusher performance, the **distance** from the quarterback *and* the **rate** at which this distance is being reduced must both be considered.
-We propose **STRAIN**, a new measure for pass-rusher effectiveness, which is computed as the ratio of the aforementioned moving rate and distance quantities.
-Our proposed metric has a nice interpretation and is inspired by the concept of "strain rate" in materials science.
-
-# Methods
-
-## Strain rate in materials science
-
-In [*materials science*](https://en.wikipedia.org/wiki/Materials_science), [*strain*](https://en.wikipedia.org/wiki/Strain_(materials_science)) is the deformation of a material from stress.
-The stretching of a rubber band is an example of strain: the longer the stretch, the larger the strain.
-Formally, let $L(t)$ be the distance between two points of interest at time $t$, and $L_0$ be the initial distance between those two points.
-Then the strain at time $t$ is defined as $\displaystyle s(t)={\frac  {L(t)-L_{0}}{L_{0}}}$.
-Notice that this measure is unitless as a ratio of two quantities having the same dimension.
-Furthermore, the [*strain rate*](https://en.wikipedia.org/wiki/Strain_rate) is the derivative of strain.
-That is, ${\displaystyle {{s}'}(t)={\frac {ds}{dt}} = {\frac {v(t)}{L_{0}}}}$, where $v(t)$ is the rate at which the two points of interest are moving away from/towards each other.
-Whereas strain is unitless, the strain rate is measured in inverse of time, usually $\text{seconds}^{-1}$.
-
-## Application to pass-rushing in football
-
-Motivated by its scientific definition, we draw an analogy between strain and pass-rushing in football.
-**Just as strain rate is a measure of deformation in materials science, a pass-rusher's efforts involve the application of force/deformation against the offensive line, with the ultimate goal of breaking through the protection to reach the quarterback.
-The players can be viewed as "particles" in some material and the defensive "particles" are attempting to exert pressure on the pocket with the aim of compressing and collapsing this pocket around the quarterback.**
-
-In order to apply strain rate to measure NFL pass-rusher effectiveness, we make modifications to how this concept is traditionally defined.
-
-Let $(x_{ijt}, y_{ijt})$ be the $(x, y)$ location on the field of player $j = 1, \cdots, J$ at frame $t = 1, \cdots, T_i$ for play $i = 1, \cdots, n$; and $(x^{QB}_{it}, y^{QB}_{it})$ be the $(x, y)$ location of the quarterback at frame $t$ during play $i$.
-
-* The distance between player $j$ and the quarterback at frame $t$ during play $i$ is $f_{d_{ij}}(t) = \sqrt{(x_{ijt} - x^{QB}_{it})^2 + (y_{ijt} - y^{QB}_{it})^2}$.
-* The rate at which player $j$ is moving towards the quarterback at frame $t$ during play $i$ is $\displaystyle f'_{d_{ij}}(t) = \frac{df_{d_{ij}}(t)}{dt}$.
-* The STRAIN for player $j$ at frame $t$ during play $i$ is
-
-$$\text{STRAIN}_{ij}(t) = \frac{-f'_{d_{ij}}(t)}{f_{d_{ij}}(t)} \,.$$
-
-(*Note that to distinguish our metric from strain and strain rate in materials science, we write it in capital letters* (STRAIN) *for the remainder of this report.*)
-
-Recall that by its materials science property, strain is generally defined to increase as the distance between two points **increases**.
-In our football setting, the two points of interest are the pass-rusher and the quarterback, and we want our metric to increase as the distance between the pass-rusher and the quarterback **decreases**.
-Thus, the negative sign in the numerator of our formula accounts for this. 
-Additionally, rather than keeping the initial distance ($L_0$ as previously denoted) between two points constant over time, we update the initial position to be the player locations at the beginning of each frame.
-This effectively gives us the STRAIN for each frame throughout a play.
-
-Since we only observe these functions discretely in increments of 10 frames/second, an estimate for our proposed metric STRAIN is
-
-$$\widehat{\text{STRAIN}}_{ij}(t) = \cfrac{-\cfrac{f_{d_{ij}}(t) - f_{d_{ij}}(t - 1)}{0.1}}{f_{d_{ij}}(t)} \,.$$
-
-Notice that this quantity increases in two ways: 1) the rate at which the rusher is moving towards the quarterback increases, and 2) the distance between the rusher and the quarterback gets smaller.
-Both of these are indications of pass-rusher effectiveness.
-Finally, our statistic STRAIN is measured in inverse seconds, similar to strain rate.
-(There is also an interesting interpretation of the reciprocal of our metric, namely, the amount of time required for the rusher to get to the quarterback at the current location and rate at any given time $t$.)
-
-# Analysis
-
-In the forthcoming analysis, we rely on the provided data of all passing plays from the first 8 weeks of the 2021 NFL season.
-We include only the frames between the ball snap and when a pass forward or QB sack is recorded for each play.
-We also remove all plays with multiple quarterbacks on the field, since we need a uniquely defined quarterback to calculate our measure.
-Using PFF scouting data, we only consider defenders with role "Pass rush" for evaluation.
-
-## Example play
-
-To illustrate STRAIN for pass-rusher evaluation, we use a play from the 2021 NFL regular season week 6 matchup between the Las Vegas Raiders and Denver Broncos, which ended with [Broncos QB Teddy Bridgewater getting sacked by Raiders DE Maxx Crosby](https://www.raiders.com/video/de-maxx-crosby-sacks-qb-teddy-bridgewater-for-a-loss-of-6-yards-nfl).
-The figure below consists of two elements:
-at top, the locations of every Las Vegas (black) and Denver (orange) player on the field during the play, with Maxx Crosby highlighted in blue;
-and at bottom, a line graph showing how Crosby's distance from the QB, velocity, and STRAIN change continuously throughout the play.
-Note that the point size for each Las Vegas defender corresponds to the estimated STRAIN in each frame as the play progresses.
-
-For the first 2.5 seconds, Crosby is being covered by a Denver player and unable to get close to the quarterback, hence the corresponding STRAIN values are virtually zero.
-Suddenly, the STRAIN increases after Crosby is freed up and charges towards Bridgewater.
-At 4 seconds after the snap, Crosby's STRAIN is 2.30, which means at his current moving rate, it will take Crosby about 0.43 (1/2.30) seconds to make the distance between him and the quarterback 0 (i.e.
-essentially sack the quarterback).
-This is consistent with the final outcome of the play, as the sack takes place at the very last frame (4.4 seconds) where the estimated STRAIN for Crosby reaches its peak at 3.96.
-
-## Positional STRAIN curves
-
-The following plot displays the average STRAIN by position for the first 40 frames (4 seconds) after the snap.
-We observe a clear difference in STRAIN between edge rushers (OLB's and DE's) and interior linemen (DT's and NT's).
-Specifically, edge rushers have higher STRAIN than interior linemen on average, as they are more easily able to approach the quarterback on the edge of the pocket versus, for instance, a nose tackle attacking the line head on.
-
-On average, STRAIN appears to increase for the first 0.5 seconds of a play, followed by a decline in the next second.
-STRAIN then increases again until around 2.5-3 seconds after the snap before trending down towards the end of the play.
-In context, this reflects the actions that a pass-rusher initially moves towards the quarterback, but is then stopped by the offensive line while the quarterback drops back.
-When the quarterback stops dropping, the rusher closes the gap and increases STRAIN, before slowing down later on.
-
-## Ranking the best pass-rushers
-
-<!--
-To determine the most effective pass rushers, we aggregate across all frames played and calculate the expected STRAIN (xSTRAIN) for each player, which is computed by simply averaging the STRAIN across all the frames of a play.
-The following tables show the top performing pass rushers (with at least 100 plays) rated by xSTRAIN for the first 8 weeks of the 2021 season.
-The top interior rusher for the first 8 weeks of the 2021 season according to our xSTRAIN metric was Aaron Donald, almost universally considered the best interior rusher in the NFL currently with an xSTRAIN of 1.67.
-Through the first 8 weeks of the 2021 season, Donald had an impressive 5 sacks and 12 quarterback hits.
-Donald Was followed by two Raiders: Solomon Thomas and Quinton Jefferson with xSTRAIN values of 1.51 and 1.46, respectively.
-Rounding out the top 5 were Chris Jones with an xSTRAIN of 1.42 and DeForest Buckner with an xSTRAIN of 1.26.
-As for edge rushers, Rashan Gary with an xSTRAIN value of 2.82 is the top of our list for the first 8 weeks of the 2021 season.
-Over this time period, Gary had 3.5 sacks and 15 quarterback hits.
-Second on our list of edge rushers is Leonard Floyd with an xSTRAIN value of 2.80 who had 6.5 sacks and 9 quarterback hits over the time period observed.
-Rounding out the top 5 are Justin Houston (2.78), Myles Garrett (2.75), and Von Miller (2.75).
-The sixth player on our rankings of edge rushers over the first 8 weeks on the 2021 season is the 2021 Defensive Player of the Year, T.J.
-Watt (2.71) who had 11.5 sacks and 18 quarterback hits in those first 8 weeks.
-Notably their is considerable overlap between our top 20 list over the first 8 weeks of the 2021 season and the Pro Football Focus edge rusher rankings prior to the 2022 season.
--->
-
-To determine the most effective pass-rushers, we compute <span style="text-decoration:overline; font-weight:bold">STRAIN</span>, which represents the average STRAIN across all frames played for each player.
-Based on the clearly distinct patterns for different positions as previously observed, we evaluate interior pass-rushers (NT's and DT's) separately from edge rushers (OLB's and DE's).
-The following tables show the top performing pass-rushers (with at least 100 plays) rated by <span style="text-decoration:overline; font-weight:bold">STRAIN</span> for the first 8 weeks of the 2021 season.
-
-Our results show many true positives with no surprises.
-Myles Garrett and TJ Watt are known as top-tier edge rushers, while Aaron Donald is undoubtedly the best interior defender in football.
-Rashan Gary, a young rusher with huge potential, ranks first among all pass-rushers according to our metric.
-Our leaderboards largely match the rankings of experts in the field (for instance, PFF's [edge](https://www.pff.com/news/nfl-2022-edge-rusher-rankings-tiers) and [interior](https://www.pff.com/news/nfl-interior-defensive-line-rankings-tiers-2022) rusher rankings after the 2021 season), lending credibility to our proposed metric as a measure of pass-rushing effectiveness.
-
-# Discussion
-
-<!--
 We have proposed a simple, effective, and easily interpretable metric for evaluating the performance of pass rushers in the NFL that we call STRAIN.
 Higher values of STRAIN indicate two important aspects of a pass rush: 1) the distance between the pass rusher and the quarterback is small and 2) the rate at which the pass rusher is moving towards the quarterback is high.
+
 One of the major advantages of STRAIN over existing measures is that it gives a measure of effectiveness of a pass rush for every single play at a much more granular level than simply the play resulted in a sack or not or the play was a "success" or not.
@@
 STRAIN is just such a measure that objectively measures the pressure that pass rushers put on a quarterback.
+
 In this report, we simply propose the new metric STRAIN; However, there are innumerable future projects that could be based on STRAIN.
@@
 treating the STRAIN curve over time as a function) in modeling how STRAIN affects offensive production though is non-trivial as the length in time of plays varies from one play to another, and we leave this as future work.
+
 In addtion, while we focused solely on evaluating the performance of pass rushers, this concept can be easily extended to the evaluation of pass blockers as a unit by simply looking at a quantity such as the total amount of STRAIN on a play or the maximum amount of STRAIN per frame aggregated across the entire play.
 Further, STRAIN could be extended to individual offensive pass blockers provided that a method of matching blockers to rushers was devised.
+
 While STRAIN is a big step forward, there are still some notable limitations to this metric.
@@
 If you are a pass rusher on the left side and the quarterback scrambles out to the left, the STRAIN value for that pass rusher can drop considerably as the distance between the quarterback and the pass rusher gets larger, which the pass rushers shouldn't necessarily be negatively affected by in a measure of pass rushing effectiveness.
+
 Finally, we believe there are other concepts in materials science that could be applied to evaluating pass rushing and pass blocking.
@@
 It is easy to imagine the pocket as an object and evaluating the succes of the pocket based on it's "resistance to being deformed".
+
 -->
-
-We have proposed STRAIN, a simple and interpretable statistic for evaluating pass-rushers, with higher STRAIN corresponds to greater pass-rushing ability.
-Compared to existing metrics such as [ESPN's Pass Rusher Win Rate (PRWR)](https://www.espn.com/nfl/story/_/id/24892208), STRAIN measures pass-rush effectiveness for every play continuously over time, which is at a much more granular level than considering whether the play resulted in a sack or converting continuous data to a binary win/loss.
-
-As for future directions, STRAIN could be utilized to collectively evaluate the performance of a pass-rushing unit by considering the total amount of STRAIN during a play, which could, in turn, be used to examine the effectiveness of different types of blitzes.
-Moreover, while we focused solely on pass-rushers, STRAIN can be easily extended to the assessment of pass-blockers *as a unit* by looking at quantities such as the total STRAIN or the maximum STRAIN per frame aggregated across the entire play.
-It is also possible to apply STRAIN to assess individual offensive pass-blockers, provided that a method of matching blockers to rushers was developed.
-
-Our metric, however, is subject to several limitations.
-As currently devised, STRAIN cannot distinguish between a pass-rusher who has to face a single blocker versus a double team.
-This problem could be addressed, at least somewhat, by having a method of matching pass-rushers to blockers as previously mentioned.
-Additionally, there are a few situations such as quarterback out-of-pocket scrambling where STRAIN does not appear to be useful.
-If a pass-rusher is on the left side and the quarterback scrambles out to the left, the STRAIN for that pass-rusher can drop considerably as the distance between the quarterback and pass-rusher increases, which the defender should not necessarily be negatively affected by in a measure of pass-rushing effectiveness.

### 2023-05-10 scripts/06_rankings.R

@@ -2,9 +2,16 @@ library(tidyverse)
 library(here)
 
 pass_rush <- read_csv(here("data", "pass_rush.csv.gz"))
+pff <- read_csv(here("data", "pffScoutingData.csv"))
+pff_hsh <- pff |> 
+  group_by(nflId) |> 
+  summarize(total_hit = sum(pff_hit, na.rm = TRUE),
+            total_hurry = sum(pff_hurry, na.rm = TRUE),
+            total_sack = sum(pff_sack, na.rm = TRUE)) |> 
+  arrange(-total_sack)
 
 test <- pass_rush |> 
-  filter(team == defensiveTeam & !is.na(strain))
+  filter(team == defensiveTeam & !is.na(strain) & !is.infinite(strain))
 test$strain <- test$slope_distance_from_QB/test$distance_from_QB
 test_summed <- test |> 
   group_by(nflId) |>
@@ -17,7 +24,8 @@ test_summed <- test |>
   ) |>
   mutate(strain_rate = 10 * strain / n) |> 
   arrange(-strain_rate) |> 
-  ungroup()
+  ungroup() |> 
+  left_join(pff_hsh)
 
 info <- test |> 
   ungroup() |> 
@@ -34,7 +42,8 @@ out <- test_summed |>
          team, 
          n_plays, 
          strain, 
-         strain_rate)
+         strain_rate,
+         contains("total_"))
 
 library(xtable)
 # edge 
@@ -46,6 +55,9 @@ edge <- out |>
          Team = team, 
          Position = officialPosition,
          Snaps = n_plays, 
+         Hits = total_hit,
+         Hurries = total_hurry,
+         Sacks = total_sack,
          "Average STRAIN" = strain_rate) |> 
   slice_head(n = 15)
 
@@ -60,7 +72,11 @@ interior <- out |>
          Team = team, 
          Position = officialPosition,
          Snaps = n_plays, 
+         Hits = total_hit,
+         Hurries = total_hurry,
+         Sacks = total_sack,
          "Average STRAIN" = strain_rate) |> 
   slice_head(n = 15)
 
 print(xtable(interior), include.rownames = FALSE)
+

### 2023-05-11 paper/03-methods.Rmd
fixed effects table

@@
 In order to apply strain rate to measure NFL pass rusher effectiveness, we make modifications to how this concept is traditionally defined.
-
 Let $(x_{ijt}, y_{ijt})$ be the $(x, y)$ location on the field of pass rusher $j = 1, \cdots, J$ at frame $t = 1, \cdots, T_i$ for play $i = 1, \cdots, n$; and $(x^{QB}_{it}, y^{QB}_{it})$ be the $(x, y)$ location of the quarterback at frame $t$ during play $i$.
@@
 \end{equation}
+
 Note that to distinguish our metric from strain and strain rate in materials science, we write it in capital letters (STRAIN) for the remainder of this manuscript.
@@
 We use random intercepts for the three player groups: (1) pass rushers as $R$, (2) nearest pass blockers as $B$, and (3) QBs as $Q$.
-The coefficients $\beta$ for additional information $X_{j,i}$ about pass rusher $j$ in play $i$ are estimated as fixed-effects. 
+The coefficients $\beta$ for additional information $X_{ij}$ about pass rusher $j$ in play $i$ are estimated as fixed-effects. 
 Our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows.
@@
 This provides us with average individual effects on STRAIN for each player while applying shrinkage, which is a useful property in this problem's context since we do not observe the same number of plays for each player.
-With regards to the fixed-effects about pass rusher $j$ in play $i$, we include the length of the play, the pass rusher position, the nearest pass blocker's position, and the number of pass blockers assigned to the pass rusher based on the PFF data.
+With regards to the fixed-effects about pass rusher $j$ in play $i$, we include the length of the play (in terms of elapsed time between ball snap and a QB event), the positions of pass rusher and nearest blocker, and the number of pass blockers assigned to the pass rusher based on the PFF data.
 Table \ref{tab:pos} shows our positional categorization for the pass rush and pass block roles.

### 2023-05-11 paper/04-results.Rmd
fixed effects table

@@
 
-[^unsuccess]: https://www.nfl.com/videos/teddy-bridgewater-courtland-sutton-barely-miss-out-on-chance-at-62-yard-td-bomb
+[^unsuccess]: A real-game footage of this play is available at https://www.nfl.com/videos/teddy-bridgewater-courtland-sutton-barely-miss-out-on-chance-at-62-yard-td-bomb
 
@@
 We clearly see that players tend to generate more STRAIN when a play ends in a hit, sack, or hurry, compared to no outcome.
-Within the three pressure metrics, it is not surprised that a sack corresponds to the highest amount of STRAIN.
-[Can we say anything about why NT has highest curve when there's a sack?]
+Within the three pressure metrics, it is not surprised that a sack corresponds to the highest amount of STRAIN, followed by a hit and hurry.
 
@@
 
+<!-- reference levels: DE (rush), T (block) -->
+
 \begin{table}
-\caption{Fixed-effects coefficients estimates for the multilevel model. \label{tab:fixedeff}}
+\caption{Fixed-effects coefficients estimates for the multilevel model for average STRAIN.
+Note that the reference pass rusher level is defensive end (DE), denoted PR:DE; and the reference pass blocker level is tackle (T), denoted PB:T.
+\label{tab:fixedeff}}
 \centering
-\begin{tabular}{rrrrrr}
+\begin{tabular}{lrrrr}
 \hline
- & estimate & se & df & $t$-statistic & $p$-value \\ 
+& estimate & se & $t$-statistic & $p$-value \\ 
 \hline
-Intercept & 0.355 & 0.010 & 332.288 & 35.688 & 0.000 \\ 
-  play\_len & -0.004 & 0.000 & 30653.793 & -38.214 & 0.000 \\ 
-  rush\_posDT & -0.081 & 0.008 & 369.713 & -10.340 & 0.000 \\ 
-  rush\_posinteriorLB & 0.058 & 0.011 & 770.441 & 5.332 & 0.000 \\ 
-  rush\_posNT & -0.106 & 0.012 & 355.999 & -9.173 & 0.000 \\ 
-  rush\_posOLB & 0.062 & 0.008 & 388.506 & 7.695 & 0.000 \\ 
-  rush\_possecondary & 0.145 & 0.012 & 1734.779 & 12.328 & 0.000 \\ 
-  block\_posG & 0.011 & 0.006 & 123.672 & 1.867 & 0.064 \\ 
-  block\_posother & 0.010 & 0.008 & 500.451 & 1.261 & 0.208 \\ 
-  block\_posT & 0.023 & 0.006 & 161.556 & 3.625 & 0.000 \\ 
-  n\_blockers & -0.064 & 0.003 & 30949.514 & -24.788 & 0.000 \\ 
+Intercept & 0.378 & 0.009 & 43.310 & 0.000 \\ 
+Length of play & -0.004 & 0.000 & -38.214 & 0.000 \\
+Number of blockers & -0.064 & 0.003 & -24.788 & 0.000 \\
+$I_{\text{\{PR:DT\}}}$ & -0.081 & 0.008 & -10.340 & 0.000 \\
+$I_{\text{\{PR:interior\}}}$ & 0.058 & 0.011 & 5.332 & 0.000 \\
+$I_{\text{\{PR:NT\}}}$ & -0.106 & 0.012 & -9.174 & 0.000 \\
+$I_{\text{\{PR:OLB\}}}$ & 0.062 & 0.008 & 7.695 & 0.000 \\
+$I_{\text{\{PR:secondary\}}}$ & 0.145 & 0.012 & 12.328 & 0.000 \\
+$I_{\text{\{PB:C\}}}$ & -0.023 & 0.006 & -3.625 & 0.000 \\
+$I_{\text{\{PB:G\}}}$ & -0.011 & 0.005 & -2.290 & 0.023 \\ 
+$I_{\text{\{PB:other\}}}$ & -0.013 & 0.007 & -1.911 & 0.056 \\
 \hline
@@
 
+<!-- (Intercept) & 0.378 & 0.009 & 43.310 & 0.000 \\  -->
+<!--   play\_len & -0.004 & 0.000 & -38.214 & 0.000 \\  -->
+<!--   rush\_posDT & -0.081 & 0.008 & -10.340 & 0.000 \\  -->
+<!--   rush\_posinteriorLB & 0.058 & 0.011 & 5.332 & 0.000 \\  -->
+<!--   rush\_posNT & -0.106 & 0.012 & -9.174 & 0.000 \\  -->
+<!--   rush\_posOLB & 0.062 & 0.008 & 7.695 & 0.000 \\  -->
+<!--   rush\_possecondary & 0.145 & 0.012 & 12.328 & 0.000 \\  -->
+<!--   block\_posC & -0.023 & 0.006 & -3.625 & 0.000 \\  -->
+<!--   block\_posG & -0.011 & 0.005 & -2.290 & 0.023 \\  -->
+<!--   block\_posother & -0.013 & 0.007 & -1.911 & 0.056 \\  -->
+<!--   n\_blockers & -0.064 & 0.003 & -24.788 & 0.000 \\  -->
+
 \begin{table}[ht]
-\caption{ICC. \label{tab:icc}}
+\caption{Intraclass correlation coefficients for the multilevel model for average STRAIN. \label{tab:icc}}
 \centering
-\begin{tabular}{rlr}
+\begin{tabular}{lr}
 \hline

### 2023-05-11 paper/05-discussion.Rmd
fixed effects table

@@
 
+rapm
+
 <!--
 
-(Stuff from BDB draft)
+from bdb submission
+
+As for future directions, STRAIN could be utilized to collectively evaluate the performance of a pass-rushing unit by considering the total amount of STRAIN during a play, which could, in turn, be used to examine the effectiveness of different types of blitzes.
+Moreover, while we focused solely on pass-rushers, STRAIN can be easily extended to the assessment of pass-blockers as a unit by looking at quantities such as the total STRAIN or the maximum STRAIN per frame aggregated across the entire play.
+It is also possible to apply STRAIN to assess individual offensive pass-blockers, provided that a method of matching blockers to rushers was developed.
+
+Our metric, however, is subject to several limitations.
+As currently devised, STRAIN cannot distinguish between a pass-rusher who has to face a single blocker versus a double team.
+This problem could be addressed, at least somewhat, by having a method of matching pass-rushers to blockers as previously mentioned.
+Additionally, there are a few situations such as quarterback out-of-pocket scrambling where STRAIN does not appear to be useful.
+If a pass-rusher is on the left side and the quarterback scrambles out to the left, the STRAIN for that pass-rusher can drop considerably as the distance between the quarterback and pass-rusher increases, which the defender should not necessarily be negatively affected by in a measure of pass-rushing effectiveness.
+-->
+
+<!--
+
+from bdb draft
 

### 2023-05-11 scripts/02_fig_crosby_curves.R

@@ -34,6 +34,7 @@ fig_crosby_curves <- play_lv_den |>
   ) |> 
   ggplot() +
   geom_line(aes(frameId_snap_corrected, value, color = name_fct, alpha = name_fct, group = name), linewidth = 1.2) +
+  geom_vline(xintercept = c(10, 20, 30, 40), linetype = "dashed", alpha = 0.3) +
   labs(x = "Time since snap (seconds)",
        y = "Feature value") +
   scale_x_continuous(breaks = seq(0, 40, 10), labels = 0:4) +

### 2023-05-11 scripts/09_fig_play_unsuccess.R

@@ -51,6 +51,7 @@ fig_crosby_curves_unsucess <- play_unsucess |>
   ) |> 
   ggplot() +
   geom_line(aes(frameId_snap_corrected, value, color = name_fct, alpha = name_fct, group = name), linewidth = 1.2) +
+  # geom_vline(xintercept = c(10, 20, 30), linetype = "dashed", alpha = 0.3) +
   labs(x = "Time since snap (seconds)",
        y = "Feature value") +
   scale_x_continuous(breaks = seq(0, 40, 10), labels = 0:4) +

### 2023-05-11 scripts/10_multilevel_mod.R

@@ -122,7 +122,10 @@ mod_df_final <- mod_df |>
          dis = sqrt((x_rush - x_block) ^ 2 + (y_rush - y_block) ^ 2)) |>
   group_by(gameId, playId, nflId) |> 
   slice_min(dis, n = 1) |> 
-  left_join(n_blockers)
+  ungroup() |> 
+  left_join(n_blockers) |> 
+  mutate(block_pos = relevel(factor(block_pos), "T"),
+         rush_pos = relevel(factor(rush_pos), "DE"))
 
 
 library(lme4)  

### 2023-05-11 paper/03-methods.Rmd
small fix to fixed effects

@@
 
-To this end, we fit a multilevel model to evaluate pass rushers' impact on the average STRAIN observed in a play, while accounting for their pass blocker, QB, and play-level information.
+To this end, we fit a multilevel model to evaluate pass rushers' impact on the average STRAIN observed in a play, while accounting for their pass blocker, QB, and play-level attributes.
 We identify the pass blocker linked with the pass rusher of interest using the data provided by PFF.
-Since there can be multiple blockers linked to a rusher, for simplicity we only consider the nearest blocker positioned to the pass rusher at the start of the play.
+Since there can be multiple blockers linked to a rusher, for simplicity, we consider the nearest blocker positioned to the pass rusher at the start of the play.
 We use random intercepts for the three player groups: (1) pass rushers as $R$, (2) nearest pass blockers as $B$, and (3) QBs as $Q$.

### 2023-05-11 paper/04-results.Rmd
small fix to fixed effects

@@
 \caption{Fixed-effects coefficients estimates for the multilevel model for average STRAIN.
-Note that the reference pass rusher level is defensive end (DE), denoted PR:DE; and the reference pass blocker level is tackle (T), denoted PB:T.
+Note that the reference pass rusher level is defensive end (DE), denoted R:DE; and the reference pass blocker level is tackle (T), denoted B:T.
 \label{tab:fixedeff}}
@@
 Number of blockers & -0.064 & 0.003 & -24.788 & 0.000 \\
-$I_{\text{\{PR:DT\}}}$ & -0.081 & 0.008 & -10.340 & 0.000 \\
-$I_{\text{\{PR:interior\}}}$ & 0.058 & 0.011 & 5.332 & 0.000 \\
-$I_{\text{\{PR:NT\}}}$ & -0.106 & 0.012 & -9.174 & 0.000 \\
-$I_{\text{\{PR:OLB\}}}$ & 0.062 & 0.008 & 7.695 & 0.000 \\
-$I_{\text{\{PR:secondary\}}}$ & 0.145 & 0.012 & 12.328 & 0.000 \\
-$I_{\text{\{PB:C\}}}$ & -0.023 & 0.006 & -3.625 & 0.000 \\
-$I_{\text{\{PB:G\}}}$ & -0.011 & 0.005 & -2.290 & 0.023 \\ 
-$I_{\text{\{PB:other\}}}$ & -0.013 & 0.007 & -1.911 & 0.056 \\
+$I_{\text{\{R:DT\}}}$ & -0.081 & 0.008 & -10.340 & 0.000 \\
+$I_{\text{\{R:interior\}}}$ & 0.058 & 0.011 & 5.332 & 0.000 \\
+$I_{\text{\{R:NT\}}}$ & -0.106 & 0.012 & -9.174 & 0.000 \\
+$I_{\text{\{R:OLB\}}}$ & 0.062 & 0.008 & 7.695 & 0.000 \\
+$I_{\text{\{R:secondary\}}}$ & 0.145 & 0.012 & 12.328 & 0.000 \\
+$I_{\text{\{B:C\}}}$ & -0.023 & 0.006 & -3.625 & 0.000 \\
+$I_{\text{\{B:G\}}}$ & -0.011 & 0.005 & -2.290 & 0.023 \\ 
+$I_{\text{\{B:other\}}}$ & -0.013 & 0.007 & -1.911 & 0.056 \\
 \hline

### 2023-05-11 paper/00-main.Rmd
fixed effects table
abstract

@@
 
-# abstract from BDB submission form
-# rewrite this
 abstract: |
-  This paper evaluates pass rusher performance in the National Football League.
-We propose STRAIN, a new measure for pass rusher effectiveness, which accounts for the distance between the pass rusher and quarterback and the rate at which this distance is being reduced.
-STRAIN is a simple and interpretable statistic, with higher STRAIN corresponds to greater pass rushing ability.
-Our metric is closely related to the concept of "strain rate" in materials science.
-We present the mathematical motivations for our metric, followed by an illustration with an example play.
-The analysis concludes with a comparison of STRAIN for different positions and our rankings of NFL's best pass-rushers according to our metric.
+  In American football, existing metrics for assessing pass rush are either discrete-time quantities or based on subjective judgement.
+Using player tracking data, we propose STRAIN, a novel metric for evaluating pass rushers in the National Football League (NFL) at the continuous-time within-play level.
+Inspired by the concept of strain rate in materials science, STRAIN is a simple and interpretable means for measuring defensive pressure in football.
+It is a directly-observed quantity as a function of two features: the distance between the pass rusher and quarterback, and the rate at which this distance is being reduced.
+Our metric possesses great predictability of pressure and stability over time.
+We use a multilevel model to isolate the contribution of every pass rusher in a play while controlling for their pass blocker matchup, the quarterback, and play-level information.
+We apply our approach to NFL data and present results for the first 8 weeks of the 2021 season.
+These include comparisons of STRAIN for different positions and play outcomes, and rankings of the NFL's best pass rushers according to our metric.
 
@@
 output: rticles::asa_article
-# header-includes:
+header-includes:
+  \usepackage{mathptmx}
   # \renewcommand{\arraystretch}{0.8}

### 2023-05-11 paper/01-introduction.Rmd
abstract

@@
 Since its inception, the BDB has chosen a different theme each year leading to new insight about evaluating different positions such as running backs, defensive backs, and special teams.
-The [2023 edition of the NFL BDB](https://www.kaggle.com/competitions/nfl-big-data-bowl-2023) asked participants to evaluate lineman on passing plays [@Howard2023NFL].
+The 2023 edition of the NFL Big Data Bowl asked participants to evaluate linemen on passing plays [@Howard2023NFL].
+
+## Linemen in American Football
+
+<!-- Importance of D-linemen, pass rush (Maybe talk about role of each position, edge vs interior, defensive formation) -->
 
@@
 Below, we provide a brief overview of existing pass rush metrics.
-
-<!-- Previous research bdb -->
-<!-- this year: linemen -->
-<!-- Importance of D-linemen, pass rush (Maybe talk about role of each position, edge vs interior, defensive formation) -->
 
@@
 Hits & A defender tackles the QB behind the line of scrimmage after the QB is able to throw a pass. \\
-Hurries & A defender pressures the QB behind the line of scrimmage such that the QB must release the ball prior sooner than they wanted to. \\
+Hurries & A defender pressures the QB behind the line of scrimmage forcing the QB to throw the ball prior sooner than intended. \\
 Pressures & $\text{Hurries} + \text{ Hits } + \text{ Sacks}$ \\

### 2023-05-11 paper/03-methods.Rmd
abstract

@@
 
-## Application to pass rush in football
+## Application to pass rush in football {#sec:app}
 
@@
 This provides us with average individual effects on STRAIN for each player while applying shrinkage, which is a useful property in this problem's context since we do not observe the same number of plays for each player.
-With regards to the fixed-effects about pass rusher $j$ in play $i$, we include the length of the play (in terms of elapsed time between ball snap and a QB event), the positions of pass rusher and nearest blocker, and the number of pass blockers assigned to the pass rusher based on the PFF data.
+With regards to the fixed-effects about pass rusher $j$ in play $i$, we include the length of the play (i.e.
+the elapsed time between the ball snap and a QB event), the positions of pass rusher and nearest blocker, and the number of pass blockers assigned to the pass rusher based on the PFF data.
 Table \ref{tab:pos} shows our positional categorization for the pass rush and pass block roles.

### 2023-05-11 paper/04-results.Rmd
abstract

@@
 
-Since STRAIN is observed continuously for every play in our data, this allows us to aggregate across all frames played and compute the average STRAIN for NFL pass rushers over the course of the eight-week sample size.
+Since STRAIN is observed continuously for every play in our data, this allows us to aggregate across all frames played and compute the average STRAIN for NFL pass rushers over the course of the eight-week sample size, as discussed in Section \ref{sec:app}.
 Based on the clearly distinct patterns for different positions as previously observed, we evaluate interior pass-rushers (NT's and DT's) separately from edge rushers (OLB's and DE's).
@@
 #| label: fig_rankings_boot
-#| fig.cap: "Top 10 pass rushers for each position (DE, DT, NT, OLB) by mean varying intercept."
+#| fig.cap: "Top 10 pass rushers for each position (DE, DT, NT, OLB) by mean varying intercept with 95% confidence intervals."
 #| fig.width: 6

### 2023-05-11 paper/00-main.Rmd
more abstract edits

@@
 title: |
-  | Bearing the STRAIN: Analyzing Pass Rush in American Football with Player Tracking Data
+  | \Large Bearing the STRAIN: Analyzing Pass Rush in American Football with Player Tracking Data
 
@@
 abstract: |
-  In American football, existing metrics for assessing pass rush are either discrete-time quantities or based on subjective judgement.
+  In American football, a pass rush is an attempt by the defensive team to disrupt the offense and prevent the quarterback (QB) from completing a pass.
+Existing metrics for assessing pass rush are either discrete-time quantities or based on subjective judgement.
 Using player tracking data, we propose STRAIN, a novel metric for evaluating pass rushers in the National Football League (NFL) at the continuous-time within-play level.
 Inspired by the concept of strain rate in materials science, STRAIN is a simple and interpretable means for measuring defensive pressure in football.
-It is a directly-observed quantity as a function of two features: the distance between the pass rusher and quarterback, and the rate at which this distance is being reduced.
+It is a directly-observed statistic as a function of two features: the distance between the pass rusher and QB, and the rate at which this distance is being reduced.
 Our metric possesses great predictability of pressure and stability over time.
-We use a multilevel model to isolate the contribution of every pass rusher in a play while controlling for their pass blocker matchup, the quarterback, and play-level information.
+We use a multilevel model to isolate the contribution of every pass rusher in a play to the average STRAIN while controlling for their pass blocker matchup, the QB, and play-level information.
 We apply our approach to NFL data and present results for the first 8 weeks of the 2021 season.
-These include comparisons of STRAIN for different positions and play outcomes, and rankings of the NFL's best pass rushers according to our metric.
+In particular, we provide comparisons of STRAIN for different defensive positions and play outcomes, and rankings of the NFL's best pass rushers according to our metric.
 
@@
 header-includes:
-  \usepackage{mathptmx}
+  # \usepackage{mathptmx}
   # \renewcommand{\arraystretch}{0.8}

### 2023-05-11 paper/01-introduction.Rmd
update intro with position paragraph

@@
 
-## Linemen in American Football
+We limit our attention in this manuscript to the defensive side of the ball.
+There are two main types of defensive linemen in American football: (1) defensive tackles (DTs) located within the interior of the line and (2) defensive ends (DEs) located along the edges (see Figure INSERT FIGURE NUMBER HERE for example formation with DTs and DEs).
+The primary purpose of both positions is to rush the QB on passing plays, with DEs displaying superiority in observed pass rushing ability [@Eager2018nfl].
+Additionally, within DTs there are nose tackles (NTs) who directly line up across from the ball at the line of scrimmage (see Figure INSERT FIGURE NUMBER HERE for example defensive formation with a NT).
+Typically, teams will employ either one NT or two DTs on the interior with DEs along either side of the defensive line.
+In addition to defensive lineman, other positions may attempt to rush the QB on blitzing plays such as outside linebackers (OLBs), interior linebackers (ILBs), and potentially members of the secondary (cornerback, free safety, strong safety) who typically play pass coverage.
 
-<!-- Importance of D-linemen, pass rush (Maybe talk about role of each position, edge vs interior, defensive formation) -->
+INSERT FORMATION FIGURE X
 
-[I think we need a couple of paragraphs here to introduce the defensive linemen position, and the role/significance of linemen (kinda similar to section 2 of rishav's paper https://arxiv.org/pdf/1906.11373.pdf)]
+INSERT FORMATION FIGURE Y
 

### 2023-05-11 paper/02-data.Rmd
fixed effects table
Some small edits in data section

@@
 In the forthcoming analysis, we rely on the data from the 2023 NFL Big Data Bowl provided by the NFL Next Gen Stats tracking system.
-The data corresponds to 8557 passing plays across 122 games in the first eight weeks of the 2021 NFL regular season. 
+The data corresponds to 8,557 passing plays across 122 games in the first eight weeks of the 2021 NFL regular season. 
 For each play, we have information on the on-field location, speed, angle, direction, and orientation of each player on the field and the football at a rate of 10 Hz, along with event annotations for each frame such as ball snap, pass forward, and quarterback sack, to name a few.
@@
 In addition, Figure \ref{fig:fig_field_raw} presents the locations of every Las Vegas (in black) and Denver (in orange) player on the field from this play at 1, 2, 3 and 4 seconds after the ball snap, with Maxx Crosby highlighted in blue.
+
+[^crosbysack]: A real-game footage of this play is available at https://www.raiders.com/video/de-maxx-crosby-sacks-qb-teddy-bridgewater-for-a-loss-of-6-yards-nfl
 
@@
 For context, Figure \ref{fig:pass_rushers_count} displays the distribution of the number of observed pass rushers involved in a play, ranging from 1 to 8 with 4 pass rushers as the most common value.
-We also leverage the PFF data to count how many hits, hurries, and sacks each pass rusher is credited with across span of collected data. 
+We also leverage the PFF data to count how many hits, hurries, and sacks each pass rusher is credited with across the span of observed data. 
 Additionally, we use the PFF player roles to identify which players were labeled as blocking each pass rusher to adjust for opponent strength, as discussed in Section \ref{sec:methods}.

### 2023-05-11 scripts/99_scratch.R
defense formation code

@@ -20,3 +20,31 @@ test_summed |>
         legend.text = element_text(size = rel(0.8)),
         axis.text = element_text(size = rel(0.8)),
         panel.grid.minor = element_blank())
+
+# defense formation
+tibble(x = c(3:5, 3:6 - 0.5, c(2, 3.7, 4.3, 6)),
+       y = c(1, 1, 1, 1.01, 1.01, 1.01, 1.01, 1.02, 1.02, 1.02, 1.02),
+       pos = c("DE", "NT", "DE", "OLB", "ILB", "ILB", "OLB",
+               "CB", "S", "S", "CB"),
+       d = "3–4 Defense") |> 
+  bind_rows(
+    tibble(x = c(3:6, c(2.5, 4.5, 6.5), c(2, 4, 5, 7)),
+           y = c(1, 1, 1, 1, 1.01, 1.01, 1.01, 1.02, 1.02, 1.02, 1.02),
+           pos = c("DE", "DT", "DT", "DE", "OLB", "MLB", "OLB",
+                   "CB", "S", "S", "CB"),
+           d = "4-3 Defense") 
+  ) |> 
+  mutate(d = fct_relevel(d, "4-3 Defense")) |> 
+  ggplot(aes(x, y)) +
+  geom_tile(fill = "white") +
+  geom_label(aes(label = pos), size = rel(4)) +
+  geom_hline(yintercept = 0.997, linetype = "dashed") +
+  facet_wrap(~ d, scales = "free", ncol = 1) +
+  coord_cartesian(ylim = c(0.99, 1.03)) +
+  theme_light() +
+  theme(strip.text = element_text(size = rel(1.2)),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        axis.title = element_blank(),
+        panel.grid = element_blank())
+

### 2023-05-12 paper/00-main.Rmd
added defensive formation figure

@@
 library(here)
+library(tidyverse)
 ```

### 2023-05-12 paper/01-introduction.Rmd
added defensive formation figure

@@
 We limit our attention in this manuscript to the defensive side of the ball.
-There are two main types of defensive linemen in American football: (1) defensive tackles (DTs) located within the interior of the line and (2) defensive ends (DEs) located along the edges (see Figure INSERT FIGURE NUMBER HERE for example formation with DTs and DEs).
+There are two main types of defensive linemen in American football: (1) defensive tackles (DTs) located within the interior of the line and (2) defensive ends (DEs) located along the edges (see Figure \ref{fig:formation} (top) for example formation with DTs and DEs).
 The primary purpose of both positions is to rush the QB on passing plays, with DEs displaying superiority in observed pass rushing ability [@Eager2018nfl].
-Additionally, within DTs there are nose tackles (NTs) who directly line up across from the ball at the line of scrimmage (see Figure INSERT FIGURE NUMBER HERE for example defensive formation with a NT).
+Additionally, within DTs there are nose tackles (NTs) who directly line up across from the ball at the line of scrimmage (see Figure \ref{fig:formation} (bottom) for example defensive formation with a NT).
 Typically, teams will employ either one NT or two DTs on the interior with DEs along either side of the defensive line.
@@
 
-INSERT FORMATION FIGURE X
-
-INSERT FORMATION FIGURE Y
+```{r formation, fig.width=5, fig.height=4, fig.cap="Two common defensive alignments in football: 4-3 defense (top) and 3-4 defense (bottom).
+The dashed line represents the line of scrimmage separating the defense and offense."}
+tibble(x = c(3:5, 3:6 - 0.5, c(2, 3.7, 4.3, 6)),
+       y = c(1, 1, 1, 1.01, 1.01, 1.01, 1.01, 1.02, 1.02, 1.02, 1.02),
+       pos = c("DE", "NT", "DE", "OLB", "ILB", "ILB", "OLB",
+               "CB", "S", "S", "CB"),
+       d = "3-4 Defense") |> 
+  bind_rows(
+    tibble(x = c(3:6, c(2.5, 4.5, 6.5), c(2, 4.1, 4.9, 7)),
+           y = c(1, 1, 1, 1, 1.01, 1.01, 1.01, 1.02, 1.02, 1.02, 1.02),
+           pos = c("DE", "DT", "DT", "DE", "OLB", "MLB", "OLB",
+                   "CB", "S", "S", "CB"),
+           d = "4-3 Defense") 
+  ) |> 
+  mutate(d = fct_relevel(d, "4-3 Defense", "3-4 Defense")) |> 
+  ggplot(aes(x, y)) +
+  geom_tile(fill = "white") +
+  geom_label(aes(label = pos), size = rel(4)) +
+  geom_hline(yintercept = 0.996, linetype = "dashed") +
+  facet_wrap(~ d, scales = "free", ncol = 1) +
+  coord_cartesian(ylim = c(0.99, 1.03)) +
+  theme_light() +
+  theme(strip.text = element_text(size = rel(1.2)),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        axis.title = element_blank(),
+        panel.grid = element_blank())
+```
 

### 2023-05-12 paper/04-results.Rmd
added defensive formation figure

@@
 #| label: fig_rankings_boot
-#| fig.cap: "Top 10 pass rushers for each position (DE, DT, NT, OLB) by mean varying intercept with 95% confidence intervals."
+#| fig.cap: "Top 10 pass rushers for each position (DE, DT, NT, OLB) by mean varying intercept with 95\\% confidence intervals."
 #| fig.width: 6

### 2023-05-12 paper/01-introduction.Rmd
more intro

@@
 
-Officially tracked by the NFL since 1982, the most well-known statistic for evaluating defensive linemen on pass rushing plays is the sack.
-A sack is recorded when a defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass.
+Perhaps the most commonly-known statistics for evaluating defensive linemen on pass rush plays are sacks, hits, and hurries, which are observed at the discrete play-level.
+Officially tracked by the NFL since 1982, a sack is recorded when a defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass.
 Sacks represent a simple binary measure of success for a rusher on a pass play.
 However, there are many intermediate outcomes for a rusher that can also be considered successes that do not result in a sack.
-Measures such a hits and hurries are collected by various outlets where hits counts the number of times a QB is hit while passing and hurries counts how often a QB has to throw the ball earlier than they would have liked to.
+Other traditional box-score statistics such hits and hurries are collected by various outlets where hits counts the number of times a QB is hit while passing and hurries counts how often a QB has to throw the ball earlier than they would have liked to.
+The sum of sacks, hits, and hurries has been defined as "pressures" and is an improvement over sacks, but suffers from problems of subjectivity (e.g.
+was the play actually a hurry?) and are still binary outcomes (i.e.
+There is a continuum of plays that aren't hurries that all counted as simply "not hurries").
+Pro football focus (PFF) comes up with a metric called "pass-rush productivity" (PRP), which is a minor modification from the aforementioned pressures metric (see Table \ref{tab:metrics}).
+In particular, PRP gives twice as much weight to a sack relative to hurries and hits, which is a slight improvement to pressures.
+However, the choice of weights is ad-hoc and still only measures binary outcomes similar to the shortcomings of previously discussed metrics.
 
@@
 
-The sum of sacks, hits, and hurries has been defined as "pressures" and is an improvement over sacks, but suffers from problems of subjectivity (e.g.
-was the play actually a hurry?) and are still binary outcomes (i.e.
-There is a continuum of plays that aren't hurries that all counted as simply "not hurries").
-
 <!-- $$\frac{(\text{Hurries} + \text{Hits})/2 + \text{Sacks}}{\text{ Pass Rush Snaps}}$$ -->
-
-Pro football focus (PFF) comes up with a metric called "pass-rush productivity" (PRP), which is a minor modification from the aforementioned pressures metric (see Table \ref{tab:metrics}).
-In particular, PRP gives twice as much weight to a sack relative to hurries and hits, which is a slight improvement to pressures.
-However, the choice of weights is ad-hoc and still only measures binary outcomes similar to the shortcomings of previously discussed metrics.
 
@@
 
-More recently metrics such as Time In Pocket (TIP) and Pass Rush Win Rate (PRWR) [@Burke2018Created] have also been proposed and are substantial improvements over the less sophisticated counting statistics.
+More recently proposed metrics such as Time In Pocket (TIP) and Pass Rush Win Rate (PRWR) [@Burke2018Created] are substantial improvements over the less sophisticated counting statistics.
 Nevertheless, these are still imperfect measures.
 TIP is highly context dependent, as it can be influenced by the type of passing route (e.g. deep versus shallow).
-PRWR is based on whether a pass rusher is able to beat their block before a fixed time (ESPN chooses 2.5 seconds) from the snap, but this is highly dependent on the rather arbitrary choice of time used to define a "win".
-In addition, once a cut-off is chosen, this metric becomes binary like most of the previously discussed measures.
+PRWR is created using player tracking data, which is at a more granular level than previous measures.
+It demonstrates whether a pass rusher is able to beat their block before a fixed time (ESPN chooses 2.5 seconds) from the snap.
+This is highly dependent on the rather arbitrary choice of time used to define a "win".
+In addition, once a cut-off is chosen, this metric converts continuous data to a win/loss indicator, becoming binary like most of the previously discussed measures.
 
@@
 
+<!-- https://www.taylorfrancis.com/chapters/edit/10.1201/9781315166070-12/forecasting-performance-college-prospects-selected-national-football-league-draft-julian-wolfson-vittorio-addona-robert-schmicker -->
+
 The literature on the performance of linemen, either offensive or defensive in football is scant.
@@
 The paper uses survival analysis to model time in pocket and completion percentage for quarterbacks, before ultimately measures a lineman's contribution to the overall offense in terms of yards gained.
+
+<!-- The academic literature on the performance of linemen, either offensive or defensive in football is scant.
+@AlamarGould2008 studied the the relationship between pass completion rate and performance of offensive linemen in terms of how successful they were at blocking.
+The data for this study was collecting essentially by hand by watching and charting the performance of each lineman on a play.
+This was done using games from the beginning 2007 season for seven teams and their first three games each.
+The authors note that there is not much research on linemen saying "One of the principal reasons for this lack of study is the absence of publicly available data that is needed in order to track the performance of a specific lineman."  -->
+
+<!-- Later, @AlamarGoldner2011 again studied offensive linemen performance.
+They analyzed 20 teams during the 2010 season for a total of 1,259 plays.
+They don't estimate the performance of an individual lineman, but rather positions on a team (e.g.
+Chicago Bears' left tackle, Pittsburgh Steelers' center, etc.).
+They then go on to use a survival analysis model to estimate time in pocket.
+The average probability of success from the survival model is then used as a covariate in a model to estimate the probability of competing a pass, and they show that there is a significant relationship between these two probabilities.
+-->
 
@@
 We first describe the player tracking data provided by 2023 Big Data Bowl in Section \ref{sec:data}.
-We then introduces the formal definition of our metric STRAIN and our modeling approach in Section \ref{sec:methods}.
+We then introduces the mathematical motivation and definition of our metric STRAIN and our modeling approach in Section \ref{sec:methods}.
 We present an real-game illustration of the proposed metric and study its performance in Section \ref{sec:results}.

### 2023-05-12 paper/00-main.Rmd
some writing on multilevel results

@@
   In American football, a pass rush is an attempt by the defensive team to disrupt the offense and prevent the quarterback (QB) from completing a pass.
-Existing metrics for assessing pass rush are either discrete-time quantities or based on subjective judgement.
+Existing metrics for assessing pass rush performance are either discrete-time quantities or based on subjective judgement.
 Using player tracking data, we propose STRAIN, a novel metric for evaluating pass rushers in the National Football League (NFL) at the continuous-time within-play level.
@@
 header-includes:
-  # \usepackage{mathptmx}
+  \usepackage{mathptmx}
   # \renewcommand{\arraystretch}{0.8}

### 2023-05-12 paper/04-results.Rmd
some writing on multilevel results

@@
 
-## Positional STRAIN curves
+## Positional STRAIN curves {#sec:positional}
 
@@
 
+[CHECK THIS, MAKE SURE WORDING IS CORRECT/NOT CONFUSING]
+
 There are a variety of quantities we are interested in from the multilevel model fit mentioned in Section \ref{sec:methods}.
 First, Table \ref{tab:fixedeff} displays the results for the fixed effects terms.
-[INSERT CONCLUSIONS ABOUT THESE VARIABLES]
+It appears that that the average STRAIN decreases by 0.042 (95% CI [-0.044, -0.040]) for every extra second in play length, and by 0.064 (95% CI [-0.069, -0.059]) every additional blocker in the offensive unit, after accounting for other covariates in the model.
+In other words, NFL pass rushers tend to generate more pressure in plays with shorter length and when facing fewer number of blockers, which totally makes sense.
+
+We observe statistically significant differences among the pass rush and pass block positions in most cases, while controlling for other variables.
+The coefficient estimates for the pass rush position terms reveal that players with positions closer to the line of scrimmage (DT and NT) appear to generate less STRAIN (relative to DE) than those that are further back, which is consistent with our results in Section \ref{sec:positional}.
+For blockers, centers are those that absorb the most pressure on average, compared to the baseline level tackles, guards, and other offensive positions.
+
+<!-- [INSERT CONCLUSIONS ABOUT THESE VARIABLES] -->
+<!-- [SAY SOMETHING ABOUT POSITION] -->
+
 Next, Table \ref{tab:icc} displays the intraclass correlation coefficients (ICC) for the three different grouped terms, describing the proportion variance explained between the group terms in comparison to the residual variance.
@@
 \begin{table}
-\caption{Fixed-effects coefficients estimates for the multilevel model for average STRAIN.
+\caption{Fixed-effects coefficient estimates for the multilevel model for average STRAIN.
 Note that the reference pass rusher level is defensive end (DE), denoted R:DE; and the reference pass blocker level is tackle (T), denoted B:T.
@@
 Intercept & 0.378 & 0.009 & 43.310 & 0.000 \\ 
-Length of play & -0.004 & 0.000 & -38.214 & 0.000 \\
+Length of play & -0.042 & 0.001 & -38.214 & 0.000 \\
 Number of blockers & -0.064 & 0.003 & -24.788 & 0.000 \\

### 2023-05-12 scripts/10_multilevel_mod.R

@@ -125,7 +125,8 @@ mod_df_final <- mod_df |>
   ungroup() |> 
   left_join(n_blockers) |> 
   mutate(block_pos = relevel(factor(block_pos), "T"),
-         rush_pos = relevel(factor(rush_pos), "DE"))
+         rush_pos = relevel(factor(rush_pos), "DE"),
+         play_len = play_len / 10)
 
 
 library(lme4)  

### 2023-05-12 paper/04-results.Rmd
more work

@@
 #| label: fig_crosby_curves
-#| fig.cap: "Changes in STRAIN, velocity, and distance from the QB for Maxx Crosby over the course of a successful pass rush play that results in a sack."
+#| fig.cap: "Changes in STRAIN, velocity, and distance from the QB for Maxx Crosby over the course of a successful pass rush play that results in a sack.
+The STRAIN for other pass rushers throughout this play is also displayed."
 #| fig.width: 4.8
@@
 #| label: fig_crosby_curves_unsucess
-#| fig.cap: "Changes in STRAIN, velocity, and distance from the QB for Maxx Crosby over the course of an unsuccessful pass rush play."
+#| fig.cap: "Changes in STRAIN, velocity, and distance from the QB for Maxx Crosby over the course of an unsuccessful pass rush play.
+The STRAIN for other pass rushers throughout this play is also displayed."
 #| fig.width: 4.8
@@
 
-[CHECK THIS, MAKE SURE WORDING IS CORRECT/NOT CONFUSING]
+[CHECK/EDIT THIS, MAKE SURE WORDING IS CORRECT/NOT CONFUSING]
 
@@
 The coefficient estimates for the pass rush position terms reveal that players with positions closer to the line of scrimmage (DT and NT) appear to generate less STRAIN (relative to DE) than those that are further back, which is consistent with our results in Section \ref{sec:positional}.
-For blockers, centers are those that absorb the most pressure on average, compared to the baseline level tackles, guards, and other offensive positions.
+For blockers, centers are those that absorb the most pressure on average, more than guards and other offensive positions (compared to the baseline level, tackles).
 
@@
 We observe different levels of variability across these top players, with the only a subset of players with intervals strictly above zero.
+
 [NEED TO THINK ABOUT WHAT MORE TO SAY HERE]

### 2023-05-12 paper/00-main.Rmd
more

--- 2023-05-12 paper/00-main.Rmd
@@
 
-**GitHub repository:** https://github.com/getstrained/intro-strain.
-
-**NFL Big Data Bowl 2023 Data:**: https://www.kaggle.com/competitions/nfl-big-data-bowl-2023/data.
+All code related to this paper is available at https://github.com/getstrained/intro-strain.
+The data provided by the NFL Big Data Bowl 2023 is available at: https://www.kaggle.com/competitions/nfl-big-data-bowl-2023/data.
 

### 2023-05-12 paper/01-introduction.Rmd
more

@@
 
-In this work, using data made available in the 2023 BDB, we present a novel approach to measure the performance of pass rushers (defensive players whose role on a given play is to put pressure on the QB).
+In this work, using data made available in the 2023 BDB, we present a novel approach to measure the performance of pass rushers.
 Relative to other aspects of American football, such as QB evaluation [@burke2019deepqb, @reyers2021quarterback], the literature on evaluating pass rushers is scarce.
@@
 
+<!-- (defensive players whose role on a given play is to put pressure on the QB) -->
+
 ## Previous Pass Rush Metrics
 
-Table \ref{tab:metrics} gives a summary of existing pass-rush metrics.
+Table \ref{tab:metrics} gives a summary of existing football metrics for pass rush 
 We now highlight what these quantities measure as well as their limitations.
 
-Perhaps the most commonly-known statistics for evaluating defensive linemen on pass rush plays are sacks, hits, and hurries, which are observed at the discrete play-level.
-Officially tracked by the NFL since 1982, a sack is recorded when a defensive player tackles the quarterback behind the line of scrimmage before the quarterback is able to throw a pass.
+[NEED TO ADD "NGS Get Off" https://fivethirtyeight.com/features/the-nfl-has-a-new-way-to-measure-the-explosiveness-of-pass-rushers/]
+
+Perhaps the most commonly-known statistics for evaluating defensive linemen on pass rush plays are sacks, hits, and hurries, which are discretely observed at the play-level.
+Officially tracked by the NFL since 1982, a sack is recorded when a defender tackles the QB behind the line of scrimmage before the QB releases a pass.
 Sacks represent a simple binary measure of success for a rusher on a pass play.
@@
 Pass Rush Win Rate & Rate at which pass rusher beats pass block within 2.5 seconds after ball snap\\
+NGS Get Off & add this\\
 \hline
@@
 
-The literature on the performance of linemen, either offensive or defensive in football is scant.
+The peer-reviewed literature on measuring the performance of football linemen (either offensive or defensive) is scant.
 @AlamarGould2008 find an association between pass completion rate and successful pass blocking by offensive linemen.
 The data for this study are collected for the first three weeks of the 2007 NFL season, manually recording whether a lineman holds a block and the time it took for quarterback to throw the football.
-Later, @AlamarGoldner2011 use manually-tracked data during the 2010 season to estimate lineman performance for different team-positions instead of individual defenders (e.g.
+@AlamarGoldner2011 later follow up by using manually-tracked data for the 2010 season to estimate lineman performance for different team-positions instead of individual defenders (e.g.
 Chicago Bears' left tackle, Pittsburgh Steelers' center).
-The paper uses survival analysis to model time in pocket and completion percentage for quarterbacks, before ultimately measures a lineman's contribution to the overall offense in terms of yards gained.
+The paper uses survival analysis to model time in pocket and completion percentage for quarterbacks, before proposing a measure for linemen's contribution to their team's passing in terms of yards gained.
+
+@Wolfson2017Forecasting comment on the two aforementioned articles that "[a]lthough these are exciting preliminary steps, there is still a long way to go before we can provide a comprehensive appraisal of the achievements of an individual lineman."
+The challenge here is fundamental, since there were not enough public data at the time to develop any meaningful metric for linemen in football, as also noted by @AlamarGould2008.
+However, with the granularity of player tracking data, we have access to data not only for the linemen but also for every player on the field.
+This provides us with a great opportunity to assess the performance of linemen in football.
 
@@
 
-<!-- [this next part needs re-wording] -->
-<!-- They then go on to use a survival analysis model to estimate time in pocket.
-The average probability of success from the survival model is then used as a covariate in a model to estimate the probability of competing a pass, and they show that there is a significant relationship between these two probabilities.
--->
+## Our Contribution
 
-## Contributions of STRAIN
+In this paper, we focus on the evaluation of defensive linemen in football.
+We devise STRAIN, a metric for measuring pass rush effectiveness, inspired by the concept of strain rate in materials science.
+Our metric gives a continuous measure of pressure for every pass rusher on the football field over the course of an entire play.
+This allows for the assessment of pass rush success even on plays that do not result in an observed outcome like a sack, hit, or hurry.
+We view this as a major step forward for accurately evaluating defensive linemen performance.
+Further, we note that although our focus in this paper is on defensive linemen, our proposed approach can be extended to the evaluation of offensive linemen in American football.
 
-All the aforementioned methods suffered from not having the data that was actually necessary to develop a meaningful metric for linemen in football, either offensive or defensive.
-Now though, thanks to the 2023 Big Data Bowl, we have a publicly available tracking data not only for the linemen, but for every player on the field.
-This allows us to meaningfully assess the performance of linemen in football.
-In this paper, we focus on defensive linemen, and propose STRAIN, a metric for pass rush evaluation inspired by the concept of strain rate in materials science.
-STRAIN gives a continuous measure of pressure for each of the defensive linemen over the course of an entire play.
-This allows for the measurement of pass rush success even on plays that do not result in an observed outcome like a sack, hit, or hurry.
-We view this as a major step forward in the accurate evaluation of defensive linemen performance.
-Further, we note than while we focus exclusively on defensive linemen here, the proposed metric STRAIN is relatively easily extended to the evaluation of offensive linemen performance.
-
-The remainder of this paper is outlined as follows.
+The remainder of this manuscript is outlined as follows.
 We first describe the player tracking data provided by 2023 Big Data Bowl in Section \ref{sec:data}.
-We then introduces the mathematical motivation and definition of our metric STRAIN and our modeling approach in Section \ref{sec:methods}.
-We present an real-game illustration of the proposed metric and study its performance in Section \ref{sec:results}.
-We close with a discussion of future directions for STRAIN in Section \ref{sec:discussion}.
+We then introduce the mathematical motivation and definition of our statistic STRAIN and our modeling approach in Section \ref{sec:methods}.
+Next, we present applications of STRAIN and study different statistical properties of the metric in Section \ref{sec:results}.
+We close with our discussion of future directions related to this work in Section \ref{sec:discussion}.
 

### 2023-05-13 paper/03-methods.Rmd
Edit methods notation and update results description

@@
 Moreover, since we observe STRAIN continuously at every tenth of a second within each play, we can then compute the average STRAIN across all frames played for every pass rusher.
-Formally, the average STRAIN, denoted by $\overline{\text{STRAIN}}$, for pass rusher $i$ over $n$ total plays with $T$ total frames is
-[CHECK THIS - is notation consistent with what we have above?]
+Formally, the average STRAIN, denoted by $\overline{\text{STRAIN}}$, for pass rusher $j$ involved in $n_j$ total plays across $T_j$ total frames is
+
 \begin{equation}
-\overline{\text{STRAIN}}_{i} = \frac{1}{n} \sum_{t=1}^T {\text{STRAIN}}_{it}.
+\overline{\text{STRAIN}}_{j} = \frac{1}{T_j} \sum_{t=1}^{T_j} \widehat{\text{STRAIN}}_{j}(t).
 \end{equation}
@@
 
-## Multilevel Model
+## Multilevel model of play-level STRAIN
 
@@
 
-While average STRAIN is a simple first step for evaluating the pass rush, the observed average STRAIN on a single play is likely due to a convolution of factors.
+While average STRAIN is a simple first step for evaluating the pass rush, we can also measure pass rusher $j$'s observed average STRAIN on a single play $i$ consisting of $T_i$ total frames,
+
+\begin{equation}
+\overline{\text{STRAIN}}_{ij} = \frac{1}{T_i} \sum_{t=1}^{T_i} \widehat{\text{STRAIN}}_{ij}(t).
+\end{equation}
+
+However, the observed average STRAIN on a single play is likely due to a convolution of factors.
 Outside of the pass rusher's ability, there is variability in the opposing strength of pass blockers across plays a pass rusher is involved.
@@
 We use random intercepts for the three player groups: (1) pass rushers as $R$, (2) nearest pass blockers as $B$, and (3) QBs as $Q$.
-The coefficients $\beta$ for additional information $X_{ij}$ about pass rusher $j$ in play $i$ are estimated as fixed-effects. 
+The coefficients $\boldsymbol{\beta}$ for additional information vector of information  $\mathbf{x_{ij}}$ about pass rusher $j$ in play $i$ are estimated as fixed-effects. 
 Our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows.
@@
 \begin{aligned}
-\overline{\text{STRAIN}}_{ij} &\sim N(R_{r[ij]} + B_{b[ij]} + Q_{q[i]} + X_{ij} \cdot \beta, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
-R_{r} &\sim N(\mu_R, \sigma^2_R), \text{ for } r = 1, \cdots, \text{ \# of pass rushers}, \\
+\overline{\text{STRAIN}}_{ij} &\sim N(R_{j[i]} + B_{b[ij]} + Q_{q[i]} + \mathbf{x_{ij}} \cdot \boldsymbol{\beta}, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
+R_{j} &\sim N(\mu_R, \sigma^2_R), \text{ for } j = 1, \cdots, \text{ \# of pass rushers}, \\
 B_{b} &\sim N(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
@@
 This provides us with average individual effects on STRAIN for each player while applying shrinkage, which is a useful property in this problem's context since we do not observe the same number of plays for each player.
-With regards to the fixed-effects about pass rusher $j$ in play $i$, we include the length of the play (i.e.
-the elapsed time between the ball snap and a QB event), the positions of pass rusher and nearest blocker, and the number of pass blockers assigned to the pass rusher based on the PFF data.
+With regards to the fixed-effects about pass rusher $j$ in play $i$, we include the length of the play (i.e., the elapsed time between the ball snap and a QB event), the positions of pass rusher and nearest blocker, and the number of pass blockers assigned to the pass rusher based on the PFF data.
 Table \ref{tab:pos} shows our positional categorization for the pass rush and pass block roles.
 We implement the model using penalized likelihood via the `lme4` package in `R` [@bateslme4; @R2023Language].
+Additionally, we generate estimates of uncertainty for the player-level random intercepts based on 95% confidence intervals generated by sampling from the posterior distribution via parametric bootstrapping with the `merTools` package [@knowlesmerTools, @gelman2006data].
 

### 2023-05-13 paper/04-results.Rmd
Edit methods notation and update results description

@@
 In addition to adjusting for opposition, our multilevel model allows us to provide some notion of uncertainty about the pass rusher's effect on the play level STRAIN.
-Figure \ref{fig:fig_rankings_boot} displays the mean varying intercept for the top ten pass rushers for each position along with 95% confidence intervals generated by sampling from the posterior distributions via parametric bootstrapping using the `merTools` package [@knowlesmerTools, @gelman2006data].
+Figure \ref{fig:fig_rankings_boot} displays the mean varying intercept for the top ten pass rushers for each position along with 95% confidence intervals. 
 We observe similar rankings based on the random intercepts in comparison to the average STRAIN results in Tables \ref{tab:edge} and \ref{tab:interior}, with Myles Garrett (DE), Aaron Donald (DT), Vita Vea (NT), and T.J.
@@
 We observe different levels of variability across these top players, with the only a subset of players with intervals strictly above zero.
-
-[NEED TO THINK ABOUT WHAT MORE TO SAY HERE]
+This is not necessarily surprising given the limited sample of data.
+Additionally, we are ignoring correlated performance between pass rushers on the same team, e.g., INSERT TWO PLAYERS ON SAME TEAM, but leave exploration of this for future work.
 

### 2023-05-13 paper/01-introduction.Rmd
more lit review on metrics

@@
 
-Table \ref{tab:metrics} gives a summary of existing football metrics for pass rush 
-We now highlight what these quantities measure as well as their limitations.
-
-[NEED TO ADD "NGS Get Off" https://fivethirtyeight.com/features/the-nfl-has-a-new-way-to-measure-the-explosiveness-of-pass-rushers/]
-
-Perhaps the most commonly-known statistics for evaluating defensive linemen on pass rush plays are sacks, hits, and hurries, which are discretely observed at the play-level.
-Officially tracked by the NFL since 1982, a sack is recorded when a defender tackles the QB behind the line of scrimmage before the QB releases a pass.
-Sacks represent a simple binary measure of success for a rusher on a pass play.
-However, there are many intermediate outcomes for a rusher that can also be considered successes that do not result in a sack.
-Other traditional box-score statistics such hits and hurries are collected by various outlets where hits counts the number of times a QB is hit while passing and hurries counts how often a QB has to throw the ball earlier than they would have liked to.
-The sum of sacks, hits, and hurries has been defined as "pressures" and is an improvement over sacks, but suffers from problems of subjectivity (e.g.
-was the play actually a hurry?) and are still binary outcomes (i.e.
-There is a continuum of plays that aren't hurries that all counted as simply "not hurries").
-Pro football focus (PFF) comes up with a metric called "pass-rush productivity" (PRP), which is a minor modification from the aforementioned pressures metric (see Table \ref{tab:metrics}).
-In particular, PRP gives twice as much weight to a sack relative to hurries and hits, which is a slight improvement to pressures.
-However, the choice of weights is ad-hoc and still only measures binary outcomes similar to the shortcomings of previously discussed metrics.
-
-<!-- (https://www.pro-football-reference.com/years/2022/advanced.htm) -->
-
-<!-- [Pro football ref has a slightly different QB pressures formula than the formula below https://www.pro-football-reference.com/about/advanced_stats.htm] -->
-
-<!-- $$\frac{(\text{Hurries} + \text{Hits})/2 + \text{Sacks}}{\text{ Pass Rush Snaps}}$$ -->
-
-<!-- (https://www.pro-football-reference.com/about/advanced_stats.htm) -->
-
-More recently proposed metrics such as Time In Pocket (TIP) and Pass Rush Win Rate (PRWR) [@Burke2018Created] are substantial improvements over the less sophisticated counting statistics.
-Nevertheless, these are still imperfect measures.
-TIP is highly context dependent, as it can be influenced by the type of passing route (e.g. deep versus shallow).
-PRWR is created using player tracking data, which is at a more granular level than previous measures.
-It demonstrates whether a pass rusher is able to beat their block before a fixed time (ESPN chooses 2.5 seconds) from the snap.
-This is highly dependent on the rather arbitrary choice of time used to define a "win".
-In addition, once a cut-off is chosen, this metric converts continuous data to a win/loss indicator, becoming binary like most of the previously discussed measures.
-
-<!-- (https://www.espn.com/nfl/story/_/id/34536376/2022-nfl-pass-rushing-run-stopping-blocking-leaderboard-win-rate-rankings-top-players-teams) -->
-
-All of these measures represent different versions of "the best we can do with the available data".
-Without granular data, it is difficult to measure effectiveness of a pass rusher so many simple metrics of success have been proposed.
+Table \ref{tab:metrics} gives a summary of existing football metrics for pass rush.
+We now highlight what these quantities describe as well as their limitations.
 
@@
 Time In Pocket & Time (in seconds) between ball snap and throw or pocket collapse for a QB \\
+NGS Get Off & Average time (in seconds) required for a defender to cross the line of scrimmage after the ball snap.\\
 Pass Rush Win Rate & Rate at which pass rusher beats pass block within 2.5 seconds after ball snap\\
-NGS Get Off & add this\\
 \hline
@@
 \end{table}
+
+Perhaps the most commonly-known statistics for evaluating defensive linemen on pass rush plays are sacks, hits, and hurries, which are discretely observed at the play-level.
+Officially tracked by the NFL since 1982, a sack is recorded when a defender tackles the QB behind the line of scrimmage before the QB releases a pass. 
+Other traditional box score statistics such as hits and hurries are collected by various outlets. 
+A hit is a collision between a defender and the opposing team's quarterback while the QB is in possession of the ball.
+A hurry represents an instance when a defender successfully disrupts without necessarily making direct contact with the QB and forces the QB to make a throw earlier than expected.
+These are all simple binary measures of pass rush outcome for any given play.
+However, for plays that do not result in the aforementioned outcomes (in particular, sack), there are still many intermediate defensive actions on the field within the play that are valuable and can be considered successes.
+
+Moreover, the sum of sacks, hits, and hurries is often defined as "pressures".
+This is an improvement over sacks, but suffers from problems of subjectivity (e.g., whether there is an actual hurry or not).
+Pro football focus (PFF) comes up with a metric called "pass-rush productivity" (PRP), which is a minor modification from the aforementioned pressures metric (see Table \ref{tab:metrics}).
+In particular, PRP gives twice as much weight to a sack relative to hurries and hits, which is a slight improvement to pressures.
+However, the choice of weights is ad-hoc and still only measures binary outcomes, similar to the shortcomings of previously discussed metrics.
+
+<!-- and are still binary outcomes (i.e. There is a continuum of plays that are not hurries that all counted as simply "not hurries"). -->
+
+<!-- (https://www.pro-football-reference.com/years/2022/advanced.htm) -->
+
+<!-- [Pro football ref has a slightly different QB pressures formula than the formula below https://www.pro-football-reference.com/about/advanced_stats.htm] -->
+
+<!-- $$\frac{(\text{Hurries} + \text{Hits})/2 + \text{Sacks}}{\text{ Pass Rush Snaps}}$$ -->
+
+<!-- (https://www.pro-football-reference.com/about/advanced_stats.htm) -->
+
+More recently proposed metrics such as time in pocket (TIP), NGS Get Off [@Hermsmeyer2021nfl], and pass rush win rate (PRWR) [@Burke2018Created] are substantial improvements over the less sophisticated counting statistics, but nevertheless are still imperfect.
+TIP refers to how long a QB can operate within the protected space behind the offensive line, known as the pocket.
+However, this measure is highly context-dependent, as it can be influenced by a number of factors such as the defensive scheme or type of passing route.
+NGS Get Off is an aggregated statistic, illustrating how quickly a defender can get past the line of scrimmage after the snap on average.
+PRWR is created using player tracking data, which is at a more granular level than previous measures.
+It demonstrates whether a pass rusher is able to beat their blocking matchup before a fixed time from the snap (2.5 seconds as chosen by ESPN).
+This depends on the rather arbitrary time threshold used to define a pass rush win.
+In addition, once a cutoff is chosen, this metric converts continuous data to a win-loss indicator, becoming binary like most of the metrics discussed above.
+
+<!-- (e.g. deep versus shallow). -->
+
+<!-- (https://www.espn.com/nfl/story/_/id/34536376/2022-nfl-pass-rushing-run-stopping-blocking-leaderboard-win-rate-rankings-top-players-teams) -->
+
+<!-- All of these measures represent different versions of "the best we can do with the available data". -->
+<!-- Without granular data, it is difficult to measure effectiveness of a pass rusher so many simple metrics of success have been proposed.  -->
 

### 2023-05-13 paper/00-main.Rmd
started on #10

@@
   warning = FALSE,
-  fig.align = "center",
-  fig.pos = "p",
-  cache = TRUE
+  fig.align = "center"
+  # fig.pos = "p",
+  # cache = TRUE
 )
@@
 
-Big data bowl, Mike
+We thank the organizers of the 2023 NFL Big Data Bowl for hosting the competition and providing access to the data.
 

### 2023-05-13 paper/01-introduction.Rmd
started on #10

@@
 
-Moreover, the sum of sacks, hits, and hurries is often defined as "pressures".
-This is an improvement over sacks, but suffers from problems of subjectivity (e.g., whether there is an actual hurry or not).
+Moreover, the sum of sacks, hits, and hurries is often defined as (QB) pressures.
+This is somewhat better than sacks, but suffers from problems of subjectivity (e.g., whether there is an actual hurry or not).
 Pro football focus (PFF) comes up with a metric called "pass-rush productivity" (PRP), which is a minor modification from the aforementioned pressures metric (see Table \ref{tab:metrics}).
-In particular, PRP gives twice as much weight to a sack relative to hurries and hits, which is a slight improvement to pressures.
+In particular, PRP gives twice as much weight to a sack relative to hurries and hits, which is a small upgrade to pressures.
 However, the choice of weights is ad-hoc and still only measures binary outcomes, similar to the shortcomings of previously discussed metrics.

### 2023-05-13 paper/05-discussion.Rmd
started on #10

@@
 # Discussion {#sec:discussion}
+
+We have proposed STRAIN, a simple and interpretable statistic for evaluating pass-rushers, with higher STRAIN corresponds to greater pass-rushing ability. 
+Compared to existing metrics, STRAIN measures pass-rush effectiveness for every play continuously over time, which is at a much more granular level than considering whether the play resulted in a sack or converting continuous data to a binary win/loss.
+
+As for future directions, while we focus solely on pass-rushers in this paper, STRAIN can be easily extended to the assessment of pass-blockers as a unit. 
+This can be accomplished by looking at quantities such as the total STRAIN or the maximum STRAIN per frame aggregated across the entire play.
+It is also possible to apply STRAIN to assess individual offensive pass-blockers, provided that a method of matching blockers to rushers was developed.
 
@@
 
-Critique
+Critique/Limitations
 
-Future work
-
-rapm
+Future work rapm
 

### 2023-05-14 paper/00-main.Rmd
more edits/writing

@@
 Our metric possesses great predictability of pressure and stability over time.
-We use a multilevel model to isolate the contribution of every pass rusher in a play to the average STRAIN while controlling for their pass blocker matchup, the QB, and play-level information.
+We also fit a multilevel model for STRAIN to understand the defensive pressure contribution of every pass rusher at the play-level.
 We apply our approach to NFL data and present results for the first 8 weeks of the 2021 season.
@@
   warning = FALSE,
-  fig.align = "center"
+  fig.align = "center",
   # fig.pos = "p",
-  # cache = TRUE
+  cache = TRUE
 )
@@
 All code related to this paper is available at https://github.com/getstrained/intro-strain.
-The data provided by the NFL Big Data Bowl 2023 is available at: https://www.kaggle.com/competitions/nfl-big-data-bowl-2023/data.
+The data provided by the NFL Big Data Bowl 2023 is available at https://www.kaggle.com/competitions/nfl-big-data-bowl-2023/data.
 

### 2023-05-14 paper/01-introduction.Rmd
more edits/writing

@@
 
-In an attempt to foster analytics and innovate the game, the National Football League (NFL) introduced their player tracking system known as [Next Gen Stats (NGS)](https://operations.nfl.com/gameday/technology/nfl-next-gen-stats/) in 2016, using radio frequency identification (RFID) chips placed in players' shoulder pads (and in the ball) to collect data 10 times per second.
-These NGS tracking data contain information about the on-field locations, speeds, and angles of all 22 players on the field (as well as the ball).
-While these data were initially only available for teams, media, and vendors, in December 2018 the NFL launched the inaugural edition of their annual [Big Data Bowl (BDB)](https://operations.nfl.com/gameday/analytics/big-data-bowl/past-big-data-bowl-recaps/) competition.
+In an attempt to foster analytics and innovate the game, the National Football League (NFL) introduced their player tracking system known as Next Gen Stats (NGS) in 2016 [@nfl2023ngs].
+NGS uses radio frequency identification (RFID) chips placed in players' shoulder pads (and in the ball) to collect data at a rate of 10 frames per second.
+The data captures real-time on-field information such as locations, speeds, and accelerations of all 22 players (as well as the ball).
+While these data were initially only available for teams, media, and vendors, in December 2018 the NFL launched the inaugural edition of their annual Big Data Bowl (BDB) competition [@nfl2023big].
 
@@
 The winners of the inaugural BDB focused on identifying receiver routes via clustering techniques [@chu2020route] and convolutional neural networks [@sterken2019routenet].
-Along with the BDB competition entries, the public release of NGS data allowed researchers to tackle (pun intended) a variety of other problems such as revisiting fourth down decision making [@lopez2020bigger], annotating pass coverage with Gaussian mixture models [@dutta2020unsupervised], and introducing a continuous-time framework to estimate within-play value [@yurko2020going].
+Along with the BDB competition entries, the public release of NGS data allowed researchers to tackle <!-- (pun intended) --> a variety of other problems such as revisiting fourth down decision making [@lopez2020bigger], annotating pass coverage with Gaussian mixture models [@dutta2020unsupervised], and introducing a continuous-time framework to estimate within-play value [@yurko2020going].
 Since its inception, the BDB has chosen a different theme each year leading to new insight about evaluating different positions such as running backs, defensive backs, and special teams.
@@
 
-We limit our attention in this manuscript to the defensive side of the ball.
+In this manuscript, we specifically focus on measuring the performance of defensive linemen in the NFL.
 There are two main types of defensive linemen in American football: (1) defensive tackles (DTs) located within the interior of the line and (2) defensive ends (DEs) located along the edges (see Figure \ref{fig:formation} (top) for example formation with DTs and DEs).
@@
 
-Moreover, the sum of sacks, hits, and hurries is often defined as (QB) pressures.
-This is somewhat better than sacks, but suffers from problems of subjectivity (e.g., whether there is an actual hurry or not).
-Pro football focus (PFF) comes up with a metric called "pass-rush productivity" (PRP), which is a minor modification from the aforementioned pressures metric (see Table \ref{tab:metrics}).
+Moreover, the sum of sacks, hits, and hurries is often defined as pressures.
+This is better than sacks to some extent, but suffers from problems of subjectivity (e.g., whether there is an actual hurry or not).
+Pro football focus (PFF) comes up with a metric called pass-rush productivity (PRP), which is a minor modification from the aforementioned pressures metric (see Table \ref{tab:metrics}).
 In particular, PRP gives twice as much weight to a sack relative to hurries and hits, which is a small upgrade to pressures.
-However, the choice of weights is ad-hoc and still only measures binary outcomes, similar to the shortcomings of previously discussed metrics.
+However, the choice of weights is ad-hoc and still only indicates binary outcomes, similar to the shortcomings of previous metrics.
 
@@
 
-More recently proposed metrics such as time in pocket (TIP), NGS Get Off [@Hermsmeyer2021nfl], and pass rush win rate (PRWR) [@Burke2018Created] are substantial improvements over the less sophisticated counting statistics, but nevertheless are still imperfect.
+More recently proposed metrics such as time in pocket (TIP), NGS get off [@Hermsmeyer2021nfl], and pass rush win rate (PRWR) [@Burke2018Created] are substantial improvements over the less sophisticated counting statistics, but nevertheless are still imperfect.
 TIP refers to how long a QB can operate within the protected space behind the offensive line, known as the pocket.
 However, this measure is highly context-dependent, as it can be influenced by a number of factors such as the defensive scheme or type of passing route.
-NGS Get Off is an aggregated statistic, illustrating how quickly a defender can get past the line of scrimmage after the snap on average.
+NGS get off is an aggregated statistic, illustrating how quickly a defender can get past the line of scrimmage after the snap on average.
 PRWR is created using player tracking data, which is at a more granular level than previous measures.
@@
 This depends on the rather arbitrary time threshold used to define a pass rush win.
-In addition, once a cutoff is chosen, this metric converts continuous data to a win-loss indicator, becoming binary like most of the metrics discussed above.
+In addition, once a cutoff is chosen, PRWR converts continuous data to a win-loss indicator, becoming dichotomous like most of the metrics discussed above.
 
@@
 The data for this study are collected for the first three weeks of the 2007 NFL season, manually recording whether a lineman holds a block and the time it took for quarterback to throw the football.
-@AlamarGoldner2011 later follow up by using manually-tracked data for the 2010 season to estimate lineman performance for different team-positions instead of individual defenders (e.g.
-Chicago Bears' left tackle, Pittsburgh Steelers' center).
+@AlamarGoldner2011 later follow up by using manually-tracked data for the 2010 season to estimate lineman performance for different team-positions instead of individual defenders (e.g., Chicago Bears' left tackle, Pittsburgh Steelers' center, etc.).
 The paper uses survival analysis to model time in pocket and completion percentage for quarterbacks, before proposing a measure for linemen's contribution to their team's passing in terms of yards gained.

### 2023-05-14 paper/02-data.Rmd
more work
more edits/writing

@@
 For each play, we have information on the on-field location, speed, angle, direction, and orientation of each player on the field and the football at a rate of 10 Hz, along with event annotations for each frame such as ball snap, pass forward, and quarterback sack, to name a few.
+
 For our investigation, we consider only the frames between the ball snap and when a pass forward or quarterback sack is recorded for each play.
@@
 Along with the tracking information, the 2023 BDB includes scouting data provided by Pro Football Focus (PFF). 
-PFF's scouting data includes manually collected player-level information at the play level, such as the player's role (e.g., whether they are a pass rusher) and credited events (e.g., player is credited with hitting the QB on the play).
+This contains manually collected player-level information at the play level, such as the player's role (e.g., whether they are a pass rusher and pass blocker) and credited events (e.g., player is credited with hitting the QB on the play).
 In this manuscript, we use the PFF data to identify 36,362 unique players designated in the "pass rush" role across all plays.
-For context, Figure \ref{fig:pass_rushers_count} displays the distribution of the number of observed pass rushers involved in a play, ranging from 1 to 8 with 4 pass rushers as the most common value.
-We also leverage the PFF data to count how many hits, hurries, and sacks each pass rusher is credited with across the span of observed data. 
-Additionally, we use the PFF player roles to identify which players were labeled as blocking each pass rusher to adjust for opponent strength, as discussed in Section \ref{sec:methods}.
+For context, Figure \ref{fig:pass_rushers_count} displays the distribution of the number of observed pass rushers involved in a play, ranging from 1 to 8 with 4 pass rushers (i.e.
+4-man rush formation) as the most common value.
+We also leverage the PFF scouting data to count how many hits, hurries, and sacks each pass rusher is credited with across the span of observed data. 
+Additionally, we use the PFF player roles to identify the blocking matchup for each pass rusher in order to adjust for opponent strength, as discussed in Section \ref{sec:methods}.
 

### 2023-05-14 paper/03-methods.Rmd
more edits/writing

@@
 
-## Strain rate in materials science
+## Motivation and Definition of STRAIN {#sec:defn}
 
@@
 Whereas strain has no units, the strain rate is measured in inverse of time, usually 1/second.
-
-## Application to pass rush in football {#sec:app}
 
@@
 \end{equation}
-This can be helpful for player evaluation, as we determine the most effective pass-rushers based on their average STRAIN values in Section \ref{sec:strainbar}.
-We also use average STRAIN to assess statistical properties of our metric in Section \ref{sec:statprop}.
+This can be helpful for player evaluation, as we determine the most effective pass rushers based on their average STRAIN values in Section \ref{sec:strainbar}.
+We also use average STRAIN to assess different statistical properties of our metric in Section \ref{sec:statprop}.
 
-## Multilevel model of play-level STRAIN
+## Multilevel Model for Play-Level STRAIN
 

### 2023-05-14 paper/04-results.Rmd
more edits/writing

@@
 
-## Example play
+## Real-Game Illustration of STRAIN
 
@@
 STRAIN then increases again until around 2.5-3 seconds after the snap before trending down towards the end of the play.
-In context, this reflects the actions that a pass-rusher initially moves towards the quarterback, but is then stopped by the offensive line while the quarterback drops back.
+In context, this reflects the actions that a pass rusher initially moves towards the quarterback, but is then stopped by the offensive line while the quarterback drops back.
 When the quarterback stops dropping, the rusher closes the gap and increases STRAIN, before slowing down later on.
@@
 
-## Ranking the best pass-rushers {#sec:strainbar}
+## Ranking the Best Pass Rushers {#sec:strainbar}
 
-Since STRAIN is observed continuously for every play in our data, this allows us to aggregate across all frames played and compute the average STRAIN for NFL pass rushers over the course of the eight-week sample size, as discussed in Section \ref{sec:app}.
-Based on the clearly distinct patterns for different positions as previously observed, we evaluate interior pass-rushers (NT's and DT's) separately from edge rushers (OLB's and DE's).
+Since STRAIN is observed continuously for every play in our data, this allows us to aggregate across all frames played and compute the average STRAIN for NFL pass rushers over the course of the eight-week sample size, as discussed in Section \ref{sec:defn}.
+Based on the clearly distinct patterns for different positions as previously observed, we evaluate interior pass rushers (NT's and DT's) separately from edge rushers (OLB's and DE's).
 Tables \ref{tab:edge} and \ref{tab:interior} are leaderboards for the NFL's best edge and interior rushers (with at least 100 plays) rated by the average STRAIN across all frames for the first eight weeks of the 2021 regular season.
@@
 Myles Garrett and TJ Watt are known as top-tier edge rushers, while Aaron Donald is undoubtedly the best interior defender in football.
-Rashan Gary, a young rusher with huge potential, ranks first among all pass-rushers according to our metric.
+Rashan Gary, a young rusher with huge potential, ranks first among all pass rushers according to our metric.
 Moreover, our leaderboards largely match the rankings of experts in the field.
 For instance, there is considerable overlap between our lists and PFF's edge [@Monson2022nfl] and interior [@Linsey2022nfl] rusher rankings released after the 2021 season.
-This ultimately lends credibility to our proposed metric STRAIN as a measure of pass-rushing effectiveness.
+This ultimately lends credibility to our proposed metric STRAIN as a measure of pass rushing effectiveness.
 
@@
 Figure \ref{fig:fig_cor_pressure} is a scatterplot of average STRAIN and pressure rate over the course the provided 8-game sample, which reveals a fairly strong correlation ($r =0.625$) between the quantities.
+Hence, defenders with high STRAIN values also tend to generate more pressure toward the quarterback.
 Furthermore, the average STRAIN for the first 4 weeks of the 2021 season is more predictive of the last 4 weeks' pressure rate ($r=0.322$) than the first 4 weeks' pressure rate ($r=0.097$), as illustrated in Figure \ref{fig:fig_predictability}.
-We also analyze the stability of STRAIN over time by comparing the average STRAIN across all frames played in first and last four weeks of the 2021 NFL regular season.
-It is apparent from Figure \ref{fig:fig_stability} that there is a strong correlation $(r = 0.854)$ for this relationship, illustrating that STRAIN is a stable quantity over the provided 8-week window.
+
+We also analyze the stability of STRAIN over time by comparing the average STRAIN across all frames played for the first and last four weeks of the 2021 NFL regular season.
+It is apparent from Figure \ref{fig:fig_stability} that there is a strong positive correlation $(r = 0.854)$ for this relationship.
+This means that STRAIN is a highly stable football metric over the provided 8-week time window, and pass rushers appear to carry their STRAIN values with them from week to week.
+Overall, STRAIN performs well in both explaining defensive pressure on the field and predicting future performance of pass rushers.
 
@@
 
-## Multilevel Model
+## Multilevel Model Results
 

### 2023-05-14 paper/04-results.Rmd
more results: players on the same team

@@
 This is not necessarily surprising given the limited sample of data.
-Additionally, we are ignoring correlated performance between pass rushers on the same team, e.g., INSERT TWO PLAYERS ON SAME TEAM, but leave exploration of this for future work.
+Additionally, we are ignoring correlated performance between pass rushers on the same team.
+For instance, the top two defensive ends, Myles Garrett and Takkarist McKinley (see the top panel of Figure \ref{fig:fig_rankings_boot}), are teammates on the Cleveland Browns.
+McKinley is a surprising name in our list and we suspect that his high rank is mostly due to playing on the same defensive unit with Garrett, who is well-known as an elite pass rusher.
+We leave the exploration of adjusting for teammates for future work.
 

### 2023-05-15 paper/01-introduction.Rmd
more results: players on the same team
more minor edits

@@
 
-In this manuscript, we specifically focus on measuring the performance of defensive linemen in the NFL.
+Our focus of this manuscript is specifically on measuring the performance of defensive linemen in the NFL.
 There are two main types of defensive linemen in American football: (1) defensive tackles (DTs) located within the interior of the line and (2) defensive ends (DEs) located along the edges (see Figure \ref{fig:formation} (top) for example formation with DTs and DEs).
@@
 These are all simple binary measures of pass rush outcome for any given play.
-However, for plays that do not result in the aforementioned outcomes (in particular, sack), there are still many intermediate defensive actions on the field within the play that are valuable and can be considered successes.
+However, for plays that do not result in the aforementioned outcomes (in particular, sack), there are still many intermediate defensive actions on the field within the play that are valuable and can be considered positive achievements.
 
 Moreover, the sum of sacks, hits, and hurries is often defined as pressures.
-This is better than sacks to some extent, but suffers from problems of subjectivity (e.g., whether there is an actual hurry or not).
+This is better than the individual counts to some extent, but suffers from problems of subjectivity (e.g., whether there is an actual hurry or not).
 Pro football focus (PFF) comes up with a metric called pass-rush productivity (PRP), which is a minor modification from the aforementioned pressures metric (see Table \ref{tab:metrics}).
@@
 However, with the granularity of player tracking data, we have access to data not only for the linemen but also for every player on the field.
-This provides us with a great opportunity to assess the performance of linemen in football.
+This provides us with a great opportunity to study and gain better insights into linemen performance in football.
 

### 2023-05-15 paper/02-data.Rmd
more minor edits

@@
 Along with the tracking information, the 2023 BDB includes scouting data provided by Pro Football Focus (PFF). 
-This contains manually collected player-level information at the play level, such as the player's role (e.g., whether they are a pass rusher and pass blocker) and credited events (e.g., player is credited with hitting the QB on the play).
+This contains manually-collected player-level information, such as the player's role (e.g., whether they are a pass rusher and pass blocker) and credited events (e.g., player is credited with hitting the QB on the play).
 In this manuscript, we use the PFF data to identify 36,362 unique players designated in the "pass rush" role across all plays.

### 2023-05-15 paper/03-methods.Rmd
more minor edits

@@
 
-## Multilevel Model for Play-Level STRAIN
+## Multilevel Model for Play-Level STRAIN {#sec:multilevel}
 
@@
 
-While average STRAIN is a simple first step for evaluating the pass rush, we can also measure pass rusher $j$'s observed average STRAIN on a single play $i$ consisting of $T_i$ total frames,
+<!-- While average STRAIN is a simple first step for evaluating the pass rush,  -->
 
+In addition to the average STRAIN over all frames played, we can also calculate pass rusher $j$'s observed average STRAIN on a single play $i$ consisting of $T_i$ total frames,
 \begin{equation}
@@
 \end{equation}
-
-However, the observed average STRAIN on a single play is likely due to a convolution of factors.
+While this aggregated measure is a simple first step for pass rush evaluation, the observed average STRAIN on a single play is likely due to a convolution of factors.
 Outside of the pass rusher's ability, there is variability in the opposing strength of pass blockers across plays a pass rusher is involved.
-Additionally, based on previous work by PFF, the opposing QB plays an important role on observed pressure in passing plays [@eager2019pressure].
+Additionally, based on previous work by PFF, the opposing QB plays an important role on the observed pressure in passing plays [@eager2019pressure].
 Thus, we need to appropriately divide the credit of an observed average STRAIN across the different players involved and other factors.
@@
 We identify the pass blocker linked with the pass rusher of interest using the data provided by PFF.
-Since there can be multiple blockers linked to a rusher, for simplicity, we consider the nearest blocker positioned to the pass rusher at the start of the play.
+Since there can be multiple blockers matching up with a rusher, for simplicity, we consider the nearest blocker positioned to the pass rusher at the start of the play.
 We use random intercepts for the three player groups: (1) pass rushers as $R$, (2) nearest pass blockers as $B$, and (3) QBs as $Q$.
-The coefficients $\boldsymbol{\beta}$ for additional information vector of information  $\mathbf{x_{ij}}$ about pass rusher $j$ in play $i$ are estimated as fixed-effects. 
+The coefficients $\boldsymbol \beta$ for covariate vector $\mathbf{x_{ij}}$ for pass rusher $j$ in play $i$ are estimated as fixed-effects. 
 Our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows.
@@
 We implement the model using penalized likelihood via the `lme4` package in `R` [@bateslme4; @R2023Language].
-Additionally, we generate estimates of uncertainty for the player-level random intercepts based on 95% confidence intervals generated by sampling from the posterior distribution via parametric bootstrapping with the `merTools` package [@knowlesmerTools, @gelman2006data].
+Additionally, we generate estimates of uncertainty for the player-level random intercepts based on 95% confidence intervals generated by sampling from the posterior distribution via parametric bootstrapping with the `merTools` package [@knowlesmerTools; @gelman2006data].
 

### 2023-05-15 paper/04-results.Rmd
more minor edits

@@
 In contrast, Figure \ref{fig:fig_crosby_curves_unsucess} shows the feature curves for an unsuccessful pass rush attempt by the Raiders defense from the same game.
-In this play, Broncos QB Teddy Bridgewater is well-protected by the offensive line and is able to release a long pass[^unsuccess].
+In this play, Broncos QB Teddy Bridgewater is well-protected by the offensive line and is able to release a long pass[^unsuccess] to a receiver.
 We see that the STRAIN generated by Crosby is relatively small over the course of this play, compared to the previous play shown in Figure \ref{fig:fig_crosby_curves} which results in a sack.
@@
 
-There are a variety of quantities we are interested in from the multilevel model fit mentioned in Section \ref{sec:methods}.
-First, Table \ref{tab:fixedeff} displays the results for the fixed effects terms.
-It appears that that the average STRAIN decreases by 0.042 (95% CI [-0.044, -0.040]) for every extra second in play length, and by 0.064 (95% CI [-0.069, -0.059]) every additional blocker in the offensive unit, after accounting for other covariates in the model.
-In other words, NFL pass rushers tend to generate more pressure in plays with shorter length and when facing fewer number of blockers, which totally makes sense.
+The results of fitting the multilevel model described in Section \ref{sec:multilevel} is displayed in Table \ref{tab:fixedeff}.
+First, we investigate the fixed-effects terms of this model.
+It appears that that the average STRAIN decreases by 0.042 (95% CI [-0.044, -0.040]) for every extra second in play length, and by 0.064 (95% CI [-0.069, -0.059]) every additional blocker in the offensive unit, after accounting for other covariates.
+In other words, NFL pass rushers tend to generate more pressure in plays with shorter length and when facing fewer number of blockers, which makes intuitive sense.
 
-We observe statistically significant differences among the pass rush and pass block positions in most cases, while controlling for other variables.
-The coefficient estimates for the pass rush position terms reveal that players with positions closer to the line of scrimmage (DT and NT) appear to generate less STRAIN (relative to DE) than those that are further back, which is consistent with our results in Section \ref{sec:positional}.
+Moreover, we observe statistically significant differences among the pass rush and pass block positions in most cases, while controlling for other variables.
+The coefficient estimates for the pass rush position terms reveal that players with positions closer to the line of scrimmage (DT and NT) appear to generate less STRAIN (relative to DE) than those lining up further back, which is consistent with our results in Section \ref{sec:positional}.
 For blockers, centers are those that absorb the most pressure on average, more than guards and other offensive positions (compared to the baseline level, tackles).

### 2023-05-15 paper/05-discussion.Rmd
Begin revising discussion

--- 2023-05-11 paper/05-discussion.Rmd
@@
 # Discussion {#sec:discussion}
+
+In this work, we have proposed STRAIN - a simple and interpretable statistic for evaluating pass-rushers, with higher STRAIN values corresponding to greater pass-rushing ability.
+We demonstrated that a pass rusher's average STRAIN is both stable and more predictive of future pressure events than using the player's previously observed pressure events.
+This is analogous to the use of exit velocity in baseball, a predictive measurement that avoids the noisy nature of observed outcomes, emphasizing how the opportunity of working with player-tracking data can lead to the development of novel metrics in American football.
+Additionally, through our multilevel modeling efforts, we observe that the pass rusher explains more variation in play-level STRAIN than the quarterback or assigned blocker.
+
+As for additional future directions, while we focus solely on pass-rushers in this paper, STRAIN can be easily extended to the assessment of pass-blockers as a unit.
+This can be accomplished by looking at quantities such as the total STRAIN or the maximum STRAIN per frame aggregated across the entire play.
+It is also possible to apply STRAIN to assess individual offensive pass-blockers, provided that a method of matching blockers to rushers was developed.
+Furthermore, compared to existing metrics, STRAIN measures pass-rush effectiveness for every play continuously over time, which is at a much more granular level than considering whether the play resulted in a binary outcome such as a sack.
+
+or converting continuous data to a binary win/loss.
 

### 2023-05-15 paper/01-introduction.Rmd
Small edits.

@@
 
-In recent years, tracking data have replaced traditional box-score statistics and play-by-play data as the state-of-the-art tool in sports analytics.
-Numerous sports have been collecting and releasing data on player and ball locations on the playing surface over the course of a game.
-This spatio-temporal multi-resolution source of data have provided exceptional opportunities for researchers to perform advanced studies at a more complex level to deepen the understanding of different sports.
+In recent years, tracking data have replaced traditional box-score statistics and play-by-play data as the state-of-the-art in sports analytics.
+Numerous sports are collecting and releasing data on player and ball locations on the playing surface over the course of a game.
+This spatio-temporal multi-resolution source of data has provided exceptional opportunities for researchers to perform advanced studies at a more granular level to deepen our understanding of different sports.
 For complete surveys on how tracking data have transformed sports analytics, see @Macdonald2020Recreating, @Baumer2023Big, and @Kovalchik2023Player.
@@
 
-In an attempt to foster analytics and innovate the game, the National Football League (NFL) introduced their player tracking system known as Next Gen Stats (NGS) in 2016 [@nfl2023ngs].
+In an attempt to further develop analytics and innovate the game, the National Football League (NFL) introduced their player tracking system known as Next Gen Stats (NGS) in 2016 [@nfl2023ngs].
 NGS uses radio frequency identification (RFID) chips placed in players' shoulder pads (and in the ball) to collect data at a rate of 10 frames per second.
@@
 The winners of the inaugural BDB focused on identifying receiver routes via clustering techniques [@chu2020route] and convolutional neural networks [@sterken2019routenet].
-Along with the BDB competition entries, the public release of NGS data allowed researchers to tackle <!-- (pun intended) --> a variety of other problems such as revisiting fourth down decision making [@lopez2020bigger], annotating pass coverage with Gaussian mixture models [@dutta2020unsupervised], and introducing a continuous-time framework to estimate within-play value [@yurko2020going].
+Along with the BDB competition entries, the public release of NGS data allowed researchers to tackle <!-- (pun intended.
+#eyeroll) --> a variety of other problems such as revisiting fourth down decision making [@lopez2020bigger], annotating pass coverage with Gaussian mixture models [@dutta2020unsupervised], and introducing a continuous-time framework to estimate within-play value [@yurko2020going].
 Since its inception, the BDB has chosen a different theme each year leading to new insight about evaluating different positions such as running backs, defensive backs, and special teams.
@@
 Other traditional box score statistics such as hits and hurries are collected by various outlets. 
-A hit is a collision between a defender and the opposing team's quarterback while the QB is in possession of the ball.
+A hit is a collision between a defender and the opposing team's quarterback while the QB is in possession of the ball.  <!-- This is a sack, right? -->
 A hurry represents an instance when a defender successfully disrupts without necessarily making direct contact with the QB and forces the QB to make a throw earlier than expected.
@@
 
-Moreover, the sum of sacks, hits, and hurries is often defined as pressures.
+In addition to these individual binary metrics, the sum of sacks, hits, and hurries is often defined as pressures.
 This is better than the individual counts to some extent, but suffers from problems of subjectivity (e.g., whether there is an actual hurry or not).
-Pro football focus (PFF) comes up with a metric called pass-rush productivity (PRP), which is a minor modification from the aforementioned pressures metric (see Table \ref{tab:metrics}).
-In particular, PRP gives twice as much weight to a sack relative to hurries and hits, which is a small upgrade to pressures.
-However, the choice of weights is ad-hoc and still only indicates binary outcomes, similar to the shortcomings of previous metrics.
+Pro football focus (PFF) defines a metric called pass-rush productivity (PRP), which is a minor modification from the aforementioned pressures metric (see Table \ref{tab:metrics}).
+In particular, PRP gives twice as much weight to a sack relative to hurries and hits, which is a small upgrade to pressures, however, the choice of weights is ad-hoc and still only copnsiders binary outcomes, similar to the shortcomings of previous metrics.
 

### 2023-05-15 paper/01-introduction.Rmd
Small edits.

@@
 It demonstrates whether a pass rusher is able to beat their blocking matchup before a fixed time from the snap (2.5 seconds as chosen by ESPN).
-This depends on the rather arbitrary time threshold used to define a pass rush win.
+However, this depends on the rather arbitrary time threshold used to define a pass rush win.
 In addition, once a cutoff is chosen, PRWR converts continuous data to a win-loss indicator, becoming dichotomous like most of the metrics discussed above.
@@
 @AlamarGoldner2011 later follow up by using manually-tracked data for the 2010 season to estimate lineman performance for different team-positions instead of individual defenders (e.g., Chicago Bears' left tackle, Pittsburgh Steelers' center, etc.).
-The paper uses survival analysis to model time in pocket and completion percentage for quarterbacks, before proposing a measure for linemen's contribution to their team's passing in terms of yards gained.
+The paper uses survival analysis to model time in pocket and completion percentage for quarterbacks before proposing a measure for linemen's contribution to their team's passing in terms of yards gained.
 
@@
 In this paper, we focus on the evaluation of defensive linemen in football.
-We devise STRAIN, a metric for measuring pass rush effectiveness, inspired by the concept of strain rate in materials science.
+We propose STRAIN, a metric for measuring pass rush effectiveness, inspired by the concept of strain rate in materials science.
 Our metric gives a continuous measure of pressure for every pass rusher on the football field over the course of an entire play.

### 2023-05-15 paper/02-data.Rmd
Small edits.

@@
 The data corresponds to 8,557 passing plays across 122 games in the first eight weeks of the 2021 NFL regular season. 
-For each play, we have information on the on-field location, speed, angle, direction, and orientation of each player on the field and the football at a rate of 10 Hz, along with event annotations for each frame such as ball snap, pass forward, and quarterback sack, to name a few.
+For each play, we have information on the on-field location, speed, angle, direction, and orientation of each player on the field and the football at a rate of 10 Hz (i.e.
+ten measurements per second), along with event annotations for each frame such as ball snap, pass forward, and quarterback sack, to name a few.
 
@@
 
-[^crosbysack]: A real-game footage of this play is available at https://www.raiders.com/video/de-maxx-crosby-sacks-qb-teddy-bridgewater-for-a-loss-of-6-yards-nfl
+[^crosbysack]: Game footage of this play is available at https://www.raiders.com/video/de-maxx-crosby-sacks-qb-teddy-bridgewater-for-a-loss-of-6-yards-nfl
 
@@
 This contains manually-collected player-level information, such as the player's role (e.g., whether they are a pass rusher and pass blocker) and credited events (e.g., player is credited with hitting the QB on the play).
-In this manuscript, we use the PFF data to identify 36,362 unique players designated in the "pass rush" role across all plays.
+In this manuscript, we use the PFF data to identify 36,362 unique pass rushing attempts by players designated in the "pass rush" role across all plays.
+<!-- Need to reword this.
+There aren't 36,362 unique players.
+It's like players by rushes or something like that.
+-->
 For context, Figure \ref{fig:pass_rushers_count} displays the distribution of the number of observed pass rushers involved in a play, ranging from 1 to 8 with 4 pass rushers (i.e.
@@
 #| label: pass_rushers_count
-#| fig.cap: "Distribution of the number of pass rushers in a play."
+#| fig.cap: "Distribution of the number of pass rushers on passing plays."
 #| fig.width: 4.5

### 2023-05-15 paper/03-methods.Rmd
Small edits.

@@
 
-In materials science, strain is the deformation of a material from stress, showing the rate of change in a material's length relative to its original length.
+In materials science, strain is the deformation of a material from stress, showing the rate of change in a material's length relative to its original length [@strain].
 Formally, let $L(t)$ be the distance between any given two points of interest within a material at time $t$, and $L_0$ be the initial distance between those two points.
@@
 \end{equation}
-*   The velocity at which pass ruher $j$ is moving towards the quarterback at frame $t$ during play $i$ is 
+*   The velocity at which pass rusher $j$ is moving towards the quarterback at frame $t$ during play $i$ is 
 \begin{equation}
@@
 Recall that based on its materials science property, an increase in strain rate is associated with an increase in the distance between two points.
-In our football setting, the two points of interest are the pass rusher and the quarterback, and we expect our metric to increase as the distance between the pass rusher and the quarterback decreases.
+In the American football setting, the two points of interest are the pass rusher and the quarterback, and we expect our metric to increase as the distance between the pass rusher and the quarterback decreases.
 Thus, the negative sign in the numerator of our formula effectively accounts for this. 
 Additionally, rather than keeping the initial distance ($L_0$ as previously denoted) between two points constant over time, we update the initial position to be the player locations at the beginning of each frame.
-This effectively gives us the STRAIN for each frame throughout a play.
+This gives us the STRAIN for each frame throughout a play.
 
@@
 \end{equation}
-Notice that this quantity increases in two ways: 1) the rate at which the rusher is moving towards the quarterback increases, and 2) the distance between the rusher and the quarterback gets smaller.
+Notice that this quantity increases in two ways: 1) the rate at which the rusher is moving towards the quarterback increases, and 2) the distance between the rusher and the quarterback decreases.
 Both of these are indications of an effective pass rush attempt.
 Finally, our statistic STRAIN is measured in inverse second, similar to strain rate.
-In addition, the reciprocal of our metric (1/STRAIN) has a straightforward interpretation:
-it illustrates the amount of time required for the rusher to get to the quarterback at the current location and rate at any given time $t$.
+Note that the reciprocal of our metric (1/STRAIN) has an interesting and straightforward interpretation: the amount of time required for the rusher to get to the quarterback at the current location and rate at any given time $t$.
 
-Moreover, since we observe STRAIN continuously at every tenth of a second within each play, we can then compute the average STRAIN across all frames played for every pass rusher.
+Moreover, since we observe STRAIN at every tenth of a second within each play, we can then compute the average STRAIN across all frames played for every pass rusher.
 Formally, the average STRAIN, denoted by $\overline{\text{STRAIN}}$, for pass rusher $j$ involved in $n_j$ total plays across $T_j$ total frames is
@@
 We implement the model using penalized likelihood via the `lme4` package in `R` [@bateslme4; @R2023Language].
-Additionally, we generate estimates of uncertainty for the player-level random intercepts based on 95% confidence intervals generated by sampling from the posterior distribution via parametric bootstrapping with the `merTools` package [@knowlesmerTools; @gelman2006data].
+Additionally, we generate estimates of uncertainty for the player-level random intercepts based on 95% confidence intervals generated by sampling from the posterior distribution <!-- Posterior distribution of what?
+--> via parametric bootstrapping with the `merTools` package [@knowlesmerTools; @gelman2006data].
 

### 2023-05-15 paper/03-methods.Rmd
Updated Strain citation.

@@
 
-In materials science, strain is the deformation of a material from stress, showing the rate of change in a material's length relative to its original length [@strain].
+In materials science, strain is the deformation of a material from stress, showing the rate of change in a material's length relative to its original length [@callister2009materials].
 Formally, let $L(t)$ be the distance between any given two points of interest within a material at time $t$, and $L_0$ be the initial distance between those two points.

### 2023-05-15 paper/01-introduction.Rmd
Updated Strain citation.
Updated hit vs sack.

@@
 Other traditional box score statistics such as hits and hurries are collected by various outlets. 
-A hit is a collision between a defender and the opposing team's quarterback while the QB is in possession of the ball.  <!-- This is a sack, right? -->
+A hit is a collision between a defender and the opposing team's quarterback after the quarterback makes a throw.  
 A hurry represents an instance when a defender successfully disrupts without necessarily making direct contact with the QB and forces the QB to make a throw earlier than expected.
@@
 In this paper, we focus on the evaluation of defensive linemen in football.
-We propose STRAIN, a metric for measuring pass rush effectiveness, inspired by the concept of strain rate in materials science.
+We propose STRAIN, a metric for measuring pass rush effectiveness, inspired by the concept of strain rate in materials science [@callister2009materials].
 Our metric gives a continuous measure of pressure for every pass rusher on the football field over the course of an entire play.

### 2023-05-15 paper/04-results.Rmd
more edits.

@@
 
-[^unsuccess]: A real-game footage of this play is available at https://www.nfl.com/videos/teddy-bridgewater-courtland-sutton-barely-miss-out-on-chance-at-62-yard-td-bomb
+[^unsuccess]: Game footage of this play is available at https://www.nfl.com/videos/teddy-bridgewater-courtland-sutton-barely-miss-out-on-chance-at-62-yard-td-bomb
 
@@
 
-Our results show many true positives with no surprises.
-Myles Garrett and TJ Watt are known as top-tier edge rushers, while Aaron Donald is undoubtedly the best interior defender in football.
-Rashan Gary, a young rusher with huge potential, ranks first among all pass rushers according to our metric.
-Moreover, our leaderboards largely match the rankings of experts in the field.
+<!-- Our results show many true positives with no surprises. -->
+Our results are mostly consistent with conventional rankings of rushers.
+For instance, Myles Garrett and TJ Watt are known as top-tier edge rushers and both show up in our top ten edge rushers.
+Additionally, Aaron Donald is nearly universally considered the best interior defender in football, and he appears at the top of our rankings of interior rushers.
+<!-- Rashan Gary, a young rusher with huge potential, ranks first among all pass rushers according to our metric over the period of time observed. -->
+Moreover, our leader boards largely match the rankings of experts in the field.
 For instance, there is considerable overlap between our lists and PFF's edge [@Monson2022nfl] and interior [@Linsey2022nfl] rusher rankings released after the 2021 season.
@@
 How much does our metric vary from week to week?
-In other words, is previous performance predictive of future performance based on our metric?
+In other words, is previous performance as measured by our metric predictive of future performance based on our metric?
 Below, we attempt the answer these questions, and the following results are for pass rushers with at least 100 snaps played during the first 8 weeks of the 2021 NFL regular season.
@@
 Watt (OLB) as the top pass rusher for their respective position.
-We observe different levels of variability across these top players, with the only a subset of players with intervals strictly above zero.
+We observe different levels of variability across these top players, with only a subset of players with intervals strictly above zero.
 This is not necessarily surprising given the limited sample of data.
@@
 For instance, the top two defensive ends, Myles Garrett and Takkarist McKinley (see the top panel of Figure \ref{fig:fig_rankings_boot}), are teammates on the Cleveland Browns.
-McKinley is a surprising name in our list and we suspect that his high rank is mostly due to playing on the same defensive unit with Garrett, who is well-known as an elite pass rusher.
+McKinley is a surprising name on our list and we suspect that his high rank is mostly due to playing on the same defensive unit with Garrett, who is well-known as an elite pass rusher.
 We leave the exploration of adjusting for teammates for future work.

### 2023-05-16 paper/05-discussion.Rmd
make additional discussion updates with more future work discussion

--- 2023-05-11 paper/05-discussion.Rmd
@@
 
-Summary
+In this work, we have proposed STRAIN - a simple and interpretable statistic for evaluating pass-rushers, with higher STRAIN values corresponding to greater pass-rushing ability.
+We demonstrated that a pass rusher's average STRAIN is both stable and more predictive of future pressure events than using the player's previously observed pressure events.
+This is analogous to the use of exit velocity in baseball, a predictive measurement that avoids the noisy nature of observed outcomes, emphasizing how the opportunity of working with player-tracking data can lead to the development of insightful metrics in American football.
 
-Critique
+Through our multilevel modeling efforts, we observe that the pass rusher explains more variation in play-level STRAIN, followed by the quarterback and then their assigned blocker.
+This is an intuitive result, consistent with previous literature on pass rushing and pressure.
+However, we ignore the fact that pass rushers can be teammates working together in the same play leading to some type of correlation structure in their STRAIN values.
+One way to address this complication, could be with a adjusted plus-minus (APM) regression approach [@matano2023augmenting], with all players accounted for in modeling a total STRAIN quantity.
 
-Future work
+We believe there are other concepts in materials science that could be applied to evaluating pass rushing and pass blocking in American football.
+One potential idea is a quantity called stress, which measures force over an area.
+Recall from intro physics that force is mass times acceleration, meaning this quantity would take into account the physical size of a pass rusher in the computation of force.
+You could then somehow divide this force over the "area" of the pocket formed by pass blockers to compute stress.
 
-rapm
+As for additional future directions, athough we focus solely on pass-rushers in this paper, STRAIN can be easily extended to the assessment of pass-blockers as a unit.
+This can be accomplished through the aforementioned APM regression or by simply looking at quantities such as the total STRAIN or the maximum STRAIN per frame aggregated across the entire play.
+It is also possible to apply STRAIN to assess individual offensive pass-blockers, provided that a method of matching blockers to rushers was developed.
+
+Furthermore, compared to existing metrics, STRAIN measures pass-rush effectiveness for every play continuously over time, which is at a much more granular level than considering whether the play resulted in a binary outcome such as a sack.
+There is ample opportunity for working with the complete STRAIN trajectories via functional data analysis techniques to help understand how STRAIN affects offensive production.
 

### 2023-05-16 paper/01-introduction.Rmd
some fixes

@@
 
-In recent years, tracking data have replaced traditional box-score statistics and play-by-play data as the state-of-the-art in sports analytics.
+In recent years, tracking data have replaced traditional box-score statistics and play-by-play data as the state of the art in sports analytics.
 Numerous sports are collecting and releasing data on player and ball locations on the playing surface over the course of a game.
@@
 
-In an attempt to further develop analytics and innovate the game, the National Football League (NFL) introduced their player tracking system known as Next Gen Stats (NGS) in 2016 [@nfl2023ngs].
+In an attempt to foster analytics and innovate the game, the National Football League (NFL) introduced their player tracking system known as Next Gen Stats (NGS) in 2016 [@nfl2023ngs].
 NGS uses radio frequency identification (RFID) chips placed in players' shoulder pads (and in the ball) to collect data at a rate of 10 frames per second.
@@
 \hline
-Sacks & A defender tackles the QB behind the line of scrimmage before the QB is able to throw a pass \\
-Hits & A defender tackles the QB behind the line of scrimmage after the QB is able to throw a pass. \\
-Hurries & A defender pressures the QB behind the line of scrimmage forcing the QB to throw the ball prior sooner than intended. \\
+Sacks & A defender tackles the QB behind the line of scrimmage before a QB throw \\
+Hits & A defender tackles the QB behind the line of scrimmage after a QB throw \\
+Hurries & A defender pressures the QB behind the line of scrimmage forcing the QB to throw the ball sooner than intended \\
 Pressures & $\text{Hurries} + \text{ Hits } + \text{ Sacks}$ \\
@@
 Time In Pocket & Time (in seconds) between ball snap and throw or pocket collapse for a QB \\
-NGS Get Off & Average time (in seconds) required for a defender to cross the line of scrimmage after the ball snap.\\
+NGS Get Off & Average time (in seconds) required for a defender to cross the line of scrimmage after the ball snap\\
 Pass Rush Win Rate & Rate at which pass rusher beats pass block within 2.5 seconds after ball snap\\
@@
 
-In addition to these individual binary metrics, the sum of sacks, hits, and hurries is often defined as pressures.
+In addition, the sum of sacks, hits, and hurries is often defined as pressures.
 This is better than the individual counts to some extent, but suffers from problems of subjectivity (e.g., whether there is an actual hurry or not).
 Pro football focus (PFF) defines a metric called pass-rush productivity (PRP), which is a minor modification from the aforementioned pressures metric (see Table \ref{tab:metrics}).
-In particular, PRP gives twice as much weight to a sack relative to hurries and hits, which is a small upgrade to pressures, however, the choice of weights is ad-hoc and still only copnsiders binary outcomes, similar to the shortcomings of previous metrics.
+In particular, PRP gives twice as much weight to a sack relative to hurries and hits, which is a small upgrade to pressures.
+However, the choice of weights is ad-hoc and still only considers binary outcomes, similar to the shortcomings of previous metrics.
 
@@
 However, this depends on the rather arbitrary time threshold used to define a pass rush win.
-In addition, once a cutoff is chosen, PRWR converts continuous data to a win-loss indicator, becoming dichotomous like most of the metrics discussed above.
+Besides, once a cutoff is chosen, PRWR converts continuous data to a win-loss indicator, becoming dichotomous like most of the metrics discussed above.
 
@@
 In this paper, we focus on the evaluation of defensive linemen in football.
-We propose STRAIN, a metric for measuring pass rush effectiveness, inspired by the concept of strain rate in materials science [@callister2009materials].
+We propose STRAIN, a metric for measuring pass rush effectiveness, inspired by the concept of strain rate in materials science.
 Our metric gives a continuous measure of pressure for every pass rusher on the football field over the course of an entire play.
@@
 The remainder of this manuscript is outlined as follows.
-We first describe the player tracking data provided by 2023 Big Data Bowl in Section \ref{sec:data}.
+We first describe the player tracking data provided by the Big Data Bowl 2023 in Section \ref{sec:data}.
 We then introduce the mathematical motivation and definition of our statistic STRAIN and our modeling approach in Section \ref{sec:methods}.

### 2023-05-16 paper/02-data.Rmd
some fixes

@@
 
-In the forthcoming analysis, we rely on the data from the 2023 NFL Big Data Bowl provided by the NFL Next Gen Stats tracking system.
+In the forthcoming analysis, we rely on the data from the NFL Big Data Bowl 2023 provided by the NFL Next Gen Stats tracking system.
 The data corresponds to 8,557 passing plays across 122 games in the first eight weeks of the 2021 NFL regular season. 
 For each play, we have information on the on-field location, speed, angle, direction, and orientation of each player on the field and the football at a rate of 10 Hz (i.e.
-ten measurements per second), along with event annotations for each frame such as ball snap, pass forward, and quarterback sack, to name a few.
+10 measurements per second), along with event annotations for each frame such as ball snap, pass forward, and quarterback sack, to name a few.
 
@@
 This contains manually-collected player-level information, such as the player's role (e.g., whether they are a pass rusher and pass blocker) and credited events (e.g., player is credited with hitting the QB on the play).
-In this manuscript, we use the PFF data to identify 36,362 unique pass rushing attempts by players designated in the "pass rush" role across all plays.
-<!-- Need to reword this.
-There aren't 36,362 unique players.
-It's like players by rushes or something like that.
--->
+In this manuscript, we use the PFF data to identify 36,362 unique pass rush attempts by players designated in the "pass rush" role across all plays.
 For context, Figure \ref{fig:pass_rushers_count} displays the distribution of the number of observed pass rushers involved in a play, ranging from 1 to 8 with 4 pass rushers (i.e.

### 2023-05-16 paper/03-methods.Rmd
some fixes

--- 2023-05-15 paper/03-methods.Rmd
@@
 
-In materials science, strain is the deformation of a material from stress, showing the rate of change in a material's length relative to its original length [@strain].
+In materials science, strain [@callister2018materials] is the deformation of a material from stress, showing the rate of change in a material's length relative to its original length.
 Formally, let $L(t)$ be the distance between any given two points of interest within a material at time $t$, and $L_0$ be the initial distance between those two points.
-The **strain** for a material at time $t$ is defined as 
+The strain for a material at time $t$ is defined as 
 \begin{equation}
@@
 
-Accordingly, the **strain rate** of a material measures the change in its deformation with respect to time.
+Accordingly, the strain rate [@callister2018materials] of a material measures the change in its deformation with respect to time.
 Mathematically, the strain rate of a material can be expressed as the derivative of its strain.
@@
 Additionally, we generate estimates of uncertainty for the player-level random intercepts based on 95% confidence intervals generated by sampling from the posterior distribution <!-- Posterior distribution of what?
---> via parametric bootstrapping with the `merTools` package [@knowlesmerTools; @gelman2006data].
+--> via parametric bootstrapping of 10,000 samples with the `merTools` package [@knowlesmerTools; @gelman2006data].
 

### 2023-05-16 paper/04-results.Rmd
some fixes

@@
 Tables \ref{tab:edge} and \ref{tab:interior} are leaderboards for the NFL's best edge and interior rushers (with at least 100 plays) rated by the average STRAIN across all frames for the first eight weeks of the 2021 regular season.
-The tables also consist of the total number of hits, hurries, and sacks for each defender.
+The tables also consist of the total number of hits, hurries, and sacks (determined from PFF scouting data) for each defender.
 
 <!-- Our results show many true positives with no surprises. -->
+<!-- Rashan Gary, a young rusher with huge potential, ranks first among all pass rushers according to our metric over the period of time observed. -->
+
 Our results are mostly consistent with conventional rankings of rushers.
-For instance, Myles Garrett and TJ Watt are known as top-tier edge rushers and both show up in our top ten edge rushers.
-Additionally, Aaron Donald is nearly universally considered the best interior defender in football, and he appears at the top of our rankings of interior rushers.
-<!-- Rashan Gary, a young rusher with huge potential, ranks first among all pass rushers according to our metric over the period of time observed. -->
-Moreover, our leader boards largely match the rankings of experts in the field.
+Notably, Myles Garrett and TJ Watt are widely recognized as top-tier edge rushers and both show up in our top edge rusher list; whereas Aaron Donald, who is undoubtedly the best interior defender in football, appears at the top of our interior rusher rankings.
+Moreover, our leaderboards largely match the rankings of experts in the field.
 For instance, there is considerable overlap between our lists and PFF's edge [@Monson2022nfl] and interior [@Linsey2022nfl] rusher rankings released after the 2021 season.
@@
 How much does our metric vary from week to week?
-In other words, is previous performance as measured by our metric predictive of future performance based on our metric?
+In other words, is previous performance predictive of future performance based on our metric?
 Below, we attempt the answer these questions, and the following results are for pass rushers with at least 100 snaps played during the first 8 weeks of the 2021 NFL regular season.

### 2023-05-16 paper/00-main.Rmd
more minor edits
more revision

@@
 title: |
-  | \Large Bearing the STRAIN: Analyzing Pass Rush in American Football with Player Tracking Data
+  | Get Strained: Analyzing Pass Rush in American Football with Player Tracking Data
 
@@
 # Straining the trenches:
+# Bearing the STRAIN
 
@@
   fig.align = "center",
-  # fig.pos = "p",
+  # fig.pos = "ht",
   cache = TRUE
@@
 
-We thank the organizers of the 2023 NFL Big Data Bowl for hosting the competition and providing access to the data.
+We thank the organizers of the NFL Big Data Bowl 2023 for hosting the competition and providing access to the data.
 

### 2023-05-16 paper/01-introduction.Rmd
more revision

@@
 NGS uses radio frequency identification (RFID) chips placed in players' shoulder pads (and in the ball) to collect data at a rate of 10 frames per second.
-The data captures real-time on-field information such as locations, speeds, and accelerations of all 22 players (as well as the ball).
+The data captures real-time on-field information such as locations, speeds, and accelerations of all 22 players (and also the football).
 While these data were initially only available for teams, media, and vendors, in December 2018 the NFL launched the inaugural edition of their annual Big Data Bowl (BDB) competition [@nfl2023big].
@@
 We propose STRAIN, a metric for measuring pass rush effectiveness, inspired by the concept of strain rate in materials science.
-Our metric gives a continuous measure of pressure for every pass rusher on the football field over the course of an entire play.
+Our statistic gives a continuous measure of pressure for every pass rusher on the football field over the course of an entire play.
 This allows for the assessment of pass rush success even on plays that do not result in an observed outcome like a sack, hit, or hurry.
 We view this as a major step forward for accurately evaluating defensive linemen performance.
-Further, we note that although our focus in this paper is on defensive linemen, our proposed approach can be extended to the evaluation of offensive linemen in American football.
+We also demonstrate that STRAIN is a stable quantity over time and predictive of defensive pressure.
+Additionally, we consider a multilevel model to estimate every pass rusher's contribution to the average STRAIN in a play while controlling for other positions like QB and pass blocker as well as various play-level information.
+We note that although our focus in this paper is on pass rushers, our proposed approach can be extended to the evaluation of pass blockers in American football.
 
@@
 We first describe the player tracking data provided by the Big Data Bowl 2023 in Section \ref{sec:data}.
-We then introduce the mathematical motivation and definition of our statistic STRAIN and our modeling approach in Section \ref{sec:methods}.
+We then introduce the mathematical motivation and definition of our measure STRAIN and our modeling approach in Section \ref{sec:methods}.
 Next, we present applications of STRAIN and study different statistical properties of the metric in Section \ref{sec:results}.

### 2023-05-16 paper/03-methods.Rmd
more revision

@@
 Formally, the average STRAIN, denoted by $\overline{\text{STRAIN}}$, for pass rusher $j$ involved in $n_j$ total plays across $T_j$ total frames is
-
 \begin{equation}
@@
 We implement the model using penalized likelihood via the `lme4` package in `R` [@bateslme4; @R2023Language].
-Additionally, we generate estimates of uncertainty for the player-level random intercepts based on 95% confidence intervals generated by sampling from the posterior distribution <!-- Posterior distribution of what?
---> via parametric bootstrapping of 10,000 samples with the `merTools` package [@knowlesmerTools; @gelman2006data].
+Additionally, we obtain interval estimates for the player-level random intercepts via parametric bootstrapping of 10,000 samples with the `merTools` package [@knowlesmerTools; @gelman2006data].
 

### 2023-05-16 paper/04-results.Rmd
more revision

@@
 
-[CHECK/EDIT THIS, MAKE SURE WORDING IS CORRECT/NOT CONFUSING]
+<!-- [CHECK/EDIT THIS, MAKE SURE WORDING IS CORRECT/NOT CONFUSING] -->
 
@@
 It appears that that the average STRAIN decreases by 0.042 (95% CI [-0.044, -0.040]) for every extra second in play length, and by 0.064 (95% CI [-0.069, -0.059]) every additional blocker in the offensive unit, after accounting for other covariates.
-In other words, NFL pass rushers tend to generate more pressure in plays with shorter length and when facing fewer number of blockers, which makes intuitive sense.
+In other words, NFL pass rushers tend to generate more pressure in plays with shorter length (e.g., blitzes) and when facing fewer number of blockers, which makes intuitive sense.
 
@@
 In addition to adjusting for opposition, our multilevel model allows us to provide some notion of uncertainty about the pass rusher's effect on the play level STRAIN.
-Figure \ref{fig:fig_rankings_boot} displays the mean varying intercept for the top ten pass rushers for each position along with 95% confidence intervals. 
+Figure \ref{fig:fig_rankings_boot} displays the mean varying intercepts for the top ten pass rushers for each position along with the corresponding 95% confidence intervals. 
 We observe similar rankings based on the random intercepts in comparison to the average STRAIN results in Tables \ref{tab:edge} and \ref{tab:interior}, with Myles Garrett (DE), Aaron Donald (DT), Vita Vea (NT), and T.J.
-Watt (OLB) as the top pass rusher for their respective position.
+Watt (OLB) as the first-ranked pass rusher for their respective position.
 We observe different levels of variability across these top players, with only a subset of players with intervals strictly above zero.
 This is not necessarily surprising given the limited sample of data.
-Additionally, we are ignoring correlated performance between pass rushers on the same team.
+Additionally, we are ignoring correlated performance between defensive linemen on the same team.
 For instance, the top two defensive ends, Myles Garrett and Takkarist McKinley (see the top panel of Figure \ref{fig:fig_rankings_boot}), are teammates on the Cleveland Browns.

### 2023-05-16 paper/05-discussion.Rmd
more revision

--- 2023-05-13 paper/05-discussion.Rmd
@@
 
-We have proposed STRAIN, a simple and interpretable statistic for evaluating pass-rushers, with higher STRAIN corresponds to greater pass-rushing ability. 
-Compared to existing metrics, STRAIN measures pass-rush effectiveness for every play continuously over time, which is at a much more granular level than considering whether the play resulted in a sack or converting continuous data to a binary win/loss.
+In this work, we have proposed STRAIN - a simple and interpretable statistic for evaluating pass rushers, with higher STRAIN values corresponding to greater pass rushing ability.
+We demonstrated that a pass rusher's average STRAIN is both stable and more predictive of future pressure events than using the player's previously observed pressure events.
+This is analogous to the use of exit velocity in baseball, a predictive measurement that avoids the noisy nature of observed outcomes, emphasizing how the opportunity of working with player-tracking data can lead to the development of insightful metrics in American football.
 
-As for future directions, while we focus solely on pass-rushers in this paper, STRAIN can be easily extended to the assessment of pass-blockers as a unit. 
-This can be accomplished by looking at quantities such as the total STRAIN or the maximum STRAIN per frame aggregated across the entire play.
-It is also possible to apply STRAIN to assess individual offensive pass-blockers, provided that a method of matching blockers to rushers was developed.
+Through multilevel modeling, we observe that the pass rusher explains more variation in play-level STRAIN, followed by the quarterback and then their assigned blocker.
+This is an intuitive result, consistent with previous literature on pass rushing and pressure.
+However, we ignore the fact that pass rushers can be teammates cooperating in the same play leading to some type of correlation structure in their STRAIN values.
+One way to address this complication could be with a adjusted plus-minus (APM) regression approach [@matano2023augmenting], with all players accounted for in modeling a total STRAIN quantity.
 
-Summary
+In addition, we believe there are other concepts in materials science that could be applied to evaluating pass rushing and pass blocking in American football.
+One potential idea is to consider a quantity called stress, which measures force over an area.
+By definition, force is the product of mass and acceleration, meaning this quantity would take into account the physical size of a pass rusher in the computation of force.
+We could then divide this force over the "area" of the pocket formed by pass blockers to compute stress.
 
-Critique/Limitations
-
-Future work rapm
+Moreover, although we focus solely on pass rushers in this paper, STRAIN can be easily extended to the assessment of pass blockers as a unit.
+This can be accomplished through the aforementioned APM regression or by simply looking at quantities such as the total STRAIN or the maximum STRAIN per frame aggregated across the entire play.
+It is also possible to apply STRAIN to assess individual offensive pass blockers, provided that a method of matching blockers to rushers was developed.
+Furthermore, compared to existing metrics, STRAIN measures pass rush effectiveness for every play continuously over time, which is at a much more granular level than considering whether the play resulted in a binary outcome such as a sack.
+There is ample opportunity for working with the complete STRAIN trajectories via functional data analysis to help understand how STRAIN affects offensive production.
 

### 2023-05-16 paper/05-discussion.Rmd
Small grammar edits.

@@
 # Discussion {#sec:discussion}
-
-In this work, we have proposed STRAIN - a simple and interpretable statistic for evaluating pass rushers, with higher STRAIN values corresponding to greater pass rushing ability.
+In this work, we have proposed STRAIN - a simple and interpretable statistic for evaluating pass rushers - with higher values corresponding to greater pass rushing ability.
 We demonstrated that a pass rusher's average STRAIN is both stable and more predictive of future pressure events than using the player's previously observed pressure events.

### 2023-05-16 paper/01-introduction.Rmd
Make some small edits to wording

@@
 Numerous sports are collecting and releasing data on player and ball locations on the playing surface over the course of a game.
-This spatio-temporal multi-resolution source of data has provided exceptional opportunities for researchers to perform advanced studies at a more granular level to deepen our understanding of different sports.
+This multi-resolution, spatio-temporal source of data has provided exceptional opportunities for researchers to perform advanced studies at a more granular level to deepen our understanding of different sports.
 For complete surveys on how tracking data have transformed sports analytics, see @Macdonald2020Recreating, @Baumer2023Big, and @Kovalchik2023Player.
@@
 Our focus of this manuscript is specifically on measuring the performance of defensive linemen in the NFL.
-There are two main types of defensive linemen in American football: (1) defensive tackles (DTs) located within the interior of the line and (2) defensive ends (DEs) located along the edges (see Figure \ref{fig:formation} (top) for example formation with DTs and DEs).
+There are two main types of defensive linemen in American football: (1) defensive tackles (DTs) located within the interior of the line and (2) defensive ends (DEs) located along the edges; see Figure \ref{fig:formation} (top) for example formation with DTs and DEs.
 The primary purpose of both positions is to rush the QB on passing plays, with DEs displaying superiority in observed pass rushing ability [@Eager2018nfl].
-Additionally, within DTs there are nose tackles (NTs) who directly line up across from the ball at the line of scrimmage (see Figure \ref{fig:formation} (bottom) for example defensive formation with a NT).
+Additionally, within DTs there are nose tackles (NTs) who directly line up across from the ball at the line of scrimmage; see Figure \ref{fig:formation} (bottom) for example defensive formation with a NT.
 Typically, teams will employ either one NT or two DTs on the interior with DEs along either side of the defensive line.
@@
 We first describe the player tracking data provided by the Big Data Bowl 2023 in Section \ref{sec:data}.
-We then introduce the mathematical motivation and definition of our measure STRAIN and our modeling approach in Section \ref{sec:methods}.
+We then introduce the mathematical motivation and definition of our measure STRAIN, followed by our modeling approach in Section \ref{sec:methods}.
 Next, we present applications of STRAIN and study different statistical properties of the metric in Section \ref{sec:results}.

### 2023-05-16 paper/02-data.Rmd
Make some small edits to wording

@@
 The data corresponds to 8,557 passing plays across 122 games in the first eight weeks of the 2021 NFL regular season. 
-For each play, we have information on the on-field location, speed, angle, direction, and orientation of each player on the field and the football at a rate of 10 Hz (i.e.
-10 measurements per second), along with event annotations for each frame such as ball snap, pass forward, and quarterback sack, to name a few.
+For each play, we have information on the on-field location, speed, angle, direction, and orientation of each player on the field and the football at a rate of 10 Hz (i.e., 10 measurements per second), along with event annotations for each frame such as ball snap, pass forward, and quarterback sack, to name a few.
 
@@
 In this manuscript, we use the PFF data to identify 36,362 unique pass rush attempts by players designated in the "pass rush" role across all plays.
-For context, Figure \ref{fig:pass_rushers_count} displays the distribution of the number of observed pass rushers involved in a play, ranging from 1 to 8 with 4 pass rushers (i.e.
-4-man rush formation) as the most common value.
+For context, Figure \ref{fig:pass_rushers_count} displays the distribution of the number of observed pass rushers involved in a play, ranging from 1 to 8 with 4 pass rushers (i.e., 4-man rush formation) as the most common value.
 We also leverage the PFF scouting data to count how many hits, hurries, and sacks each pass rusher is credited with across the span of observed data. 

### 2023-05-16 paper/03-methods.Rmd
Make some small edits to wording

@@
 \begin{aligned}
-\overline{\text{STRAIN}}_{ij} &\sim N(R_{j[i]} + B_{b[ij]} + Q_{q[i]} + \mathbf{x_{ij}} \cdot \boldsymbol{\beta}, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
+\overline{\text{STRAIN}}_{ij} &\sim N(R_{j[i]} + B_{b[ij]} + Q_{q[i]} + \mathbf{x_{ij}} \boldsymbol{\beta}, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
 R_{j} &\sim N(\mu_R, \sigma^2_R), \text{ for } j = 1, \cdots, \text{ \# of pass rushers}, \\

### 2023-05-16 paper/04-results.Rmd
Make some small edits to wording

@@
 Additionally, we are ignoring correlated performance between defensive linemen on the same team.
-For instance, the top two defensive ends, Myles Garrett and Takkarist McKinley (see the top panel of Figure \ref{fig:fig_rankings_boot}), are teammates on the Cleveland Browns.
+For instance, the top two defensive ends, Myles Garrett and Takkarist McKinley (see the top panel of Figure \ref{fig:fig_rankings_boot}), were teammates on the Cleveland Browns during this period of time.
 McKinley is a surprising name on our list and we suspect that his high rank is mostly due to playing on the same defensive unit with Garrett, who is well-known as an elite pass rusher.

### 2023-05-17 paper/00-main.Rmd
arxiv submission

@@
 title: |
-  | Get Strained: Analyzing Pass Rush in American Football with Player Tracking Data
+  | Here Comes the STRAIN: Analyzing Defensive Pass Rush in American Football with Player Tracking Data
 
@@
     | Department of Mathematics and Statistics
+    | Center for Data Science and Consulting
     | Loyola University Chicago
@@
 keywords:
-- tracking data
-- football
-- pass rush
-- statistics in sports
+- American football
+- defensive linemen
+- multilevel model
+- player tracking data
 
@@
 We also fit a multilevel model for STRAIN to understand the defensive pressure contribution of every pass rusher at the play-level.
-We apply our approach to NFL data and present results for the first 8 weeks of the 2021 season.
+We apply our approach to NFL data and present results for the first eight weeks of the 2021 regular season.
 In particular, we provide comparisons of STRAIN for different defensive positions and play outcomes, and rankings of the NFL's best pass rushers according to our metric.
@@
 bibliography: bibliography.bib
-biblio-style: chicago
 output: rticles::asa_article
-header-includes:
-  # \usepackage{mathptmx}
-  # \renewcommand{\arraystretch}{0.8}
 ---
@@
   fig.align = "center",
-  # fig.pos = "ht",
-  cache = TRUE
+  fig.pos = "p"
+  #cache = TRUE
 )

### 2023-05-17 paper/01-introduction.Rmd
arxiv submission

@@
 Numerous sports are collecting and releasing data on player and ball locations on the playing surface over the course of a game.
-This multi-resolution, spatio-temporal source of data has provided exceptional opportunities for researchers to perform advanced studies at a more granular level to deepen our understanding of different sports.
+This multiresolution spatiotemporal source of data has provided exceptional opportunities for researchers to perform advanced studies at a more granular level to deepen our understanding of different sports.
 For complete surveys on how tracking data have transformed sports analytics, see @Macdonald2020Recreating, @Baumer2023Big, and @Kovalchik2023Player.
@@
 NGS uses radio frequency identification (RFID) chips placed in players' shoulder pads (and in the ball) to collect data at a rate of 10 frames per second.
-The data captures real-time on-field information such as locations, speeds, and accelerations of all 22 players (and also the football).
+The data captures real-time on-field information such as locations, speeds, and accelerations of all 22 players (and the football).
 While these data were initially only available for teams, media, and vendors, in December 2018 the NFL launched the inaugural edition of their annual Big Data Bowl (BDB) competition [@nfl2023big].
@@
 
-```{r formation, fig.width=5, fig.height=4, fig.cap="Two common defensive alignments in football: 4-3 defense (top) and 3-4 defense (bottom).
+```{r formation, fig.width=5, fig.height=4, fig.pos="t", fig.cap="Two common defensive alignments in football: 4-3 defense (top) and 3-4 defense (bottom).
 The dashed line represents the line of scrimmage separating the defense and offense."}
@@
 
-In this work, using data made available in the 2023 BDB, we present a novel approach to measure the performance of pass rushers.
+In this work, using data made available in the BDB 2023, we present a novel approach to measure the performance of pass rushers.
 Relative to other aspects of American football, such as QB evaluation [@burke2019deepqb, @reyers2021quarterback], the literature on evaluating pass rushers is scarce.
@@
 A hit is a collision between a defender and the opposing team's quarterback after the quarterback makes a throw.
-A hurry represents an instance when a defender successfully disrupts without necessarily making direct contact with the QB and forces the QB to make a throw earlier than expected.
+A hurry represents an instance when a defender successfully disrupts without necessarily making direct contact with the QB and forces the QB to throw the football earlier than expected.
 These are all simple binary measures of pass rush outcome for any given play.
@@
 This is better than the individual counts to some extent, but suffers from problems of subjectivity (e.g., whether there is an actual hurry or not).
-Pro football focus (PFF) defines a metric called pass-rush productivity (PRP), which is a minor modification from the aforementioned pressures metric (see Table \ref{tab:metrics}).
+Pro Football Focus (PFF) defines a metric called pass-rush productivity (PRP), which is a minor modification from the aforementioned pressures metric (see Table \ref{tab:metrics}).
 In particular, PRP gives twice as much weight to a sack relative to hurries and hits, which is a small upgrade to pressures.
@@
 Additionally, we consider a multilevel model to estimate every pass rusher's contribution to the average STRAIN in a play while controlling for other positions like QB and pass blocker as well as various play-level information.
-We note that although our focus in this paper is on pass rushers, our proposed approach can be extended to the evaluation of pass blockers in American football.
+We note that although our focus in this paper is on pass rushers, our approach can be extended to the evaluation of pass blockers in American football.
 

### 2023-05-17 paper/04-results.Rmd
arxiv submission

@@
 In this play, Broncos QB Teddy Bridgewater is well-protected by the offensive line and is able to release a long pass[^unsuccess] to a receiver.
-We see that the STRAIN generated by Crosby is relatively small over the course of this play, compared to the previous play shown in Figure \ref{fig:fig_crosby_curves} which results in a sack.
+We see that the STRAIN generated by Crosby is relatively small over the course of this play, compared to the previous play (as shown in Figure \ref{fig:fig_crosby_curves}) which results in a sack.
 
@@
 
-## Positional STRAIN curves {#sec:positional}
+## Positional STRAIN Curves {#sec:positional}
 
@@
 Figure \ref{fig:fig_pos_curves} displays the average STRAIN by position for the first 40 frames (4 seconds) after the snap.
-We observe a clear difference in STRAIN between edge rushers (OLB's and DE's) and interior linemen (DT's and NT's).
+We observe a clear difference in STRAIN between edge rushers (OLBs and DEs) and interior linemen (DTs and NTs).
 Specifically, edge rushers have higher STRAIN than interior linemen on average, as they are more easily able to approach the quarterback on the edge of the pocket versus, for instance, a nose tackle attacking the line head on.
@@
 Since STRAIN is observed continuously for every play in our data, this allows us to aggregate across all frames played and compute the average STRAIN for NFL pass rushers over the course of the eight-week sample size, as discussed in Section \ref{sec:defn}.
-Based on the clearly distinct patterns for different positions as previously observed, we evaluate interior pass rushers (NT's and DT's) separately from edge rushers (OLB's and DE's).
+Based on the clearly distinct patterns for different positions as previously observed, we evaluate interior pass rushers (NTs and DTs) separately from edge rushers (OLBs and DEs).
 Tables \ref{tab:edge} and \ref{tab:interior} are leaderboards for the NFL's best edge and interior rushers (with at least 100 plays) rated by the average STRAIN across all frames for the first eight weeks of the 2021 regular season.
@@
 In other words, is previous performance predictive of future performance based on our metric?
-Below, we attempt the answer these questions, and the following results are for pass rushers with at least 100 snaps played during the first 8 weeks of the 2021 NFL regular season.
+Below, we attempt the answer these questions, and the following results are for pass rushers with at least 100 snaps played during the first eight weeks of the 2021 NFL regular season.
 
@@
 In particular, we look at how well our metric correlates with a simple measure of pressure rate, defined as the total hits, sacks and hurries per snap.
-Figure \ref{fig:fig_cor_pressure} is a scatterplot of average STRAIN and pressure rate over the course the provided 8-game sample, which reveals a fairly strong correlation ($r =0.625$) between the quantities.
+Figure \ref{fig:fig_cor_pressure} is a scatterplot of average STRAIN and pressure rate over the course the provided eight-game sample, which reveals a fairly strong correlation ($r =0.625$) between the quantities.
 Hence, defenders with high STRAIN values also tend to generate more pressure toward the quarterback.
-Furthermore, the average STRAIN for the first 4 weeks of the 2021 season is more predictive of the last 4 weeks' pressure rate ($r=0.322$) than the first 4 weeks' pressure rate ($r=0.097$), as illustrated in Figure \ref{fig:fig_predictability}.
+Furthermore, the average STRAIN for the first four weeks of the 2021 season is more predictive of the last four weeks' pressure rate ($r=0.322$) than the first four weeks' pressure rate ($r=0.097$), as illustrated in Figure \ref{fig:fig_predictability}.
 
@@
 It is apparent from Figure \ref{fig:fig_stability} that there is a strong positive correlation $(r = 0.854)$ for this relationship.
-This means that STRAIN is a highly stable football metric over the provided 8-week time window, and pass rushers appear to carry their STRAIN values with them from week to week.
+This means that STRAIN is a highly stable football metric over the provided eight-week time window, and pass rushers appear to carry their STRAIN values with them from week to week.
 Overall, STRAIN performs well in both explaining defensive pressure on the field and predicting future performance of pass rushers.
@@
 #| label: fig_cor_pressure
-#| fig.cap: "Relationship between average STRAIN and pressure rate (total hits, sacks, and hurries per snap) over the first 8 weeks of the 2021 NFL season.
+#| fig.cap: "Relationship between average STRAIN and pressure rate (total hits, sacks, and hurries per snap) over the first eight weeks of the 2021 NFL season.
 There is a moderately strong association between average STRAIN and pressure rate ($r =0.625$).
-Results shown here are for pass rushers with at least 100 snaps play over the 8-week data."
+Results shown here are for pass rushers with at least 100 snaps played over the eight-week data."
 #| fig.width: 4
@@
 #| label: fig_predictability
-#| fig.cap: "Relationships between pressure rate for the last 4 weeks of the 2021 NFL season and first 4 weeks' pressure rate (left) and average STRAIN (right).
+#| fig.cap: "Relationships between pressure rate for the last 4 weeks of the 2021 NFL season and first four weeks' pressure rate (left) and average STRAIN (right).
 STRAIN is more predictive ($r=0.322$) of future pressure rate than previous pressure rate itself ($r=0.097$).
-Results shown here are for pass rushers with at least 100 snaps play over the 8-week data."
+Results shown here are for pass rushers with at least 100 snaps played over the eight-week data."
 #| fig.width: 6.5
@@
 #| label: fig_stability
-#| fig.cap: "Relationship between average STRAIN for the last and first 4 weeks of the 2021 NFL season.
+#| fig.cap: "Relationship between average STRAIN for the last and first four weeks of the 2021 NFL season.
 A strong linear correlation ($r = 0.854$) demonstrates that STRAIN is a highly stable metric over time.
-Results shown here are for pass rushers with at least 100 snaps play over the 8-week data."
+Results shown here are for pass rushers with at least 100 snaps played over the eight-week data."
 #| fig.width: 4
@@
 Moreover, we observe statistically significant differences among the pass rush and pass block positions in most cases, while controlling for other variables.
-The coefficient estimates for the pass rush position terms reveal that players with positions closer to the line of scrimmage (DT and NT) appear to generate less STRAIN (relative to DE) than those lining up further back, which is consistent with our results in Section \ref{sec:positional}.
+The coefficient estimates for the pass rush position terms reveal that players with positions closer to the line of scrimmage (DTs and NTs) appear to generate less STRAIN (relative to DEs) than those lining up further back, which is consistent with our results in Section \ref{sec:positional}.
 For blockers, centers are those that absorb the most pressure on average, more than guards and other offensive positions (compared to the baseline level, tackles).

### 2023-05-17 paper/05-discussion.Rmd
arxiv submission

@@
 # Discussion {#sec:discussion}
-In this work, we have proposed STRAIN - a simple and interpretable statistic for evaluating pass rushers - with higher values corresponding to greater pass rushing ability.
+In this work, we have proposed STRAIN---a simple and interpretable statistic for evaluating pass rushers---with higher values corresponding to greater pass rushing ability.
 We demonstrated that a pass rusher's average STRAIN is both stable and more predictive of future pressure events than using the player's previously observed pressure events.
-This is analogous to the use of exit velocity in baseball, a predictive measurement that avoids the noisy nature of observed outcomes, emphasizing how the opportunity of working with player-tracking data can lead to the development of insightful metrics in American football.
+This is analogous to the use of exit velocity in baseball, a predictive measurement that avoids the noisy nature of observed outcomes, emphasizing how the opportunity of working with player tracking data can lead to the development of insightful metrics in American football.
 
@@
 
-Moreover, although we focus solely on pass rushers in this paper, STRAIN can be easily extended to the assessment of pass blockers as a unit.
+Moreover, although we focus solely on pass rushers in this paper, STRAIN can also be applied to the assessment of pass blockers as a unit.
 This can be accomplished through the aforementioned APM regression or by simply looking at quantities such as the total STRAIN or the maximum STRAIN per frame aggregated across the entire play.
@@
 Furthermore, compared to existing metrics, STRAIN measures pass rush effectiveness for every play continuously over time, which is at a much more granular level than considering whether the play resulted in a binary outcome such as a sack.
-There is ample opportunity for working with the complete STRAIN trajectories via functional data analysis to help understand how STRAIN affects offensive production.
+There is ample opportunity for working with the complete STRAIN trajectories via functional data analysis to better understand the impact of STRAIN on offensive production in American football.
 

### 2023-07-05 paper/response-reviewers.Rmd
initialized revision files

---
title: "Response to Reviewers"
output: pdf_document
fontsize: 11pt
---

We thank the editor, associate editor, and two reviewers for their critical assessment of our paper.
Below, we provide a point-by-point response to their comments.

# Reviewer 1

**Model assumptions**

> I really liked the multi-level model for STRAIN. That said, the model encodes some strong exchangeability assumptions that, in my opinion, are worth probing. Insofar as (i) pass blockers are nested within teams and (ii) different teams deploy different pass protection schemes/techniques, have you considered an elaborated model in which pass blocker effects are first shrunk towards a team-specific average, which themselves are shrunk to a league-wide average? Perhaps more importantly, as you noted in Section 4.3, the model ignores correlated performance between defensive linemen on the same team. To account for this, you could, similarly nest pass rushers within teams. Doing so, however, introduces positive dependence between the $R_j$'s and it is not obvious to me that two teammates' strains on the same play should be positively correlated. We might observe negative dependence, for instance, if pass blockers concentrated more on blocking one player over another. At the very least, you should discuss the assumptions underpinning the multi-level model. A clear demonstrate that the proposed model provides a better fit to the data than alternatives would, of course, be much better!

> Relatedly, is there a reason to model the average STRAIN across a play as opposed to the frame-level STRAIN values? The two example plays reveal considerable variation in the frame-to-frame variability. I worry that simply averaging over all frames in a play might obscure this variation.

**Strain curves**

> I was a bit confused by the STRAIN curves in Figure 7. Are these curves based on a single play? Are they averages of the observed $\overline{\text{STRAIN}}_{ij}$'s values across all plays and all players within positions? Or are they averages of the estimated $\overline{\text{STRAIN}}$ values from the fitted multi-level model? If these curves are based solely on the observed $\overline{\text{STRAIN}}$ values, to what extent can the positional differences be explained by the fact that different pass rush positions tend to be blocked by different offensive line positions? Since multi-level model allows you to adjusting for the position of the nearest pass blocker, it would be interesting to see whether the differences observed between the curves persist after such adjustment.

**Minor comments**

> The $f_{d_{ij}}(t)$ notation in Equations (3)–(5) seems cumbersome. What is $f$ and why is indexed by a distance? Why not just use $d_{ij}(t)$ for the distance?

> Fixed effects in the multi-level model: it would be helpful to describe what variables are recorded in $\mathbf{x}$ when you introduce the multi-level model instead of in a table towards the end.

# Reviewer 2

> Page 3: It seems the author is defining anyone who lines up on the line of scrimmage outside a DE. This is not always agreed upon, I would recommend the authors make how they're defining that clear. Is it per play? Or is one player defined as always being a DE?

> Page 13: other fixed effects – this part feels a little rushed. I'd like to see more on why you chose those fixed effects.

> Page 13: mixed effects model – the way this is written, if a quarterback tends to have more pass rushers or less pass rushers than an average QB, there may be some effect on his random effect of average STRAIN that is due to the this effect

> I know time to throw is a fixed effect here, but are we worried that the causal relationship might be that average STRAIN acts on time to throw and not the other way around?

> Why is this random effects model focused on 1 rusher x 1 blocker x 1 qb at a time and not try to put all the players who are present on the field at the same time in the same model? This current format will likely have a lot of teammate effects bleeding into the estimate of another teammate due to being on the field so often together. ( you mention this briefly on page 24 and again in the conclusion – could add to the literature of APM metrics in football (Sabin, 2021).

# Associate Editor

1. The first comment that warrants serious consideration is whether the average strain, across an entire play, is the best way to summarize a pash rushers effect on a particular play. Furthermore, and as just one example, would this summary consistently capture the pash rusher effect across plays of different length?

2. Summarizing the pass rush / pass blocker scheme to a one rusher vs. one defender was also highlighted as a place of concern by both referees. Team pass blocking schemes, multiple pass blockers, and/or changing blockers over the course of the play could all be influential here.

# Editor

1. Avoid abbreviations except obvious ones like NFL and QB. Your paper is short enough (without tables/figures) that we can afford not to use them. This also relates to readability for our readers who are not football experts. See remark #6 below.

2. Reduce the number of tables and figures--you currently have 19 combined. My usual limit is 6-7, but I realize this submission may need to use more in the manuscript. A supplementary article can be used to house additional figures and additional work required to complete this revision. Please see recent issues of TAS for guidance on constructing the supplement. Please submit your revised article and supplement as separate .pdf files.

3. Do not use footnotes on pages 7 and 16. Either absorb into the text (somewhere) if critical or remove them.

4. You might consider using different plotting symbols in Figure 2, for example, when referring to the blue and orange dots. I'm thinking ahead if you want to avoid having to pay for extra charges which relate to the use of color. Same with Figure 4 and others. I'll leave this to you.

5. Only number those equations which are referred to later in the exposition.

6. Overall comment: Not all of our readers will be football experts like in sports analytics journal submissions. Please try to revise with this in mind. Use language that will make your paper inviting for all readers. Have any NFL teams used your methods? You might mention this in the Discussion section if appropriate.

### 2023-07-05 paper/supplemental.Rmd
initialized revision files

---
title: |
  | Supplementary materials for
  | "Here Comes the STRAIN: Analyzing Defensive Pass Rush in American Football with Player Tracking Data"
blinded: 0
bibliography: bibliography.bib
output: rticles::asa_article
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.pos = "p"
)
library(here)
library(tidyverse)
```

### 2023-07-07 scripts/02_fig_crosby_curves.R
prep bw figures

@@ -49,3 +49,47 @@ fig_crosby_curves <- play_lv_den |>
         axis.text = element_text(size = rel(0.8)),
         legend.text = element_text(size = rel(0.8)),
         legend.title = element_text(size = rel(1)))
+
+fig_crosby_curves_bw <- play_lv_den |> 
+  left_join(pff) |>
+  filter(pff_role == "Pass Rush", nflId != 47889) |> 
+  select(name = displayName, frameId_snap_corrected, value = strain) |> 
+  mutate(name = str_c("strain_", name)) |> 
+  bind_rows(pivot_longer(crosby, !frameId_snap_corrected)) |> 
+  mutate(
+    name = case_when(
+      name == "distance_from_QB" ~ "Distance from QB (yards)",
+      name == "slope_distance_from_QB" ~ "Velocity (yards/second)",
+      name == "strain" ~ "STRAIN (1/second)",
+      TRUE ~ as.character(name)
+    ),
+    name_fct = ifelse(str_detect(name, "strain_"), "STRAIN of other rushers", name),
+    name_fct = factor(
+      name_fct,
+      levels = c(
+        "Distance from QB (yards)",
+        "Velocity (yards/second)",
+        "STRAIN (1/second)",
+        "STRAIN of other rushers"
+      )
+    )
+  ) |> 
+  ggplot() +
+  geom_line(aes(frameId_snap_corrected, value, alpha = name_fct, group = name, linetype = name_fct), linewidth = 1.2) +
+  # geom_vline(xintercept = c(10, 20, 30, 40), linetype = "dashed", alpha = 0.3) +
+  labs(x = "Time since snap (seconds)",
+       y = "Feature value") +
+  scale_x_continuous(breaks = seq(0, 40, 10), labels = 0:4) +
+  # scale_color_manual(values = c("red", "#FFCC33" , "#1143E2", "#CDCDCD"), guide = guide_legend(order = 1)) +
+  scale_linetype_manual(values = c("dashed", "dotted" , "solid", "solid"),
+                     guide = guide_legend(order = 1)) +
+  scale_alpha_manual(values = c(1, 1, 1, 0.1),
+                     guide = guide_legend(order = 1)) +
+  labs(linetype = NULL,
+       alpha = NULL) +
+  theme_light() +
+  theme(axis.title = element_text(size = rel(1)),
+        axis.text = element_text(size = rel(0.8)),
+        legend.text = element_text(size = rel(0.8)),
[…]

### 2023-07-07 scripts/03_fig_pos_curves.R

@@ -43,3 +43,28 @@ fig_pos_curves <- pass_rush |>
         axis.text = element_text(size = rel(0.8)),
         legend.title = element_text(size = rel(1)),
         legend.text = element_text(size = rel(0.8)))
+
+fig_pos_curves_bw <- pass_rush |> 
+  filter(team == defensiveTeam,
+         officialPosition %in% c("OLB", "NT", "DE", "DT"),
+         frameId_snap_corrected <= 40) |> 
+  group_by(officialPosition, frameId_snap_corrected) |> 
+  summarise(mn = mean(strain, na.rm = TRUE)) |> 
+  bind_rows(mutate(avg_strain_all, officialPosition = "Average")) |>
+  mutate(officialPosition = factor(officialPosition, levels = c("OLB", "DE", "Average", "DT", "NT"))) |> 
+  ggplot(aes(frameId_snap_corrected, mn, 
+             color = officialPosition, 
+             group = officialPosition, 
+             linetype = officialPosition)) +
+  geom_smooth(se = FALSE, span = 0.3, linewidth = 1) +
+  scale_x_continuous(breaks = seq(0, 40, 10), labels = 0:4) +
+  scale_linetype_manual(values = c("solid", "dashed", "twodash", "dotdash", "dotted"), name = "Position") +
+  scale_color_manual(values = c("black", "black", "gray90", "gray60", "gray60"), name = "Position") +
+  labs(y = "STRAIN",
+       x = "Time since snap (seconds)") +
+  theme_light() +
+  theme(legend.key.width = unit(2, "cm"),
+        axis.title = element_text(size = rel(1)),
+        axis.text = element_text(size = rel(0.8)),
+        legend.title = element_text(size = rel(1)),
+        legend.text = element_text(size = rel(0.8)))

### 2023-07-07 scripts/04_fig_cor_pressure.R

@@ -49,3 +49,30 @@ fig_cor_pressure <- test_summed |>
         legend.text = element_text(size = rel(0.8)),
         axis.text = element_text(size = rel(0.8)),
         panel.grid.minor = element_blank())
+
+fig_cor_pressure_bw <- test_summed |> 
+  left_join(hsh) |> 
+  filter(n_plays >= 100 & !is.infinite(strain)) |> 
+  mutate(pres = (hurries + sacks + hits) / n_plays) |>
+  mutate(pos = factor(pos, levels = c("OLB", "DE", "DT", "NT"))) |> 
+  ggplot(aes(strain_rate, pres)) +
+  geom_smooth(method = "lm", se = FALSE, linewidth = 1.5, color = "black", alpha = 0.5) +
+  geom_point(aes(color = pos, group = name, shape = pos, fill = pos), size = 2, alpha = 0.8) +
+  expand_limits(x = 2.7, y = c(0.02, 0.12)) +
+  scale_x_continuous(breaks = seq(0.5, 2.5, 1)) +
+  scale_y_continuous(breaks = seq(0, 0.12, 0.04)) +
+  #scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107", "#004D40")) +
+  scale_color_manual(values = c("gray20", "gray20", "gray80", "gray80")) +
+  scale_fill_manual(values = c("gray20", "gray20", "gray80", "gray80")) +
+  scale_shape_manual(values = c(21, 24, 22, 25)) +
+  labs(y = "Pressure rate",
+       x = "Average STRAIN",
+       color = "Position",
+       fill = "Position",
+       shape = "Position") +
+  theme_light() +
+  theme(axis.title = element_text(size = rel(1)),
+        legend.title = element_text(size = rel(1)),
+        legend.text = element_text(size = rel(0.8)),
+        axis.text = element_text(size = rel(0.8)),
+        panel.grid.minor = element_blank())

### 2023-07-07 scripts/05_fig_stability.R

@@ -57,3 +57,25 @@ fig_stability <- strain_first |>
           axis.text = element_text(size = rel(0.8)),
           panel.grid.minor = element_blank())
 
+fig_stability_bw <- strain_first |> 
+  full_join(strain_last) |>
+  mutate(pos = factor(pos, levels = c("OLB", "DE", "DT", "NT"))) |> 
+  ggplot(aes(first_strain_rate, last_strain_rate)) +
+  geom_smooth(method = "lm", se = FALSE, linewidth = 1.5, color = "black", alpha = 0.5) +
+  geom_point(aes(color = pos, group = name, fill = pos, shape = pos), size = 2, alpha = 0.8) +
+  #scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107", "#004D40")) +
+  scale_color_manual(values = c("gray20", "gray20", "gray80", "gray80")) +
+  scale_fill_manual(values = c("gray20", "gray20", "gray80", "gray80")) +
+  scale_shape_manual(values = c(21, 24, 22, 25)) +
+  labs(x = "Average STRAIN (first 4 weeks)",
+       y = "Average STRAIN (last 4 weeks)",
+       color = "Position",
+       fill = "Position",
+       shape = "Position") +
+  theme_light() +
+  theme(axis.title = element_text(size = rel(1)),
+        legend.title = element_text(size = rel(1)),
+        legend.text = element_text(size = rel(0.8)),
+        axis.text = element_text(size = rel(0.8)),
+        panel.grid.minor = element_blank())
+

### 2023-07-07 scripts/07_fig_predictability.R

@@ -96,3 +96,61 @@ strain_hsh <- hsh_last |>
         panel.grid.minor = element_blank())
 
 fig_predictability <- cowplot::plot_grid(hsh_hsh, strain_hsh, rel_widths = c(2.1, 2.4))
+
+
+# bw ----------------------------------------------------------------------
+
+hsh_hsh_bw <- hsh_first |> 
+  full_join(hsh_last) |>
+  left_join(select(players, nflId, officialPosition)) |>
+  mutate(officialPosition = factor(officialPosition, levels = c("OLB", "DE", "DT", "NT"))) |> 
+  ggplot(aes(first_hsh, last_hsh)) +
+  geom_smooth(method = "lm", se = FALSE, linewidth = 1.5, color = "black", alpha = 0.5) +
+  geom_point(aes(color = officialPosition, group = nflId, fill = officialPosition, shape = officialPosition), size = 2, alpha = 0.8) +
+  #scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107", "#004D40")) +
+  scale_color_manual(values = c("gray20", "gray20", "gray80", "gray80")) +
+  scale_fill_manual(values = c("gray20", "gray20", "gray80", "gray80")) +
+  scale_shape_manual(values = c(21, 24, 22, 25)) +
+  labs(x = "Pressure rate (first 4 weeks)",
+       y = "Pressure rate (last 4 weeks)",
+       color = "Position",
+       fill = "Position",
+       shape = "Position") +
+  theme_light() +
+  theme(axis.title = element_text(size = rel(1)),
+        legend.title = element_text(size = rel(1)),
+        legend.text = element_text(size = rel(0.8)),
+        legend.position = "none",
+        axis.text = element_text(size = rel(0.8)),
+        panel.grid.minor = element_blank())
+
+
+#0.3216974
+
+strain_hsh_bw <- hsh_last |> 
+  left_join(select(players, nflId, officialPosition, displayName)) |> 
+  select(name = displayName, pos = officialPosition, last_hsh) |> 
+  full_join(strain_first) |>
+  mutate(pos = factor(pos, levels = c("OLB", "DE", "DT", "NT"))) |> 
+  ggplot(aes(first_strain_rate, last_hsh)) +
+  geom_smooth(method = "lm", se = FALSE, linewidth = 1.5, color = "black", alpha = 0.5) +
[…]

### 2023-07-07 scripts/08_fig_outcome_curves.R

@@ -82,3 +82,30 @@ fig_outcome_curves <- hit |>
         legend.text = element_text(size = rel(0.8)))
 
 
+fig_outcome_curves_bw <- hit |> 
+  bind_rows(sack) |> 
+  bind_rows(hurry) |> 
+  bind_rows(none) |> 
+  mutate(officialPosition = factor(officialPosition, levels = c("OLB", "DE", "DT", "NT")),
+         outcome = factor(outcome, levels = c("Sack", "Hit", "Hurry", "None"))) |> 
+  ggplot(aes(frameId_snap_corrected, mn, 
+             color = officialPosition, 
+             group = officialPosition,
+             linetype = officialPosition)) +
+  geom_smooth(se = FALSE, span = 0.4, linewidth = 1) +
+  scale_x_continuous(breaks = seq(0, 40, 10), labels = 0:4) +
+  scale_linetype_manual(values = c("solid", "dashed", "dotdash", "dotted"), name = "Position") +
+  scale_color_manual(values = c("black", "black", "gray60", "gray60"), name = "Position") +
+  expand_limits(y = 2.3) +
+  labs(y = "STRAIN",
+       x = "Time since snap (seconds)") +
+  theme_light() +
+  facet_wrap(~ outcome, nrow = 1) +
+  theme(legend.key.width = unit(2, "cm"),
+        legend.position = "bottom",
+        axis.title = element_text(size = rel(1)),
+        axis.text = element_text(size = rel(0.8)),
+        legend.title = element_text(size = rel(1)),
+        legend.text = element_text(size = rel(0.8)))
+
+

### 2023-07-07 paper/03-methods.Rmd
arxiv submission
fixed distance notation (#11)

@@
 \begin{equation}
-s(t)= \frac{L(t)-L_{0}}{L_{0}}.
+\varepsilon(t)= \frac{L(t) - L_{0}}{L_{0}}.
 \end{equation}
@@
 \begin{equation}
-s'(t) = \frac{ds}{dt} = \frac{v(t)}{L_{0}},
+\varepsilon'(t) = \frac{d \varepsilon}{dt} = \frac{v(t)}{L_{0}},
 \end{equation}
@@
 \begin{equation}
-f_{d_{ij}}(t) = \sqrt{(x_{ijt} - x^{QB}_{it})^2 + (y_{ijt} - y^{QB}_{it})^2}.
+s_{ij}(t) = \sqrt{(x_{ijt} - x^{QB}_{it})^2 + (y_{ijt} - y^{QB}_{it})^2}.
 \end{equation}
@@
 \begin{equation}
-f'_{d_{ij}}(t) = \frac{df_{d_{ij}}(t)}{dt}.
+v_{ij}(t) = s'_{ij}(t) = \frac{ds_{ij}(t)}{dt}.
 \end{equation}
@@
 \begin{equation}
-\text{STRAIN}_{ij}(t) = \frac{-f'_{d_{ij}}(t)}{f_{d_{ij}}(t)}.
+\text{STRAIN}_{ij}(t) = \frac{v_{ij}(t)}{s_{ij}(t)}.
 \end{equation}
@@
 \begin{equation}
-\widehat{\text{STRAIN}}_{ij}(t) = \cfrac{-\cfrac{f_{d_{ij}}(t) - f_{d_{ij}}(t - 1)}{0.1}}{f_{d_{ij}}(t)}.
+\widehat{\text{STRAIN}}_{ij}(t) = \cfrac{-\cfrac{s_{ij}(t) - s_{ij}(t - 1)}{0.1}}{s_{ij}(t)}.
 \end{equation}
@@
 We use random intercepts for the three player groups: (1) pass rushers as $R$, (2) nearest pass blockers as $B$, and (3) QBs as $Q$.
-The coefficients $\boldsymbol \beta$ for covariate vector $\mathbf{x_{ij}}$ for pass rusher $j$ in play $i$ are estimated as fixed-effects. 
+The coefficients $\boldsymbol \beta$ associated with covariate vector $\mathbf{x_{ij}}$ for pass rusher $j$ in play $i$ are estimated as fixed-effects. 
 Our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows.

### 2023-07-08 scripts/10_multilevel_mod.R
new multilevel model code

@@ -126,7 +126,8 @@ mod_df_final <- mod_df |>
   left_join(n_blockers) |> 
   mutate(block_pos = relevel(factor(block_pos), "T"),
          rush_pos = relevel(factor(rush_pos), "DE"),
-         play_len = play_len / 10)
+         play_len = play_len / 10) |> 
+  left_join(select(plays, gameId:playId, possessionTeam:defensiveTeam))
 
 
 library(lme4)  
@@ -212,7 +213,50 @@ fig_rankings_boot <- strain_eff |>
   geom_point(aes(y = mean)) +
   geom_errorbar(aes(ymin = mean - 2 * sd,
                     ymax = mean + 2 * sd)) +
-  facet_wrap(~officialPosition, ncol = 1, scales = "free_y") +
+  facet_wrap(~ officialPosition, ncol = 1, scales = "free_y") +
+  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
+  coord_flip() +
+  theme_light() +
+  labs(x = NULL, y = "intercept")
+
+# revised model -----------------------------------------------------------
+
+strain_fit_new <- lmer(
+  avg_strain ~ (1 | nflId) + (1 | blocker_id) + (1 | possessionTeam) + (1 | defensiveTeam) + 
+    play_len + rush_pos + block_pos + n_blockers,
+  data = mod_df_final
+)
+
+summary(strain_fit_new)$coef
+broom.mixed::tidy(strain_fit_new, conf.int=TRUE)
+
+
+strain_fit_new |> 
+  VarCorr() |> 
+  as_tibble() |> 
+  mutate(icc = vcov / sum(vcov)) |> 
+  select(grp, icc)
+
+strain_eff_new <- merTools::REsim(strain_fit_new, n.sims = 10000, seed = 101)
+
+# strain_eff_new |>
+#   merTools::plotREsim()
+
+strain_eff_new |> 
+  as_tibble() |> 
+  filter(groupFctr == "nflId") |> 
+  mutate(nflId = as.double(groupID)) |> 
+  left_join(players) |> 
+  filter(nflId %in% filter(pass_rush_snaps, n_plays >= 100)$nflId) |> 
+  filter(officialPosition %in% c("DE", "OLB", "DT", "NT")) |> 
+  group_by(officialPosition) |> 
+  arrange(desc(mean)) |>
+  slice(1:10) |>
+  ggplot(aes(x = reorder(displayName, mean))) +
[…]

### 2023-07-08 scripts/10_multilevel_mod.R
leave one week out cv for multilevel models

@@ -24,6 +24,7 @@ library(here)
 #   summarize(n_plays = n_distinct(playId))
 # write_rds(pass_block_snaps, here("data", "pass_block_snaps.rds"))
 
+games <- read_csv(here("data", "games.csv"))
 plays <- read_csv(here("data", "plays.csv"))
 players <- read_csv(here("data", "players.csv"))
 pff <- read_csv(here("data", "pffScoutingData.csv"))
@@ -231,6 +232,14 @@ summary(strain_fit_new)$coef
 broom.mixed::tidy(strain_fit_new, conf.int=TRUE)
 
 
+
+strain_fit_new |> 
+  ranef() |> 
+  pluck("possessionTeam") |> 
+  as.data.frame() |> 
+  arrange(`(Intercept)`)
+
+
 strain_fit_new |> 
   VarCorr() |> 
   as_tibble() |> 
@@ -261,3 +270,118 @@ strain_eff_new |>
   coord_flip() +
   theme_light() +
   labs(x = NULL, y = "intercept")
+
+
+
+
+
+
+
+
+
+# lowocv ------------------------------------------------------------------
+
+mod_df_final <- mod_df_final |> 
+  left_join(select(games, gameId, week))
+
+strain_loocv_initial <- function(w) {
+  
+  train <- filter(mod_df_final, week != w)
+  test <- filter(mod_df_final, week == w)
+  
+  fit <- lmer(
+    avg_strain ~ (1 | nflId) + (1 | blocker_id) + (1 | passer) + play_len + rush_pos + block_pos + n_blockers,
+    data = train
+  )
+  
+  out <- tibble(
+    pred = predict(fit, newdata = test, allow.new.levels = TRUE),
+    obs = pull(test, avg_strain),
+    week = w
+  )
+  return(out)
+}
+
+strain_loocv_initial_df <- map(1:8, strain_loocv_initial) |> 
+  list_rbind()
+
+library(yardstick)
+strain_loocv_initial_rmse <- strain_loocv_initial_df |>
+  group_by(week) |> 
+  rmse(obs, pred)
+
+
+# strain_loocv_initial_df |>
+#   mutate(resid = obs - pred,
+#          sq_err = resid ^ 2) |>
+#   group_by(week) |>
+#   summarize(mse = mean(sq_err, na.rm = TRUE),
+#             # lower = quantile(sq_err, 0.025, na.rm = TRUE),
[…]

### 2023-07-08 scripts/10_multilevel_mod.R
changed lowo mse plot to dodged position

@@ -260,7 +260,7 @@ strain_eff_new |>
   filter(officialPosition %in% c("DE", "OLB", "DT", "NT")) |> 
   group_by(officialPosition) |> 
   arrange(desc(mean)) |>
-  slice(1:10) |>
+  slice(1:8) |>
   ggplot(aes(x = reorder(displayName, mean))) +
   geom_point(aes(y = mean)) +
   geom_errorbar(aes(ymin = mean - 2 * sd,
@@ -311,35 +311,6 @@ strain_loocv_initial_rmse <- strain_loocv_initial_df |>
   rmse(obs, pred)
 
 
-# strain_loocv_initial_df |>
-#   mutate(resid = obs - pred,
-#          sq_err = resid ^ 2) |>
-#   group_by(week) |>
-#   summarize(mse = mean(sq_err, na.rm = TRUE),
-#             # lower = quantile(sq_err, 0.025, na.rm = TRUE),
-#             # upper = quantile(sq_err, 0.975, na.rm = TRUE),
-#             se = sd(sq_err, na.rm = TRUE) / sqrt(n()),
-#             type = "no team eff") |>
-#   bind_rows(
-#     strain_loocv_new_df |>
-#       mutate(resid = obs - pred,
-#              sq_err = resid ^ 2) |>
-#       group_by(week) |>
-#       summarize(mse = mean(sq_err, na.rm = TRUE),
-#                 # lower = quantile(sq_err, 0.025, na.rm = TRUE),
-#                 # upper = quantile(sq_err, 0.975, na.rm = TRUE),
-#                 se = sd(sq_err, na.rm = TRUE) / sqrt(n()),
-#                 type = "with team eff")
-#   ) |>
-#   mutate(lower = mse - 2*se, upper = mse + 2*se) |> 
-#   ggplot(aes(week, mse, color = type)) +
-#   #geom_col() +
-#   geom_point() +
-#   geom_line() +
-#   geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
-#   facet_wrap(~ type)
-
-
 strain_loocv_new <- function(w) {
   
   train <- filter(mod_df_final, week != w)
@@ -383,5 +354,36 @@ mutate(strain_loocv_initial_rmse, team_eff = "No") |>
        y = "RMSE",
        color = "Team effects?") +
   scale_x_continuous(breaks = 1:8) +
-  theme_light() 
-  
+  theme_light()
+
+
+strain_loocv_initial_df |>
[…]

### 2023-07-08 scripts/10_multilevel_mod.R
compared min/max strain vs avg strain

@@ -379,8 +379,8 @@ strain_loocv_initial_df |>
   ) |>
   mutate(lower = mse - 2*se, upper = mse + 2*se) |>
   ggplot(aes(week, mse, color = type)) +
-  geom_errorbar(aes(ymin = lower, ymax = upper), position = position_dodge(width = 0.5), width = 0.5, linewidth = 1) +
-  geom_point(position = position_dodge(width = 0.5), size = 2.5) +
+  geom_errorbar(aes(ymin = lower, ymax = upper), position = position_dodge(width = 0.5), width = 0.5) +
+  geom_point(position = position_dodge(width = 0.5), size = 2) +
   scale_x_continuous(breaks = 1:8) +
   labs(x = "Left-out week",
        y = "MSE",

### 2023-07-08 scripts/13_strain_min_max.R

@@ -1,0 +1,33 @@
+library(tidyverse)
+library(here)
+
+games <- read_csv(here("data", "games.csv"))
+source(here("scripts", "04_fig_cor_pressure.R"))
+
+min_100_snaps <- test |> 
+  select(nflId, team, displayName, officialPosition) |> 
+  distinct() |> 
+  right_join(test_summed) |> 
+  filter(n_plays >= 100 & !is.infinite(strain)) |> 
+  pull(displayName)
+
+strain_first <- test |>
+  left_join(select(games, gameId, week)) |> 
+  filter(displayName %in% min_100_snaps, week %in% 1:4, !is.infinite(strain)) |> 
+  group_by(nflId) |>
+  summarize(mean_strain = mean(strain),
+            min_strain = min(strain),
+            max_strain = max(strain))
+
+strain_last <- test |>
+  left_join(select(games, gameId, week)) |> 
+  filter(displayName %in% min_100_snaps, week %in% 5:8, !is.infinite(strain)) |> 
+  group_by(nflId) |>
+  summarize(mean_strain = mean(strain),
+            min_strain = min(strain),
+            max_strain = max(strain))
+
+
+cor(strain_first$mean_strain, strain_last$mean_strain)
+cor(strain_first$min_strain, strain_last$min_strain)
+cor(strain_first$max_strain, strain_last$max_strain)

### 2023-07-10 paper/supplemental.Rmd
Move previous multilevel model description to supplement

@@
 ```
+
+## Old Multilevel Model for Play-Level STRAIN {#sec:multilevel}
+
+We additionally fit a multilevel model to evaluate pass rushers' impact on the average STRAIN observed in a play, while accounting for their pass blocker, QB, and play-level attributes.
+This version of our model has random intercepts for the three player groups: (1) pass rushers as $R$, (2) nearest pass blockers as $B$, and (3) QBs as $Q$.
+As before, the coefficients $\boldsymbol \beta$ associated with covariate vector $\mathbf{x_{ij}}$ for pass rusher $j$ in play $i$ are estimated as fixed-effects. 
+This version of our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows.
+\begin{equation}
+\begin{aligned}
+\overline{\text{STRAIN}}_{ij} &\sim N(R_{j[i]} + B_{b[ij]} + Q_{q[i]} + \mathbf{x_{ij}} \boldsymbol{\beta}, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
+R_{j} &\sim N(\mu_R, \sigma^2_R), \text{ for } j = 1, \cdots, \text{ \# of pass rushers}, \\
+B_{b} &\sim N(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
+Q_{q} &\sim N(\mu_Q, \sigma^2_Q), \text{ for } q = 1, \cdots, \text{ \# of QBs}.
+\end{aligned}
+\end{equation}
+
+We consider a normal distribution to obtain the random intercepts for each player toward their respective position group means.

### 2023-07-11 paper/03-methods.Rmd
update fixed effects description #11 and add a brief description about resampling strategy for boostrapping uncertainty for random effects

@@
 While this aggregated measure is a simple first step for pass rush evaluation, the observed average STRAIN on a single play is likely due to a convolution of factors.
-Outside of the pass rusher's ability, there is variability in the opposing strength of pass blockers across plays a pass rusher is involved.
+Outside of the pass rusher's ability, there is variability in the opposing strength of pass blockers across plays a pass rusher is involved - both at the individual and team level.
 Additionally, based on previous work by PFF, the opposing QB plays an important role on the observed pressure in passing plays [@eager2019pressure].
-Thus, we need to appropriately divide the credit of an observed average STRAIN across the different players involved and other factors.
+We also suspect that pass rusher's STRAIN is affected by their defensive teammates as well.
+Thus, we need to appropriately divide the credit of an observed average STRAIN across the different players and team involved, amongst other factors.
 
-To this end, we fit a multilevel model to evaluate pass rushers' impact on the average STRAIN observed in a play, while accounting for their pass blocker, QB, and play-level attributes.
+To this end, we fit a multilevel model to evaluate pass rushers' impact on the average STRAIN observed in a play, while accounting for their team on defense, the opposing team on offense, and their assigned pass blocker.
 We identify the pass blocker linked with the pass rusher of interest using the data provided by PFF.
 Since there can be multiple blockers matching up with a rusher, for simplicity, we consider the nearest blocker positioned to the pass rusher at the start of the play.
-We use random intercepts for the three player groups: (1) pass rushers as $R$, (2) nearest pass blockers as $B$, and (3) QBs as $Q$.
-The coefficients $\boldsymbol \beta$ associated with covariate vector $\mathbf{x_{ij}}$ for pass rusher $j$ in play $i$ are estimated as fixed-effects. 
+We use random intercepts for the two player groups: pass rushers as $R$ and nearest pass blockers as $B$, as well as for the two team groups: defense $D$ and offense $O$.
+We also account for attributes about pass rusher $j$ in play $i$ through the covariate vector $\mathbf{x_{ij}}$, and estimate their respective coefficients $\boldsymbol \beta$ as fixed-effects. 
 Our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows.
@@
 \begin{aligned}
-\overline{\text{STRAIN}}_{ij} &\sim N(R_{j[i]} + B_{b[ij]} + Q_{q[i]} + \mathbf{x_{ij}} \boldsymbol{\beta}, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
+\overline{\text{STRAIN}}_{ij} &\sim N(R_{j[i]} + B_{b[ij]} + D_{d[i]} + O_{o[i]} + \mathbf{x_{ij}} \boldsymbol{\beta}, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
 R_{j} &\sim N(\mu_R, \sigma^2_R), \text{ for } j = 1, \cdots, \text{ \# of pass rushers}, \\
 B_{b} &\sim N(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
-Q_{q} &\sim N(\mu_Q, \sigma^2_Q), \text{ for } q = 1, \cdots, \text{ \# of QBs}.
+D_{d} &\sim N(\mu_D, \sigma^2_D), \text{ for } d = 1, \cdots, \text{ \# of teams}, \\
+O_{o} &\sim N(\mu_O, \sigma^2_O), \text{ for } o = 1, \cdots, \text{ \# of teams}.
 \end{aligned}
@@
 
-We consider a normal distribution to obtain the random intercepts for each player toward their respective position group means.
-This provides us with average individual effects on STRAIN for each player while applying shrinkage, which is a useful property in this problem's context since we do not observe the same number of plays for each player.
-With regards to the fixed-effects about pass rusher $j$ in play $i$, we include the length of the play (i.e., the elapsed time between the ball snap and a QB event), the positions of pass rusher and nearest blocker, and the number of pass blockers assigned to the pass rusher based on the PFF data.
+We consider a normal distribution to shrink the random intercepts for each player and team toward their respective group means, which is a useful property this problem's context since we do not observe the same number of plays for each player.
+For the team effects, this provides us with the average defense and offense team-level effects on a pass rusher's STRAIN.
+Due to the nested nature of players on teams, the individual pass rusher and blocker random intercepts reflect the respective player's effect relative to their team effects.
+We implement the model using penalized likelihood via the `lme4` package in `R` [@bateslme4; @R2023Language].
+
+With regards to the fixed-effects about pass rusher $j$ in play $i$, we include a variety of features that likely contribute to variation in STRAIN.
+First, we adjust for the position of both the pass rusher and nearest blocker to account for any positional effects.
 Table \ref{tab:pos} shows our positional categorization for the pass rush and pass block roles.
-We implement the model using penalized likelihood via the `lme4` package in `R` [@bateslme4; @R2023Language].
-Additionally, we obtain interval estimates for the player-level random intercepts via parametric bootstrapping of 10,000 samples with the `merTools` package [@knowlesmerTools; @gelman2006data].
+These are encoded as indicator variables with defensive ends and tackles as the reference groups for the pass rushers and blockers respectively.
+We also adjust for the number of pass blockers on the play, since teams may decide to employ a more protective scheme that could lower the observed STRAIN.
+Finally, we also adjust for the length of the play (i.e., the elapsed time between the ball snap and a quarterback event).
+We expect a longer display to display lower STRAIN since this would indicate a quarterback has more time to throw the ball.
+NEED TO THINK ABOUT THIS MORE....
+
+<!--Additionally, we obtain interval estimates for the player-level random intercepts via parametric bootstrapping of 10,000 samples with the `merTools` package [@knowlesmerTools; @gelman2006data].-->
 
@@
 ```
+
+In order to provide a measure of uncertainty regarding about our random effects, we use a bootstrapping strategy similar to the approach in @yurko2019nflwar.
+Specifically, we resample team drives within games.
+This preserves the fact that team schedules are fixed, but allows for random variation in player usage since this is dependent on team decision making.
+By resampling plays within the same drive together, our sampling is generating realistic simulated data in comparison to sampling individual plays.
+For each resampled dataset, we fit the aforementioned multilevel model to generate a distribution of estimates for the considered player and team effects.

### 2023-07-11 scripts/12_fig_pass_rushers_count.R
number of blockers code

@@ -17,3 +17,40 @@ pass_rushers_count <- pff_full |>
   scale_x_continuous(breaks = 1:8) +
   labs(x = "Number of pass rushers",
        y = "Count")
+
+pass_blockers_count <- pff_full |> 
+  filter(pff_role == "Pass Block") |> 
+  count(gameId, playId) |> 
+  count(n) |> 
+  ggplot(aes(n, nn)) +
+  geom_col() +
+  scale_x_continuous(breaks = 1:8) +
+  labs(x = "Number of pass blockers",
+       y = "Count")
+
+
+rushers_blockers <- pff_full |> 
+  filter(pff_role == "Pass Rush") |> 
+  count(gameId, playId, name = "n_rushers") |> 
+  full_join(
+    pff_full |> 
+      filter(pff_role == "Pass Block") |> 
+      count(gameId, playId, name = "n_blockers")
+  )
+
+rushers_blockers |> 
+  select(contains("n_")) |> 
+  drop_na() |> 
+  cor()
+
+rushers_blockers |> 
+  ggplot(aes(n_rushers, n_blockers)) +
+  geom_point() +
+  geom_smooth(method = "lm", se = FALSE)
+
+
+rushers_blockers |> 
+  count(n_rushers, n_blockers, sort = TRUE) |> 
+  ggplot(aes(n_rushers, n_blockers)) +
+  geom_tile(aes(fill = n), color = "white") +
+  coord_fixed()

### 2023-07-13 scripts/10_multilevel_mod.R
set up bootstrap (#15)

@@ -102,6 +102,10 @@ mod_df <- pff |>
 # mod_df |> 
 #   count(block_pos)
 
+n_rushers <- pff |> 
+  filter(pff_role == "Pass Rush") |> 
+  count(gameId, playId, name = "n_rushers")
+
 # some rushers are assigned 4 blockers 
 n_blockers <- mod_df |> 
   count(gameId, playId, nflId, name = "n_blockers")
@@ -125,6 +129,7 @@ mod_df_final <- mod_df |>
   slice_min(dis, n = 1) |> 
   ungroup() |> 
   left_join(n_blockers) |> 
+  left_join(n_rushers) |> 
   mutate(block_pos = relevel(factor(block_pos), "T"),
          rush_pos = relevel(factor(rush_pos), "DE"),
          play_len = play_len / 10) |> 
@@ -222,9 +227,9 @@ fig_rankings_boot <- strain_eff |>
 
 # revised model -----------------------------------------------------------
 
+# play_len
 strain_fit_new <- lmer(
-  avg_strain ~ (1 | nflId) + (1 | blocker_id) + (1 | possessionTeam) + (1 | defensiveTeam) + 
-    play_len + rush_pos + block_pos + n_blockers,
+  avg_strain ~ (1 | nflId) + (1 | blocker_id) + (1 | possessionTeam) + (1 | defensiveTeam) + rush_pos + block_pos + n_blockers + n_rushers,
   data = mod_df_final
 )
 
@@ -232,6 +237,11 @@ summary(strain_fit_new)$coef
 broom.mixed::tidy(strain_fit_new, conf.int=TRUE)
 
 
+strain_fit_new |> 
+  ranef() |> 
+  pluck("defensiveTeam") |> 
+  as.data.frame() |> 
+  arrange(-`(Intercept)`)
 
 strain_fit_new |> 
   ranef() |> 
@@ -317,8 +327,7 @@ strain_loocv_new <- function(w) {
   test <- filter(mod_df_final, week == w)
   
   fit <- lmer(
-    avg_strain ~ (1 | nflId) + (1 | blocker_id) + (1 | possessionTeam) + (1 | defensiveTeam) + 
-      play_len + rush_pos + block_pos + n_blockers,
+    avg_strain ~ (1 | nflId) + (1 | blocker_id) + (1 | possessionTeam) + (1 | defensiveTeam) + rush_pos + block_pos + n_blockers + n_rushers,
     data = mod_df_final
   )
   
[…]

### 2023-07-13 scripts/10_multilevel_mod.R
removed unnecessary code

@@ -236,7 +236,6 @@ strain_fit_new <- lmer(
 summary(strain_fit_new)$coef
 broom.mixed::tidy(strain_fit_new, conf.int=TRUE)
 
-
 strain_fit_new |> 
   ranef() |> 
   pluck("defensiveTeam") |> 
@@ -249,44 +248,11 @@ strain_fit_new |>
   as.data.frame() |> 
   arrange(`(Intercept)`)
 
-
 strain_fit_new |> 
   VarCorr() |> 
   as_tibble() |> 
   mutate(icc = vcov / sum(vcov)) |> 
   select(grp, icc)
-
-strain_eff_new <- merTools::REsim(strain_fit_new, n.sims = 10000, seed = 101)
-
-# strain_eff_new |>
-#   merTools::plotREsim()
-
-strain_eff_new |> 
-  as_tibble() |> 
-  filter(groupFctr == "nflId") |> 
-  mutate(nflId = as.double(groupID)) |> 
-  left_join(players) |> 
-  filter(nflId %in% filter(pass_rush_snaps, n_plays >= 100)$nflId) |> 
-  filter(officialPosition %in% c("DE", "OLB", "DT", "NT")) |> 
-  group_by(officialPosition) |> 
-  arrange(desc(mean)) |>
-  slice(1:8) |>
-  ggplot(aes(x = reorder(displayName, mean))) +
-  geom_point(aes(y = mean)) +
-  geom_errorbar(aes(ymin = mean - 2 * sd,
-                    ymax = mean + 2 * sd)) +
-  facet_wrap(~ officialPosition, ncol = 1, scales = "free_y") +
-  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
-  coord_flip() +
-  theme_light() +
-  labs(x = NULL, y = "intercept")
-
-
-
-
-
-
-
 
 
 # lowocv ------------------------------------------------------------------
@@ -400,7 +366,6 @@ strain_loocv_initial_df |>
 
 # resampling --------------------------------------------------------------
 
-
 strain_boot <- function(b) {
   boot_df <- mod_df_final |> 
     group_by(gameId, playId) |> 
@@ -430,12 +395,10 @@ strain_boot <- function(b) {
 library(furrr)
 plan(multisession, workers = 5)
 set.seed(101)
-tictoc::tic()
 n_boots <- 1000
 strain_boot_eff <- seq_len(n_boots) |> 
[…]

### 2023-07-13 paper/04-results.Rmd
Updated language about the positionsl STRAIN curves.

@@
 Figure \ref{fig:fig_pos_curves} displays the average STRAIN by position for the first 40 frames (4 seconds) after the snap.
+For each frame, this curve is calculated by averaging across all players of a particular position (where position is defined by pro football focus) and all plays that lasted at least as long as the particular frame of interest meaning that the sample size used to compute these curves diminishes for frames that are chronologically later than frames at the beginning of the play.
+
 We observe a clear difference in STRAIN between edge rushers (OLBs and DEs) and interior linemen (DTs and NTs).

### 2023-07-13 paper/04-results.Rmd
Updated caption on figure 7.
NFl not PFF.

--- 2023-05-17 paper/04-results.Rmd
@@
 #| label: fig_pos_curves
-#| fig.cap: "STRAIN curves for different positions.
+#| fig.cap: "STRAIN curves for different positions computed by averaging STRAIN values across all players of a particular position and all plays.
 Edge rushers (OLB and DE) tend to generate more STRAIN than interior rushers (DT and NT).
@@
 Figure \ref{fig:fig_pos_curves} displays the average STRAIN by position for the first 40 frames (4 seconds) after the snap.
+For each frame, this curve is calculated by averaging across all players of a particular position (where position is defined by the NFL) and all plays that lasted at least as long as the particular frame of interest meaning that the sample size used to compute these curves diminishes for frames that are chronologically later than frames at the beginning of the play.
+
 We observe a clear difference in STRAIN between edge rushers (OLBs and DEs) and interior linemen (DTs and NTs).

### 2023-07-14 scripts/03_fig_pos_curves.R
corrected for strain per second

@@ -16,7 +16,11 @@ avg_strain_all <- pass_rush |>
   filter(team == defensiveTeam,
          frameId_snap_corrected <= 40) |> 
   group_by(frameId_snap_corrected) |> 
-  summarise(mn = mean(strain, na.rm = TRUE))
+  summarize(
+    strain = sum(strain),
+    n = n()
+  ) |>
+  mutate(mn = 10 * strain / n)
 
 
 fig_pos_curves <- pass_rush |> 
@@ -24,7 +28,11 @@ fig_pos_curves <- pass_rush |>
          officialPosition %in% c("OLB", "NT", "DE", "DT"),
          frameId_snap_corrected <= 40) |> 
   group_by(officialPosition, frameId_snap_corrected) |> 
-  summarise(mn = mean(strain, na.rm = TRUE)) |> 
+  summarize(
+    strain = sum(strain),
+    n = n()
+  ) |>
+  mutate(mn = 10 * strain / n) |> 
   bind_rows(mutate(avg_strain_all, officialPosition = "Average")) |>
   mutate(officialPosition = factor(officialPosition, levels = c("OLB", "DE", "DT", "NT", "Average"))) |> 
   ggplot(aes(frameId_snap_corrected, mn, 
@@ -49,7 +57,11 @@ fig_pos_curves_bw <- pass_rush |>
          officialPosition %in% c("OLB", "NT", "DE", "DT"),
          frameId_snap_corrected <= 40) |> 
   group_by(officialPosition, frameId_snap_corrected) |> 
-  summarise(mn = mean(strain, na.rm = TRUE)) |> 
+  summarize(
+    strain = sum(strain),
+    n = n()
+  ) |>
+  mutate(mn = 10 * strain / n) |> 
   bind_rows(mutate(avg_strain_all, officialPosition = "Average")) |>
   mutate(officialPosition = factor(officialPosition, levels = c("OLB", "DE", "Average", "DT", "NT"))) |> 
   ggplot(aes(frameId_snap_corrected, mn, 

### 2023-07-14 scripts/10_multilevel_mod.R

@@ -47,7 +47,11 @@ pass_rush <- read_csv(here("data", "pass_rush.csv.gz"))
 get_avg_strain <- pass_rush |> 
   filter(team == defensiveTeam, !is.infinite(strain)) |> 
   group_by(gameId, playId, nflId) |> 
-  summarize(avg_strain = mean(strain, na.rm = TRUE)) |>
+  summarize(
+    strain = sum(strain),
+    n = n()
+  ) |>
+  mutate(avg_strain = 10 * strain / n) |> 
   ungroup()
 
 # get_avg_strain |> 
@@ -133,7 +137,9 @@ mod_df_final <- mod_df |>
   mutate(block_pos = relevel(factor(block_pos), "T"),
          rush_pos = relevel(factor(rush_pos), "DE"),
          play_len = play_len / 10) |> 
-  left_join(select(plays, gameId:playId, possessionTeam:defensiveTeam))
+  left_join(select(plays, gameId:playId, down:defensiveTeam, absoluteYardlineNumber))
+
+
 
 
 library(lme4)  
@@ -229,24 +235,25 @@ fig_rankings_boot <- strain_eff |>
 
 # play_len
 strain_fit_new <- lmer(
-  avg_strain ~ (1 | nflId) + (1 | blocker_id) + (1 | possessionTeam) + (1 | defensiveTeam) + rush_pos + block_pos + n_blockers + n_rushers,
+  avg_strain ~ (1 | nflId) + (1 | blocker_id) + (1 | possessionTeam) + (1 | defensiveTeam) + 
+    rush_pos + block_pos + n_blockers + down * ydstogo + absoluteYardlineNumber,
   data = mod_df_final
 )
 
 summary(strain_fit_new)$coef
 broom.mixed::tidy(strain_fit_new, conf.int=TRUE)
 
-strain_fit_new |> 
-  ranef() |> 
-  pluck("defensiveTeam") |> 
-  as.data.frame() |> 
-  arrange(-`(Intercept)`)
-
-strain_fit_new |> 
-  ranef() |> 
-  pluck("possessionTeam") |> 
-  as.data.frame() |> 
-  arrange(`(Intercept)`)
+# strain_fit_new |> 
+#   ranef() |> 
+#   pluck("defensiveTeam") |> 
+#   as.data.frame() |> 
+#   arrange(-`(Intercept)`)
+# 
+# strain_fit_new |> 
+#   ranef() |> 
+#   pluck("possessionTeam") |> 
+#   as.data.frame() |> 
[…]

### 2023-07-14 scripts/14_play_context.eda.R

@@ -1,0 +1,51 @@
+library(tidyverse)
+library(here)
+
+pass_rush <- read_csv(here("data", "pass_rush.csv.gz"))
+
+get_avg_strain <- pass_rush |> 
+  filter(team == defensiveTeam, !is.infinite(strain)) |> 
+  group_by(gameId, playId, nflId) |> 
+  summarize(avg_strain = mean(strain, na.rm = TRUE)) |>
+  ungroup()
+
+games <- read_csv(here("data", "games.csv"))
+plays <- read_csv(here("data", "plays.csv"))
+
+
+
+#table(plays$down)
+
+get_avg_strain |> 
+  left_join(plays) |> 
+  mutate(down = factor(down)) |> 
+  ggplot(aes(down, avg_strain)) +
+  geom_boxplot()
+
+get_avg_strain |> 
+  left_join(plays) |> 
+  mutate(down = factor(down)) |> 
+  ggplot(aes(avg_strain, down)) +
+  ggridges::geom_density_ridges(scale = 1, quantile_lines = TRUE, quantiles = 0.5)
+
+
+get_avg_strain |> 
+  left_join(plays) |> 
+  ggplot(aes(yardsToGo, avg_strain, color = factor(down))) +
+  geom_point(alpha = 0.5) +
+  geom_smooth(method = "lm") +
+  facet_wrap(~ down, nrow = 1)
+
+get_avg_strain |> 
+  left_join(plays) |> 
+  ggplot(aes(absoluteYardlineNumber, avg_strain)) +
+  geom_point(alpha = 0.2) +
+  geom_smooth(method = "lm")
+
+
+plays |> 
+  count(yardlineNumber)
+
+
+
+

### 2023-07-14 paper/04-results.Rmd
saved bootstrap output

@@
 
-\begin{table}
-\caption{Fixed-effects coefficient estimates for the multilevel model for average STRAIN.
+<!-- \begin{table} -->
+<!-- \caption{Fixed-effects coefficient estimates for the multilevel model for average STRAIN.
 Note that the reference pass rusher level is defensive end (DE), denoted R:DE; and the reference pass blocker level is tackle (T), denoted B:T.
-\label{tab:fixedeff}}
-\centering
-\begin{tabular}{lrrrr}
-\hline
-& estimate & se & $t$-statistic & $p$-value \\ 
-\hline
-Intercept & 0.378 & 0.009 & 43.310 & 0.000 \\ 
-Length of play & -0.042 & 0.001 & -38.214 & 0.000 \\
-Number of blockers & -0.064 & 0.003 & -24.788 & 0.000 \\
-$I_{\text{\{R:DT\}}}$ & -0.081 & 0.008 & -10.340 & 0.000 \\
-$I_{\text{\{R:interior\}}}$ & 0.058 & 0.011 & 5.332 & 0.000 \\
-$I_{\text{\{R:NT\}}}$ & -0.106 & 0.012 & -9.174 & 0.000 \\
-$I_{\text{\{R:OLB\}}}$ & 0.062 & 0.008 & 7.695 & 0.000 \\
-$I_{\text{\{R:secondary\}}}$ & 0.145 & 0.012 & 12.328 & 0.000 \\
-$I_{\text{\{B:C\}}}$ & -0.023 & 0.006 & -3.625 & 0.000 \\
-$I_{\text{\{B:G\}}}$ & -0.011 & 0.005 & -2.290 & 0.023 \\ 
-$I_{\text{\{B:other\}}}$ & -0.013 & 0.007 & -1.911 & 0.056 \\
-\hline
-\end{tabular}
-\end{table}
+\label{tab:fixedeff}} -->
+<!-- \centering -->
+<!-- \begin{tabular}{lrrrr} -->
+<!-- \hline -->
+<!-- & estimate & se & $t$-statistic & $p$-value \\  -->
+<!-- \hline -->
+<!-- Intercept & 0.378 & 0.009 & 43.310 & 0.000 \\  -->
+<!-- Length of play & -0.042 & 0.001 & -38.214 & 0.000 \\ -->
+<!-- Number of blockers & -0.064 & 0.003 & -24.788 & 0.000 \\ -->
+<!-- $I_{\text{\{R:DT\}}}$ & -0.081 & 0.008 & -10.340 & 0.000 \\ -->
+<!-- $I_{\text{\{R:interior\}}}$ & 0.058 & 0.011 & 5.332 & 0.000 \\ -->
+<!-- $I_{\text{\{R:NT\}}}$ & -0.106 & 0.012 & -9.174 & 0.000 \\ -->
+<!-- $I_{\text{\{R:OLB\}}}$ & 0.062 & 0.008 & 7.695 & 0.000 \\ -->
+<!-- $I_{\text{\{R:secondary\}}}$ & 0.145 & 0.012 & 12.328 & 0.000 \\ -->
+<!-- $I_{\text{\{B:C\}}}$ & -0.023 & 0.006 & -3.625 & 0.000 \\ -->
+<!-- $I_{\text{\{B:G\}}}$ & -0.011 & 0.005 & -2.290 & 0.023 \\  -->
+<!-- $I_{\text{\{B:other\}}}$ & -0.013 & 0.007 & -1.911 & 0.056 \\ -->
+<!-- \hline -->
+<!-- \end{tabular} -->
+<!-- \end{table} -->
 
@@
 <!--   n\_blockers & -0.064 & 0.003 & -24.788 & 0.000 \\  -->
+
+\begin{table}
+\caption{Fixed-effects coefficient estimates for the multilevel model for average STRAIN.
+Note that the reference down level is first down, the reference pass rusher level is defensive end (DE), denoted R:DE; the reference pass blocker level is tackle (T), denoted B:T.
+\label{tab:fixedeff}}
+\centering
+\begin{tabular}{lrrrr}
+\hline
+& estimate & se & $t$-statistic & $p$-value \\ 
+\hline
+Intercept & 1.800 & 0.099 & 18.172 & 0.000 \\ 
+Number of blockers & -0.737 & 0.026 & -28.415 & 0.000 \\
+Yards to go & 0.049 & 0.003 & 14.901 & 0.000 \\
+Current yardline & 0.000 & 0.000 & 0.184 & 0.854 \\
+$I_{\text{\{2nd down\}}}$ & 0.514 & 0.029 & 17.744 & 0.000 \\
+$I_{\text{\{3rd down\}}}$ & 0.918 & 0.032 & 28.361 & 0.000 \\
+$I_{\text{\{4th down\}}}$ & 0.465 & 0.077 & 6.064 & 0.000 \\
+$I_{\text{\{2pt conversion\}}}$ & -0.138 & 0.180 & -0.770 & 0.441 \\
+$I_{\text{\{R:DT\}}}$ & -0.717 & 0.068 & -10.566 & 0.000 \\
+$I_{\text{\{R:interior\}}}$ & 0.561 & 0.100 & 5.614 & 0.000 \\
+$I_{\text{\{R:NT\}}}$ & -0.943 & 0.102 & -9.234 & 0.000 \\
+$I_{\text{\{R:OLB\}}}$ & 0.557 & 0.073 & 7.674 & 0.000 \\
+$I_{\text{\{R:secondary\}}}$ & 1.425 & 0.111 & 12.790 & 0.000 \\
+$I_{\text{\{B:C\}}}$ & -0.298 & 0.059 & -5.024 & 0.000 \\
+$I_{\text{\{B:G\}}}$ & -0.155 & 0.047 & -3.285 & 0.001 \\ 
+$I_{\text{\{B:other\}}}$ & -0.126 & 0.065 & -1.951 & 0.051 \\
+\hline
+\end{tabular}
+\end{table}
+
+```{r, eval=FALSE}
+                       Estimate Std. Error t value Pr(>|t|)
+(Intercept)               1.800      0.099  18.172    0.000
+rush_posDT               -0.717      0.068 -10.566    0.000
+rush_posinteriorLB        0.561      0.100   5.614    0.000
+rush_posNT               -0.943      0.102  -9.234    0.000
+rush_posOLB               0.557      0.073   7.674    0.000
+rush_possecondary         1.425      0.111  12.790    0.000
+block_posC               -0.298      0.059  -5.024    0.000
+block_posG               -0.155      0.047  -3.285    0.001
+block_posother           -0.126      0.065  -1.951    0.051
+n_blockers               -0.737      0.026 -28.415    0.000
+down0                    -0.138      0.180  -0.770    0.441
+down2                     0.514      0.029  17.744    0.000
+down3                     0.918      0.032  28.361    0.000
+down4                     0.465      0.077   6.064    0.000
+yardsToGo                 0.049      0.003  14.901    0.000
+absoluteYardlineNumber    0.000      0.000   0.184    0.854
+```
 

### 2023-07-14 scripts/08_fig_outcome_curves.R

@@ -20,7 +20,11 @@ hit <- pass_rush |>
          officialPosition %in% c("OLB", "NT", "DE", "DT"),
          frameId_snap_corrected <= 40) |> 
   group_by(officialPosition, frameId_snap_corrected) |> 
-  summarise(mn = mean(strain, na.rm = TRUE)) |> 
+  summarize(
+    strain = sum(strain),
+    n = n()
+  ) |>
+  mutate(mn = 10 * strain / n) |> 
   mutate(outcome = "Hit")
 
 sack <- pass_rush |> 
@@ -30,7 +34,11 @@ sack <- pass_rush |>
          officialPosition %in% c("OLB", "NT", "DE", "DT"),
          frameId_snap_corrected <= 40) |> 
   group_by(officialPosition, frameId_snap_corrected) |> 
-  summarise(mn = mean(strain, na.rm = TRUE)) |> 
+  summarize(
+    strain = sum(strain),
+    n = n()
+  ) |>
+  mutate(mn = 10 * strain / n) |> 
   mutate(outcome = "Sack")
 
 
@@ -41,7 +49,11 @@ hurry <- pass_rush |>
          officialPosition %in% c("OLB", "NT", "DE", "DT"),
          frameId_snap_corrected <= 40) |> 
   group_by(officialPosition, frameId_snap_corrected) |> 
-  summarise(mn = mean(strain, na.rm = TRUE)) |> 
+  summarize(
+    strain = sum(strain),
+    n = n()
+  ) |>
+  mutate(mn = 10 * strain / n) |> 
   mutate(outcome = "Hurry")
 
 none <- pass_rush |> 
@@ -52,7 +64,11 @@ none <- pass_rush |>
          officialPosition %in% c("OLB", "NT", "DE", "DT"),
          frameId_snap_corrected <= 40) |> 
   group_by(officialPosition, frameId_snap_corrected) |> 
-  summarise(mn = mean(strain, na.rm = TRUE)) |> 
+  summarize(
+    strain = sum(strain),
+    n = n()
+  ) |>
+  mutate(mn = 10 * strain / n) |> 
   mutate(outcome = "None")
 
 

### 2023-07-14 scripts/10_multilevel_mod.R

@@ -59,17 +59,23 @@ get_avg_strain <- pass_rush |>
 #   geom_histogram()
 
 # get qb and play info from nflfastR
-d <- nflfastR::load_pbp(2021)
+# d <- nflfastR::load_pbp(2021)
+# 
+# pass_plays <- plays |> 
+#   transmute(game_play_id = str_c(gameId, playId, sep = "_")) |> 
+#   pull()
+# 
+# get_qb <- d |> 
+#   select(gameId = old_game_id, playId = play_id, passer, half_seconds_remaining,
+#          yardline_100, down, ydstogo, posteam_timeouts_remaining, defteam_timeouts_remaining) |> 
+#   filter(str_c(gameId, playId, sep = "_") %in% pass_plays) |> 
+#   mutate(gameId = as.double(gameId))
 
-pass_plays <- plays |> 
-  transmute(game_play_id = str_c(gameId, playId, sep = "_")) |> 
-  pull()
 
-get_qb <- d |> 
-  select(gameId = old_game_id, playId = play_id, passer, half_seconds_remaining,
-         yardline_100, down, ydstogo, posteam_timeouts_remaining, defteam_timeouts_remaining) |> 
-  filter(str_c(gameId, playId, sep = "_") %in% pass_plays) |> 
-  mutate(gameId = as.double(gameId))
+get_qb <- pff |> 
+  filter(pff_role == "Pass") |> 
+  select(gameId, playId, passer = nflId)
+
 
 get_play_len <- pass_rush |> 
   distinct(gameId, playId, snap_frame, end_frame) |> 
@@ -137,7 +143,8 @@ mod_df_final <- mod_df |>
   mutate(block_pos = relevel(factor(block_pos), "T"),
          rush_pos = relevel(factor(rush_pos), "DE"),
          play_len = play_len / 10) |> 
-  left_join(select(plays, gameId:playId, down:defensiveTeam, absoluteYardlineNumber))
+  left_join(select(plays, gameId:playId, down:defensiveTeam, absoluteYardlineNumber)) |>
+  mutate(down = factor(down, levels = c(1, 0, 2:4)))
 
 
 
@@ -233,15 +240,18 @@ fig_rankings_boot <- strain_eff |>
 
 # revised model -----------------------------------------------------------
 
-# play_len
+# library(lmerTest)
+
 strain_fit_new <- lmer(
[…]

### 2023-07-15 paper/03-methods.Rmd
revision edits: strain curves and bootstrap results

@@
 
-Accordingly, the strain rate [@callister2018materials] of a material measures the change in its deformation with respect to time.
+Accordingly, the strain rate of a material measures the change in its deformation with respect to time.
 Mathematically, the strain rate of a material can be expressed as the derivative of its strain.
@@
 B_{b} &\sim N(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
-D_{d} &\sim N(\mu_D, \sigma^2_D), \text{ for } d = 1, \cdots, \text{ \# of teams}, \\
-O_{o} &\sim N(\mu_O, \sigma^2_O), \text{ for } o = 1, \cdots, \text{ \# of teams}.
+D_{d} &\sim N(\mu_D, \sigma^2_D), \text{ for } d = 1, \cdots, \text{ \# of defensive teams}, \\
+O_{o} &\sim N(\mu_O, \sigma^2_O), \text{ for } o = 1, \cdots, \text{ \# of offensive teams}.
 \end{aligned}
@@
 
-We consider a normal distribution to shrink the random intercepts for each player and team toward their respective group means, which is a useful property this problem's context since we do not observe the same number of plays for each player.
+We consider a normal distribution to shrink the random intercepts for each player and team toward their respective group means.
+This is a useful property since we do not observe the same number of plays for each player. 
 For the team effects, this provides us with the average defense and offense team-level effects on a pass rusher's STRAIN.
@@
 
-With regards to the fixed-effects about pass rusher $j$ in play $i$, we include a variety of features that likely contribute to variation in STRAIN.
+In order to provide a measure of uncertainty for our random effects, we use a bootstrapping strategy similar to the approach in @yurko2019nflwar.
+Specifically, we resample team drives within games, which preserves the fact that team schedules are fixed but allows for random variation in player usage since this is dependent on team decision making.
+By resampling plays within the same drive together, our sampling is generating realistic simulated data in comparison to sampling individual plays.
+For each bootstrapped dataset, we fit the aforementioned multilevel model to obtain a distribution of estimates for the considered player and team effects.
+
+As for the fixed-effects about pass rusher $j$ in play $i$, we include a variety of features that likely contribute to variation in STRAIN.
 First, we adjust for the position of both the pass rusher and nearest blocker to account for any positional effects.
 Table \ref{tab:pos} shows our positional categorization for the pass rush and pass block roles.
-These are encoded as indicator variables with defensive ends and tackles as the reference groups for the pass rushers and blockers respectively.
+These are encoded as indicator variables with defensive ends and tackles as the reference levels for the pass rushers and blockers, respectively.
 We also adjust for the number of pass blockers on the play, since teams may decide to employ a more protective scheme that could lower the observed STRAIN.
-Finally, we also adjust for the length of the play (i.e., the elapsed time between the ball snap and a quarterback event).
+Finally, we control for play-context covariates with respect to the offensive team.
+These include the current down (first, second, third, fourth, or two-point conversion), yards to go for a first down, and current yardline (i.e., distance from the possession team's goal line).
+[TODO: add explanation why play context is a proxy for time]
+
+<!-- Finally, we also adjust for the length of the play (i.e., the elapsed time between the ball snap and a quarterback event).
 We expect a longer display to display lower STRAIN since this would indicate a quarterback has more time to throw the ball.
 NEED TO THINK ABOUT THIS MORE....
+-->
 
-<!--Additionally, we obtain interval estimates for the player-level random intercepts via parametric bootstrapping of 10,000 samples with the `merTools` package [@knowlesmerTools; @gelman2006data].-->
+<!-- Additionally, we obtain interval estimates for the player-level random intercepts via parametric bootstrapping of 10,000 samples with the `merTools` package [@knowlesmerTools; @gelman2006data].
+-->
 
@@
 ```
-
-In order to provide a measure of uncertainty regarding about our random effects, we use a bootstrapping strategy similar to the approach in @yurko2019nflwar.
-Specifically, we resample team drives within games.
-This preserves the fact that team schedules are fixed, but allows for random variation in player usage since this is dependent on team decision making.
-By resampling plays within the same drive together, our sampling is generating realistic simulated data in comparison to sampling individual plays.
-For each resampled dataset, we fit the aforementioned multilevel model to generate a distribution of estimates for the considered player and team effects.

### 2023-07-15 paper/04-results.Rmd
revision edits: strain curves and bootstrap results

@@
 #| label: fig_pos_curves
-#| fig.cap: "STRAIN curves for different positions computed by averaging STRAIN values across all players of a particular position and all plays.
+#| fig.cap: "STRAIN curves for different positions.
 Edge rushers (OLB and DE) tend to generate more STRAIN than interior rushers (DT and NT).
@@
 
+<!-- 
+too wordy and confusing imo...
+For each frame, this curve is calculated by averaging across all players of a particular position (where position is defined by the NFL) and all plays that lasted at least as long as the particular frame of interest meaning that the sample size used to compute these curves diminishes for frames that are chronologically later than frames at the beginning of the play.
+-->
+
 Figure \ref{fig:fig_pos_curves} displays the average STRAIN by position for the first 40 frames (4 seconds) after the snap.
-For each frame, this curve is calculated by averaging across all players of a particular position (where position is defined by the NFL) and all plays that lasted at least as long as the particular frame of interest meaning that the sample size used to compute these curves diminishes for frames that are chronologically later than frames at the beginning of the play.
-
+For each position, the curve is based on the average of the observed STRAIN at each frame across all plays and players within the position.
 We observe a clear difference in STRAIN between edge rushers (OLBs and DEs) and interior linemen (DTs and NTs).
@@
 In addition to adjusting for opposition, our multilevel model allows us to provide some notion of uncertainty about the pass rusher's effect on the play level STRAIN.
-Figure \ref{fig:fig_rankings_boot} displays the mean varying intercepts for the top ten pass rushers for each position along with the corresponding 95% confidence intervals. 
+Figure \ref{fig:fig_rankings_boot} displays the varying intercept distributions for the top ten pass rushers for each position.
 We observe similar rankings based on the random intercepts in comparison to the average STRAIN results in Tables \ref{tab:edge} and \ref{tab:interior}, with Myles Garrett (DE), Aaron Donald (DT), Vita Vea (NT), and T.J.
 Watt (OLB) as the first-ranked pass rusher for their respective position.
-We observe different levels of variability across these top players, with only a subset of players with intervals strictly above zero.
+We observe different levels of variability across these top players, with only a subset of players with intercept ranges strictly above zero.
 This is not necessarily surprising given the limited sample of data.
@@
 #| label: fig_rankings_boot
-#| fig.cap: "Top 10 pass rushers for each position (DE, DT, NT, OLB) by mean varying intercept with 95\\% confidence intervals."
+#| fig.cap: "Distributions of player effects for top 10 pass rushers in each position (DE, DT, NT, OLB) by median varying intercept"
 #| fig.width: 6
-#| fig.height: 8
+#| fig.height: 5
 source(here("scripts", "10_multilevel_mod.R"))

### 2023-07-15 scripts/10_multilevel_mod.R

@@ -213,30 +213,30 @@ strain_fit |>
   mutate(icc = vcov / sum(vcov)) |> 
   select(grp, icc)
 
-strain_eff <- merTools::REsim(strain_fit, n.sims = 10000, seed = 101)
+# strain_eff <- merTools::REsim(strain_fit, n.sims = 10000, seed = 101)
 
 # strain_eff |> 
 #   merTools::plotREsim()
 
-fig_rankings_boot <- strain_eff |> 
-  as_tibble() |> 
-  filter(groupFctr == "nflId") |> 
-  mutate(nflId = as.double(groupID)) |> 
-  left_join(players) |> 
-  filter(nflId %in% filter(pass_rush_snaps, n_plays >= 100)$nflId) |> 
-  filter(officialPosition %in% c("DE", "OLB", "DT", "NT")) |> 
-  group_by(officialPosition) |> 
-  arrange(desc(mean)) |>
-  slice(1:10) |>
-  ggplot(aes(x = reorder(displayName, mean))) +
-  geom_point(aes(y = mean)) +
-  geom_errorbar(aes(ymin = mean - 2 * sd,
-                    ymax = mean + 2 * sd)) +
-  facet_wrap(~ officialPosition, ncol = 1, scales = "free_y") +
-  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
-  coord_flip() +
-  theme_light() +
-  labs(x = NULL, y = "intercept")
+# fig_rankings_boot_old <- strain_eff |> 
+#   as_tibble() |> 
+#   filter(groupFctr == "nflId") |> 
+#   mutate(nflId = as.double(groupID)) |> 
+#   left_join(players) |> 
+#   filter(nflId %in% filter(pass_rush_snaps, n_plays >= 100)$nflId) |> 
+#   filter(officialPosition %in% c("DE", "OLB", "DT", "NT")) |> 
+#   group_by(officialPosition) |> 
+#   arrange(desc(mean)) |>
+#   slice(1:10) |>
+#   ggplot(aes(x = reorder(displayName, mean))) +
+#   geom_point(aes(y = mean)) +
+#   geom_errorbar(aes(ymin = mean - 2 * sd,
+#                     ymax = mean + 2 * sd)) +
+#   facet_wrap(~ officialPosition, ncol = 1, scales = "free_y") +
+#   geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
+#   coord_flip() +
+#   theme_light() +
+#   labs(x = NULL, y = "intercept")
 
[…]

### 2023-07-15 paper/03-methods.Rmd
removed equation numbering (#14)

@@
 The strain for a material at time $t$ is defined as 
-\begin{equation}
+\begin{equation*}
 \varepsilon(t)= \frac{L(t) - L_{0}}{L_{0}}.
-\end{equation}
+\end{equation*}
 Notice that this measure is unitless due to being a ratio of two quantities having the same dimension.
@@
 That is,
-\begin{equation}
+\begin{equation*}
 \varepsilon'(t) = \frac{d \varepsilon}{dt} = \frac{v(t)}{L_{0}},
-\end{equation}
+\end{equation*}
 where $v(t)$ is the velocity at which the two points of interest within the material are moving away from or towards each other.
@@
 *   The distance between pass rusher $j$ and the quarterback at frame $t$ during play $i$ is 
-\begin{equation}
+\begin{equation*}
 s_{ij}(t) = \sqrt{(x_{ijt} - x^{QB}_{it})^2 + (y_{ijt} - y^{QB}_{it})^2}.
-\end{equation}
+\end{equation*}
 *   The velocity at which pass rusher $j$ is moving towards the quarterback at frame $t$ during play $i$ is 
-\begin{equation}
+\begin{equation*}
 v_{ij}(t) = s'_{ij}(t) = \frac{ds_{ij}(t)}{dt}.
-\end{equation}
+\end{equation*}
 *   The STRAIN for pass rusher $j$ at frame $t$ during play $i$ is 
-\begin{equation}
+\begin{equation*}
 \text{STRAIN}_{ij}(t) = \frac{v_{ij}(t)}{s_{ij}(t)}.
-\end{equation}
+\end{equation*}
 
@@
 Since we only observe the distance and velocity quantities discretely in increments of 10 frames/second, a point estimate for our proposed metric STRAIN for pass rusher $j$ at frame $t$ during play $i$ is
-\begin{equation}
+\begin{equation*}
 \widehat{\text{STRAIN}}_{ij}(t) = \cfrac{-\cfrac{s_{ij}(t) - s_{ij}(t - 1)}{0.1}}{s_{ij}(t)}.
-\end{equation}
+\end{equation*}
 Notice that this quantity increases in two ways: 1) the rate at which the rusher is moving towards the quarterback increases, and 2) the distance between the rusher and the quarterback decreases.
@@
 Formally, the average STRAIN, denoted by $\overline{\text{STRAIN}}$, for pass rusher $j$ involved in $n_j$ total plays across $T_j$ total frames is
-\begin{equation}
+\begin{equation*}
 \overline{\text{STRAIN}}_{j} = \frac{1}{T_j} \sum_{t=1}^{T_j} \widehat{\text{STRAIN}}_{j}(t).
-\end{equation}
+\end{equation*}
 This can be helpful for player evaluation, as we determine the most effective pass rushers based on their average STRAIN values in Section \ref{sec:strainbar}.
@@
 In addition to the average STRAIN over all frames played, we can also calculate pass rusher $j$'s observed average STRAIN on a single play $i$ consisting of $T_i$ total frames,
-\begin{equation}
+\begin{equation*}
 \overline{\text{STRAIN}}_{ij} = \frac{1}{T_i} \sum_{t=1}^{T_i} \widehat{\text{STRAIN}}_{ij}(t).
-\end{equation}
+\end{equation*}
 While this aggregated measure is a simple first step for pass rush evaluation, the observed average STRAIN on a single play is likely due to a convolution of factors.
@@
 Our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows.
-\begin{equation}
+\begin{equation*}
 \begin{aligned}
@@
 \end{aligned}
-\end{equation}
+\end{equation*}
 

### 2023-07-15 paper/00-main.Rmd
one minor typo
added new model results to paper

@@
   In American football, a pass rush is an attempt by the defensive team to disrupt the offense and prevent the quarterback (QB) from completing a pass.
-Existing metrics for assessing pass rush performance are either discrete-time quantities or based on subjective judgement.
+Existing metrics for assessing pass rush performance are either discrete-time quantities or based on subjective judgment.
 Using player tracking data, we propose STRAIN, a novel metric for evaluating pass rushers in the National Football League (NFL) at the continuous-time within-play level.
@@
 bibliography: bibliography.bib
+biblio-style: chicago
 output: rticles::asa_article
@@
   warning = FALSE,
-  fig.align = "center",
-  fig.pos = "p"
-  #cache = TRUE
+  fig.align = "center"
+  # fig.pos = "p",
+  # cache = TRUE
 )
@@
 \newpage
+\spacingset{1.45}
 

### 2023-07-15 paper/04-results.Rmd
added new model results to paper

@@
 In particular, we look at how well our metric correlates with a simple measure of pressure rate, defined as the total hits, sacks and hurries per snap.
-Figure \ref{fig:fig_cor_pressure} is a scatterplot of average STRAIN and pressure rate over the course the provided eight-game sample, which reveals a fairly strong correlation ($r =0.625$) between the quantities.
+Figure \ref{fig:fig_cor_pressure} is a scatterplot of average STRAIN and pressure rate over the course the provided eight-game sample, which reveals a fairly strong correlation ($r =0.6255$) between the quantities.
 Hence, defenders with high STRAIN values also tend to generate more pressure toward the quarterback.
-Furthermore, the average STRAIN for the first four weeks of the 2021 season is more predictive of the last four weeks' pressure rate ($r=0.322$) than the first four weeks' pressure rate ($r=0.097$), as illustrated in Figure \ref{fig:fig_predictability}.
+Furthermore, the average STRAIN for the first four weeks of the 2021 season is more predictive of the last four weeks' pressure rate ($r=0.3217$) than the first four weeks' pressure rate ($r=0.0965$), as illustrated in Figure \ref{fig:fig_predictability}.
 
 We also analyze the stability of STRAIN over time by comparing the average STRAIN across all frames played for the first and last four weeks of the 2021 NFL regular season.
-It is apparent from Figure \ref{fig:fig_stability} that there is a strong positive correlation $(r = 0.854)$ for this relationship.
+It is apparent from Figure \ref{fig:fig_stability} that there is a strong positive correlation $(r = 0.8545)$ for this relationship.
 This means that STRAIN is a highly stable football metric over the provided eight-week time window, and pass rushers appear to carry their STRAIN values with them from week to week.
@@
 #| fig.cap: "Relationship between average STRAIN and pressure rate (total hits, sacks, and hurries per snap) over the first eight weeks of the 2021 NFL season.
-There is a moderately strong association between average STRAIN and pressure rate ($r =0.625$).
+There is a moderately strong association between average STRAIN and pressure rate ($r =0.6255$).
 Results shown here are for pass rushers with at least 100 snaps played over the eight-week data."
@@
 #| fig.cap: "Relationships between pressure rate for the last 4 weeks of the 2021 NFL season and first four weeks' pressure rate (left) and average STRAIN (right).
-STRAIN is more predictive ($r=0.322$) of future pressure rate than previous pressure rate itself ($r=0.097$).
+STRAIN is more predictive ($r=0.3217$) of future pressure rate than previous pressure rate itself ($r=0.0965$).
 Results shown here are for pass rushers with at least 100 snaps played over the eight-week data."
@@
 #| fig.cap: "Relationship between average STRAIN for the last and first four weeks of the 2021 NFL season.
-A strong linear correlation ($r = 0.854$) demonstrates that STRAIN is a highly stable metric over time.
+A strong linear correlation ($r = 0.8545$) demonstrates that STRAIN is a highly stable metric over time.
 Results shown here are for pass rushers with at least 100 snaps played over the eight-week data."
@@
 \hline
-Intercept & 1.800 & 0.099 & 18.172 & 0.000 \\ 
-Number of blockers & -0.737 & 0.026 & -28.415 & 0.000 \\
-Yards to go & 0.049 & 0.003 & 14.901 & 0.000 \\
-Current yardline & 0.000 & 0.000 & 0.184 & 0.854 \\
-$I_{\text{\{2nd down\}}}$ & 0.514 & 0.029 & 17.744 & 0.000 \\
-$I_{\text{\{3rd down\}}}$ & 0.918 & 0.032 & 28.361 & 0.000 \\
-$I_{\text{\{4th down\}}}$ & 0.465 & 0.077 & 6.064 & 0.000 \\
-$I_{\text{\{2pt conversion\}}}$ & -0.138 & 0.180 & -0.770 & 0.441 \\
-$I_{\text{\{R:DT\}}}$ & -0.717 & 0.068 & -10.566 & 0.000 \\
-$I_{\text{\{R:interior\}}}$ & 0.561 & 0.100 & 5.614 & 0.000 \\
-$I_{\text{\{R:NT\}}}$ & -0.943 & 0.102 & -9.234 & 0.000 \\
-$I_{\text{\{R:OLB\}}}$ & 0.557 & 0.073 & 7.674 & 0.000 \\
-$I_{\text{\{R:secondary\}}}$ & 1.425 & 0.111 & 12.790 & 0.000 \\
-$I_{\text{\{B:C\}}}$ & -0.298 & 0.059 & -5.024 & 0.000 \\
-$I_{\text{\{B:G\}}}$ & -0.155 & 0.047 & -3.285 & 0.001 \\ 
-$I_{\text{\{B:other\}}}$ & -0.126 & 0.065 & -1.951 & 0.051 \\
+Intercept & 1.8004 & 0.0991 & 18.1723 & 0.0000 \\ 
+Number of blockers & -0.7366 & 0.0259 & -28.4153 & 0.0000 \\
+Yards to go & 0.0491 & 0.0033 & 14.9013 & 0.0000 \\
+Current yardline & 0.0001 & 0.0005 & 0.1844 & 0.8537 \\
+$I_{\text{\{2nd down\}}}$ & 0.5143 & 0.0290 & 17.7439 & 0.0000 \\
+$I_{\text{\{3rd down\}}}$ & 0.9176 & 0.0324 & 28.3613 & 0.0000 \\
+$I_{\text{\{4th down\}}}$ & 0.4649 & 0.0767 & 6.0637 & 0.0000 \\
+$I_{\text{\{2pt conversion\}}}$ & -0.1384 & 0.1797 & -0.7699 & 0.4413 \\
+$I_{\text{\{R:DT\}}}$ & -0.7166 & 0.0678 & -10.5658 & 0.0000 \\
+$I_{\text{\{R:interior\}}}$ & 0.5611 & 0.0999 & 5.6143 & 0.0000 \\
+$I_{\text{\{R:NT\}}}$ & -0.9426 & 0.1021 & -9.2344 & 0.0000 \\
+$I_{\text{\{R:OLB\}}}$ & 0.5574 & 0.0726 & 7.6742 & 0.0000 \\
+$I_{\text{\{R:secondary\}}}$ & 1.4252 & 0.1114 & 12.7899 & 0.0000 \\
+$I_{\text{\{B:C\}}}$ & -0.2983 & 0.0594 & -5.0237 & 0.0000 \\
+$I_{\text{\{B:G\}}}$ & -0.1546 & 0.0471 & -3.2847 & 0.0000 \\ 
+$I_{\text{\{B:other\}}}$ & -0.1260 & 0.0646 & -1.9508 & 0.0514 \\
 \hline
@@
 
-```{r, eval=FALSE}
-                       Estimate Std. Error t value Pr(>|t|)
-(Intercept)               1.800      0.099  18.172    0.000
-rush_posDT               -0.717      0.068 -10.566    0.000
-rush_posinteriorLB        0.561      0.100   5.614    0.000
-rush_posNT               -0.943      0.102  -9.234    0.000
-rush_posOLB               0.557      0.073   7.674    0.000
-rush_possecondary         1.425      0.111  12.790    0.000
-block_posC               -0.298      0.059  -5.024    0.000
-block_posG               -0.155      0.047  -3.285    0.001
-block_posother           -0.126      0.065  -1.951    0.051
-n_blockers               -0.737      0.026 -28.415    0.000
-down0                    -0.138      0.180  -0.770    0.441
-down2                     0.514      0.029  17.744    0.000
-down3                     0.918      0.032  28.361    0.000
-down4                     0.465      0.077   6.064    0.000
-yardsToGo                 0.049      0.003  14.901    0.000
-absoluteYardlineNumber    0.000      0.000   0.184    0.854
-```
+<!-- \begin{table}[ht] -->
+<!-- \caption{Intraclass correlation coefficients for the multilevel model for average STRAIN. \label{tab:icc}} -->
+<!-- \centering -->
+<!-- \begin{tabular}{lr} -->
+<!-- \hline -->
+<!-- Group & ICC \\  -->
+<!-- \hline -->
+<!-- Pass rusher & 0.057 \\  -->
+<!-- Pass blocker & 0.012 \\  -->
+<!-- Quarterback & 0.023 \\  -->
+<!-- Residual & 0.907 \\  -->
+<!-- \hline -->
+<!-- \end{tabular} -->
+<!-- \end{table} -->
 
@@
 \hline
-Pass rusher & 0.057 \\ 
-Pass blocker & 0.012 \\ 
-Quarterback & 0.023 \\ 
-Residual & 0.907 \\ 
+Pass rusher & 0.0365 \\ 
+Pass blocker & 0.0098 \\ 
+Defensive team & 0.0134 \\ 
+Offensive team & 0.0143 \\
+Residual & 0.9260 \\ 
 \hline
@@
 #| label: fig_rankings_boot
-#| fig.cap: "Distributions of player effects for top 10 pass rushers in each position (DE, DT, NT, OLB) by median varying intercept"
-#| fig.width: 6
+#| fig.cap: "Distributions of player effects for top 10 pass rushers in each position (DE, DT, NT, OLB) by median varying intercept."
+#| fig.width: 6.3
 #| fig.height: 5

### 2023-07-15 paper/supplemental.Rmd
added new model results to paper

@@
 
-## Old Multilevel Model for Play-Level STRAIN {#sec:multilevel}
+# Old Multilevel Model for Play-Level STRAIN {#sec:multilevel}
 
@@
 We consider a normal distribution to obtain the random intercepts for each player toward their respective position group means.
+
+# Team effects justification
+
+We perform leave-one-week-out cross-validation (LOWO CV)
+
+We evaluate the model performance based on mean squared error (MSE)
+
+```{r}
+source(here("scripts", "10_multilevel_mod.R"))
+fig_loocv_team_eff
+```

### 2023-07-15 scripts/10_multilevel_mod.R

@@ -346,7 +346,7 @@ mutate(strain_loocv_initial_rmse, team_eff = "No") |>
   theme_light()
 
 
-strain_loocv_initial_df |>
+fig_loocv_team_eff <- strain_loocv_initial_df |>
   mutate(resid = obs - pred,
          sq_err = resid ^ 2) |>
   group_by(week) |>

### 2023-07-15 scripts/15_fig_features_combined.R
planning for supplemental and figure adjustments

@@ -1,0 +1,73 @@
+library(here)
+source(here("scripts", "01_fig_field.R"))
+pff <- read_csv(here("data", "pffScoutingData.csv"))
+play_unsucess <- read_csv(here("data", "play_unsucess.csv.gz"))
+
+
+crosby_sack <- play_lv_den |> 
+  ungroup() |> 
+  filter(nflId == 47889) |>
+  select(frameId_snap_corrected, distance_from_QB, slope_distance_from_QB, strain)
+
+feat_sack <- play_lv_den |> 
+  left_join(pff) |>
+  filter(pff_role == "Pass Rush", nflId != 47889) |> 
+  select(name = displayName, frameId_snap_corrected, value = strain) |> 
+  mutate(name = str_c("strain_", name)) |> 
+  bind_rows(pivot_longer(crosby_sack, !frameId_snap_corrected)) |> 
+  mutate(play = "Successful pass rush")
+
+
+
+crosby_unsuccess <- play_unsucess |> 
+  filter(nflId == 47889) |>
+  select(frameId_snap_corrected, distance_from_QB, slope_distance_from_QB, strain)
+
+feat_unsuccess <- play_unsucess |> 
+  left_join(pff) |>
+  filter(pff_role == "Pass Rush", nflId != 47889) |> 
+  select(name = displayName, frameId_snap_corrected, value = strain) |> 
+  mutate(name = str_c("strain_", name)) |> 
+  bind_rows(pivot_longer(crosby_unsuccess, !frameId_snap_corrected)) |> 
+  mutate(play = "Unsuccessful pass rush")
+
+
+feat_sack |> 
+  bind_rows(feat_unsuccess) |> 
+  mutate(
+    name = case_when(
+      name == "distance_from_QB" ~ "Distance from QB (yards)",
+      name == "slope_distance_from_QB" ~ "Velocity (yards/second)",
+      name == "strain" ~ "STRAIN (1/second)",
+      TRUE ~ as.character(name)
+    ),
+    name_fct = ifelse(str_detect(name, "strain_"), "STRAIN of other rushers", name),
+    name_fct = factor(
+      name_fct,
+      levels = c(
+        "Distance from QB (yards)",
+        "Velocity (yards/second)",
+        "STRAIN (1/second)",
+        "STRAIN of other rushers"
+      )
+    )
+  ) |> 
+  ggplot() +
+  geom_line(aes(frameId_snap_corrected, value, color = name_fct, alpha = name_fct, group = name), linewidth = 1.2) +
[…]

### 2023-07-16 paper/05-discussion.Rmd
more discussion (#14)

@@
 # Discussion {#sec:discussion}
+
 In this work, we have proposed STRAIN---a simple and interpretable statistic for evaluating pass rushers---with higher values corresponding to greater pass rushing ability.
-We demonstrated that a pass rusher's average STRAIN is both stable and more predictive of future pressure events than using the player's previously observed pressure events.
+STRAIN is a model-free metric which can easily be integrated into any data pipeline without much computational cost.
+Visualizations of STRAIN can be useful and intuitive for coaches and broadcasters in various aspects ranging from gameplan preparation to in-game real-time play analysis.
+We demonstrate that a pass rusher's average STRAIN is both stable and more predictive of future pressure events than using the player's previously observed pressure events.
 This is analogous to the use of exit velocity in baseball, a predictive measurement that avoids the noisy nature of observed outcomes, emphasizing how the opportunity of working with player tracking data can lead to the development of insightful metrics in American football.

### 2023-07-16 paper/02-data.Rmd
shitton of work on supplement and figures

@@
 
-Table \ref{tab:track} displays a tracking data example for a play from the 2021 NFL regular season week six matchup between the Las Vegas Raiders and Denver Broncos, which ends with Broncos QB Teddy Bridgewater getting sacked by Raiders DE Maxx Crosby[^crosbysack].
+Table \ref{tab:track} displays a tracking data example for a [play](https://www.raiders.com/video/de-maxx-crosby-sacks-qb-teddy-bridgewater-for-a-loss-of-6-yards-nfl) from the 2021 NFL regular season week six matchup between the Las Vegas Raiders and Denver Broncos, which ends with Broncos QB Teddy Bridgewater getting sacked by Raiders DE Maxx Crosby[^crosbysack].
 In addition, Figure \ref{fig:fig_field_raw} presents the locations of every Las Vegas (in black) and Denver (in orange) player on the field from this play at 1, 2, 3 and 4 seconds after the ball snap, with Maxx Crosby highlighted in blue.
 
-[^crosbysack]: Game footage of this play is available at https://www.raiders.com/video/de-maxx-crosby-sacks-qb-teddy-bridgewater-for-a-loss-of-6-yards-nfl
+<!-- [^crosbysack]: Game footage of this play is available at https://www.raiders.com/video/de-maxx-crosby-sacks-qb-teddy-bridgewater-for-a-loss-of-6-yards-nfl -->
 

### 2023-07-16 paper/04-results.Rmd
shitton of work on supplement and figures

@@
 ```
+
+```{r}
+#| label: fig_features_combined
+#| fig.cap: "Changes in STRAIN, velocity, and distance from the QB for Maxx Crosby over the course of a successful pass rush play that results in a sack (left) and an unsuccessful pass rush play which results in a long pass attempt (right).
+The STRAIN for other pass rushers throughout each play is also displayed."
+#| fig.width: 6
+#| fig.height: 3.6
+source(here("scripts", "15_fig_features_combined.R"))
+fig_features_combined_bw
+```
+
+Figure \ref{fig:fig_features_combined}
 

### 2023-07-16 paper/supplement.Rmd
shitton of work on supplement and figures

---
title: |
  | Supplementary materials for
  | "Here Comes the STRAIN: Analyzing Defensive Pass Rush in American Football with Player Tracking Data"
blinded: 0
bibliography: bibliography.bib
biblio-style: chicago
output: rticles::asa_article
header-includes:
  \date{}
---

\spacingset{1.45}

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
library(here)
library(tidyverse)
```

# Visualizations of Player Tracking Data

Figure \ref{fig:fig_field} shows snapshots of a play from the week six game between the Las Vegas Raiders and Denver Broncos in the 2021 NFL regular season.
The play outcome is a sack made by Raiders defensive end Maxx Crosby on Broncos quarterback Teddy Bridgewater.
The locations of every Las Vegas (in black, on defense) and Denver (in orange, on offense) player on the field at 1, 2, 3, and 4 seconds after the ball snap is displayed, with Crosby highlighted in blue.
Note that the point size for each Las Vegas defender corresponds to their estimated STRAIN as the play progresses.
This play is represented by Table [INSERT NUMBER - tracking data table] and Figure [INSERT NUMBER - line graph] in the paper.

```{r}
#| label: fig_field
#| fig.cap: "A display of the player tracking data for a play during the Las Vegas Raiders (defense, in black) versus Denver Broncos (offense, in orange) NFL game on October 17, 2021. Raiders DE Maxx Crosby is highlighted in blue. For each Raiders defender, the point size indicates their individual STRAIN value, with larger points suggesting larger STRAIN. Snapshots are captured at 1, 2, 3, and 4 seconds after the ball snap."
#| fig.width: 6.6
#| fig.height: 4.4
source(here("scripts", "01_fig_field.R"))
fig_field
```

# Distribution of number of pass rushers

Figure \ref{fig:pass_rushers_count} displays the distribution of the number of observed pass rushers involved in a play.
It ranges from 1 to 8, with 4 pass rushers (i.e., 4-man rush formation) as the most common value.

```{r}
#| label: pass_rushers_count
#| fig.cap: "Distribution of the number of pass rushers on passing plays."
#| fig.width: 4
#| fig.height: 3
source(here("scripts", "12_fig_pass_rushers_count.R"))
pass_rushers_count
```

# Pass rusher rankings based on average STRAIN

Since STRAIN is observed continuously for every play in our data, this allows us to aggregate across all frames played and compute the average STRAIN for NFL pass rushers over the course of the eight-week sample size, as discussed in Section \ref{sec:defn}.
Based on the clearly distinct patterns for different positions as previously observed, we evaluate interior pass rushers (NTs and DTs) separately from edge rushers (OLBs and DEs).
Tables \ref{tab:edge} and \ref{tab:interior} are leaderboards for the NFL's best edge and interior rushers (with at least 100 plays) rated by the average STRAIN across all frames for the first eight weeks of the 2021 regular season.
The tables also consist of the total number of hits, hurries, and sacks (determined from PFF scouting data) for each defender.

<!-- Our results show many true positives with no surprises. -->
<!-- Rashan Gary, a young rusher with huge potential, ranks first among all pass rushers according to our metric over the period of time observed. -->

[TODO: rewrite this, move some of this to analyze the ggridges figure]

Our results are mostly consistent with conventional rankings of rushers.
Notably, Myles Garrett and TJ Watt are widely recognized as top-tier edge rushers and both show up in our top edge rusher list; whereas Aaron Donald, who is undoubtedly the best interior defender in football, appears at the top of our interior rusher rankings.
Moreover, our leaderboards largely match the rankings of experts in the field.
For instance, there is considerable overlap between our lists and PFF's edge [@Monson2022nfl] and interior [@Linsey2022nfl] rusher rankings released after the 2021 season.
This ultimately lends credibility to our proposed metric STRAIN as a measure of pass rushing effectiveness.

\begin{table}
\caption{Top 10 edge rushers (with at least 100 snaps played) according to the average STRAIN across all frames played. \label{tab:edge}}
\centering
\begin{tabular}{rlllrrrrr}
\hline
Rank & Player & Team & Position & Snaps & Hits & Hurries & Sacks & $\overline{\text{STRAIN}}$ \\
\hline
1 & Rashan Gary & GB & OLB & 176 & 10 & 25 & 5 & 2.82 \\ 
2 & Leonard Floyd & LA & OLB & 185 & 2 & 25 & 8 & 2.80 \\ 
3 & Justin Houston & BAL & OLB & 132 & 8 & 8 & 4 & 2.78 \\ 
4 & Myles Garrett & CLE & DE & 197 & 9 & 29 & 12 & 2.75 \\ 
5 & Von Miller & DEN & OLB & 145 & 4 & 21 & 5 & 2.75 \\ 
6 & T.J. Watt & PIT & OLB & 147 & 6 & 9 & 8 & 2.71 \\ 
7 & Yannick Ngakoue & LV & DE & 175 & 6 & 20 & 4 & 2.70 \\ 
8 & Alex Highsmith & PIT & OLB & 129 & 4 & 7 & 2 & 2.65 \\ 
9 & Preston Smith & GB & OLB & 124 & 4 & 8 & 2 & 2.61 \\ 
10 & Randy Gregory & DAL & DE & 134 & 7 & 19 & 5 & 2.58 \\ 
\hline
\end{tabular}
\end{table}

<!-- 11 & Joey Bosa & LAC & OLB & 160 & 5 & 21 & 4 & 2.58 \\  -->
<!-- 12 & Darrell Taylor & SEA & DE & 107 & 5 & 9 & 3 & 2.57 \\  -->
<!-- 13 & Josh Sweat & PHI & DE & 159 & 4 & 14 & 5 & 2.57 \\  -->
<!-- 14 & Maxx Crosby & LV & DE & 198 & 12 & 30 & 7 & 2.56 \\  -->
<!-- 15 & Markus Golden & ARI & OLB & 164 & 5 & 14 & 5 & 2.50 \\  -->

\begin{table}
\caption{Top 10 interior rushers (with at least 100 snaps played) according to the average STRAIN across all frames played. \label{tab:interior}}
\centering
\begin{tabular}{rlllrrrrr}
\hline
Rank & Player & Team & Position & Snaps & Hits & Hurries & Sacks & $\overline{\text{STRAIN}}$ \\
\hline
1 & Aaron Donald & LA & DT & 239 & 8 & 24 & 6 & 1.67 \\ 
2 & Solomon Thomas & LV & DT & 115 & 7 & 11 & 3 & 1.51 \\ 
3 & Quinton Jefferson & LV & DT & 144 & 6 & 8 & 3 & 1.46 \\ 
4 & Chris Jones & KC & DT & 139 & 3 & 18 & 3 & 1.42 \\ 
5 & DeForest Buckner & IND & DT & 198 & 4 & 18 & 4 & 1.26 \\ 
6 & Cameron Heyward & PIT & DT & 188 & 2 & 22 & 3 & 1.25 \\ 
7 & Javon Hargrave & PHI & DT & 156 & 6 & 15 & 5 & 1.24 \\ 
8 & Jerry Tillery & LAC & DT & 171 & 4 & 7 & 3 & 1.16 \\ 
9 & Ed Oliver & BUF & DT & 133 & 4 & 12 & 1 & 1.15 \\ 
10 & Osa Odighizuwa & DAL & DT & 162 & 3 & 18 & 3 & 1.13 \\ 
\hline
\end{tabular}
\end{table}

<!-- 11 & Greg Gaines & LA & NT & 111 & 2 & 13 & 2 & 1.11 \\  -->
<!-- 12 & Leonard Williams & NYG & DT & 226 & 4 & 14 & 6 & 1.03 \\  -->
<!-- 13 & Christian Barmore & NE & DT & 166 & 5 & 17 & 1 & 1.02 \\  -->
<!-- 14 & Vita Vea & TB & NT & 184 & 6 & 12 & 1 & 1.01 \\  -->
<!-- 15 & B.J. Hill & CIN & DT & 123 & 3 & 4 & 3 & 0.96 \\  -->

# Comparison of multilevel models for play-level STRAIN

## Model without team effects

We additionally consider a multilevel model to evaluate pass rushers' impact on the average STRAIN observed in a play, while accounting for their pass blocker, QB, and play-level attributes.
The average STRAIN by pass rusher $j$ on play $i$ is modeled as follows.
\begin{equation*}
\begin{aligned}
\overline{\text{STRAIN}}_{ij} &\sim N(R_{j[i]} + B_{b[ij]} + Q_{q[i]} + \mathbf{x_{ij}} \boldsymbol{\beta}, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
R_{j} &\sim N(\mu_R, \sigma^2_R), \text{ for } j = 1, \cdots, \text{ \# of pass rushers}, \\
B_{b} &\sim N(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
Q_{q} &\sim N(\mu_Q, \sigma^2_Q), \text{ for } q = 1, \cdots, \text{ \# of QBs}.
\end{aligned}
\end{equation*}
This version of our model contains random intercepts for three player groups: (1) pass rushers as $R$, (2) nearest pass blockers as $B$, and (3) QBs as $Q$.
As before, the coefficients $\boldsymbol \beta$ associated with covariate vector $\mathbf{x_{ij}}$ for pass rusher $j$ in play $i$ are estimated as fixed-effects. 
The variables included as fixed effects are the same as those from the model in the paper: position of pass rusher and blocker, number of blockers in a play, and play-context information (down, yards to go, and current yardline).
We consider a normal distribution to obtain the random intercepts for each player toward their respective position group means.

## Model comparison

We perform leave-one-week-out cross-validation (LOWO CV) to compare the multilevel models with and without team effects for play-level STRAIN.
For example, we train each model using data from weeks one through seven, and perform evaluation for week eight.
This procedure is repeated for every single holdout week, and evaluate the model performance based on mean squared error (MSE).
Figure \ref{fig:fig_loocv_team_eff} shows the week-by-week holdout MSE for the aforementioned multilevel models.
We clearly see that a smaller MSE is achieved for each left-out week when accounting for team effects.
Furthermore, the model with team effects yields a lower overall holdout MSE ($4.1882$) than the one without random intercepts for teams ($4.4138$).

```{r}
#| label: fig_loocv_team_eff
#| fig.cap: "Comparison of leave-one-week-out cross-validation MSE with $\\pm 2$ standard error bounds for multilevel models with and without team effects."
#| fig.width: 6
#| fig.height: 3
source(here("scripts", "10_multilevel_mod.R"))
fig_loocv_team_eff
```

### 2023-07-16 scripts/10_multilevel_mod.R

@@ -274,99 +274,103 @@ strain_fit_new |>
 
 # lowocv ------------------------------------------------------------------
 
-mod_df_final <- mod_df_final |> 
-  left_join(select(games, gameId, week))
-
-strain_loocv_initial <- function(w) {
-  
-  train <- filter(mod_df_final, week != w)
-  test <- filter(mod_df_final, week == w)
-  
-  fit <- lmer(
-    avg_strain ~ (1 | nflId) + (1 | blocker_id) + (1 | passer) + play_len + rush_pos + block_pos + n_blockers,
-    data = train
-  )
-  
-  out <- tibble(
-    pred = predict(fit, newdata = test, allow.new.levels = TRUE),
-    obs = pull(test, avg_strain),
-    week = w
-  )
-  return(out)
-}
-
-strain_loocv_initial_df <- map(1:8, strain_loocv_initial) |> 
-  list_rbind()
-
-library(yardstick)
-strain_loocv_initial_rmse <- strain_loocv_initial_df |>
-  group_by(week) |> 
-  rmse(obs, pred)
-
-
-strain_loocv_new <- function(w) {
-  
-  train <- filter(mod_df_final, week != w)
-  test <- filter(mod_df_final, week == w)
-  
-  fit <- lmer(
-    avg_strain ~ (1 | nflId) + (1 | blocker_id) + (1 | possessionTeam) + (1 | defensiveTeam) + 
-      rush_pos + block_pos + n_blockers + down + yardsToGo + absoluteYardlineNumber,
-    data = mod_df_final
-  )
-  
-  out <- tibble(
-    pred = predict(fit, newdata = test, allow.new.levels = TRUE),
-    obs = pull(test, avg_strain),
-    week = w
-  )
-  return(out)
-}
-
-strain_loocv_new_df <- map(1:8, strain_loocv_new) |> 
-  list_rbind()
-
-strain_loocv_new_rmse <- strain_loocv_new_df |>
-  group_by(week) |> 
-  rmse(obs, pred)
-
-mean(strain_loocv_initial_rmse$.estimate ^ 2)
-mean(strain_loocv_new_rmse$.estimate ^ 2)
-
-
-mutate(strain_loocv_initial_rmse, team_eff = "No") |>
-  bind_rows(mutate(strain_loocv_new_rmse, team_eff = "Yes")) |> 
-  ggplot(aes(week, .estimate, color = team_eff)) +
-  geom_point() +
-  geom_line() +
[…]

### 2023-07-16 scripts/12_fig_pass_rushers_count.R

@@ -13,7 +13,7 @@ pass_rushers_count <- pff_full |>
   count(gameId, playId) |> 
   count(n) |> 
   ggplot(aes(n, nn)) +
-  geom_col() +
+  geom_col(fill = "gray") +
   scale_x_continuous(breaks = 1:8) +
   labs(x = "Number of pass rushers",
        y = "Count")

### 2023-07-16 scripts/15_fig_features_combined.R

@@ -32,23 +32,23 @@ feat_unsuccess <- play_unsucess |>
   mutate(play = "Unsuccessful pass rush")
 
 
-feat_sack |> 
+fig_features_combined <- feat_sack |> 
   bind_rows(feat_unsuccess) |> 
   mutate(
     name = case_when(
-      name == "distance_from_QB" ~ "Distance from QB (yards)",
-      name == "slope_distance_from_QB" ~ "Velocity (yards/second)",
-      name == "strain" ~ "STRAIN (1/second)",
+      name == "distance_from_QB" ~ "Distance from QB\n(yards)",
+      name == "slope_distance_from_QB" ~ "Velocity\n(yards/second)",
+      name == "strain" ~ "STRAIN\n(1/second)",
       TRUE ~ as.character(name)
     ),
-    name_fct = ifelse(str_detect(name, "strain_"), "STRAIN of other rushers", name),
+    name_fct = ifelse(str_detect(name, "strain_"), "STRAIN of\nother rushers", name),
     name_fct = factor(
       name_fct,
       levels = c(
-        "Distance from QB (yards)",
-        "Velocity (yards/second)",
-        "STRAIN (1/second)",
-        "STRAIN of other rushers"
+        "Distance from QB\n(yards)",
+        "Velocity\n(yards/second)",
+        "STRAIN\n(1/second)",
+        "STRAIN of\nother rushers"
       )
     )
   ) |> 
@@ -69,5 +69,54 @@ feat_sack |>
         axis.text = element_text(size = rel(0.8)),
         strip.text = element_text(size = rel(1)),
         legend.position = "bottom",
-        legend.text = element_text(size = rel(0.8)),
-        legend.title = element_text(size = rel(1)))
+        legend.text = element_text(size = rel(0.9)),
+        panel.spacing = unit(2, "lines"),
+        legend.box.spacing = unit(0.25, "lines"))
+
+
+
+
+
+
+fig_features_combined_bw <- feat_sack |> 
+  bind_rows(feat_unsuccess) |> 
+  mutate(
+    name = case_when(
+      name == "distance_from_QB" ~ "Distance from QB\n(yards)",
+      name == "slope_distance_from_QB" ~ "Velocity\n(yards/second)",
+      name == "strain" ~ "STRAIN\n(1/second)",
+      TRUE ~ as.character(name)
+    ),
+    name_fct = ifelse(str_detect(name, "strain_"), "STRAIN of\nother rushers", name),
[…]

### 2023-07-16 paper/04-results.Rmd
removed footnotes (#14)

--- 2023-07-15 paper/04-results.Rmd
@@
 In contrast, Figure \ref{fig:fig_crosby_curves_unsucess} shows the feature curves for an unsuccessful pass rush attempt by the Raiders defense from the same game.
-In this play, Broncos QB Teddy Bridgewater is well-protected by the offensive line and is able to release a long pass[^unsuccess] to a receiver.
+In this [play](https://www.nfl.com/videos/teddy-bridgewater-courtland-sutton-barely-miss-out-on-chance-at-62-yard-td-bomb), Broncos QB Teddy Bridgewater is well-protected by the offensive line and is able to release a long pass to a receiver.
 We see that the STRAIN generated by Crosby is relatively small over the course of this play, compared to the previous play (as shown in Figure \ref{fig:fig_crosby_curves}) which results in a sack.
 
-[^unsuccess]: Game footage of this play is available at https://www.nfl.com/videos/teddy-bridgewater-courtland-sutton-barely-miss-out-on-chance-at-62-yard-td-bomb
+<!-- [^unsuccess]: Game footage of this play is available at https://www.nfl.com/videos/teddy-bridgewater-courtland-sutton-barely-miss-out-on-chance-at-62-yard-td-bomb -->
 
@@
 ```
+
+<!-- ```{r} -->
+<!-- #| label: fig_features_combined -->
+<!-- #| fig.cap: "Changes in STRAIN, velocity, and distance from the QB for Maxx Crosby over the course of a successful pass rush play that results in a sack (left) and an unsuccessful pass rush play which results in a long pass attempt (right).
+The STRAIN for other pass rushers throughout each play is also displayed." -->
+<!-- #| fig.width: 6 -->
+<!-- #| fig.height: 3.6 -->
+<!-- source(here("scripts", "15_fig_features_combined.R")) -->
+<!-- fig_features_combined_bw -->
+<!-- ``` -->
+
+<!-- Figure \ref{fig:fig_features_combined} -->
 

### 2023-07-16 paper/04-results.Rmd
edited icc table and text

@@
 
-Next, Table \ref{tab:icc} displays the intraclass correlation coefficients (ICC) for the three different grouped terms, describing the proportion variance explained between the group terms in comparison to the residual variance.
-While the residual variance is unsurprisingly the largest value, between the three group terms, we observe the largest ICC for pass rushers, followed by the QB, and the blocker.
+Next, Table \ref{tab:icc} displays the intraclass correlation coefficients (ICC) for the four different groupings, describing the proportion variance explained between the group terms in comparison to the residual variance.
+While the residual variance is unsurprisingly the largest value, between the player and team factors, we observe the largest ICC for pass rushers, followed by the offensive and defensive teams, and the blocker.
 This emphasizes how STRAIN is mostly attributable to pass rushers, but is necessary to adjust for opposition and other factors.
@@
 
+<!-- \begin{table}[ht] -->
+<!-- \caption{Intraclass correlation coefficients for the multilevel model for average STRAIN. \label{tab:icc}} -->
+<!-- \centering -->
+<!-- \begin{tabular}{lr} -->
+<!-- \hline -->
+<!-- Group & ICC \\  -->
+<!-- \hline -->
+<!-- Pass rusher & 0.0365 \\  -->
+<!-- Pass blocker & 0.0098 \\  -->
+<!-- Defensive team & 0.0134 \\  -->
+<!-- Offensive team & 0.0143 \\ -->
+<!-- Residual & 0.9260 \\  -->
+<!-- \hline -->
+<!-- \end{tabular} -->
+<!-- \end{table} -->
+
 \begin{table}[ht]
@@
 \centering
-\begin{tabular}{lr}
+\begin{tabular}{cccccc}
 \hline
-Group & ICC \\ 
+\text{ } & Pass rusher & Pass blocker & Defensive team & Offensive team & Residual \\ 
 \hline
-Pass rusher & 0.0365 \\ 
-Pass blocker & 0.0098 \\ 
-Defensive team & 0.0134 \\ 
-Offensive team & 0.0143 \\
-Residual & 0.9260 \\ 
+ICC & 0.0365 & 0.0098 & 0.0134 & 0.0143 & 0.9260 \\ 
 \hline

### 2023-07-17 paper/04-results.Rmd
analysis of play context fixed effects

@@
 First, we investigate the fixed-effects terms of this model.
-It appears that that the average STRAIN decreases by 0.042 (95% CI [-0.044, -0.040]) for every extra second in play length, and by 0.064 (95% CI [-0.069, -0.059]) every additional blocker in the offensive unit, after accounting for other covariates.
-In other words, NFL pass rushers tend to generate more pressure in plays with shorter length (e.g., blitzes) and when facing fewer number of blockers, which makes intuitive sense.
+It appears that that the average STRAIN decreases by 0.7366 (95% CI [-0.7875, -0.6858]) for every additional blocker in the offensive unit, after accounting for other covariates.
+In other words, NFL pass rushers tend to generate more pressure when facing fewer number of blockers, which makes intuitive sense.
+
+As for play context, defensive pressure appears to increase when more yardage is required for the offense to reach a first down and for later plays within a set of down, when all other predictors are held constant.
+In particular, every extra yard in the distance needed for a first down is associated with a 0.0491 increase (95% CI [0.0426, 0.0555]) in the average STRAIN.
+Relative to first down situations, higher amount of defensive pressure seem to happen in plays that come after (second, third, and fourth downs).
+In addition, we have insufficient evidence for a relationship between average STRAIN and the current yardline on the field for the offensive team.
 
 Moreover, we observe statistically significant differences among the pass rush and pass block positions in most cases, while controlling for other variables.
-The coefficient estimates for the pass rush position terms reveal that players with positions closer to the line of scrimmage (DTs and NTs) appear to generate less STRAIN (relative to DEs) than those lining up further back, which is consistent with our results in Section \ref{sec:positional}.
+The coefficient estimates for the pass rush position terms reveal that players with positions closer to the line of scrimmage (DTs and NTs) tend to generate less STRAIN (relative to DEs) than those lining up further back, which is consistent with our results in Section \ref{sec:positional}.
 For blockers, centers are those that absorb the most pressure on average, more than guards and other offensive positions (compared to the baseline level, tackles).
@@
 
+[TODO: edit the above? McKinley still there but we did control for team effects]
+
 ```{r}

### 2023-07-17 paper/02-data.Rmd
fixed abbreviations (#14)

@@
 
-Table \ref{tab:track} displays a tracking data example for a [play](https://www.raiders.com/video/de-maxx-crosby-sacks-qb-teddy-bridgewater-for-a-loss-of-6-yards-nfl) from the 2021 NFL regular season week six matchup between the Las Vegas Raiders and Denver Broncos, which ends with Broncos QB Teddy Bridgewater getting sacked by Raiders DE Maxx Crosby[^crosbysack].
+Table \ref{tab:track} displays a tracking data example for a [play](https://www.raiders.com/video/de-maxx-crosby-sacks-qb-teddy-bridgewater-for-a-loss-of-6-yards-nfl) from the 2021 NFL regular season week six matchup between the Las Vegas Raiders and Denver Broncos, which ends with Broncos quarterback Teddy Bridgewater getting sacked by Raiders defensive end Maxx Crosby.
 In addition, Figure \ref{fig:fig_field_raw} presents the locations of every Las Vegas (in black) and Denver (in orange) player on the field from this play at 1, 2, 3 and 4 seconds after the ball snap, with Maxx Crosby highlighted in blue.
@@
 \caption{Example of tracking data for a play during the Las Vegas Raiders versus Denver Broncos NFL game on October 17, 2021.
-The data shown here are for Raiders DE Maxx Crosby, and the frames included are between the ball snap and when the sack by Crosby is recorded.
+The data shown here are for Raiders defensive end Maxx Crosby, and the frames included are between the ball snap and when the sack by Crosby is recorded.
 \label{tab:track}}
@@
 #| fig.cap: "A display of the player tracking data for a play during the Las Vegas Raiders (defense, in black) versus Denver Broncos (offense, in orange) NFL game on October 17, 2021.
-Raiders DE Maxx Crosby is highlighted in blue.
+Raiders defensive end Maxx Crosby is highlighted in blue.
 Snapshots are captured at 1, 2, 3, and 4 seconds after the ball snap."

### 2023-07-17 paper/04-results.Rmd
fixed abbreviations (#14)

@@
 #| label: fig_crosby_curves
-#| fig.cap: "Changes in STRAIN, velocity, and distance from the QB for Maxx Crosby over the course of a successful pass rush play that results in a sack.
+#| fig.cap: "Changes in STRAIN, velocity, and distance from the quarterback for Maxx Crosby over the course of a successful pass rush play that results in a sack.
 The STRAIN for other pass rushers throughout this play is also displayed."
@@
 We observe that for the first two seconds, Crosby is being blocked by a Denver offensive lineman and unable to get close to the quarterback, hence the corresponding STRAIN values are virtually zero. 
-Suddenly at around three seconds after the snap, the STRAIN for Crosby starts to increase after the Raiders DE is freed up and charges towards Bridgewater. 
+Suddenly at around three seconds after the snap, the STRAIN for Crosby starts to increase after the Raiders defensive end is freed up and charges towards Bridgewater. 
 At 4 seconds after the snap, Crosby's STRAIN is 2.30, which means at his current moving rate, it will take Crosby about 0.43 (1/2.30) seconds to make the distance between him and the quarterback 0 (i.e.
@@
 #| label: fig_crosby_curves_unsucess
-#| fig.cap: "Changes in STRAIN, velocity, and distance from the QB for Maxx Crosby over the course of an unsuccessful pass rush play.
+#| fig.cap: "Changes in STRAIN, velocity, and distance from the quarterback for Maxx Crosby over the course of an unsuccessful pass rush play.
 The STRAIN for other pass rushers throughout this play is also displayed."
@@
 <!-- #| label: fig_features_combined -->
-<!-- #| fig.cap: "Changes in STRAIN, velocity, and distance from the QB for Maxx Crosby over the course of a successful pass rush play that results in a sack (left) and an unsuccessful pass rush play which results in a long pass attempt (right).
+<!-- #| fig.cap: "Changes in STRAIN, velocity, and distance from the quarterback for Maxx Crosby over the course of a successful pass rush play that results in a sack (left) and an unsuccessful pass rush play which results in a long pass attempt (right).
 The STRAIN for other pass rushers throughout each play is also displayed." -->
@@
 #| fig.cap: "STRAIN curves for different positions.
-Edge rushers (OLB and DE) tend to generate more STRAIN than interior rushers (DT and NT).
+Edge rushers (outside linebackers and defensive ends) tend to generate more STRAIN than interior rushers (defensive tackles and nose tackles).
 The dotted gray line represents the average STRAIN curve for all players without accounting for position."
@@
 For each position, the curve is based on the average of the observed STRAIN at each frame across all plays and players within the position.
-We observe a clear difference in STRAIN between edge rushers (OLBs and DEs) and interior linemen (DTs and NTs).
+We observe a clear difference in STRAIN between edge rushers (outside linebackers and defensive ends) and interior linemen (defensive tackles and nose tackles).
 Specifically, edge rushers have higher STRAIN than interior linemen on average, as they are more easily able to approach the quarterback on the edge of the pocket versus, for instance, a nose tackle attacking the line head on.
@@
 Since STRAIN is observed continuously for every play in our data, this allows us to aggregate across all frames played and compute the average STRAIN for NFL pass rushers over the course of the eight-week sample size, as discussed in Section \ref{sec:defn}.
-Based on the clearly distinct patterns for different positions as previously observed, we evaluate interior pass rushers (NTs and DTs) separately from edge rushers (OLBs and DEs).
+Based on the clearly distinct patterns for different positions as previously observed, we evaluate interior pass rushers (nose tackles and defensive tackles) separately from edge rushers (outside linebackers and defensive ends).
 Tables \ref{tab:edge} and \ref{tab:interior} are leaderboards for the NFL's best edge and interior rushers (with at least 100 plays) rated by the average STRAIN across all frames for the first eight weeks of the 2021 regular season.
@@
 
-<!-- From @Franks2016Meta (github https://github.com/afranks86/meta-analytics) -->
-
-<!--
-
-Discrimination: Does the metric reliably differentiate between players?
-
-Independence: Does the metric provide new information?
-
--->
-
 Next, we examine different statistical properties of our proposed metric STRAIN.
@@
 
-<!-- [CHECK/EDIT THIS, MAKE SURE WORDING IS CORRECT/NOT CONFUSING] -->
-
 The results of fitting the multilevel model described in Section \ref{sec:multilevel} is displayed in Table \ref{tab:fixedeff}.
@@
 Moreover, we observe statistically significant differences among the pass rush and pass block positions in most cases, while controlling for other variables.
-The coefficient estimates for the pass rush position terms reveal that players with positions closer to the line of scrimmage (DTs and NTs) tend to generate less STRAIN (relative to DEs) than those lining up further back, which is consistent with our results in Section \ref{sec:positional}.
+The coefficient estimates for the pass rush position terms reveal that players with positions closer to the line of scrimmage (defensive tackles and nose tackles) tend to generate less STRAIN (relative to defensive ends) than those lining up further back, which is consistent with our results in Section \ref{sec:positional}.
 For blockers, centers are those that absorb the most pressure on average, more than guards and other offensive positions (compared to the baseline level, tackles).
-
-<!-- [INSERT CONCLUSIONS ABOUT THESE VARIABLES] -->
-<!-- [SAY SOMETHING ABOUT POSITION] -->
 
@@
 
-\begin{table}[ht]
+\begin{table}
 \caption{Intraclass correlation coefficients for the multilevel model for average STRAIN. \label{tab:icc}}
@@
 Figure \ref{fig:fig_rankings_boot} displays the varying intercept distributions for the top ten pass rushers for each position.
-We observe similar rankings based on the random intercepts in comparison to the average STRAIN results in Tables \ref{tab:edge} and \ref{tab:interior}, with Myles Garrett (DE), Aaron Donald (DT), Vita Vea (NT), and T.J.
-Watt (OLB) as the first-ranked pass rusher for their respective position.
+We obtain similar rankings based on the random intercepts in comparison to the average STRAIN results in Tables \ref{tab:edge} and \ref{tab:interior}, with Myles Garrett (defensive end), T.J.
+Watt (outside linebacker), Aaron Donald (defensive tackle), and Vita Vea (nose tackle) as the first-ranked pass rusher for their respective position.
 We observe different levels of variability across these top players, with only a subset of players with intercept ranges strictly above zero.
 This is not necessarily surprising given the limited sample of data.
+
+[TODO: edit the below? McKinley still there but we did control for team effects]
+
 Additionally, we are ignoring correlated performance between defensive linemen on the same team.
@@
 
-[TODO: edit the above? McKinley still there but we did control for team effects]
-
 ```{r}
 #| label: fig_rankings_boot
-#| fig.cap: "Distributions of player effects for top 10 pass rushers in each position (DE, DT, NT, OLB) by median varying intercept."
+#| fig.cap: "Distributions of player effects for top 10 pass rushers in each position (defensive end, outside linebacker, defensive tackle, nose tackle) by median varying intercept."
 #| fig.width: 6.3
-#| fig.height: 5
+#| fig.height: 6
 source(here("scripts", "10_multilevel_mod.R"))

### 2023-07-17 paper/04-results.Rmd
more edits on supplement

@@
 
+<!-- Our results are mostly consistent with conventional rankings of rushers, with Myles Garrett (defensive end), T.J.
+Watt (outside linebacker), Aaron Donald (defensive tackle), and Vita Vea (nose tackle) as the first-ranked pass rusher for their respective position.
+-->
+<!-- Moreover, our leaderboards largely match the rankings of experts in the field. -->
+<!-- For instance, there is considerable overlap between our lists and PFF's edge [@Monson2022nfl] and interior [@Linsey2022nfl] rusher rankings released after the 2021 season. -->
+<!-- This ultimately lends credibility to our proposed metric STRAIN as a measure of pass rushing effectiveness. -->
+
 In addition to adjusting for opposition, our multilevel model allows us to provide some notion of uncertainty about the pass rusher's effect on the play level STRAIN.

### 2023-07-17 paper/supplement.Rmd
more edits on supplement

@@
 \spacingset{1.45}
+
+# Summary
+
+This document contains supplementary materials for the article ""Here Comes the STRAIN: Analyzing Defensive Pass Rush in American Football with Player Tracking Data".
+We provide additional figures and tables, and justification for the multilevel model presented in the paper.
 
@@
 Note that the point size for each Las Vegas defender corresponds to their estimated STRAIN as the play progresses.
-This play is represented by Table [INSERT NUMBER - tracking data table] and Figure [INSERT NUMBER - line graph] in the paper.
+This play is represented in Table [INSERT NUMBER - tracking data table] and Figure [INSERT NUMBER - line graph] in the paper.
 
@@
 
-Since STRAIN is observed continuously for every play in our data, this allows us to aggregate across all frames played and compute the average STRAIN for NFL pass rushers over the course of the eight-week sample size, as discussed in Section \ref{sec:defn}.
-Based on the clearly distinct patterns for different positions as previously observed, we evaluate interior pass rushers (NTs and DTs) separately from edge rushers (OLBs and DEs).
+Since STRAIN is observed continuously for every play in our data, this allows us to aggregate across all frames played and compute the average STRAIN for NFL pass rushers over the course of the eight-week sample size.
+Based on the clearly distinct patterns for different positions as shown in Figure [INSERT strain curve figure] of the article, we evaluate interior pass rushers (NTs and DTs) separately from edge rushers (OLBs and DEs).
 Tables \ref{tab:edge} and \ref{tab:interior} are leaderboards for the NFL's best edge and interior rushers (with at least 100 plays) rated by the average STRAIN across all frames for the first eight weeks of the 2021 regular season.
@@
 
-Our results are mostly consistent with conventional rankings of rushers.
-Notably, Myles Garrett and TJ Watt are widely recognized as top-tier edge rushers and both show up in our top edge rusher list; whereas Aaron Donald, who is undoubtedly the best interior defender in football, appears at the top of our interior rusher rankings.
-Moreover, our leaderboards largely match the rankings of experts in the field.
-For instance, there is considerable overlap between our lists and PFF's edge [@Monson2022nfl] and interior [@Linsey2022nfl] rusher rankings released after the 2021 season.
-This ultimately lends credibility to our proposed metric STRAIN as a measure of pass rushing effectiveness.
+We observe similar rankings based on the average STRAIN in comparison to the random intercepts results in the paper.
+Notably, Myles Garrett and TJ Watt are widely recognized as top-tier edge rushers and both show up in our top edge rusher list.
+Aaron Donald, who is undoubtedly the best interior defender in football, appears at the top of our interior rusher rankings.
+
+<!-- Our results are mostly consistent with conventional rankings of rushers. -->
+<!-- Moreover, our leaderboards largely match the rankings of experts in the field. -->
+<!-- For instance, there is considerable overlap between our lists and PFF's edge [@Monson2022nfl] and interior [@Linsey2022nfl] rusher rankings released after the 2021 season. -->
+<!-- This ultimately lends credibility to our proposed metric STRAIN as a measure of pass rushing effectiveness. -->
 
@@
 
-# Comparison of multilevel models for play-level STRAIN
+# Comparison of multilevel models
 
-## Model without team effects
+## Modeling play-level STRAIN without team effects
 

### 2023-07-17 paper/03-methods.Rmd
fixed abbreviations (#14)
updated fixed effects paragraph (#11)

@@
 \begin{equation*}
-\text{STRAIN}_{ij}(t) = \frac{v_{ij}(t)}{s_{ij}(t)}.
+\text{STRAIN}_{ij}(t) = \frac{- v_{ij}(t)}{s_{ij}(t)}.
 \end{equation*}
@@
 These include the current down (first, second, third, fourth, or two-point conversion), yards to go for a first down, and current yardline (i.e., distance from the possession team's goal line).
-[TODO: add explanation why play context is a proxy for time]
+We account for play-context since these variables impact a team's designed play, which may result in a play with low or high STRAIN regardless of the pass rusher's role.
+For instance, a team may call a short pass that is intended to be thrown early which could limit the potential STRAIN on a play.
+Or a team may need to throw a deep pass which would require more time and create more STRAIN potential.
+We do not account for time directly in the model due to the concern that the time it takes for a quarterback to throw the ball is itself a function of both the play call and pressure from pass rushers.
+Thus, since we do not know the designed play call, we condition on the play context to adjust for play-level differences attributing to a pass rusher's STRAIN.
 

### 2023-07-17 paper/05-discussion.Rmd
Update discussion on rapm model with sabin reference (#12)

@@
 
-Through multilevel modeling, we observe that the pass rusher explains more variation in play-level STRAIN, followed by the quarterback and then their assigned blocker.
+Through multilevel modeling, we observe that the pass rusher explains more variation in play-level STRAIN, followed by the possession team, the pass rushing team, and finally the pass rusher's assigned blocker.
 This is an intuitive result, consistent with previous literature on pass rushing and pressure.
-However, we ignore the fact that pass rushers can be teammates cooperating in the same play leading to some type of correlation structure in their STRAIN values.
-One way to address this complication could be with a adjusted plus-minus (APM) regression approach [@matano2023augmenting], with all players accounted for in modeling a total STRAIN quantity.
+However, our model is limited in that we only account for players directly involved as the pass rusher or nearest blocker, on top of team-level effects.
+The nested structure of our model enforces positive dependence between pass rushers on the same team, which may or may not be true.
+For instance, if one player is known to be an elite pass rusher then the opposing blocking team may focus their blocking efforts on this player.
+This could leave the path open for another player to rush the quarterback with ease, leading to high STRAIN values.
+In order to capture this type of behavior, we could consider modeling an aggregate STRAIN across pass rushers with a regularized adjusted plus-minus (RAPM) regression approach.
+The use of RAPM techniques has been successfully demonstrated in American football by @sabin2021estimating, which accounts for all players on the field with Bayesian hierarchical models.
+We leave this type of analysis for future work, which will require careful consideration of informative prior information [@matano2023augmenting].
 
@@
 Moreover, although we focus solely on pass rushers in this paper, STRAIN can also be applied to the assessment of pass blockers as a unit.
-This can be accomplished through the aforementioned APM regression or by simply looking at quantities such as the total STRAIN or the maximum STRAIN per frame aggregated across the entire play.
+This can be accomplished through the aforementioned RAPM regression or by simply looking at quantities such as the total STRAIN or the maximum STRAIN per frame aggregated across the entire play.
 It is also possible to apply STRAIN to assess individual offensive pass blockers, provided that a method of matching blockers to rushers was developed.

### 2023-07-17 paper/05-discussion.Rmd
small edits to rapm paragraph text

@@
 
-Through multilevel modeling, we observe that the pass rusher explains more variation in play-level STRAIN, followed by the possession team, the pass rushing team, and finally the pass rusher's assigned blocker.
+Through multilevel modeling, we observe that the pass rusher explains more variation in their play-level STRAIN, followed by the possession team, the pass rushing team, and finally the pass rusher's assigned blocker.
 This is an intuitive result, consistent with previous literature on pass rushing and pressure.
 However, our model is limited in that we only account for players directly involved as the pass rusher or nearest blocker, on top of team-level effects.
-The nested structure of our model enforces positive dependence between pass rushers on the same team, which may or may not be true.
+The nested structure of our model enforces positive dependence between pass rushers on the same team, which may not be true.
 For instance, if one player is known to be an elite pass rusher then the opposing blocking team may focus their blocking efforts on this player.
-This could leave the path open for another player to rush the quarterback with ease, leading to high STRAIN values.
+This could leave the path open for another player to rush the quarterback with ease, resulting in higher STRAIN values.
 In order to capture this type of behavior, we could consider modeling an aggregate STRAIN across pass rushers with a regularized adjusted plus-minus (RAPM) regression approach.
 The use of RAPM techniques has been successfully demonstrated in American football by @sabin2021estimating, which accounts for all players on the field with Bayesian hierarchical models.
-We leave this type of analysis for future work, which will require careful consideration of informative prior information [@matano2023augmenting].
+We leave this type of analysis for future work, which will require careful consideration of available informative priors [@matano2023augmenting].
 

### 2023-07-17 paper/05-discussion.Rmd
update discussion a bit about frame level variation (#11)

@@
 Furthermore, compared to existing metrics, STRAIN measures pass rush effectiveness for every play continuously over time, which is at a much more granular level than considering whether the play resulted in a binary outcome such as a sack.
-There is ample opportunity for working with the complete STRAIN trajectories via functional data analysis to better understand the impact of STRAIN on offensive production in American football.
+Indeed, visualizations of STRAIN curves across moments of time within plays reveal variability that simple averaging may obscure.
+There is ample opportunity for working with the complete STRAIN trajectories via temporal modeling and functional data analysis techniques to better understand the impact of STRAIN on offensive production in American football.
 

### 2023-07-17 paper/01-introduction.Rmd
address page 3 comment (#12)

@@
 Our focus of this manuscript is specifically on measuring the performance of defensive linemen in the NFL.
-There are two main types of defensive linemen in American football: (1) defensive tackles (DTs) located within the interior of the line and (2) defensive ends (DEs) located along the edges; see Figure \ref{fig:formation} (top) for example formation with DTs and DEs.
+There are two main types of defensive linemen in American football: (1) defensive tackles (DTs) and (2) defensive ends (DEs).
+Typically, DTs and DEs are located within the interior of the line and along the edges, respectively; see Figure \ref{fig:formation} (top) for example formation with DTs and DEs.
 The primary purpose of both positions is to rush the QB on passing plays, with DEs displaying superiority in observed pass rushing ability [@Eager2018nfl].
-Additionally, within DTs there are nose tackles (NTs) who directly line up across from the ball at the line of scrimmage; see Figure \ref{fig:formation} (bottom) for example defensive formation with a NT.
-Typically, teams will employ either one NT or two DTs on the interior with DEs along either side of the defensive line.
-In addition to defensive lineman, other positions may attempt to rush the QB on blitzing plays such as outside linebackers (OLBs), interior linebackers (ILBs), and potentially members of the secondary (cornerback, free safety, strong safety) who typically play pass coverage.
+Additionally, within DTs there are nose tackles (NTs) who directly line up across from the ball at the line of scrimmage; see Figure \ref{fig:formation} (bottom) for example defensive scheme with a NT.
+NFL teams often employ either one NT or two DTs on the interior with DEs along either side of the defensive line.
+In addition to defensive lineman, other positions may attempt to rush the QB on blitzing plays such as outside linebackers (OLBs), interior linebackers (ILBs), and potentially members of the secondary (cornerback, free safety, strong safety) whose primary role is pass coverage.
+Note that apart from the formations shown in Figure \ref{fig:formation}, defensive linemen can have the flexibility to line up differently.
+For example, a defensive end, depending on their opposing matchup, may not necessarily be positioned toward the outside of the line of scrimmage.
 

### 2023-07-17 paper/04-results.Rmd
address page 3 comment (#12)

@@
 Watt (outside linebacker), Aaron Donald (defensive tackle), and Vita Vea (nose tackle) as the first-ranked pass rusher for their respective position.
-We observe different levels of variability across these top players, with only a subset of players with intercept ranges strictly above zero.
+Additionally, we observe different levels of variability across these top players, with only a subset of players with intercept ranges strictly above zero.
 This is not necessarily surprising given the limited sample of data.
 
-[TODO: edit the below? McKinley still there but we did control for team effects]
+<!-- [TODO: edit the below? McKinley still there but we did control for team effects] -->
 
-Additionally, we are ignoring correlated performance between defensive linemen on the same team.
-For instance, the top two defensive ends, Myles Garrett and Takkarist McKinley (see the top panel of Figure \ref{fig:fig_rankings_boot}), were teammates on the Cleveland Browns during this period of time.
-McKinley is a surprising name on our list and we suspect that his high rank is mostly due to playing on the same defensive unit with Garrett, who is well-known as an elite pass rusher.
-We leave the exploration of adjusting for teammates for future work.
+<!-- Additionally, we are ignoring correlated performance between defensive linemen on the same team. -->
+<!-- For instance, the top two defensive ends, Myles Garrett and Takkarist McKinley (see the top panel of Figure \ref{fig:fig_rankings_boot}), were teammates on the Cleveland Browns during this period of time.
+-->
+<!-- McKinley is a surprising name on our list and we suspect that his high rank is mostly due to playing on the same defensive unit with Garrett, who is well-known as an elite pass rusher. -->
+<!-- We leave the exploration of adjusting for teammates for future work. -->
 

### 2023-07-17 paper/05-discussion.Rmd
address page 3 comment (#12)

@@
 
-Through multilevel modeling, we observe that the pass rusher explains more variation in their play-level STRAIN, followed by the possession team, the pass rushing team, and finally the pass rusher's assigned blocker.
+Through multilevel modeling, we observe that the pass rusher explains more variation in their play-level STRAIN, followed by the possession and pass rushing teams, and finally the pass rusher's assigned blocker.
 This is an intuitive result, consistent with previous literature on pass rushing and pressure.
@@
 This could leave the path open for another player to rush the quarterback with ease, resulting in higher STRAIN values.
+As shown in Figure \ref{fig:fig_rankings_boot}, the first- and fourth-ranked defensive ends Myles Garrett and Takkarist McKinley were teammates on the Cleveland Browns during this period of time.
+McKinley is a surprising name on our list and we suspect that his high rank is mostly due to playing on the same defensive unit with Garrett, who is well-known as an great pass rusher.
 In order to capture this type of behavior, we could consider modeling an aggregate STRAIN across pass rushers with a regularized adjusted plus-minus (RAPM) regression approach.

### 2023-07-17 paper/00-main.Rmd
shitton of work on supplement and figures
more edits... almost done?

@@
   fig.align = "center"
-  # fig.pos = "p",
+  # fig.pos = "t"
   # cache = TRUE

### 2023-07-17 paper/01-introduction.Rmd
more edits... almost done?

@@
 Note that apart from the formations shown in Figure \ref{fig:formation}, defensive linemen can have the flexibility to line up differently.
-For example, a defensive end, depending on their opposing matchup, may not necessarily be positioned toward the outside of the line of scrimmage.
+For example, a defensive end, depending on the opposing matchup, may not necessarily be positioned toward the outside of the line of scrimmage.
 
-```{r formation, fig.width=5, fig.height=4, fig.pos="t", fig.cap="Two common defensive alignments in football: 4-3 defense (top) and 3-4 defense (bottom).
+```{r formation, fig.width=4.8, fig.height=3.5, fig.pos="t", fig.cap="Two common defensive alignments in football: 4-3 defense (top) and 3-4 defense (bottom).
 The dashed line represents the line of scrimmage separating the defense and offense."}
@@
   geom_label(aes(label = pos), size = rel(4)) +
-  geom_hline(yintercept = 0.996, linetype = "dashed") +
+  geom_hline(yintercept = 0.995, linetype = "dashed") +
   facet_wrap(~ d, scales = "free", ncol = 1) +
@@
 
-In this work, using data made available in the BDB 2023, we present a novel approach to measure the performance of pass rushers.
-Relative to other aspects of American football, such as QB evaluation [@burke2019deepqb, @reyers2021quarterback], the literature on evaluating pass rushers is scarce.
+In this work, using data made available in the Big Data Bowl 2023, we present a novel approach to measure the performance of pass rushers.
+Relative to other aspects of American football, such as QB evaluation [@burke2019deepqb; @reyers2021quarterback], the literature on evaluating pass rushers is scarce.
 Below, we provide a brief overview of existing pass rush metrics.

### 2023-07-17 paper/03-methods.Rmd
more edits... almost done?

@@
 We account for play-context since these variables impact a team's designed play, which may result in a play with low or high STRAIN regardless of the pass rusher's role. 
-For instance, a team may call a short pass that is intended to be thrown early which could limit the potential STRAIN on a play.
-Or a team may need to throw a deep pass which would require more time and create more STRAIN potential.
+For instance, a team may call a short pass that is intended to be thrown early which could limit the amount of STRAIN on a play. 
+Or a team may need to throw a deep pass which would require more time and potentially create more STRAIN.
 We do not account for time directly in the model due to the concern that the time it takes for a quarterback to throw the ball is itself a function of both the play call and pressure from pass rushers.

### 2023-07-17 paper/04-results.Rmd
more edits... almost done?

@@
 
+```{r}
+#| label: fig_rankings_boot
+#| fig.cap: "Distributions of player effects for top 10 pass rushers in each position (defensive end, outside linebacker, defensive tackle, nose tackle) by median varying intercept."
+#| fig.width: 6.3
+#| fig.height: 6
+source(here("scripts", "10_multilevel_mod.R"))
+fig_rankings_boot
+```
+
 <!-- We display the random intercepts for pass rusher in Table [INSERT TABLE REFERENCE] alongside their unadjusted STRAIN values.  -->
@@
 Watt (outside linebacker), Aaron Donald (defensive tackle), and Vita Vea (nose tackle) as the first-ranked pass rusher for their respective position.
-Additionally, we observe different levels of variability across these top players, with only a subset of players with intercept ranges strictly above zero.
+Additionally, we observe different levels of variability across these top players, with only a subset of players having intercept distributions strictly above zero.
 This is not necessarily surprising given the limited sample of data.
@@
 <!-- We leave the exploration of adjusting for teammates for future work. -->
-
-```{r}
-#| label: fig_rankings_boot
-#| fig.cap: "Distributions of player effects for top 10 pass rushers in each position (defensive end, outside linebacker, defensive tackle, nose tackle) by median varying intercept."
-#| fig.width: 6.3
-#| fig.height: 6
-source(here("scripts", "10_multilevel_mod.R"))
-fig_rankings_boot
-```

### 2023-07-17 paper/05-discussion.Rmd
more edits... almost done?

@@
 This is analogous to the use of exit velocity in baseball, a predictive measurement that avoids the noisy nature of observed outcomes, emphasizing how the opportunity of working with player tracking data can lead to the development of insightful metrics in American football.
+Through multilevel modeling, we observe that the pass rusher explains more variation in their play-level STRAIN, followed by the possession and pass rushing teams, and finally the pass rusher's assigned blocker.
+This is an intuitive result, consistent with previous literature on pass rushing and defensive pressure.
 
-Through multilevel modeling, we observe that the pass rusher explains more variation in their play-level STRAIN, followed by the possession and pass rushing teams, and finally the pass rusher's assigned blocker.
-This is an intuitive result, consistent with previous literature on pass rushing and pressure.
-However, our model is limited in that we only account for players directly involved as the pass rusher or nearest blocker, on top of team-level effects.
+Our multilevel model, however, is subject to several limitations.
+We only account for players directly involved as the pass rusher or nearest blocker, on top of team-level effects.
 The nested structure of our model enforces positive dependence between pass rushers on the same team, which may not be true.
@@
 This could leave the path open for another player to rush the quarterback with ease, resulting in higher STRAIN values.
-As shown in Figure \ref{fig:fig_rankings_boot}, the first- and fourth-ranked defensive ends Myles Garrett and Takkarist McKinley were teammates on the Cleveland Browns during this period of time.
+As shown in Figure \ref{fig:fig_rankings_boot}, the first- and fourth-ranked defensive ends Myles Garrett and Takkarist McKinley were teammates on the Cleveland Browns during the first eight weeks of the 2021 NFL season.
 McKinley is a surprising name on our list and we suspect that his high rank is mostly due to playing on the same defensive unit with Garrett, who is well-known as an great pass rusher.

### 2023-07-18 paper/03-methods.Rmd
updated bootstrap and more edits

@@
 \end{equation*}
-While this aggregated measure is a simple first step for pass rush evaluation, the observed average STRAIN on a single play is likely due to a convolution of factors.
-Outside of the pass rusher's ability, there is variability in the opposing strength of pass blockers across plays a pass rusher is involved - both at the individual and team level.
-Additionally, based on previous work by PFF, the opposing QB plays an important role on the observed pressure in passing plays [@eager2019pressure].
+While this aggregated measure is a simple first step for pass rush evaluation, the observed average STRAIN on a single play is likely due to numerous factors.
+Besides the pass rusher's ability, there is variability in the opposing strength of pass blockers across plays a pass rusher is involved, both at the individual and team levels.
+Thus, we need to appropriately divide the credit of an observed average STRAIN across the different players and team involved, amongst other factors.
+
+<!-- Additionally, based on previous work by PFF, the opposing QB plays an important role on the observed pressure in passing plays [@eager2019pressure].
 We also suspect that pass rusher's STRAIN is affected by their defensive teammates as well.
-Thus, we need to appropriately divide the credit of an observed average STRAIN across the different players and team involved, amongst other factors.
+-->
 
@@
 
-We consider a normal distribution to shrink the random intercepts for each player and team toward their respective group means.
+In detail, we consider a normal distribution to shrink the random intercepts for each player and team toward their respective group means.
 This is a useful property since we do not observe the same number of plays for each player. 
@@
 These are encoded as indicator variables with defensive ends and tackles as the reference levels for the pass rushers and blockers, respectively.
-We also adjust for the number of pass blockers on the play, since teams may decide to employ a more protective scheme that could lower the observed STRAIN.
+We also account for the number of pass blockers on the play, since teams may decide to employ a more protective scheme that could lower the observed STRAIN.
 Finally, we control for play-context covariates with respect to the offensive team.
 These include the current down (first, second, third, fourth, or two-point conversion), yards to go for a first down, and current yardline (i.e., distance from the possession team's goal line). 
-We account for play-context since these variables impact a team's designed play, which may result in a play with low or high STRAIN regardless of the pass rusher's role. 
+We consider play-context information since these variables impact a team's designed play, which may result in a play with low or high STRAIN regardless of the pass rusher's role. 
 For instance, a team may call a short pass that is intended to be thrown early which could limit the amount of STRAIN on a play. 

### 2023-07-18 paper/04-results.Rmd
updated bootstrap and more edits

--- 2023-07-17 paper/04-results.Rmd
@@
 In particular, every extra yard in the distance needed for a first down is associated with a 0.0491 increase (95% CI [0.0426, 0.0555]) in the average STRAIN.
-Relative to first down situations, higher amount of defensive pressure seem to happen in plays that come after (second, third, and fourth downs).
+Relative to first down situations, higher amount of pressure seem to happen in plays that come after (second, third, and fourth downs).
 In addition, we have insufficient evidence for a relationship between average STRAIN and the current yardline on the field for the offensive team.
@@
 
-In addition to adjusting for opposition, our multilevel model allows us to provide some notion of uncertainty about the pass rusher's effect on the play level STRAIN.
+Aside from adjusting for opposition, our multilevel model allows us to provide some notion of uncertainty about the pass rusher's effect on the play level STRAIN.
 Figure \ref{fig:fig_rankings_boot} displays the varying intercept distributions for the top ten pass rushers for each position.
@@
 Watt (outside linebacker), Aaron Donald (defensive tackle), and Vita Vea (nose tackle) as the first-ranked pass rusher for their respective position.
-Additionally, we observe different levels of variability across these top players, with only a subset of players with intercept ranges strictly above zero.
+Additionally, we observe different levels of variability across these top defenders, with only a subset of players having intercept distributions strictly above zero.
 This is not necessarily surprising given the limited sample of data.

### 2023-07-18 scripts/10_multilevel_mod.R

@@ -405,21 +405,49 @@ strain_boot <- function(b) {
   rushers <- res |> 
     pluck("nflId") |> 
     as_tibble() |> 
-    transmute(nflId = rownames(res$nflId),
+    transmute(who = rownames(res$nflId),
               intercept = `(Intercept)`,
-              boot = b)
-  return(rushers)
+              eff = "rusher")
+  
+  blockers <- res |> 
+    pluck("blocker_id") |> 
+    as_tibble() |> 
+    transmute(who = rownames(res$blocker_id),
+              intercept = `(Intercept)`,
+              eff = "blocker")
+  
+  defense <- res |> 
+    pluck("defensiveTeam") |> 
+    as_tibble() |> 
+    transmute(who = rownames(res$defensiveTeam),
+              intercept = `(Intercept)`,
+              eff = "defense")
+  
+  offense <- res |> 
+    pluck("possessionTeam") |> 
+    as_tibble() |> 
+    transmute(who = rownames(res$possessionTeam),
+              intercept = `(Intercept)`,
+              eff = "offense")
+  
+  out <- rushers |> 
+    bind_rows(blockers) |> 
+    bind_rows(defense) |> 
+    bind_rows(offense) |> 
+    mutate(boot = b)
+  
+  return(out)
 }
 
 # library(furrr)
 # plan(multisession, workers = 7)
 # n_boots <- 1000
-# strain_boot_eff <- seq_len(n_boots) |> 
-#   future_map(strain_boot, 
+# strain_boot_eff <- seq_len(n_boots) |>
+#   future_map(strain_boot,
 #              .options = furrr_options(seed = 101),
 #              .progress = TRUE)
-#   
-# strain_boot_eff <- strain_boot_eff |> 
+# 
+# strain_boot_eff <- strain_boot_eff |>
 #   list_rbind()
 # 
 # write_rds(strain_boot_eff, here("data", "strain_boot_eff.rds"))
@@ -427,6 +455,8 @@ strain_boot <- function(b) {
 strain_boot_eff <- read_rds(here("data", "strain_boot_eff.rds"))
 
 top_rushers <- strain_boot_eff |> 
+  filter(eff == "rusher") |> 
+  rename(nflId = who) |> 
   group_by(nflId) |> 
   summarize(med_intercept = median(intercept, na.rm = TRUE)) |> 
   filter(nflId %in% filter(pass_rush_snaps, n_plays >= 100)$nflId) |> 
[…]

### 2023-07-18 paper/01-introduction.Rmd
I added a description of what the abbreviations meant in the figure in response to the comment about how this needs to reach as general an audience as possible and some people may not be aware of what these abbreviations mean.

@@
 ```{r formation, fig.width=4.8, fig.height=3.5, fig.pos="t", fig.cap="Two common defensive alignments in football: 4-3 defense (top) and 3-4 defense (bottom).
-The dashed line represents the line of scrimmage separating the defense and offense."}
+The dashed line represents the line of scrimmage separating the defense and offense.
+Cornerbacks (CB) and safeties (S) are primarily involved in defending against passing plays.
+Outside linbackers (OLB) and inside linebackers (ILB) generally play directly behind the defensive line and are involved in defending against passing and rushing plays.
+Defensive tackles (DT), defensive ends (DE), and nose tackles (NT) primarily rush the quarterback on passing plays and attempt to stop the rusher as quickly as possible on running plays."}
 tibble(x = c(3:5, 3:6 - 0.5, c(2, 3.7, 4.3, 6)),

### 2023-07-18 paper/01-introduction.Rmd
Changed "The paper" to "they"
removed "rate of"

@@
 @AlamarGoldner2011 later follow up by using manually-tracked data for the 2010 season to estimate lineman performance for different team-positions instead of individual defenders (e.g., Chicago Bears' left tackle, Pittsburgh Steelers' center, etc.).
-The paper uses survival analysis to model time in pocket and completion percentage for quarterbacks before proposing a measure for linemen's contribution to their team's passing in terms of yards gained.
+They used survival analysis to model time in pocket and completion percentage for quarterbacks before proposing a measure for linemen's contribution to their team's passing in terms of yards gained.
 

### 2023-07-18 paper/03-methods.Rmd
Changed "The paper" to "they"
removed "rate of"

@@
 
-In materials science, strain [@callister2018materials] is the deformation of a material from stress, showing the rate of change in a material's length relative to its original length.
+In materials science, strain [@callister2018materials] is the deformation of a material from stress, showing the change in a material's length relative to its original length.
 Formally, let $L(t)$ be the distance between any given two points of interest within a material at time $t$, and $L_0$ be the initial distance between those two points.

### 2023-07-18 paper/03-methods.Rmd
Changed "dimension" to "unit"
Changes "second" to "seconds"

@@
 \end{equation*}
-Notice that this measure is unitless due to being a ratio of two quantities having the same dimension.
+Notice that this measure is unitless due to being a ratio of two quantities having the same units.
 
@@
 where $v(t)$ is the velocity at which the two points of interest within the material are moving away from or towards each other.
-Whereas strain has no units, the strain rate is measured in inverse of time, usually 1/second.
+Whereas strain has no units, the strain rate is measured in inverse of time, usually 1/seconds.
 

### 2023-07-20 paper/01-introduction.Rmd
revise statement about multilevel model in intro

@@
 We also demonstrate that STRAIN is a stable quantity over time and predictive of defensive pressure.
-Additionally, we consider a multilevel model to estimate every pass rusher's contribution to the average STRAIN in a play while controlling for other positions like QB and pass blocker as well as various play-level information.
+Additionally, we consider a multilevel model to estimate every pass rusher's contribution to the average STRAIN in a play while controlling for team and various play-level information.
 We note that although our focus in this paper is on pass rushers, our approach can be extended to the evaluation of pass blockers in American football.

### 2023-07-20 paper/03-methods.Rmd
Updated mean STRAIN notation.

@@
 \end{equation*}
+
 Notice that this quantity increases in two ways: 1) the rate at which the rusher is moving towards the quarterback increases, and 2) the distance between the rusher and the quarterback decreases.
@@
 Moreover, since we observe STRAIN at every tenth of a second within each play, we can then compute the average STRAIN across all frames played for every pass rusher.
-Formally, the average STRAIN, denoted by $\overline{\text{STRAIN}}$, for pass rusher $j$ involved in $n_j$ total plays across $T_j$ total frames is
+Formally, the average STRAIN, denoted by $\overline{\text{STRAIN}}$, for pass rusher $j$ involved in $n_j$ total plays across $\sum_{i\in Z_j} T_i$ total frames where $Z_j$ is a set of the play indices where player $j$ was involved is
 \begin{equation*}
-\overline{\text{STRAIN}}_{j} = \frac{1}{T_j} \sum_{t=1}^{T_j} \widehat{\text{STRAIN}}_{j}(t).
-\end{equation*}
+\overline{\text{STRAIN}}_{j} = \frac{1}{\sum_{i\in Z_j} T_i} \sum_{i\in Z_j} \sum_{t=1}^{T_i} \widehat{\text{STRAIN}}_{ij}(t).
+\end{equation*}.
+
 This can be helpful for player evaluation, as we determine the most effective pass rushers based on their average STRAIN values in Section \ref{sec:strainbar}.

### 2023-07-20 paper/00-main.Rmd
revision submitted to arxiv

@@
   fig.align = "center"
-  # fig.pos = "t"
+  # fig.pos = "p"
   # cache = TRUE
@@
 
-\bigskip
+<!-- \bigskip -->
+\newpage
 \begin{center}

### 2023-07-20 paper/01-introduction.Rmd
revision submitted to arxiv

@@
 
-In an attempt to foster analytics and innovate the game, the National Football League (NFL) introduced their player tracking system known as Next Gen Stats (NGS) in 2016 [@nfl2023ngs].
-NGS uses radio frequency identification (RFID) chips placed in players' shoulder pads (and in the ball) to collect data at a rate of 10 frames per second.
+In an attempt to foster analytics and innovate the game, the National Football League (NFL) introduced their player tracking system known as Next Gen Stats in 2016 [@nfl2023ngs].
+Next Gen Stats uses radio frequency identification (RFID) chips placed in players' shoulder pads (and in the ball) to collect data at a rate of 10 frames per second.
 The data captures real-time on-field information such as locations, speeds, and accelerations of all 22 players (and the football).
-While these data were initially only available for teams, media, and vendors, in December 2018 the NFL launched the inaugural edition of their annual Big Data Bowl (BDB) competition [@nfl2023big].
+While these data were initially only available for teams, media, and vendors, in December 2018 the NFL launched the inaugural edition of their annual Big Data Bowl competition [@nfl2023big].
 
-The first BDB led to several contributions largely focused on offensive performance evaluation.
+The first Big Data Bowl led to several contributions largely focused on offensive performance evaluation.
 For example, one group of finalists introduced an approach for modeling the hypothetical completion probability of a pass aiding in the evaluation of quarterback (QB) decision making [@deshpande2020expected].
-The winners of the inaugural BDB focused on identifying receiver routes via clustering techniques [@chu2020route] and convolutional neural networks [@sterken2019routenet].
-Along with the BDB competition entries, the public release of NGS data allowed researchers to tackle <!-- (pun intended.
+The winners of the inaugural Big Data Bowl focused on identifying receiver routes via clustering techniques [@chu2020route] and convolutional neural networks [@sterken2019routenet].
+Along with the competition entries, the public release of NGS data allowed researchers to tackle <!-- (pun intended.
 #eyeroll) --> a variety of other problems such as revisiting fourth down decision making [@lopez2020bigger], annotating pass coverage with Gaussian mixture models [@dutta2020unsupervised], and introducing a continuous-time framework to estimate within-play value [@yurko2020going].
-Since its inception, the BDB has chosen a different theme each year leading to new insight about evaluating different positions such as running backs, defensive backs, and special teams.
+Since its inception, the Big Data Bowl has chosen a different theme each year leading to new insight about evaluating different positions such as running backs, defensive backs, and special teams.
 The 2023 edition of the NFL Big Data Bowl asked participants to evaluate linemen on passing plays [@Howard2023NFL].
@@
 Our focus of this manuscript is specifically on measuring the performance of defensive linemen in the NFL.
-There are two main types of defensive linemen in American football: (1) defensive tackles (DTs) and (2) defensive ends (DEs).
-Typically, DTs and DEs are located within the interior of the line and along the edges, respectively; see Figure \ref{fig:formation} (top) for example formation with DTs and DEs.
-The primary purpose of both positions is to rush the QB on passing plays, with DEs displaying superiority in observed pass rushing ability [@Eager2018nfl].
-Additionally, within DTs there are nose tackles (NTs) who directly line up across from the ball at the line of scrimmage; see Figure \ref{fig:formation} (bottom) for example defensive scheme with a NT.
-NFL teams often employ either one NT or two DTs on the interior with DEs along either side of the defensive line.
-In addition to defensive lineman, other positions may attempt to rush the QB on blitzing plays such as outside linebackers (OLBs), interior linebackers (ILBs), and potentially members of the secondary (cornerback, free safety, strong safety) whose primary role is pass coverage.
+There are two main types of defensive linemen in American football: defensive tackles and defensive ends.
+Typically, these positions are located within the interior of the line and along the edges, respectively; see Figure \ref{fig:formation} (top) for example formation with defensive tackles and defensive ends.
+The primary purpose of both positions is to rush the QB on passing plays, with defensive ends displaying superiority in observed pass rushing ability [@Eager2018nfl].
+Additionally, within defensive tackles there are nose tackles who directly line up across from the ball at the line of scrimmage; see Figure \ref{fig:formation} (bottom) for example defensive scheme with a nose tackle.
+NFL teams often employ either one nose tackle or two defensive tackles on the interior with defensive ends along either side of the defensive line.
+Besides defensive lineman, other positions may attempt to rush the QB on blitzing plays such as outside linebackers, interior linebackers, and potentially members of the secondary (cornerbacks, free safeties, and strong safeties) whose primary role is pass coverage.
 Note that apart from the formations shown in Figure \ref{fig:formation}, defensive linemen can have the flexibility to line up differently.
@@
 The dashed line represents the line of scrimmage separating the defense and offense.
-Cornerbacks (CB) and safeties (S) are primarily involved in defending against passing plays.
-Outside linbackers (OLB) and inside linebackers (ILB) generally play directly behind the defensive line and are involved in defending against passing and rushing plays.
-Defensive tackles (DT), defensive ends (DE), and nose tackles (NT) primarily rush the quarterback on passing plays and attempt to stop the rusher as quickly as possible on running plays."}
+Defensive tackles (DT), defensive ends (DE), and nose tackles (NT) primarily rush the QB on passing plays and attempt to stop the ball carrier as quickly as possible on running plays.
+Outside linebackers (OLB) and inside linebackers (ILB) usually play directly behind the defensive line and are involved in defending against passing and rushing plays.
+Cornerbacks (CB) and safeties (S) generally involve in defending against passing plays."}
 tibble(x = c(3:5, 3:6 - 0.5, c(2, 3.7, 4.3, 6)),
@@
 In this work, using data made available in the Big Data Bowl 2023, we present a novel approach to measure the performance of pass rushers.
-Relative to other aspects of American football, such as QB evaluation [@burke2019deepqb; @reyers2021quarterback], the literature on evaluating pass rushers is scarce.
+Relative to other aspects of American football, such as quarterback evaluation [@burke2019deepqb; @reyers2021quarterback], the literature on evaluating pass rushers is scarce.
 Below, we provide a brief overview of existing pass rush metrics.
@@
 We now highlight what these quantities describe as well as their limitations.
+
+Perhaps the most commonly-known statistics for evaluating defensive linemen on pass rush plays are sacks, hits, and hurries, which are discretely observed at the play-level.
+Officially tracked by the NFL since 1982, a sack is recorded when a defender tackles the QB behind the line of scrimmage before the QB releases a pass.
+Other traditional box score statistics such as hits and hurries are collected by various outlets.
+A hit is a collision between a defender and the opposing team's quarterback after the quarterback makes a throw.
+A hurry represents an instance when a defender successfully disrupts without necessarily making direct contact with the QB and forces the QB to throw the football earlier than expected.
+These are all simple binary measures of pass rush outcome for any given play.
+However, for plays that do not result in the aforementioned outcomes (in particular, sack), there are still many intermediate defensive actions on the field within the play that are valuable and can be considered positive achievements.
+
+In addition, the sum of sacks, hits, and hurries is often defined as pressures.
+This is better than the individual counts to some extent, but suffers from problems of subjectivity (e.g., whether there is an actual hurry or not).
+Pro Football Focus (PFF) defines a metric called pass-rush productivity (PRP), which is a minor modification from the aforementioned pressures metric (see Table \ref{tab:metrics}).
+In particular, PRP gives twice as much weight to a sack relative to hurries and hits, which is a small upgrade to pressures.
+However, the choice of weights is ad-hoc and still only considers binary outcomes, similar to the shortcomings of previous metrics.
 
@@
 
-Perhaps the most commonly-known statistics for evaluating defensive linemen on pass rush plays are sacks, hits, and hurries, which are discretely observed at the play-level.
-Officially tracked by the NFL since 1982, a sack is recorded when a defender tackles the QB behind the line of scrimmage before the QB releases a pass.
-Other traditional box score statistics such as hits and hurries are collected by various outlets.
-A hit is a collision between a defender and the opposing team's quarterback after the quarterback makes a throw.
-A hurry represents an instance when a defender successfully disrupts without necessarily making direct contact with the QB and forces the QB to throw the football earlier than expected.
-These are all simple binary measures of pass rush outcome for any given play.
-However, for plays that do not result in the aforementioned outcomes (in particular, sack), there are still many intermediate defensive actions on the field within the play that are valuable and can be considered positive achievements.
-
-In addition, the sum of sacks, hits, and hurries is often defined as pressures.
-This is better than the individual counts to some extent, but suffers from problems of subjectivity (e.g., whether there is an actual hurry or not).
-Pro Football Focus (PFF) defines a metric called pass-rush productivity (PRP), which is a minor modification from the aforementioned pressures metric (see Table \ref{tab:metrics}).
-In particular, PRP gives twice as much weight to a sack relative to hurries and hits, which is a small upgrade to pressures.
-However, the choice of weights is ad-hoc and still only considers binary outcomes, similar to the shortcomings of previous metrics.
-
 <!-- and are still binary outcomes (i.e. There is a continuum of plays that are not hurries that all counted as simply "not hurries"). -->
@@
 
-More recently proposed metrics such as time in pocket (TIP), NGS get off [@Hermsmeyer2021nfl], and pass rush win rate (PRWR) [@Burke2018Created] are substantial improvements over the less sophisticated counting statistics, but nevertheless are still imperfect.
-TIP refers to how long a QB can operate within the protected space behind the offensive line, known as the pocket.
+More recently proposed metrics such as time in pocket, NGS get off [@Hermsmeyer2021nfl], and pass rush win rate [@Burke2018Created] are substantial improvements over the less sophisticated counting statistics, but nevertheless are still imperfect.
+Time in pocket refers to how long a QB can operate within the protected space behind the offensive line, known as the pocket.
 However, this measure is highly context-dependent, as it can be influenced by a number of factors such as the defensive scheme or type of passing route.
 NGS get off is an aggregated statistic, illustrating how quickly a defender can get past the line of scrimmage after the snap on average.
-PRWR is created using player tracking data, which is at a more granular level than previous measures.
+Pass rush win rate is created using player tracking data, which is at a more granular level than previous measures.
 It demonstrates whether a pass rusher is able to beat their blocking matchup before a fixed time from the snap (2.5 seconds as chosen by ESPN).
 However, this depends on the rather arbitrary time threshold used to define a pass rush win.
-Besides, once a cutoff is chosen, PRWR converts continuous data to a win-loss indicator, becoming dichotomous like most of the metrics discussed above.
+Besides, once a cutoff is chosen, pass rush win rate converts continuous data to a win-loss indicator, becoming dichotomous like most of the metrics discussed above.
 
@@
 @AlamarGoldner2011 later follow up by using manually-tracked data for the 2010 season to estimate lineman performance for different team-positions instead of individual defenders (e.g., Chicago Bears' left tackle, Pittsburgh Steelers' center, etc.).
-They used survival analysis to model time in pocket and completion percentage for quarterbacks before proposing a measure for linemen's contribution to their team's passing in terms of yards gained.
+This work uses survival analysis to model time in pocket and completion percentage for quarterbacks before proposing a measure for linemen's contribution to their team's passing in terms of yards gained.
 
@@
 We also demonstrate that STRAIN is a stable quantity over time and predictive of defensive pressure.
-Additionally, we consider a multilevel model to estimate every pass rusher's contribution to the average STRAIN in a play while controlling for team and various play-level information.
+Additionally, we consider a multilevel model to estimate every pass rusher's contribution to the average STRAIN in a play while controlling for player positions, team, and various play-level information.
 We note that although our focus in this paper is on pass rushers, our approach can be extended to the evaluation of pass blockers in American football.

### 2023-07-20 paper/02-data.Rmd
revision submitted to arxiv

@@
 
-Along with the tracking information, the 2023 BDB includes scouting data provided by Pro Football Focus (PFF). 
+```{r}
+#| label: pass_rushers_count
+#| fig.cap: "Distribution of the number of pass rushers on passing plays."
+#| fig.width: 3
+#| fig.height: 2.5
+source(here("scripts", "12_fig_pass_rushers_count.R"))
+pass_rushers_count
+```
+
+Along with the tracking information, the Big Data Bowl 2023 includes scouting data provided by Pro Football Focus (PFF). 
 This contains manually-collected player-level information, such as the player's role (e.g., whether they are a pass rusher and pass blocker) and credited events (e.g., player is credited with hitting the QB on the play).
@@
 For context, Figure \ref{fig:pass_rushers_count} displays the distribution of the number of observed pass rushers involved in a play, ranging from 1 to 8 with 4 pass rushers (i.e., 4-man rush formation) as the most common value.
-We also leverage the PFF scouting data to count how many hits, hurries, and sacks each pass rusher is credited with across the span of observed data. 
+We also leverage this scouting data to count how many hits, hurries, and sacks each pass rusher is credited with across the span of observed data. 
 Additionally, we use the PFF player roles to identify the blocking matchup for each pass rusher in order to adjust for opponent strength, as discussed in Section \ref{sec:methods}.
-
-```{r}
-#| label: pass_rushers_count
-#| fig.cap: "Distribution of the number of pass rushers on passing plays."
-#| fig.width: 4.5
-#| fig.height: 3
-source(here("scripts", "12_fig_pass_rushers_count.R"))
-pass_rushers_count
-```

### 2023-07-20 paper/03-methods.Rmd
revision submitted to arxiv

--- 2023-07-18 paper/03-methods.Rmd
@@
 \end{equation*}
-Notice that this measure is unitless due to being a ratio of two quantities having the same units.
+Notice that this measure is unitless due to being a ratio of two quantities having the same unit.
 
@@
 where $v(t)$ is the velocity at which the two points of interest within the material are moving away from or towards each other.
-Whereas strain has no units, the strain rate is measured in inverse of time, usually 1/seconds.
+Whereas strain has no units, the strain rate is measured in inverse of time, usually inverse second.
 
@@
 Moreover, since we observe STRAIN at every tenth of a second within each play, we can then compute the average STRAIN across all frames played for every pass rusher.
-Formally, the average STRAIN, denoted by $\overline{\text{STRAIN}}$, for pass rusher $j$ involved in $n_j$ total plays across $T_j$ total frames is
+Formally, the average STRAIN, denoted by $\overline{\text{STRAIN}}$, for pass rusher $j$ involved in $n_j$ total plays across $\sum_{i\in Z_j} T_i$ total frames, where $Z_j$ is the set of all plays with pass rusher $j$'s involvement, is
 \begin{equation*}
-\overline{\text{STRAIN}}_{j} = \frac{1}{T_j} \sum_{t=1}^{T_j} \widehat{\text{STRAIN}}_{j}(t).
+\overline{\text{STRAIN}}_{j} = \frac{1}{\sum_{i\in Z_j} T_i} \sum_{i\in Z_j} \sum_{t=1}^{T_i} \widehat{\text{STRAIN}}_{ij}(t).
 \end{equation*}
@@
 To this end, we fit a multilevel model to evaluate pass rushers' impact on the average STRAIN observed in a play, while accounting for their team on defense, the opposing team on offense, and their assigned pass blocker.
-We identify the pass blocker linked with the pass rusher of interest using the data provided by PFF.
+We identify the pass blocker linked with the pass rusher of interest using the scouting data provided by PFF as mentioned in Section \ref{sec:data}.
 Since there can be multiple blockers matching up with a rusher, for simplicity, we consider the nearest blocker positioned to the pass rusher at the start of the play.
@@
 Specifically, we resample team drives within games, which preserves the fact that team schedules are fixed but allows for random variation in player usage since this is dependent on team decision making.
-By resampling plays within the same drive together, our sampling is generating realistic simulated data in comparison to sampling individual plays.
+By resampling plays within the same drive together, this allows us to generate realistic simulated data in comparison to sampling individual plays.
 For each bootstrapped dataset, we fit the aforementioned multilevel model to obtain a distribution of estimates for the considered player and team effects.

### 2023-07-20 scripts/10_multilevel_mod.R

@@ -478,7 +478,8 @@ fig_rankings_boot <- strain_boot_eff |>
   mutate(officialPosition = factor(officialPosition, levels = c("DE", "OLB", "DT", "NT"))) |> 
   ggplot(aes(intercept, reorder(displayName, med_intercept))) +
   geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
-  geom_density_ridges(quantile_lines = TRUE, 
+  geom_density_ridges(fill = "lightgray",
+                      quantile_lines = TRUE, 
                       quantiles = 0.5, 
                       rel_min_height = 0.01,
                       scale = 1.2) +

### 2023-07-20 scripts/12_fig_pass_rushers_count.R

@@ -16,7 +16,8 @@ pass_rushers_count <- pff_full |>
   geom_col(fill = "gray") +
   scale_x_continuous(breaks = 1:8) +
   labs(x = "Number of pass rushers",
-       y = "Count")
+       y = "Count") +
+  theme(panel.grid.minor = element_blank())
 
 pass_blockers_count <- pff_full |> 
   filter(pff_role == "Pass Block") |> 

### 2023-07-21 arxiv/01-introduction.Rmd
reorganized directory

--- 2023-07-20 paper/01-introduction.Rmd
@@
 This is better than the individual counts to some extent, but suffers from problems of subjectivity (e.g., whether there is an actual hurry or not).
-Pro Football Focus (PFF) defines a metric called pass-rush productivity (PRP), which is a minor modification from the aforementioned pressures metric (see Table \ref{tab:metrics}).
-In particular, PRP gives twice as much weight to a sack relative to hurries and hits, which is a small upgrade to pressures.
+Pro Football Focus (PFF) defines a metric called pass-rush productivity, which is a minor modification from the aforementioned pressures metric (see Table \ref{tab:metrics}).
+In particular, pass-rush productivity gives twice as much weight to a sack relative to hurries and hits, which is a small upgrade to pressures.
 However, the choice of weights is ad-hoc and still only considers binary outcomes, similar to the shortcomings of previous metrics.

### 2023-07-21 scripts/15_fig_features_combined.R

@@ -112,11 +112,11 @@ fig_features_combined_bw <- feat_sack |>
   labs(linetype = NULL,
        alpha = NULL) +
   theme_light() +
-  theme(axis.title = element_text(size = rel(1)),
-        axis.text = element_text(size = rel(0.8)),
-        strip.text = element_text(size = rel(1)),
+  theme(axis.title = element_text(size = rel(0.9)),
+        axis.text = element_text(size = rel(0.7)),
+        strip.text = element_text(size = rel(0.9)),
         legend.position = "bottom",
-        legend.text = element_text(size = rel(0.9)),
+        legend.text = element_text(size = rel(0.75)),
         panel.spacing = unit(2, "lines"),
         legend.box.spacing = unit(0.25, "lines"),
         legend.key.width = unit(1.5, "cm"))

### 2023-07-21 journal/response.Rmd
added journal folder

---
title: "Response to Reviewers"
output: pdf_document
fontsize: 10pt
header-includes:
 \usepackage{mathptmx}
---

We thank the editor, associate editor, and two reviewers for their critical assessment of our paper.
Below, we provide a point-by-point response to their comments.
Editor/reviewer comments are in normal font and our responses are bold-faced.

# Reviewer 1

**Model assumptions**

* I really liked the multi-level model for STRAIN. That said, the model encodes some strong exchangeability assumptions that, in my opinion, are worth probing. Insofar as (i) pass blockers are nested within teams and (ii) different teams deploy different pass protection schemes/techniques, have you considered an elaborated model in which pass blocker effects are first shrunk towards a team-specific average, which themselves are shrunk to a league-wide average? Perhaps more importantly, as you noted in Section 4.3, the model ignores correlated performance between defensive linemen on the same team. To account for this, you could, similarly nest pass rushers within teams. Doing so, however, introduces positive dependence between the $R_j$'s and it is not obvious to me that two teammates' strains on the same play should be positively correlated. We might observe negative dependence, for instance, if pass blockers concentrated more on blocking one player over another. At the very least, you should discuss the assumptions underpinning the multi-level model. A clear demonstrate that the proposed model provides a better fit to the data than alternatives would, of course, be much better!

> **We thank the reviewer for the positive feedback and for their feedback about the multi-level model. We agree with the reviewer's concern and have updated the model to account for both pass rush and blocking team effects, with nested players. We demonstrate in our supplement that the team model with nested player effects performs better out-of-sample in comparison to our previous model specification.  Coupled with this change, in order to provide uncertainty about the new combined team and player effects, we switched from a parametric bootstrap to a new resampling approach which we believe provides more insight about the model estimates without the rigid assumptions of parametric bootstrapping. We recognize limitations of our initial modeling efforts of STRAIN, and have included further discussion of a RAPM-style approach to address these concerns in Section 5.**

* Relatedly, is there a reason to model the average STRAIN across a play as opposed to the frame-level STRAIN values? The two example plays reveal considerable variation in the frame-to-frame variability. I worry that simply averaging over all frames in a play might obscure this variation.

> **This is a great point raised by the reviewer. Our STRAIN curve figures clearly demonstrate frame-level variability that averaging may hide. We have updated the final paragraph of Section 5 to emphasize this point more clearly. However, we leave modeling frame-level STRAIN for future work due to the increased complexity of that problem and its necessary techniques in relation to this manuscript's purpose of introducing STRAIN in American football.**

**Strain curves**

* I was a bit confused by the STRAIN curves in Figure 7. Are these curves based on a single play? Are they averages of the observed $\overline{\text{STRAIN}}_{ij}$'s values across all plays and all players within positions? Or are they averages of the estimated $\overline{\text{STRAIN}}$ values from the fitted multi-level model? If these curves are based solely on the observed $\overline{\text{STRAIN}}$ values, to what extent can the positional differences be explained by the fact that different pass rush positions tend to be blocked by different offensive line positions? Since multi-level model allows you to adjusting for the position of the nearest pass blocker, it would be interesting to see whether the differences observed between the curves persist after such adjustment.

> **The STRAIN curves are not based on the multilevel and instead are based on the observed average STRAIN values for each position. We have added a sentence to clarify this at the start of section 4.2. It now says "For each position, the curve is based on the average of the observed STRAIN at each frame across all plays and players within the position."**

**Minor comments**

* The $f_{d_{ij}}(t)$ notation in Equations (3)–(5) seems cumbersome. What is $f$ and why is indexed by a distance? Why not just use $d_{ij}(t)$ for the distance?

> **We have changed the distance notation from $f_{d_{ij}}(t)$ to $s_{ij}(t)$. We chose $s_{ij}(t)$ instead of $d_{ij}(t)$ (as suggested by the reviewer) since $s_{ij}(t)$ aligns with the common notation for distance/position in physics, in addition to avoiding notational confusion since we already have another $d$ notation for differentiation.**

* Fixed effects in the multi-level model: it would be helpful to describe what variables are recorded in $\mathbf{x}$ when you introduce the multi-level model instead of in a table towards the end.

> **We have added more descriptions of the fixed effects in our multilevel model. See the last paragraph of Section 3.2 in our revised paper.**

# Reviewer 2

* Page 3: It seems the author is defining anyone who lines up on the line of scrimmage outside a DE. This is not always agreed upon, I would recommend the authors make how they're defining that clear. Is it per play? Or is one player defined as always being a DE?

> **We thank the reviewer for the recommendation. Here, we are showing where a DE typically (not always) lines up, with regards to the two defensive schemes in Figure 1 (of the original manuscript). In the revised version, this part has been moved to the supplement; see section A. We have added clarification about the positions and it now reads "Note that apart from the formations shown in Figure 1 [of the supplement], defensive linemen can have the flexibility to line up differently. For example, a defensive end, depending on their opposing matchup, may not necessarily be positioned toward the outside of the line of scrimmage."**

* Page 13: other fixed effects – this part feels a little rushed. I'd like to see more on why you chose those fixed effects.

> **We have added more descriptions of the fixed effects in our multilevel model. See the last paragraph of Section 3.2 in our revised paper.**

* Page 13: mixed effects model – the way this is written, if a quarterback tends to have more pass rushers or less pass rushers than an average QB, there may be some effect on his random effect of average STRAIN that is due to the this effect

> **We thank the reviewer for the comment and agree with their concern. We have updated the model to no longer directly account for time due to the directionality concern of this relationship. Instead, we now adjust for play-level context: down, yards to go, and yards from endzone - which may impact the type of play call leading to variation in the intended time of the play. We have updated the last paragraph of Section 3.2 accordingly with justification for the play-context features.**

* Why is this random effects model focused on 1 rusher x 1 blocker x 1 qb at a time and not try to put all the players who are present on the field at the same time in the same model? This current format will likely have a lot of teammate effects bleeding into the estimate of another teammate due to being on the field so often together. ( you mention this briefly on page 24 and again in the conclusion – could add to the literature of APM metrics in football (Sabin, 2021).

> **We appreciate the reviewer's feedback and motivation for a regularized adjusted plus-minus (RAPM) style approach for modeling STRAIN. Since our response variable is a pass rusher's average STRAIN for a play, we are not necessarily interested in estimating effects for every player since the interpretation of the coefficient would be at a player-play level. However, we have updated text in Section 5 about modeling an aggregate STRAIN using a RAPM style approach similar to Sabin (2021). Since the primary purpose of this manuscript was to introduce STRAIN and its usefulness in measuring pass rush ability, we leave the RAPM approach for future work in a separate manuscript.**

# Associate Editor

1. The first comment that warrants serious consideration is whether the average strain, across an entire play, is the best way to summarize a pash rushers effect on a particular play. Furthermore, and as just one example, would this summary consistently capture the pash rusher effect across plays of different length?

> **In addition to the average STRAIN, we also considered other summaries, namely, minimum STRAIN and maximum STRAIN. We chose average STRAIN since it is a highly stable quantity over time, as demonstrated in Figure 6. On the other hand, we also compute the correlations in minimum and maximum STRAIN between the first and last four weeks of the 2021 NFL season, and they are both much lower compared to average STRAIN. 
Further, we agree with the comment that our analysis did not account for the fact that play length can differ. This problem can be addressed with a time-to-event analysis. Since the main goal of this paper is to introduce STRAIN as a metric for evaluating pass rushers in football, we leave this for future work.**

[TODO: edit the time to event response]

2. Summarizing the pass rush / pass blocker scheme to a one rusher vs. one defender was also highlighted as a place of concern by both referees. Team pass blocking schemes, multiple pass blockers, and/or changing blockers over the course of the play could all be influential here.

> **We agree with the reviewers on this point that a pass rusher being assigned to exactly one blocker is overly simplistic. However, we believe that the main purpose of this paper is to introduce STRAIN and its usefulness in measuring pass rush ability. Our work focuses entirely on defensive linemen, but in other research that is actively being worked on we attempt to extend STRAIN to evaluate offensive linemen. This requires the problem of assigning each pass rusher to blockers be handled much more rigorously, and we aim to develop a probabilistic approach to match each pass blocker to different defensive pass rushers at every moment within a play. But again, we leave the exploration of this for future work.**

# Editor

1. Avoid abbreviations except obvious ones like NFL and QB. Your paper is short enough (without tables/figures) that we can afford not to use them. This also relates to readability for our readers who are not football experts. See remark #6 below.

> **We have replaced non-obvious abbreviated words/phrases with their full versions.**

2. Reduce the number of tables and figures--you currently have 19 combined. My usual limit is 6--7, but I realize this submission may need to use more in the manuscript. A supplementary article can be used to house additional figures and additional work required to complete this revision. Please see recent issues of TAS for guidance on constructing the supplement. Please submit your revised article and supplement as separate .pdf files.

> **We have reduced the number of tables and figures, and added a supplement.**

3. Do not use footnotes on pages 7 and 16. Either absorb into the text (somewhere) if critical or remove them.

> **We have removed the footnotes.**

4. You might consider using different plotting symbols in Figure 2, for example, when referring to the blue and orange dots. I'm thinking ahead if you want to avoid having to pay for extra charges which relate to the use of color. Same with Figure 4 and others. I'll leave this to you.

> **We have reformatted the figures with different plotting symbols and color schemes. The figures are no longer in color.**

5. Only number those equations which are referred to later in the exposition.

> **We have removed the equation number for those that are not referenced in the text.**

6. Overall comment: Not all of our readers will be football experts like in sports analytics journal submissions. Please try to revise with this in mind. Use language that will make your paper inviting for all readers. Have any NFL teams used your methods? You might mention this in the Discussion section if appropriate.

> **We have revised the paper to make it more readable for a broader audience. While we do not have any direct knowledge of any NFL team implementing this method in the evaluation of offensive or defensive linemen, we do know that NFL teams are at least aware of our proposed metric, as this was a finalist entry in the NFL Big Data Bowl 2023 competition. We have added to Section 5 a brief discussion of how our metric can be easily adapted by football data engineering teams, coaches and broadcasters. It now says "STRAIN is a model-free metric which can easily be integrated into any data pipeline without much computational cost. Visualizations of STRAIN can be useful and intuitive for coaches and broadcasters in various aspects ranging from gameplan preparation to in-game real-time play analysis.**

### 2023-07-21 journal/01-introduction.Rmd
added journal folder

--- 2023-07-20 paper/01-introduction.Rmd
@@
 Our focus of this manuscript is specifically on measuring the performance of defensive linemen in the NFL.
-There are two main types of defensive linemen in American football: defensive tackles and defensive ends.
-Typically, these positions are located within the interior of the line and along the edges, respectively; see Figure \ref{fig:formation} (top) for example formation with defensive tackles and defensive ends.
-The primary purpose of both positions is to rush the QB on passing plays, with defensive ends displaying superiority in observed pass rushing ability [@Eager2018nfl].
-Additionally, within defensive tackles there are nose tackles who directly line up across from the ball at the line of scrimmage; see Figure \ref{fig:formation} (bottom) for example defensive scheme with a nose tackle.
-NFL teams often employ either one nose tackle or two defensive tackles on the interior with defensive ends along either side of the defensive line.
-Besides defensive lineman, other positions may attempt to rush the QB on blitzing plays such as outside linebackers, interior linebackers, and potentially members of the secondary (cornerbacks, free safeties, and strong safeties) whose primary role is pass coverage.
-Note that apart from the formations shown in Figure \ref{fig:formation}, defensive linemen can have the flexibility to line up differently.
-For example, a defensive end, depending on the opposing matchup, may not necessarily be positioned toward the outside of the line of scrimmage.
-
-```{r formation, fig.width=4.8, fig.height=3.5, fig.pos="t", fig.cap="Two common defensive alignments in football: 4-3 defense (top) and 3-4 defense (bottom).
-The dashed line represents the line of scrimmage separating the defense and offense.
-Defensive tackles (DT), defensive ends (DE), and nose tackles (NT) primarily rush the QB on passing plays and attempt to stop the ball carrier as quickly as possible on running plays.
-Outside linebackers (OLB) and inside linebackers (ILB) usually play directly behind the defensive line and are involved in defending against passing and rushing plays.
-Cornerbacks (CB) and safeties (S) generally involve in defending against passing plays."}
-tibble(x = c(3:5, 3:6 - 0.5, c(2, 3.7, 4.3, 6)),
-       y = c(1, 1, 1, 1.01, 1.01, 1.01, 1.01, 1.02, 1.02, 1.02, 1.02),
-       pos = c("DE", "NT", "DE", "OLB", "ILB", "ILB", "OLB",
-               "CB", "S", "S", "CB"),
-       d = "3-4 Defense") |> 
-  bind_rows(
-    tibble(x = c(3:6, c(2.5, 4.5, 6.5), c(2, 4.1, 4.9, 7)),
-           y = c(1, 1, 1, 1, 1.01, 1.01, 1.01, 1.02, 1.02, 1.02, 1.02),
-           pos = c("DE", "DT", "DT", "DE", "OLB", "MLB", "OLB",
-                   "CB", "S", "S", "CB"),
-           d = "4-3 Defense") 
-  ) |> 
-  mutate(d = fct_relevel(d, "4-3 Defense", "3-4 Defense")) |> 
-  ggplot(aes(x, y)) +
-  geom_tile(fill = "white") +
-  geom_label(aes(label = pos), size = rel(4)) +
-  geom_hline(yintercept = 0.995, linetype = "dashed") +
-  facet_wrap(~ d, scales = "free", ncol = 1) +
-  coord_cartesian(ylim = c(0.99, 1.03)) +
-  theme_light() +
-  theme(strip.text = element_text(size = rel(1.2)),
-        axis.text = element_blank(),
-        axis.ticks = element_blank(),
-        axis.title = element_blank(),
-        panel.grid = element_blank())
-```
-
-In this work, using data made available in the Big Data Bowl 2023, we present a novel approach to measure the performance of pass rushers.
+Using data made available in the Big Data Bowl 2023, we present a novel approach to measure the performance of pass rushers.
 Relative to other aspects of American football, such as quarterback evaluation [@burke2019deepqb; @reyers2021quarterback], the literature on evaluating pass rushers is scarce.
@@
 
-Table \ref{tab:metrics} gives a summary of existing football metrics for pass rush.
-We now highlight what these quantities describe as well as their limitations.
+<!-- Table \ref{tab:metrics} gives a summary of existing football metrics for pass rush. -->
+<!-- We now highlight what these quantities describe as well as their limitations. -->
 
@@
 This is better than the individual counts to some extent, but suffers from problems of subjectivity (e.g., whether there is an actual hurry or not).
-Pro Football Focus (PFF) defines a metric called pass-rush productivity (PRP), which is a minor modification from the aforementioned pressures metric (see Table \ref{tab:metrics}).
-In particular, PRP gives twice as much weight to a sack relative to hurries and hits, which is a small upgrade to pressures.
+Pro Football Focus (PFF) defines a metric called pass-rush productivity, which is a minor modification from the aforementioned pressures metric.<!-- (see Table \ref{tab:metrics}). -->
+In particular, pass-rush productivity gives twice as much weight to a sack relative to hurries and hits, which is a small upgrade to pressures.
 However, the choice of weights is ad-hoc and still only considers binary outcomes, similar to the shortcomings of previous metrics.
 
-\begin{table}
-\caption{A summary of previously-existing pass rush metrics. \label{tab:metrics}}
-\centering
-\begin{tabular}{p{0.27\textwidth}p{0.67\textwidth}}
-\hline
-Metric & Description \\ 
-\hline
-Sacks & A defender tackles the QB behind the line of scrimmage before a QB throw \\
-Hits & A defender tackles the QB behind the line of scrimmage after a QB throw \\
-Hurries & A defender pressures the QB behind the line of scrimmage forcing the QB to throw the ball sooner than intended \\
-Pressures & $\text{Hurries} + \text{ Hits } + \text{ Sacks}$ \\
-Pass-Rush Productivity & $\displaystyle \frac{(\text{Hurries} + \text{Hits})/2 + \text{Sacks}}{\text{ Pass Rush Snaps}}$\\
-Time In Pocket & Time (in seconds) between ball snap and throw or pocket collapse for a QB \\
-NGS Get Off & Average time (in seconds) required for a defender to cross the line of scrimmage after the ball snap\\
-Pass Rush Win Rate & Rate at which pass rusher beats pass block within 2.5 seconds after ball snap\\
-\hline
-\end{tabular}
-\end{table}
+<!-- \begin{table} -->
+<!-- \caption{A summary of previously-existing pass rush metrics. \label{tab:metrics}} -->
+<!-- \centering -->
+<!-- \begin{tabular}{p{0.27\textwidth}p{0.67\textwidth}} -->
+<!-- \hline -->
+<!-- Metric & Description \\  -->
+<!-- \hline -->
+<!-- Sacks & A defender tackles the QB behind the line of scrimmage before a QB throw \\ -->
+<!-- Hits & A defender tackles the QB behind the line of scrimmage after a QB throw \\ -->
+<!-- Hurries & A defender pressures the QB behind the line of scrimmage forcing the QB to throw the ball sooner than intended \\ -->
+<!-- Pressures & $\text{Hurries} + \text{ Hits } + \text{ Sacks}$ \\ -->
+<!-- Pass-Rush Productivity & $\displaystyle \frac{(\text{Hurries} + \text{Hits})/2 + \text{Sacks}}{\text{ Pass Rush Snaps}}$\\ -->
+<!-- Time In Pocket & Time (in seconds) between ball snap and throw or pocket collapse for a QB \\ -->
+<!-- NGS Get Off & Average time (in seconds) required for a defender to cross the line of scrimmage after the ball snap\\ -->
+<!-- Pass Rush Win Rate & Rate at which pass rusher beats pass block within 2.5 seconds after ball snap\\ -->
+<!-- \hline -->
+<!-- \end{tabular} -->
+<!-- \end{table} -->
 

### 2023-07-21 journal/02-data.Rmd
added journal folder

--- 2023-07-20 paper/02-data.Rmd
@@
 For each play, we have information on the on-field location, speed, angle, direction, and orientation of each player on the field and the football at a rate of 10 Hz (i.e., 10 measurements per second), along with event annotations for each frame such as ball snap, pass forward, and quarterback sack, to name a few.
+Table \ref{tab:track} displays a tracking data example for a [play](https://www.raiders.com/video/de-maxx-crosby-sacks-qb-teddy-bridgewater-for-a-loss-of-6-yards-nfl) from the 2021 NFL regular season week six matchup between the Las Vegas Raiders and Denver Broncos, which ends with Broncos quarterback Teddy Bridgewater getting sacked by Raiders defensive end Maxx Crosby.
 
@@
 
-Table \ref{tab:track} displays a tracking data example for a [play](https://www.raiders.com/video/de-maxx-crosby-sacks-qb-teddy-bridgewater-for-a-loss-of-6-yards-nfl) from the 2021 NFL regular season week six matchup between the Las Vegas Raiders and Denver Broncos, which ends with Broncos quarterback Teddy Bridgewater getting sacked by Raiders defensive end Maxx Crosby.
-In addition, Figure \ref{fig:fig_field_raw} presents the locations of every Las Vegas (in black) and Denver (in orange) player on the field from this play at 1, 2, 3 and 4 seconds after the ball snap, with Maxx Crosby highlighted in blue.
+<!-- In addition, Figure \ref{fig:fig_field_raw} presents the locations of every Las Vegas (in black) and Denver (in orange) player on the field from this play at 1, 2, 3 and 4 seconds after the ball snap, with Maxx Crosby highlighted in blue.
+-->
 
@@
 
-```{r}
-#| label: fig_field_raw
-#| fig.cap: "A display of the player tracking data for a play during the Las Vegas Raiders (defense, in black) versus Denver Broncos (offense, in orange) NFL game on October 17, 2021.
+<!-- ```{r} -->
+<!-- #| label: fig_field_raw -->
+<!-- #| fig.cap: "A display of the player tracking data for a play during the Las Vegas Raiders (defense, in black) versus Denver Broncos (offense, in orange) NFL game on October 17, 2021.
 Raiders defensive end Maxx Crosby is highlighted in blue.
-Snapshots are captured at 1, 2, 3, and 4 seconds after the ball snap."
-#| fig.width: 6.6
-#| fig.height: 4.4
-source(here("scripts", "11_fig_field_raw.R"))
-fig_field_raw
-```
-
-```{r}
-#| label: pass_rushers_count
-#| fig.cap: "Distribution of the number of pass rushers on passing plays."
-#| fig.width: 3
-#| fig.height: 2.5
-source(here("scripts", "12_fig_pass_rushers_count.R"))
-pass_rushers_count
-```
+Snapshots are captured at 1, 2, 3, and 4 seconds after the ball snap." -->
+<!-- #| fig.width: 6.6 -->
+<!-- #| fig.height: 4.4 -->
+<!-- source(here("scripts", "11_fig_field_raw.R")) -->
+<!-- fig_field_raw -->
+<!-- ``` -->
 
@@
 In this manuscript, we use the PFF data to identify 36,362 unique pass rush attempts by players designated in the "pass rush" role across all plays.
-For context, Figure \ref{fig:pass_rushers_count} displays the distribution of the number of observed pass rushers involved in a play, ranging from 1 to 8 with 4 pass rushers (i.e., 4-man rush formation) as the most common value.
 We also leverage this scouting data to count how many hits, hurries, and sacks each pass rusher is credited with across the span of observed data. 
 Additionally, we use the PFF player roles to identify the blocking matchup for each pass rusher in order to adjust for opponent strength, as discussed in Section \ref{sec:methods}.
+
+<!-- For context, Figure \ref{fig:pass_rushers_count} displays the distribution of the number of observed pass rushers involved in a play, ranging from 1 to 8 with 4 pass rushers (i.e., 4-man rush formation) as the most common value.
+-->
+
+<!-- ```{r} -->
+<!-- #| label: pass_rushers_count -->
+<!-- #| fig.cap: "Distribution of the number of pass rushers on passing plays." -->
+<!-- #| fig.width: 3 -->
+<!-- #| fig.height: 2.5 -->
+<!-- source(here("scripts", "12_fig_pass_rushers_count.R")) -->
+<!-- pass_rushers_count -->
+<!-- ``` -->

### 2023-07-21 journal/03-methods.Rmd
added journal folder

--- 2023-07-20 paper/03-methods.Rmd
@@
 \end{equation*}
-This can be helpful for player evaluation, as we determine the most effective pass rushers based on their average STRAIN values in Section \ref{sec:strainbar}.
-We also use average STRAIN to assess different statistical properties of our metric in Section \ref{sec:statprop}.
+This can be helpful for metric assessment, as we use average STRAIN to examine different statistical properties of our measure in Section \ref{sec:statprop}.
+
+<!-- This can be helpful for player evaluation, as we determine the most effective pass rushers based on their average STRAIN values in Section \ref{sec:strainbar}. -->
+<!-- We also use average STRAIN to assess different statistical properties of our metric in Section \ref{sec:statprop}. -->
 
@@
 First, we adjust for the position of both the pass rusher and nearest blocker to account for any positional effects.
-Table \ref{tab:pos} shows our positional categorization for the pass rush and pass block roles.
-These are encoded as indicator variables with defensive ends and tackles as the reference levels for the pass rushers and blockers, respectively.
+Our positional categorization for pass rushers include defensive end, defensive tackle, nose tackle, outside linebacker, interior linebacker (middle linebacker and inside linebacker), and Secondary (cornerback and safety).
+As for pass blockers, we consider center, guard, tackle, and other (tight end, running back, fullback and wide receiver).
+Note that these positions are encoded as indicator variables with defensive ends and tackles as the reference levels for the pass rushers and blockers, respectively.
 We also account for the number of pass blockers on the play, since teams may decide to employ a more protective scheme that could lower the observed STRAIN.
@@
 
-\begin{table}
-\caption{Position groupings for pass rushers and blockers. \label{tab:pos}}
-\centering
-\begin{tabular}{ll}
-\hline
-Role & Position \\ 
-\hline
-Pass rush & Defensive end \\ 
-Pass rush & Defensive tackle \\ 
-Pass rush & Nose tackle \\ 
-Pass rush & Outside linebacker \\ 
-Pass rush & Interior linebacker (middle linebacker, inside linebacker) \\ 
-Pass rush & Secondary (cornerback, free safety, strong safety) \\
-\hline
-Pass block & Center \\ 
-Pass block & Guard \\ 
-Pass block & Tackle \\ 
-Pass block & Other (tight end, running back, fullback, wide receiver) \\ 
-\hline
-\end{tabular}
-\end{table}
+<!-- \begin{table} -->
+<!-- \caption{Position groupings for pass rushers and blockers. \label{tab:pos}} -->
+<!-- \centering -->
+<!-- \begin{tabular}{ll} -->
+<!-- \hline -->
+<!-- Role & Position \\  -->
+<!-- \hline -->
+<!-- Pass rush & Defensive end \\  -->
+<!-- Pass rush & Defensive tackle \\  -->
+<!-- Pass rush & Nose tackle \\  -->
+<!-- Pass rush & Outside linebacker \\  -->
+<!-- Pass rush & Interior linebacker (middle linebacker, inside linebacker) \\  -->
+<!-- Pass rush & Secondary (cornerback, free safety, strong safety) \\ -->
+<!-- \hline -->
+<!-- Pass block & Center \\  -->
+<!-- Pass block & Guard \\  -->
+<!-- Pass block & Tackle \\  -->
+<!-- Pass block & Other (tight end, running back, fullback, wide receiver) \\  -->
+<!-- \hline -->
+<!-- \end{tabular} -->
+<!-- \end{table} -->
 
-```{r, eval=FALSE}
-rush <- tibble(
-  Role = "Pass rush",
-  Position = c("Defensive end", "Defensive tackle", "Nose tackle", "Outside linebacker",
-               "Interior linebacker (middle and inside linebackers)", 
-               "Secondary (cornerback, free safety, and strong safety)")
-)
-block <- tibble(
-  Role = "Pass block",
-  Position = c("Center", "Guard", "Tackle", "Other (tight end, running back, fullback, wide receiver)")
-)
-rush |> 
-  bind_rows(block) |> 
-  xtable()
-```
+<!-- ```{r, eval=FALSE} -->
+<!-- rush <- tibble( -->
+<!--   Role = "Pass rush", -->
+<!--   Position = c("Defensive end", "Defensive tackle", "Nose tackle", "Outside linebacker", -->
+<!--                "Interior linebacker (middle and inside linebackers)",  -->
+<!--                "Secondary (cornerback, free safety, and strong safety)") -->
+<!-- ) -->
+<!-- block <- tibble( -->
+<!--   Role = "Pass block", -->
+<!--   Position = c("Center", "Guard", "Tackle", "Other (tight end, running back, fullback, wide receiver)") -->
+<!-- ) -->
+<!-- rush |>  -->
+<!--   bind_rows(block) |>  -->
+<!--   xtable() -->
+<!-- ``` -->

### 2023-07-21 journal/04-results.Rmd
added journal folder

--- 2023-07-18 paper/04-results.Rmd
@@
 
-```{r}
-#| label: fig_field
-#| fig.cap: "A display of the player tracking data for a play during the Las Vegas Raiders (defense, in black) versus Denver Broncos (offense, in orange) NFL game on October 17, 2021.
+<!-- ```{r} -->
+<!-- #| label: fig_field -->
+<!-- #| fig.cap: "A display of the player tracking data for a play during the Las Vegas Raiders (defense, in black) versus Denver Broncos (offense, in orange) NFL game on October 17, 2021.
 Raiders DE Maxx Crosby is highlighted in blue.
 For each Raiders defender, the point size indicates their individual STRAIN value, with larger points suggesting larger STRAIN.
-Snapshots are captured at 1, 2, 3, and 4 seconds after the ball snap."
-#| fig.width: 6.6
-#| fig.height: 4.4
-source(here("scripts", "01_fig_field.R"))
-fig_field
-```
+Snapshots are captured at 1, 2, 3, and 4 seconds after the ball snap." -->
+<!-- #| fig.width: 6.6 -->
+<!-- #| fig.height: 4.4 -->
+<!-- source(here("scripts", "01_fig_field.R")) -->
+<!-- fig_field -->
+<!-- ``` -->
 
-```{r}
-#| label: fig_crosby_curves
-#| fig.cap: "Changes in STRAIN, velocity, and distance from the quarterback for Maxx Crosby over the course of a successful pass rush play that results in a sack.
-The STRAIN for other pass rushers throughout this play is also displayed."
-#| fig.width: 4.8
-#| fig.height: 2.7
-source(here("scripts", "02_fig_crosby_curves.R"))
-fig_crosby_curves
-```
+<!-- Figure \ref{fig:fig_field} shows an updated version of Figure \ref{fig:fig_field_raw}, with the point size for each Las Vegas defender corresponding to the estimated STRAIN in the selected frames as the play progresses.
+-->
 
 To illustrate our proposed metric STRAIN for pass rush evaluation, we use the same play from the Las Vegas Raiders and Denver Broncos game as mentioned in Section \ref{sec:data}.
-Figure \ref{fig:fig_field} shows an updated version of Figure \ref{fig:fig_field_raw}, with the point size for each Las Vegas defender corresponding to the estimated STRAIN in the selected frames as the play progresses.
-This is accompanied by Figure \ref{fig:fig_crosby_curves}, which is a line graph showing how Crosby's distance from the quarterback, velocity, and STRAIN change continuously throughout the play.
-
+Figure \ref{fig:fig_features_combined} (left) is a line graph showing how Crosby's distance from the quarterback, velocity, and STRAIN change continuously throughout the play.
 We observe that for the first two seconds, Crosby is being blocked by a Denver offensive lineman and unable to get close to the quarterback, hence the corresponding STRAIN values are virtually zero. 
-Suddenly at around three seconds after the snap, the STRAIN for Crosby starts to increase after the Raiders defensive end is freed up and charges towards Bridgewater. 
+Suddenly at around three seconds after the snap, the STRAIN for Crosby starts to increase after the Raiders defensive end is freed up and charges towards the QB. 
 At 4 seconds after the snap, Crosby's STRAIN is 2.30, which means at his current moving rate, it will take Crosby about 0.43 (1/2.30) seconds to make the distance between him and the quarterback 0 (i.e.
@@
 
-Moreover, Figure \ref{fig:fig_crosby_curves} clearly demonstrates the interactions between the features for Maxx Crosby.
+Moreover, we clearly see the interactions between the features for Maxx Crosby from Figure \ref{fig:fig_features_combined} (left).
 Here, a higher STRAIN generally corresponds to faster moving rate towards the quarterback.
@@
 
-In contrast, Figure \ref{fig:fig_crosby_curves_unsucess} shows the feature curves for an unsuccessful pass rush attempt by the Raiders defense from the same game.
+In contrast, Figure \ref{fig:fig_features_combined} (right) shows the feature curves for an unsuccessful pass rush attempt by the Raiders defense from the same game.
 In this [play](https://www.nfl.com/videos/teddy-bridgewater-courtland-sutton-barely-miss-out-on-chance-at-62-yard-td-bomb), Broncos QB Teddy Bridgewater is well-protected by the offensive line and is able to release a long pass to a receiver.
-We see that the STRAIN generated by Crosby is relatively small over the course of this play, compared to the previous play (as shown in Figure \ref{fig:fig_crosby_curves}) which results in a sack.
+We see that the STRAIN generated by Crosby is relatively small over the course of this play, compared to the previous play which results in a sack.
+
+```{r}
+#| label: fig_features_combined
+#| fig.cap: "Changes in STRAIN, velocity, and distance from the quarterback for Maxx Crosby over the course of a successful pass rush play that results in a sack (left) and an unsuccessful pass rush play which results in a long pass attempt (right).
+The STRAIN for other pass rushers throughout each play is also displayed."
+#| fig.width: 6.1
+#| fig.height: 3.4
+source(here("scripts", "15_fig_features_combined.R"))
+fig_features_combined_bw
+```
 
@@
 
-```{r}
-#| label: fig_crosby_curves_unsucess
-#| fig.cap: "Changes in STRAIN, velocity, and distance from the quarterback for Maxx Crosby over the course of an unsuccessful pass rush play.
-The STRAIN for other pass rushers throughout this play is also displayed."
-#| fig.width: 4.8
-#| fig.height: 2.7
-source(here("scripts", "09_fig_play_unsuccess.R"))
-fig_crosby_curves_unsucess
-```
+<!-- ```{r} -->
+<!-- #| label: fig_crosby_curves_unsucess -->
+<!-- #| fig.cap: "Changes in STRAIN, velocity, and distance from the quarterback for Maxx Crosby over the course of an unsuccessful pass rush play.
+The STRAIN for other pass rushers throughout this play is also displayed." -->
+<!-- #| fig.width: 4.8 -->
+<!-- #| fig.height: 2.7 -->
+<!-- source(here("scripts", "09_fig_play_unsuccess.R")) -->
+<!-- fig_crosby_curves_unsucess -->
+<!-- ``` -->
 
@@
 #| label: fig_pos_curves
-#| fig.cap: "STRAIN curves for different positions.
-Edge rushers (outside linebackers and defensive ends) tend to generate more STRAIN than interior rushers (defensive tackles and nose tackles).
-The dotted gray line represents the average STRAIN curve for all players without accounting for position."
-#| fig.width: 4.5
+#| fig.cap: "STRAIN curves for different positions, along with the average STRAIN curve for all players without accounting for position.
+Edge rushers (outside linebackers and defensive ends) tend to generate more STRAIN than interior rushers (defensive tackles and nose tackles)."
+#| fig.width: 5
 #| fig.height: 3
 source(here("scripts", "03_fig_pos_curves.R"))
-fig_pos_curves
+fig_pos_curves_bw
 ```
@@
 source(here("scripts", "08_fig_outcome_curves.R"))
-fig_outcome_curves
+fig_outcome_curves_bw
 ```
@@
 On average, STRAIN appears to increase for the first 0.5 seconds of a play, followed by a decline in the next second.
-STRAIN then increases again until around 2.5-3 seconds after the snap before trending down towards the end of the play.
+STRAIN then increases again until around 2.5--3 seconds after the snap before trending down towards the end of the play.
 In context, this reflects the actions that a pass rusher initially moves towards the quarterback, but is then stopped by the offensive line while the quarterback drops back.
@@
 We clearly see that players tend to generate more STRAIN when a play ends in a hit, sack, or hurry, compared to no outcome.
-Within the three pressure metrics, it is not surprised that a sack corresponds to the highest amount of STRAIN, followed by a hit and hurry.
+Within the three pressure metrics, it is not surprising that a sack corresponds to the highest amount of STRAIN, followed by a hit and hurry.
 
-## Ranking the Best Pass Rushers {#sec:strainbar}
+<!-- ## Ranking the Best Pass Rushers {#sec:strainbar} -->
 
-Since STRAIN is observed continuously for every play in our data, this allows us to aggregate across all frames played and compute the average STRAIN for NFL pass rushers over the course of the eight-week sample size, as discussed in Section \ref{sec:defn}.
-Based on the clearly distinct patterns for different positions as previously observed, we evaluate interior pass rushers (nose tackles and defensive tackles) separately from edge rushers (outside linebackers and defensive ends).
-Tables \ref{tab:edge} and \ref{tab:interior} are leaderboards for the NFL's best edge and interior rushers (with at least 100 plays) rated by the average STRAIN across all frames for the first eight weeks of the 2021 regular season.
-The tables also consist of the total number of hits, hurries, and sacks (determined from PFF scouting data) for each defender.
+<!-- Since STRAIN is observed continuously for every play in our data, this allows us to aggregate across all frames played and compute the average STRAIN for NFL pass rushers over the course of the eight-week sample size, as discussed in Section \ref{sec:defn}.
+-->
+<!-- Based on the clearly distinct patterns for different positions as previously observed, we evaluate interior pass rushers (nose tackles and defensive tackles) separately from edge rushers (outside linebackers and defensive ends).
+-->
+<!-- Tables \ref{tab:edge} and \ref{tab:interior} are leaderboards for the NFL's best edge and interior rushers (with at least 100 plays) rated by the average STRAIN across all frames for the first eight weeks of the 2021 regular season.
+-->
+<!-- The tables also consist of the total number of hits, hurries, and sacks (determined from PFF scouting data) for each defender. -->
 
-<!-- Our results show many true positives with no surprises. -->
-<!-- Rashan Gary, a young rusher with huge potential, ranks first among all pass rushers according to our metric over the period of time observed. -->
+<!-- <!-- Our results show many true positives with no surprises. -->
+<!-- <!-- Rashan Gary, a young rusher with huge potential, ranks first among all pass rushers according to our metric over the period of time observed. -->
 
-Our results are mostly consistent with conventional rankings of rushers.
-Notably, Myles Garrett and TJ Watt are widely recognized as top-tier edge rushers and both show up in our top edge rusher list; whereas Aaron Donald, who is undoubtedly the best interior defender in football, appears at the top of our interior rusher rankings.
-Moreover, our leaderboards largely match the rankings of experts in the field.
-For instance, there is considerable overlap between our lists and PFF's edge [@Monson2022nfl] and interior [@Linsey2022nfl] rusher rankings released after the 2021 season.
-This ultimately lends credibility to our proposed metric STRAIN as a measure of pass rushing effectiveness.
+<!-- Our results are mostly consistent with conventional rankings of rushers. -->
+<!-- Notably, Myles Garrett and TJ Watt are widely recognized as top-tier edge rushers and both show up in our top edge rusher list; whereas Aaron Donald, who is undoubtedly the best interior defender in football, appears at the top of our interior rusher rankings.
+-->
+<!-- Moreover, our leaderboards largely match the rankings of experts in the field. -->
+<!-- For instance, there is considerable overlap between our lists and PFF's edge [@Monson2022nfl] and interior [@Linsey2022nfl] rusher rankings released after the 2021 season. -->
+<!-- This ultimately lends credibility to our proposed metric STRAIN as a measure of pass rushing effectiveness. -->
 
-\begin{table}
-\caption{Top 15 edge rushers (with at least 100 snaps played) according to the average STRAIN across all frames played. \label{tab:edge}}
-\centering
-\begin{tabular}{rlllrrrrr}
-\hline
-Rank & Player & Team & Position & Snaps & Hits & Hurries & Sacks & $\overline{\text{STRAIN}}$ \\
-\hline
-1 & Rashan Gary & GB & OLB & 176 & 10 & 25 & 5 & 2.82 \\ 
-2 & Leonard Floyd & LA & OLB & 185 & 2 & 25 & 8 & 2.80 \\ 
-3 & Justin Houston & BAL & OLB & 132 & 8 & 8 & 4 & 2.78 \\ 
-4 & Myles Garrett & CLE & DE & 197 & 9 & 29 & 12 & 2.75 \\ 
-5 & Von Miller & DEN & OLB & 145 & 4 & 21 & 5 & 2.75 \\ 
-6 & T.J. Watt & PIT & OLB & 147 & 6 & 9 & 8 & 2.71 \\ 
-7 & Yannick Ngakoue & LV & DE & 175 & 6 & 20 & 4 & 2.70 \\ 
-8 & Alex Highsmith & PIT & OLB & 129 & 4 & 7 & 2 & 2.65 \\ 
-9 & Preston Smith & GB & OLB & 124 & 4 & 8 & 2 & 2.61 \\ 
-10 & Randy Gregory & DAL & DE & 134 & 7 & 19 & 5 & 2.58 \\ 
-11 & Joey Bosa & LAC & OLB & 160 & 5 & 21 & 4 & 2.58 \\ 
-12 & Darrell Taylor & SEA & DE & 107 & 5 & 9 & 3 & 2.57 \\ 
-13 & Josh Sweat & PHI & DE & 159 & 4 & 14 & 5 & 2.57 \\ 
-14 & Maxx Crosby & LV & DE & 198 & 12 & 30 & 7 & 2.56 \\ 
-15 & Markus Golden & ARI & OLB & 164 & 5 & 14 & 5 & 2.50 \\ 
-\hline
-\end{tabular}
-\end{table}
+<!-- \begin{table} -->
+<!-- \caption{Top 15 edge rushers (with at least 100 snaps played) according to the average STRAIN across all frames played. \label{tab:edge}} -->
+<!-- \centering -->
+<!-- \begin{tabular}{rlllrrrrr} -->
+<!-- \hline -->
+<!-- Rank & Player & Team & Position & Snaps & Hits & Hurries & Sacks & $\overline{\text{STRAIN}}$ \\ -->
+<!-- \hline -->
+<!-- 1 & Rashan Gary & GB & OLB & 176 & 10 & 25 & 5 & 2.82 \\  -->
+<!-- 2 & Leonard Floyd & LA & OLB & 185 & 2 & 25 & 8 & 2.80 \\  -->
+<!-- 3 & Justin Houston & BAL & OLB & 132 & 8 & 8 & 4 & 2.78 \\  -->
+<!-- 4 & Myles Garrett & CLE & DE & 197 & 9 & 29 & 12 & 2.75 \\  -->
+<!-- 5 & Von Miller & DEN & OLB & 145 & 4 & 21 & 5 & 2.75 \\  -->
+<!-- 6 & T.J. Watt & PIT & OLB & 147 & 6 & 9 & 8 & 2.71 \\  -->
+<!-- 7 & Yannick Ngakoue & LV & DE & 175 & 6 & 20 & 4 & 2.70 \\  -->
+<!-- 8 & Alex Highsmith & PIT & OLB & 129 & 4 & 7 & 2 & 2.65 \\  -->
+<!-- 9 & Preston Smith & GB & OLB & 124 & 4 & 8 & 2 & 2.61 \\  -->
+<!-- 10 & Randy Gregory & DAL & DE & 134 & 7 & 19 & 5 & 2.58 \\  -->
+<!-- 11 & Joey Bosa & LAC & OLB & 160 & 5 & 21 & 4 & 2.58 \\  -->
+<!-- 12 & Darrell Taylor & SEA & DE & 107 & 5 & 9 & 3 & 2.57 \\  -->
+<!-- 13 & Josh Sweat & PHI & DE & 159 & 4 & 14 & 5 & 2.57 \\  -->
+<!-- 14 & Maxx Crosby & LV & DE & 198 & 12 & 30 & 7 & 2.56 \\  -->
+<!-- 15 & Markus Golden & ARI & OLB & 164 & 5 & 14 & 5 & 2.50 \\  -->
+<!-- \hline -->
+<!-- \end{tabular} -->
+<!-- \end{table} -->
 
-\begin{table}
-\caption{Top 15 interior rushers (with at least 100 snaps played) according to the average STRAIN across all frames played. \label{tab:interior}}
-\centering
-\begin{tabular}{rlllrrrrr}
-\hline
-Rank & Player & Team & Position & Snaps & Hits & Hurries & Sacks & $\overline{\text{STRAIN}}$ \\
-\hline
-1 & Aaron Donald & LA & DT & 239 & 8 & 24 & 6 & 1.67 \\ 
-2 & Solomon Thomas & LV & DT & 115 & 7 & 11 & 3 & 1.51 \\ 
-3 & Quinton Jefferson & LV & DT & 144 & 6 & 8 & 3 & 1.46 \\ 
-4 & Chris Jones & KC & DT & 139 & 3 & 18 & 3 & 1.42 \\ 
-5 & DeForest Buckner & IND & DT & 198 & 4 & 18 & 4 & 1.26 \\ 
-6 & Cameron Heyward & PIT & DT & 188 & 2 & 22 & 3 & 1.25 \\ 
-7 & Javon Hargrave & PHI & DT & 156 & 6 & 15 & 5 & 1.24 \\ 
-8 & Jerry Tillery & LAC & DT & 171 & 4 & 7 & 3 & 1.16 \\ 
-9 & Ed Oliver & BUF & DT & 133 & 4 & 12 & 1 & 1.15 \\ 
-10 & Osa Odighizuwa & DAL & DT & 162 & 3 & 18 & 3 & 1.13 \\ 
-11 & Greg Gaines & LA & NT & 111 & 2 & 13 & 2 & 1.11 \\ 
-12 & Leonard Williams & NYG & DT & 226 & 4 & 14 & 6 & 1.03 \\ 
-13 & Christian Barmore & NE & DT & 166 & 5 & 17 & 1 & 1.02 \\ 
-14 & Vita Vea & TB & NT & 184 & 6 & 12 & 1 & 1.01 \\ 
-15 & B.J. Hill & CIN & DT & 123 & 3 & 4 & 3 & 0.96 \\ 
-\hline
-\end{tabular}
-\end{table}
+<!-- \begin{table} -->
+<!-- \caption{Top 15 interior rushers (with at least 100 snaps played) according to the average STRAIN across all frames played. \label{tab:interior}} -->
+<!-- \centering -->
+<!-- \begin{tabular}{rlllrrrrr} -->
+<!-- \hline -->
+<!-- Rank & Player & Team & Position & Snaps & Hits & Hurries & Sacks & $\overline{\text{STRAIN}}$ \\ -->
+<!-- \hline -->
+<!-- 1 & Aaron Donald & LA & DT & 239 & 8 & 24 & 6 & 1.67 \\  -->
+<!-- 2 & Solomon Thomas & LV & DT & 115 & 7 & 11 & 3 & 1.51 \\  -->
+<!-- 3 & Quinton Jefferson & LV & DT & 144 & 6 & 8 & 3 & 1.46 \\  -->
+<!-- 4 & Chris Jones & KC & DT & 139 & 3 & 18 & 3 & 1.42 \\  -->
+<!-- 5 & DeForest Buckner & IND & DT & 198 & 4 & 18 & 4 & 1.26 \\  -->
+<!-- 6 & Cameron Heyward & PIT & DT & 188 & 2 & 22 & 3 & 1.25 \\  -->
+<!-- 7 & Javon Hargrave & PHI & DT & 156 & 6 & 15 & 5 & 1.24 \\  -->
+<!-- 8 & Jerry Tillery & LAC & DT & 171 & 4 & 7 & 3 & 1.16 \\  -->
+<!-- 9 & Ed Oliver & BUF & DT & 133 & 4 & 12 & 1 & 1.15 \\  -->
+<!-- 10 & Osa Odighizuwa & DAL & DT & 162 & 3 & 18 & 3 & 1.13 \\  -->
+<!-- 11 & Greg Gaines & LA & NT & 111 & 2 & 13 & 2 & 1.11 \\  -->
+<!-- 12 & Leonard Williams & NYG & DT & 226 & 4 & 14 & 6 & 1.03 \\  -->
+<!-- 13 & Christian Barmore & NE & DT & 166 & 5 & 17 & 1 & 1.02 \\  -->
+<!-- 14 & Vita Vea & TB & NT & 184 & 6 & 12 & 1 & 1.01 \\  -->
+<!-- 15 & B.J. Hill & CIN & DT & 123 & 3 & 4 & 3 & 0.96 \\  -->
+<!-- \hline -->
+<!-- \end{tabular} -->
+<!-- \end{table} -->
 
@@
 source(here("scripts", "04_fig_cor_pressure.R"))
-fig_cor_pressure
+fig_cor_pressure_bw
 ```
@@
 source(here("scripts", "07_fig_predictability.R"))
-fig_predictability
+fig_predictability_bw
 ```
@@
 source(here("scripts", "05_fig_stability.R"))
-fig_stability
+fig_stability_bw
 ```
@@
 
-Next, Table \ref{tab:icc} displays the intraclass correlation coefficients (ICC) for the four different groupings, describing the proportion variance explained between the group terms in comparison to the residual variance.
-While the residual variance is unsurprisingly the largest value, between the player and team factors, we observe the largest ICC for pass rushers, followed by the offensive and defensive teams, and the blocker.
+Next, we examine the intraclass correlation coefficients (ICC) for the four different groupings
+in our multilevel model.
+This descriptive statistic illustrates the proportion variance explained between the group terms in comparison to the residual variance.
+While the residual variance is unsurprisingly the highest value (0.9260), between the player and team factors, we observe the largest ICC for pass rushers (0.0365), followed by the offensive (0.0143) and defensive teams (0.0134), and finally the blocker (0.0098).
 This emphasizes how STRAIN is mostly attributable to pass rushers, but is necessary to adjust for opposition and other factors.
@@
 
-\begin{table}
-\caption{Intraclass correlation coefficients for the multilevel model for average STRAIN. \label{tab:icc}}
-\centering
-\begin{tabular}{cccccc}
-\hline
-\text{ } & Pass rusher & Pass blocker & Defensive team & Offensive team & Residual \\ 
-\hline
-ICC & 0.0365 & 0.0098 & 0.0134 & 0.0143 & 0.9260 \\ 
-\hline
-\end{tabular}
-\end{table}
+<!-- \begin{table} -->
+<!-- \caption{Intraclass correlation coefficients for the multilevel model for average STRAIN. \label{tab:icc}} -->
+<!-- \centering -->
+<!-- \begin{tabular}{cccccc} -->
+<!-- \hline -->
+<!-- \text{ } & Pass rusher & Pass blocker & Defensive team & Offensive team & Residual \\  -->
+<!-- \hline -->
+<!-- ICC & 0.0365 & 0.0098 & 0.0134 & 0.0143 & 0.9260 \\  -->
+<!-- \hline -->
+<!-- \end{tabular} -->
+<!-- \end{table} -->
 
@@
 Figure \ref{fig:fig_rankings_boot} displays the varying intercept distributions for the top ten pass rushers for each position.
-We obtain similar rankings based on the random intercepts in comparison to the average STRAIN results in Tables \ref{tab:edge} and \ref{tab:interior}, with Myles Garrett (defensive end), T.J.
+Our rankings based on the random intercepts are mostly consistent with conventional rankings of rushers, with Myles Garrett (defensive end), T.J.
 Watt (outside linebacker), Aaron Donald (defensive tackle), and Vita Vea (nose tackle) as the first-ranked pass rusher for their respective position.
+Garrett and Watt are widely recognized as top-tier edge rushers, whereas Donald is undoubtedly the best interior defender in football.
+Moreover, our leaderboards largely match the rankings of experts in the field.
+For instance, there is considerable overlap between our lists and PFF's edge [@Monson2022nfl] and interior [@Linsey2022nfl] rusher rankings released after the 2021 season.
+This ultimately lends credibility to our proposed metric STRAIN as a measure of pass rushing effectiveness.
 Additionally, we observe different levels of variability across these top defenders, with only a subset of players having intercept distributions strictly above zero.

### 2023-07-21 journal/05-discussion.Rmd
added journal folder

--- 2023-07-17 paper/05-discussion.Rmd
@@
 This is analogous to the use of exit velocity in baseball, a predictive measurement that avoids the noisy nature of observed outcomes, emphasizing how the opportunity of working with player tracking data can lead to the development of insightful metrics in American football.
-Through multilevel modeling, we observe that the pass rusher explains more variation in their play-level STRAIN, followed by the possession and pass rushing teams, and finally the pass rusher's assigned blocker.
+Through multilevel modeling, we observe that the pass rusher explains more variation in their play-level STRAIN, followed by the possession team and pass rushing team, and finally the pass rusher's assigned blocker.
 This is an intuitive result, consistent with previous literature on pass rushing and defensive pressure.
@@
 The nested structure of our model enforces positive dependence between pass rushers on the same team, which may not be true.
-For instance, if one player is known to be an elite pass rusher then the opposing blocking team may focus their blocking efforts on this player. 
+For instance, if one defender is known to be an elite pass rusher then the opposing team may focus their blocking efforts on this player. 
 This could leave the path open for another player to rush the quarterback with ease, resulting in higher STRAIN values.
 As shown in Figure \ref{fig:fig_rankings_boot}, the first- and fourth-ranked defensive ends Myles Garrett and Takkarist McKinley were teammates on the Cleveland Browns during the first eight weeks of the 2021 NFL season.
-McKinley is a surprising name on our list and we suspect that his high rank is mostly due to playing on the same defensive unit with Garrett, who is well-known as an great pass rusher.
+McKinley is a surprising name on our list and we suspect that his high rank is mostly due to playing on the same defensive unit with Garrett, who is highly-regarded as an great pass rusher.
 In order to capture this type of behavior, we could consider modeling an aggregate STRAIN across pass rushers with a regularized adjusted plus-minus (RAPM) regression approach.

### 2023-07-21 journal/supplement.Rmd
added journal folder

---
title: |
  | Supplement to "Here Comes the STRAIN: Analyzing Defensive Pass Rush in American Football with Player Tracking Data"
blinded: 0
bibliography: bibliography.bib
biblio-style: chicago
authors: 
- name: Quang Nguyen
  # thanks: Corresponding author's email
  affiliation: |
    | Department of Statistics & Data Science
    | Carnegie Mellon University
- name: Ronald Yurko
  affiliation: |
    | Department of Statistics & Data Science
    | Carnegie Mellon University
- name: Gregory J. Matthews
  affiliation: |
    | Department of Mathematics and Statistics
    | Center for Data Science and Consulting
    | Loyola University Chicago
output: rticles::asa_article
header-includes:
  \date{}
  \def\thesection{\Alph{section}}
---

\spacingset{1.45}

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
library(here)
library(tidyverse)
```

# Summary {-}

This document contains supplementary materials for the article "Here Comes the STRAIN: Analyzing Defensive Pass Rush in American Football with Player Tracking Data".
We provide an overview of defensive positions in football, additional figures and tables, and justification for the multilevel model presented in the paper.

# Defensive Linemen in Football: Common Terms

There are two main types of defensive linemen in American football: defensive tackles and defensive ends.
Typically, these positions are located within the interior of the line and along the edges, respectively; with main role of rushing the quarterback (QB) on passing plays.
See Figure \ref{fig:formation} (top) for example formation with defensive tackles and defensive ends.
Additionally, within defensive tackles there are nose tackles who directly line up across from the ball at the line of scrimmage.
See Figure \ref{fig:formation} (bottom) for example defensive scheme with a nose tackle.
NFL teams often employ either one nose tackle or two defensive tackles on the interior with defensive ends along either side of the defensive line.
Besides defensive lineman, other positions may attempt to rush the QB on blitzing plays such as outside linebackers, interior linebackers, and potentially members of the secondary (cornerbacks, free safeties, and strong safeties) whose primary role is pass coverage.
Note that apart from the formations shown in Figure \ref{fig:formation}, defensive linemen can have the flexibility to line up differently.
For example, a defensive end, depending on the opposing matchup, may not necessarily be positioned toward the outside of the line of scrimmage.

```{r formation, fig.width=5, fig.height=4, fig.cap="Two common defensive alignments in football: 4-3 defense (top) and 3-4 defense (bottom). The dashed line represents the line of scrimmage separating the defense and offense. Defensive tackles (DT), defensive ends (DE), and nose tackles (NT) primarily rush the QB on passing plays and attempt to stop the ball carrier as quickly as possible on running plays. Outside linebackers (OLB) and inside linebackers (ILB) usually play directly behind the defensive line and are involved in defending against passing and rushing plays. Cornerbacks (CB) and safeties (S) generally involve in defending against passing plays."}
tibble(x = c(3:5, 3:6 - 0.5, c(2, 3.7, 4.3, 6)),
       y = c(1, 1, 1, 1.01, 1.01, 1.01, 1.01, 1.02, 1.02, 1.02, 1.02),
       pos = c("DE", "NT", "DE", "OLB", "ILB", "ILB", "OLB",
               "CB", "S", "S", "CB"),
       d = "3-4 Defense") |> 
  bind_rows(
    tibble(x = c(3:6, c(2.5, 4.5, 6.5), c(2, 4.1, 4.9, 7)),
           y = c(1, 1, 1, 1, 1.01, 1.01, 1.01, 1.02, 1.02, 1.02, 1.02),
           pos = c("DE", "DT", "DT", "DE", "OLB", "MLB", "OLB",
                   "CB", "S", "S", "CB"),
           d = "4-3 Defense") 
  ) |> 
  mutate(d = fct_relevel(d, "4-3 Defense", "3-4 Defense")) |> 
  ggplot(aes(x, y)) +
  geom_tile(fill = "white") +
  geom_label(aes(label = pos), size = rel(4)) +
  geom_hline(yintercept = 0.996, linetype = "dashed") +
  facet_wrap(~ d, scales = "free", ncol = 1) +
  coord_cartesian(ylim = c(0.99, 1.03)) +
  theme_light() +
  theme(strip.text = element_text(size = rel(1.2)),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank())
```

# Summary of Previous Pass Rush Metrics

Table \ref{tab:metrics} gives a summary of existing football metrics for pass rush, as discussed in Section 1.1 of the paper.

\begin{table}
\caption{A summary of previously-existing pass rush metrics. \label{tab:metrics}}
\centering
\begin{tabular}{p{0.27\textwidth}p{0.67\textwidth}}
\hline
Metric & Description \\
\hline
Sacks & A defender tackles the QB behind the line of scrimmage before a QB throw \\
Hits & A defender tackles the QB behind the line of scrimmage after a QB throw \\
Hurries & A defender pressures the QB behind the line of scrimmage forcing the QB to throw the ball sooner than intended \\
Pressures & $\text{Hurries} + \text{ Hits } + \text{ Sacks}$ \\
Pass-Rush Productivity & $\displaystyle \frac{(\text{Hurries} + \text{Hits})/2 + \text{Sacks}}{\text{ Pass Rush Snaps}}$\\
Time In Pocket & Time (in seconds) between ball snap and throw or pocket collapse for a QB \\
NGS Get Off & Average time (in seconds) required for a defender to cross the line of scrimmage after the ball snap\\
Pass Rush Win Rate & Rate at which pass rusher beats pass block within 2.5 seconds after ball snap\\
\hline
\end{tabular}
\end{table}

# Visualizing Player Tracking Data

Figure \ref{fig:fig_field} shows snapshots of a play from the week six game between the Las Vegas Raiders and Denver Broncos in the 2021 NFL regular season.
The play outcome is a sack made by Raiders defensive end Maxx Crosby on Broncos quarterback Teddy Bridgewater.
The locations of every Las Vegas (in black, on defense) and Denver (in orange, on offense) player on the field at 1, 2, 3, and 4 seconds after the ball snap is displayed, with Crosby highlighted in blue.
Note that the point size for each Las Vegas player corresponds to their estimated STRAIN as the play progresses.
This play is represented in Table 1 and Figure 1 (left) of the paper.

```{r}
#| label: fig_field
#| fig.cap: "A display of the player tracking data for a play during the Las Vegas Raiders (defense, in black) versus Denver Broncos (offense, in orange) NFL game on October 17, 2021. Raiders DE Maxx Crosby is highlighted in blue. For each Raiders defender, the point size indicates their individual STRAIN value, with larger points suggesting larger STRAIN. Snapshots are captured at 1, 2, 3, and 4 seconds after the ball snap."
#| fig.width: 6.6
#| fig.height: 4.4
source(here("scripts", "01_fig_field.R"))
fig_field
```

# Distribution of Number of Pass Rushers

Figure \ref{fig:pass_rushers_count} displays the distribution of the number of observed pass rushers involved in a play.
The number of pass rusher ranges from 1 to 8, with 4 pass rushers (i.e., 4-man rush formation) as the most common value.

```{r}
#| label: pass_rushers_count
#| fig.cap: "Distribution of the number of pass rushers on passing plays."
#| fig.width: 4
#| fig.height: 3
source(here("scripts", "12_fig_pass_rushers_count.R"))
pass_rushers_count
```

# Pass Rusher Rankings Based on Average STRAIN

Since STRAIN is observed continuously for every play in our data, this allows us to aggregate across all frames played and compute the average STRAIN for NFL pass rushers over the course of the eight-week sample size.
Based on the clearly distinct patterns for different positions as shown in Figure 2 of the article, we evaluate interior pass rushers (NTs and DTs) separately from edge rushers (OLBs and DEs).
Tables \ref{tab:edge} and \ref{tab:interior} are leaderboards for the NFL's best edge and interior rushers (with at least 100 plays) rated by the average STRAIN across all frames for the first eight weeks of the 2021 regular season.
The tables also consist of the total number of hits, hurries, and sacks (determined from PFF scouting data) for each defender.
We observe similar rankings based on the average STRAIN in comparison to the random intercepts results from Section 4.4 of the paper.
Notably, Myles Garrett and T.J. Watt both show up in our top edge rusher list based on average STRAIN, whereas Aaron Donald appears at the top of our interior rusher rankings.

\begin{table}
\caption{Top 15 edge rushers (with at least 100 snaps played) according to the average STRAIN across all frames played. \label{tab:edge}}
\centering
\begin{tabular}{rlllrrrrr}
\hline
Rank & Player & Team & Position & Snaps & Hits & Hurries & Sacks & $\overline{\text{STRAIN}}$ \\
\hline
1 & Rashan Gary & GB & OLB & 176 & 10 & 25 & 5 & 2.82 \\ 
2 & Leonard Floyd & LA & OLB & 185 & 2 & 25 & 8 & 2.80 \\ 
3 & Justin Houston & BAL & OLB & 132 & 8 & 8 & 4 & 2.78 \\ 
4 & Myles Garrett & CLE & DE & 197 & 9 & 29 & 12 & 2.75 \\ 
5 & Von Miller & DEN & OLB & 145 & 4 & 21 & 5 & 2.75 \\ 
6 & T.J. Watt & PIT & OLB & 147 & 6 & 9 & 8 & 2.71 \\ 
7 & Yannick Ngakoue & LV & DE & 175 & 6 & 20 & 4 & 2.70 \\ 
8 & Alex Highsmith & PIT & OLB & 129 & 4 & 7 & 2 & 2.65 \\ 
9 & Preston Smith & GB & OLB & 124 & 4 & 8 & 2 & 2.61 \\ 
10 & Randy Gregory & DAL & DE & 134 & 7 & 19 & 5 & 2.58 \\
11 & Joey Bosa & LAC & OLB & 160 & 5 & 21 & 4 & 2.58 \\
12 & Darrell Taylor & SEA & DE & 107 & 5 & 9 & 3 & 2.57 \\
13 & Josh Sweat & PHI & DE & 159 & 4 & 14 & 5 & 2.57 \\
14 & Maxx Crosby & LV & DE & 198 & 12 & 30 & 7 & 2.56 \\
15 & Markus Golden & ARI & OLB & 164 & 5 & 14 & 5 & 2.50 \\
\hline
\end{tabular}
\end{table}

\begin{table}
\caption{Top 15 interior rushers (with at least 100 snaps played) according to the average STRAIN across all frames played. \label{tab:interior}}
\centering
\begin{tabular}{rlllrrrrr}
\hline
Rank & Player & Team & Position & Snaps & Hits & Hurries & Sacks & $\overline{\text{STRAIN}}$ \\
\hline
1 & Aaron Donald & LA & DT & 239 & 8 & 24 & 6 & 1.67 \\ 
2 & Solomon Thomas & LV & DT & 115 & 7 & 11 & 3 & 1.51 \\ 
3 & Quinton Jefferson & LV & DT & 144 & 6 & 8 & 3 & 1.46 \\ 
4 & Chris Jones & KC & DT & 139 & 3 & 18 & 3 & 1.42 \\ 
5 & DeForest Buckner & IND & DT & 198 & 4 & 18 & 4 & 1.26 \\ 
6 & Cameron Heyward & PIT & DT & 188 & 2 & 22 & 3 & 1.25 \\ 
7 & Javon Hargrave & PHI & DT & 156 & 6 & 15 & 5 & 1.24 \\ 
8 & Jerry Tillery & LAC & DT & 171 & 4 & 7 & 3 & 1.16 \\ 
9 & Ed Oliver & BUF & DT & 133 & 4 & 12 & 1 & 1.15 \\ 
10 & Osa Odighizuwa & DAL & DT & 162 & 3 & 18 & 3 & 1.13 \\ 
11 & Greg Gaines & LA & NT & 111 & 2 & 13 & 2 & 1.11 \\
12 & Leonard Williams & NYG & DT & 226 & 4 & 14 & 6 & 1.03 \\
13 & Christian Barmore & NE & DT & 166 & 5 & 17 & 1 & 1.02 \\
14 & Vita Vea & TB & NT & 184 & 6 & 12 & 1 & 1.01 \\
15 & B.J. Hill & CIN & DT & 123 & 3 & 4 & 3 & 0.96 \\
\hline
\end{tabular}
\end{table}

# Comparison of Multilevel Models

## Modeling Play-Level STRAIN Without Team Effects

We additionally consider a multilevel model to evaluate the impact of pass rushers on the average STRAIN observed in a play, while accounting for their pass blocker, QB, and play-level attributes.
The average STRAIN by pass rusher $j$ on play $i$ is modeled as follows.
\begin{equation*}
\begin{aligned}
\overline{\text{STRAIN}}_{ij} &\sim N(R_{j[i]} + B_{b[ij]} + Q_{q[i]} + \mathbf{x_{ij}} \boldsymbol{\beta}, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
R_{j} &\sim N(\mu_R, \sigma^2_R), \text{ for } j = 1, \cdots, \text{ \# of pass rushers}, \\
B_{b} &\sim N(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
Q_{q} &\sim N(\mu_Q, \sigma^2_Q), \text{ for } q = 1, \cdots, \text{ \# of QBs}.
\end{aligned}
\end{equation*}
This model version contains random intercepts for three player groups: (1) pass rushers as $R$, (2) nearest pass blockers as $B$, and (3) QBs as $Q$.
We consider a normal distribution to obtain the random intercepts for each player toward their respective position group means.
As before, the coefficients $\boldsymbol \beta$ associated with covariate vector $\mathbf{x_{ij}}$ for pass rusher $j$ in play $i$ are estimated as fixed-effects. 
The variables included as fixed effects are the same as those in the model from Section 3.2 of the paper: position of pass rusher and blocker, number of blockers in a play, and play-context information (down, yards to go, and current yardline).

## Model Comparison

We perform leave-one-week-out cross-validation to compare the multilevel models with and without team effects for play-level STRAIN.
For example, we train each model using data from weeks one through seven, and perform evaluation for week eight.
This procedure is repeated for every single holdout week, and we evaluate the model performance based on mean squared error (MSE).
Figure \ref{fig:fig_loocv_team_eff} shows the week-by-week holdout MSE for the aforementioned multilevel models.
We clearly see that a smaller MSE is achieved for each left-out week when accounting for team effects.
Furthermore, the model with team effects yields a lower overall holdout MSE ($4.1882$) than the one without random intercepts for teams ($4.4138$).
This provides a quantitative justification for including team effects in our multilevel model.

```{r}
#| label: fig_loocv_team_eff
#| fig.cap: "Comparison of leave-one-week-out cross-validation MSE with $\\pm 2$ standard error bounds for multilevel models with and without team effects."
#| fig.width: 6
#| fig.height: 3
source(here("scripts", "10_multilevel_mod.R"))
fig_loocv_team_eff
```

### 2023-07-21 scripts/03_fig_pos_curves.R

@@ -71,7 +71,7 @@ fig_pos_curves_bw <- pass_rush |>
   geom_smooth(se = FALSE, span = 0.3, linewidth = 1) +
   scale_x_continuous(breaks = seq(0, 40, 10), labels = 0:4) +
   scale_linetype_manual(values = c("solid", "dashed", "twodash", "dotdash", "dotted"), name = "Position") +
-  scale_color_manual(values = c("black", "black", "gray90", "gray60", "gray60"), name = "Position") +
+  scale_color_manual(values = rev(c("black", "black", "gray90", "gray60", "gray60")), name = "Position") +
   labs(y = "STRAIN",
        x = "Time since snap (seconds)") +
   theme_light() +

### 2023-07-21 scripts/04_fig_cor_pressure.R

@@ -62,8 +62,8 @@ fig_cor_pressure_bw <- test_summed |>
   scale_x_continuous(breaks = seq(0.5, 2.5, 1)) +
   scale_y_continuous(breaks = seq(0, 0.12, 0.04)) +
   #scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107", "#004D40")) +
-  scale_color_manual(values = c("gray20", "gray20", "gray80", "gray80")) +
-  scale_fill_manual(values = c("gray20", "gray20", "gray80", "gray80")) +
+  scale_color_manual(values = rev(c("gray20", "gray20", "gray80", "gray80"))) +
+  scale_fill_manual(values = rev(c("gray20", "gray20", "gray80", "gray80"))) +
   scale_shape_manual(values = c(21, 24, 22, 25)) +
   labs(y = "Pressure rate",
        x = "Average STRAIN",

### 2023-07-21 scripts/05_fig_stability.R

@@ -64,8 +64,8 @@ fig_stability_bw <- strain_first |>
   geom_smooth(method = "lm", se = FALSE, linewidth = 1.5, color = "black", alpha = 0.5) +
   geom_point(aes(color = pos, group = name, fill = pos, shape = pos), size = 2, alpha = 0.8) +
   #scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107", "#004D40")) +
-  scale_color_manual(values = c("gray20", "gray20", "gray80", "gray80")) +
-  scale_fill_manual(values = c("gray20", "gray20", "gray80", "gray80")) +
+  scale_color_manual(values = rev(c("gray20", "gray20", "gray80", "gray80"))) +
+  scale_fill_manual(values = rev(c("gray20", "gray20", "gray80", "gray80"))) +
   scale_shape_manual(values = c(21, 24, 22, 25)) +
   labs(x = "Average STRAIN (first 4 weeks)",
        y = "Average STRAIN (last 4 weeks)",

### 2023-07-21 scripts/07_fig_predictability.R

@@ -108,8 +108,8 @@ hsh_hsh_bw <- hsh_first |>
   geom_smooth(method = "lm", se = FALSE, linewidth = 1.5, color = "black", alpha = 0.5) +
   geom_point(aes(color = officialPosition, group = nflId, fill = officialPosition, shape = officialPosition), size = 2, alpha = 0.8) +
   #scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107", "#004D40")) +
-  scale_color_manual(values = c("gray20", "gray20", "gray80", "gray80")) +
-  scale_fill_manual(values = c("gray20", "gray20", "gray80", "gray80")) +
+  scale_color_manual(values = rev(c("gray20", "gray20", "gray80", "gray80"))) +
+  scale_fill_manual(values = rev(c("gray20", "gray20", "gray80", "gray80"))) +
   scale_shape_manual(values = c(21, 24, 22, 25)) +
   labs(x = "Pressure rate (first 4 weeks)",
        y = "Pressure rate (last 4 weeks)",
@@ -136,8 +136,8 @@ strain_hsh_bw <- hsh_last |>
   geom_smooth(method = "lm", se = FALSE, linewidth = 1.5, color = "black", alpha = 0.5) +
   geom_point(aes(color = pos, group = name, fill = pos, shape = pos), size = 2, alpha = 0.8) +
   #scale_color_manual(values = c("#D81B60", "#1E88E5", "#FFC107", "#004D40")) +
-  scale_color_manual(values = c("gray20", "gray20", "gray80", "gray80")) +
-  scale_fill_manual(values = c("gray20", "gray20", "gray80", "gray80")) +
+  scale_color_manual(values = rev(c("gray20", "gray20", "gray80", "gray80"))) +
+  scale_fill_manual(values = rev(c("gray20", "gray20", "gray80", "gray80"))) +
   scale_shape_manual(values = c(21, 24, 22, 25)) +
   labs(x = "Average STRAIN (first 4 weeks)",
        # y = "Pressure rate (last 4 weeks)",

### 2023-07-21 scripts/08_fig_outcome_curves.R

@@ -111,7 +111,7 @@ fig_outcome_curves_bw <- hit |>
   geom_smooth(se = FALSE, span = 0.4, linewidth = 1) +
   scale_x_continuous(breaks = seq(0, 40, 10), labels = 0:4) +
   scale_linetype_manual(values = c("solid", "dashed", "dotdash", "dotted"), name = "Position") +
-  scale_color_manual(values = c("black", "black", "gray60", "gray60"), name = "Position") +
+  scale_color_manual(values = rev(c("black", "black", "gray60", "gray60")), name = "Position") +
   expand_limits(y = 2.3) +
   labs(y = "STRAIN",
        x = "Time since snap (seconds)") +

### 2023-07-21 journal/03-methods.Rmd
minor edits; ready for review

@@
 We use random intercepts for the two player groups: pass rushers as $R$ and nearest pass blockers as $B$, as well as for the two team groups: defense $D$ and offense $O$.
-We also account for attributes about pass rusher $j$ in play $i$ through the covariate vector $\mathbf{x_{ij}}$, and estimate their respective coefficients $\boldsymbol \beta$ as fixed-effects. 
+We also account for attributes about pass rusher $j$ in play $i$ through the covariate vector $\mathbf{x_{ij}}$, and estimate their respective coefficients $\boldsymbol \beta$ as fixed effects. 
 Our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows.
@@
 
-As for the fixed-effects about pass rusher $j$ in play $i$, we include a variety of features that likely contribute to variation in STRAIN.
+As for the fixed effects about pass rusher $j$ in play $i$, we include a variety of features that likely contribute to variation in STRAIN.
 First, we adjust for the position of both the pass rusher and nearest blocker to account for any positional effects.

### 2023-07-21 journal/supplement.Rmd
minor edits; ready for review

@@
 
-\spacingset{1.45}
+\spacingset{1.5}
 
@@
 Note that the point size for each Las Vegas player corresponds to their estimated STRAIN as the play progresses.
-This play is represented in Table 1 and Figure 1 (left) of the paper.
+This play is illustrated with Table 1 and Figure 1 (left) of the paper.
 
@@
 #| fig.cap: "Distribution of the number of pass rushers on passing plays."
-#| fig.width: 4
-#| fig.height: 3
+#| fig.width: 3.2
+#| fig.height: 2.7
 source(here("scripts", "12_fig_pass_rushers_count.R"))
@@
 We consider a normal distribution to obtain the random intercepts for each player toward their respective position group means.
-As before, the coefficients $\boldsymbol \beta$ associated with covariate vector $\mathbf{x_{ij}}$ for pass rusher $j$ in play $i$ are estimated as fixed-effects. 
-The variables included as fixed effects are the same as those in the model from Section 3.2 of the paper: position of pass rusher and blocker, number of blockers in a play, and play-context information (down, yards to go, and current yardline).
+As before, the coefficients $\boldsymbol \beta$ associated with covariate vector $\mathbf{x_{ij}}$ for pass rusher $j$ in play $i$ are estimated as fixed effects. 
+The variables included as fixed effects are the same as those in the model from Section 3.2 of the paper, namely, position of pass rusher and blocker, number of blockers in a play, and play-context information (down, yards to go, and current yardline).
 
@@
 Furthermore, the model with team effects yields a lower overall holdout MSE ($4.1882$) than the one without random intercepts for teams ($4.4138$).
-This provides a quantitative justification for including team effects in our multilevel model.
+The out-of-sample performance justifies the inclusion of team effects in our multilevel model.
 

### 2023-07-21 journal/response.Rmd
updated response

@@
 On the other hand, we also compute the correlations in minimum and maximum STRAIN between the first and last four weeks of the 2021 NFL season, and they are both much lower compared to average STRAIN.
-Further, we agree with the comment that our analysis did not account for the fact that play length can differ.
-This problem can be addressed with a time-to-event analysis.
-Since the main goal of this paper is to introduce STRAIN as a metric for evaluating pass rushers in football, we leave this for future work.**
+Further, we recognize the limitation that our analysis did not account for the fact that play length can differ.
+This problem can be addressed with a time-to-event model.
+Although we agree that this is an important consideration, since the main goal of this paper is to introduce STRAIN as a metric for evaluating pass rushers in football, we leave this for future work.
 

### 2023-07-21 journal/response.Rmd
bold text

@@
 This problem can be addressed with a time-to-event model.
-Although we agree that this is an important consideration, since the main goal of this paper is to introduce STRAIN as a metric for evaluating pass rushers in football, we leave this for future work.
+Although we agree that this is an important consideration, since the main goal of this paper is to introduce STRAIN as a metric for evaluating pass rushers in football, we leave this for future work.**
 

### 2023-07-22 journal/response.Rmd
edit time to event response part

--- 2023-07-21 journal/response.Rmd
@@
 Further, we recognize the limitation that our analysis did not account for the fact that play length can differ.
-This problem can be addressed with a time-to-event model.
-Although we agree that this is an important consideration, since the main goal of this paper is to introduce STRAIN as a metric for evaluating pass rushers in football, we leave this for future work.
-
-[TODO: edit the time to event response]
+Although we agree that this is an important consideration, since the main goal of this paper is to introduce STRAIN as a metric for evaluating pass rushers in football, we leave consideration of time-to-event modeling for future work.**
 

### 2023-07-22 journal/response.Rmd
last revision before submission

@@
 output: pdf_document
-fontsize: 10pt
+fontsize: 11pt
 header-includes:
@@
 In the revised version, this part has been moved to the supplement; see section A.
-We have added clarification about the positions and it now reads "Note that apart from the formations shown in Figure 1 [of the supplement], defensive linemen can have the flexibility to line up differently.
+We have added clarification about the positions and it now reads "Note that apart from the formations shown in Figure A.1 [of the supplement], defensive linemen can have the flexibility to line up differently.
 For example, a defensive end, depending on their opposing matchup, may not necessarily be positioned toward the outside of the line of scrimmage."**
@@
 Further, we recognize the limitation that our analysis did not account for the fact that play length can differ.
-Although we agree that this is an important consideration, since the main goal of this paper is to introduce STRAIN as a metric for evaluating pass rushers in football, we leave consideration of time-to-event modeling for future work.**
+Although we agree that this is an important consideration, which requires time-to-event modeling, since the main goal of this paper is to introduce STRAIN as a metric for evaluating pass rushers in football, we leave this problem for future work.**
 

### 2023-07-22 journal/01-introduction.Rmd
last revision before submission

@@
 
-Our focus of this manuscript is specifically on measuring the performance of defensive linemen in the NFL.
+Our focus of this manuscript is specifically on measuring the performance of NFL defensive linemen (see Section A of the Supplement for more information on the defensive positions in football).
 Using data made available in the Big Data Bowl 2023, we present a novel approach to measure the performance of pass rushers.
@@
 
-<!-- Table \ref{tab:metrics} gives a summary of existing football metrics for pass rush. -->
-<!-- We now highlight what these quantities describe as well as their limitations. -->
+Table B.1 of the Supplement gives a summary of existing football metrics for pass rush.
+We now highlight what these quantities describe as well as their limitations.
 
@@
 This is better than the individual counts to some extent, but suffers from problems of subjectivity (e.g., whether there is an actual hurry or not).
-Pro Football Focus (PFF) defines a metric called pass-rush productivity, which is a minor modification from the aforementioned pressures metric.<!-- (see Table \ref{tab:metrics}). -->
+Pro Football Focus (PFF) defines a metric called pass-rush productivity, which is a minor modification from the aforementioned pressures metric.
 In particular, pass-rush productivity gives twice as much weight to a sack relative to hurries and hits, which is a small upgrade to pressures.

### 2023-07-22 journal/02-data.Rmd
last revision before submission

@@
 For each play, we have information on the on-field location, speed, angle, direction, and orientation of each player on the field and the football at a rate of 10 Hz (i.e., 10 measurements per second), along with event annotations for each frame such as ball snap, pass forward, and quarterback sack, to name a few.
-Table \ref{tab:track} displays a tracking data example for a [play](https://www.raiders.com/video/de-maxx-crosby-sacks-qb-teddy-bridgewater-for-a-loss-of-6-yards-nfl) from the 2021 NFL regular season week six matchup between the Las Vegas Raiders and Denver Broncos, which ends with Broncos quarterback Teddy Bridgewater getting sacked by Raiders defensive end Maxx Crosby.
 
@@
 
+Throughout this paper, we use a [play](https://www.raiders.com/video/de-maxx-crosby-sacks-qb-teddy-bridgewater-for-a-loss-of-6-yards-nfl) from the 2021 NFL regular season week six matchup between the Las Vegas Raiders and Denver Broncos as a case study for examining pass rush in football.
+Table C.1 of the Supplement displays a tracking data example for the aforementioned play, which ends with Broncos quarterback Teddy Bridgewater getting sacked by Raiders defensive end Maxx Crosby.
+
 <!-- In addition, Figure \ref{fig:fig_field_raw} presents the locations of every Las Vegas (in black) and Denver (in orange) player on the field from this play at 1, 2, 3 and 4 seconds after the ball snap, with Maxx Crosby highlighted in blue.
@@
 <!-- [^crosbysack]: Game footage of this play is available at https://www.raiders.com/video/de-maxx-crosby-sacks-qb-teddy-bridgewater-for-a-loss-of-6-yards-nfl -->
-
-\begin{table}
-\caption{Example of tracking data for a play during the Las Vegas Raiders versus Denver Broncos NFL game on October 17, 2021.
-The data shown here are for Raiders defensive end Maxx Crosby, and the frames included are between the ball snap and when the sack by Crosby is recorded.
-\label{tab:track}}
-\centering
-\begin{tabular}{ccccccccc}
-\hline
-frameId & x & y & s & a & dis & o & dir & event \\ 
-\hline
-7 & 67.68 & 29.89 & 0.34 & 1.57 & 0.04 & 124.86 & 88.21 & ball\_snap \\ 
-8 & 67.76 & 29.89 & 0.69 & 2.13 & 0.08 & 124.07 & 89.59 & None \\ 
-\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\ 
-50 & 73.67 & 25.06 & 4.19 & 2.62 & 0.42 & 134.21 & 125.26 & qb\_sack \\ 
-\hline
-\end{tabular}
-\end{table}
 
@@
 In this manuscript, we use the PFF data to identify 36,362 unique pass rush attempts by players designated in the "pass rush" role across all plays.
+For context, Figure D.1 of the Supplement displays the distribution of the number of observed pass rushers involved in a play, ranging from 1 to 8 with 4 pass rushers (i.e., 4-man rush formation) as the most common value.
 We also leverage this scouting data to count how many hits, hurries, and sacks each pass rusher is credited with across the span of observed data. 

### 2023-07-22 journal/03-methods.Rmd
last revision before submission

@@
 \end{equation*}
-This can be helpful for metric assessment, as we use average STRAIN to examine different statistical properties of our measure in Section \ref{sec:statprop}.
+This can be helpful in determining the most effective pass rushers; see Section E of the Supplement for our pass rusher evaluation based on average STRAIN.
+We also use average STRAIN to assess different statistical properties of our metric in Section \ref{sec:statprop}.
 
@@
 Due to the nested nature of players on teams, the individual pass rusher and blocker random intercepts reflect the respective player's effect relative to their team effects.
+Furthermore, our proposed model demonstrates better out-of-sample predictive performance than a model without random intercepts for teams, as detailed in Section F of the Supplement.
 We implement the model using penalized likelihood via the `lme4` package in `R` [@bateslme4; @R2023Language].

### 2023-07-22 journal/04-results.Rmd
minor edits; ready for review
last revision before submission

@@
 
-To illustrate our proposed metric STRAIN for pass rush evaluation, we use the same play from the Las Vegas Raiders and Denver Broncos game as mentioned in Section \ref{sec:data}.
+To illustrate our proposed metric STRAIN for pass rush evaluation, we use the example play from the Las Vegas Raiders and Denver Broncos game as mentioned in Section \ref{sec:data}.
 Figure \ref{fig:fig_features_combined} (left) is a line graph showing how Crosby's distance from the quarterback, velocity, and STRAIN change continuously throughout the play.
+In addition, Figure C.1 of the Supplement presents the locations of every Las Vegas and Denver player
+on the field from this play at 1, 2, 3 and 4 seconds after the ball snap, with the point size for each Las Vegas defender corresponding to the estimated STRAIN at the selected frames.
+
 We observe that for the first two seconds, Crosby is being blocked by a Denver offensive lineman and unable to get close to the quarterback, hence the corresponding STRAIN values are virtually zero. 
@@
 The results of fitting the multilevel model described in Section \ref{sec:multilevel} is displayed in Table \ref{tab:fixedeff}.
-First, we investigate the fixed-effects terms of this model.
+First, we investigate the fixed effects terms of this model.
 It appears that that the average STRAIN decreases by 0.7366 (95% CI [-0.7875, -0.6858]) for every additional blocker in the offensive unit, after accounting for other covariates.
@@
 \begin{table}
-\caption{Fixed-effects coefficient estimates for the multilevel model for average STRAIN.
+\caption{Fixed effects coefficient estimates for the multilevel model for average STRAIN.
 Note that the reference down level is first down, the reference pass rusher level is defensive end (DE), denoted R:DE; the reference pass blocker level is tackle (T), denoted B:T.

### 2023-07-22 journal/supplement.Rmd
last revision before submission

@@
   \def\thesection{\Alph{section}}
+  \counterwithin{figure}{section}
+  \counterwithin{table}{section}
 ---
 
-\spacingset{1.5}
+\spacingset{1.45}
 
@@
 
-Table \ref{tab:metrics} gives a summary of existing football metrics for pass rush, as discussed in Section 1.1 of the paper.
+Table \ref{tab:metrics} summarizes previous pass rush statistics in football, to accompany Section 1.1 of the paper.
 
@@
 
+\newpage
+
 # Visualizing Player Tracking Data
 
-Figure \ref{fig:fig_field} shows snapshots of a play from the week six game between the Las Vegas Raiders and Denver Broncos in the 2021 NFL regular season.
+Table \ref{tab:track} shows an example of player tracking data for a play from the week six game between the Las Vegas Raiders and Denver Broncos in the 2021 NFL regular season.
 The play outcome is a sack made by Raiders defensive end Maxx Crosby on Broncos quarterback Teddy Bridgewater.
-The locations of every Las Vegas (in black, on defense) and Denver (in orange, on offense) player on the field at 1, 2, 3, and 4 seconds after the ball snap is displayed, with Crosby highlighted in blue.
-Note that the point size for each Las Vegas player corresponds to their estimated STRAIN as the play progresses.
-This play is illustrated with Table 1 and Figure 1 (left) of the paper.
+Figure \ref{fig:fig_field} shows different snapshots of this play, where the locations of every Las Vegas (in black, on defense) and Denver (in orange, on offense) player on the field at 1, 2, 3, and 4 seconds after the ball snap is displayed, with Crosby highlighted in blue.
+Note that the point size for each Las Vegas defender represents their STRAIN value at the different snapshots.
+
+\begin{table}
+\caption{Example of tracking data for a play during the Las Vegas Raiders versus Denver Broncos NFL game on October 17, 2021.
+The variables include frame identifier for each play (\textbf{frameId}); player position on the field (\textbf{x} and \textbf{y} coordinates); speed (\textbf{s}, in yards/second); acceleration (\textbf{a}, in yards/second$^2$); distance traveled from previous frame (\textbf{dis}, in yards); orientation (\textbf{o}, between 0 and 360 degrees); angle of motion (\textbf{dir}, between 0 and 360 degrees); and event annotation for each frame (\textbf{event}).
+The data shown here are for Raiders defensive end Maxx Crosby, and the frames included are between the ball snap and when the sack by Crosby is recorded.
+\label{tab:track}}
+\centering
+\begin{tabular}{ccccccccc}
+\hline
+\textbf{frameId} & \textbf{x} & \textbf{y} & \textbf{s} & \textbf{a} & \textbf{dis} & \textbf{o} & \textbf{dir} & \textbf{event} \\ 
+\hline
+7 & 67.68 & 29.89 & 0.34 & 1.57 & 0.04 & 124.86 & 88.21 & ball\_snap \\ 
+8 & 67.76 & 29.89 & 0.69 & 2.13 & 0.08 & 124.07 & 89.59 & None \\ 
+\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\ 
+50 & 73.67 & 25.06 & 4.19 & 2.62 & 0.42 & 134.21 & 125.26 & qb\_sack \\ 
+\hline
+\end{tabular}
+\end{table}
 
@@
 
-Figure \ref{fig:pass_rushers_count} displays the distribution of the number of observed pass rushers involved in a play.
-The number of pass rusher ranges from 1 to 8, with 4 pass rushers (i.e., 4-man rush formation) as the most common value.
+Figure \ref{fig:pass_rushers_count} is a histogram of the number of pass rushers across every play in our data.
+The pass rusher count ranges from 1 to 8, with 4 being the mode of the distribution.
 
@@
 
-# Comparison of Multilevel Models
-
-## Modeling Play-Level STRAIN Without Team Effects
+# Multilevel Model Without Team Effects
 
@@
 
-## Model Comparison
-
 We perform leave-one-week-out cross-validation to compare the multilevel models with and without team effects for play-level STRAIN.
@@
 Furthermore, the model with team effects yields a lower overall holdout MSE ($4.1882$) than the one without random intercepts for teams ($4.4138$).
-The out-of-sample performance justifies the inclusion of team effects in our multilevel model.
+The out-of-sample performance justifies the inclusion of team effects in our proposed multilevel model in the paper.
 

### 2023-07-25 journal/04-results.Rmd
added number of bootstraps

@@
 #| label: fig_rankings_boot
-#| fig.cap: "Distributions of player effects for top 10 pass rushers in each position (defensive end, outside linebacker, defensive tackle, nose tackle) by median varying intercept."
+#| fig.cap: "Distributions of player effects (obtained from 1,000 bootstrap samples) for top 10 pass rushers in each position (defensive end, outside linebacker, defensive tackle, nose tackle) by median varying intercept."
 #| fig.width: 6.3

### 2023-07-25 journal/05-discussion.Rmd
added number of bootstraps

@@
 As shown in Figure \ref{fig:fig_rankings_boot}, the first- and fourth-ranked defensive ends Myles Garrett and Takkarist McKinley were teammates on the Cleveland Browns during the first eight weeks of the 2021 NFL season.
-McKinley is a surprising name on our list and we suspect that his high rank is mostly due to playing on the same defensive unit with Garrett, who is highly-regarded as an great pass rusher.
+McKinley is a surprising name on our list and we suspect that his high rank is mostly due to playing on the same defensive unit with Garrett, who is highly-regarded as a great pass rusher.
 In order to capture this type of behavior, we could consider modeling an aggregate STRAIN across pass rushers with a regularized adjusted plus-minus (RAPM) regression approach.

### 2023-07-27 scripts/10_multilevel_mod.R
proofread + small fix

@@ -384,9 +384,18 @@ fig_loocv_team_eff <- read_rds(here("data", "strain_loocv_results.rds")) |>
 
 # resampling --------------------------------------------------------------
 
+pbp21 <- nflfastR::load_pbp(2021)
+
+get_drive <- pbp21 |> 
+  select(gameId = old_game_id, playId = play_id, driveId = drive) |> 
+  mutate(gameId = as.double(gameId))
+
+mod_df_final <- mod_df_final |> 
+  left_join(get_drive)
+
 strain_boot <- function(b) {
   boot_df <- mod_df_final |> 
-    group_by(gameId, playId) |> 
+    group_by(gameId, driveId) |> 
     nest() |> 
     ungroup() |> 
     group_by(gameId) |> 
@@ -449,7 +458,7 @@ strain_boot <- function(b) {
 # 
 # strain_boot_eff <- strain_boot_eff |>
 #   list_rbind()
-# 
+
 # write_rds(strain_boot_eff, here("data", "strain_boot_eff.rds"))
 
 strain_boot_eff <- read_rds(here("data", "strain_boot_eff.rds"))

### 2023-07-29 arxiv/05-discussion.Rmd
reflect preprint changes

--- 2023-07-25 journal/05-discussion.Rmd
@@
 This can be accomplished through the aforementioned RAPM regression or by simply looking at quantities such as the total STRAIN or the maximum STRAIN per frame aggregated across the entire play.
-It is also possible to apply STRAIN to assess individual offensive pass blockers, provided that a method of matching blockers to rushers was developed.
+It is also possible to apply STRAIN to assess individual offensive pass blockers, provided that a method of matching blockers to rushers is developed.
 Furthermore, compared to existing metrics, STRAIN measures pass rush effectiveness for every play continuously over time, which is at a much more granular level than considering whether the play resulted in a binary outcome such as a sack.
